\documentclass[10pt]{article}

% Be sure to use PDF Latex
\pdfoutput=1

% links
\usepackage[bookmarks,bookmarksdepth=2, colorlinks=true, linkcolor=blue,citecolor=red, urlcolor=blue]{hyperref}


\usepackage{fullpage}

\usepackage[latin1]{inputenc}
\usepackage{mystyle}
\usepackage{wrapfig}
\usepackage{notations_ot} % for OT chapter
\usepackage{tikz}



% Draft
\newcommand{\todo}[1]{\textbf{\textcolor{red}{[ToDo: #1]}}}
\newcommand{\blue}[1]{\textcolor{blue}{ #1}}
\renewcommand{\todo}[1]{}
\newcommand{\removed}[1]{}

\newcommand{\dims}{d}


\graphicspath{{./figures/}}


\newcommand{\myauthor}[3]{#1\protect\footnote{#2, \protect\url{#3}}}


\title{Optimal Transport for Machine Learners\\\textit{Course notes}} 

\author{%
\begin{tabular}{c}
	Gabriel Peyr{\'e} \\ CNRS and ENS, PSL Universit\'e \\
	 \url{gabriel.peyre@ens.fr}
\end{tabular}
}


\date{\today}

%%

\begin{document}

\maketitle

\begin{abstract}
Optimal Transport is a foundational mathematical theory that connects optimization, partial differential equations, and probability. It offers a powerful framework for comparing probability distributions and has recently become an important tool in machine learning, especially for designing and evaluating generative models. These course notes cover the fundamental mathematical aspects of OT, including the Monge and Kantorovich formulations, Brenier's theorem, the dual and dynamic formulations, the Bures metric on Gaussian distributions, and gradient flows. It also introduces numerical methods such as linear programming, semi-discrete solvers, and entropic regularization. Applications in machine learning include topics like training neural networks via gradient flows, token dynamics in transformers, and the structure of GANs and diffusion models. These notes focus primarily on mathematical content rather than deep learning techniques. \\
\noindent\textbf{Disclaimer:} These notes are not intended to be a definitive resource on optimal transport. They should be viewed as an intermediate reference positioned between two extremal points: the book by Peyr\'e and Cuturi \cite{peyre2019computational}, which focuses on computational aspects, and the book by Santambrogio \cite{SantambrogioBook}, which emphasizes theoretical foundations.
\end{abstract}

% \tableofcontents

\input{sections/matching}
\input{sections/monge}
\input{sections/kantorovitch}
\input{sections/sinkhorn}
\input{sections/dual}
\input{sections/semidiscr-w1}
\input{sections/dual-norms}
\input{sections/sinkhorn-advanced}
\input{sections/extensions}
\input{sections/grad-flows}

% \nocite{*}

\bibliographystyle{plain}
\bibliography{all}

\end{document}
