\documentclass[10pt,a4paper,twoside,openany]{book}

% Be sure to use PDF Latex
\pdfoutput=1

\usepackage[T1]{fontenc}
\usepackage[a4paper,inner=28mm,outer=24mm,top=26mm,bottom=28mm,headheight=14pt,headsep=8mm]{geometry}
\usepackage{newpxtext}
\usepackage{microtype}
\usepackage{fancyhdr}
\usepackage[explicit]{titlesec}
\usepackage{makeidx}
\usepackage{longtable}

\usepackage[latin1]{inputenc}
\usepackage{mystyle}
\let\otmltransp\transp
\let\transp\undefined
\usepackage{newpxmath}
\setcounter{localmathalphabets}{0}
\let\newpxmathtransp\transp
\let\transp\otmltransp
\usepackage[scaled=.95]{inconsolata}
\usepackage{wrapfig}
\usepackage{notations_ot} % for OT chapter
\usepackage{tikz}

\definecolor{otblue}{HTML}{1F4E79}
\definecolor{otred}{HTML}{A43D3D}
\definecolor{otgray}{HTML}{5F6670}

\titleformat{\chapter}[display]
	{\normalfont\bfseries\color{otblue}}
	{\filleft\MakeUppercase{\chaptertitlename}\ \thechapter}
	{1ex}
	{\titlerule[0.8pt]\vspace{1.3ex}\huge\raggedright #1}
	[\vspace{1ex}\titlerule]
\titleformat{name=\chapter,numberless}[display]
	{\normalfont\bfseries\color{otblue}}
	{}
	{0pt}
	{\titlerule[0.8pt]\vspace{1.3ex}\huge\raggedright #1}
	[\vspace{1ex}\titlerule]
\titleformat{\section}
	{\normalfont\Large\bfseries\color{otblue}}
	{\thesection}{.75em}{#1}
\titleformat{\subsection}
	{\normalfont\large\bfseries\color{otgray}}
	{\thesubsection}{.75em}{#1}
\titlespacing*{\chapter}{0pt}{-8pt}{28pt}
\titlespacing*{\section}{0pt}{2.8ex plus .8ex minus .2ex}{1.2ex plus .2ex}
\titlespacing*{\subsection}{0pt}{2.2ex plus .6ex minus .2ex}{.8ex plus .2ex}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\small\thepage}
\fancyhead[RE]{\small\nouppercase{\leftmark}}
\fancyhead[LO]{\small\nouppercase{\rightmark}}
\renewcommand{\headrulewidth}{0.4pt}
\fancypagestyle{plain}{%
	\fancyhf{}%
	\fancyhead[LE,RO]{\small\thepage}%
	\renewcommand{\headrulewidth}{0pt}%
}
\renewcommand{\chaptermark}[1]{\markboth{\thechapter\quad #1}{}}
\renewcommand{\sectionmark}[1]{\markright{\thesection\quad #1}}
\setcounter{tocdepth}{1}
\setcounter{secnumdepth}{2}

% links; load hyperref after most packages so float anchors are patched once.
\usepackage[bookmarks,bookmarksdepth=2, colorlinks=true, linkcolor=blue,citecolor=red, urlcolor=blue]{hyperref}
\renewcommand{\a}{\VectMode{a}}
\renewcommand{\b}{\VectMode{b}}
\renewcommand{\c}{c}
\renewcommand{\d}{\ins{d}}
\renewcommand{\H}{H}
\renewcommand{\P}{\VectMode{P}}
\renewcommand{\S}{\VectMode{S}}
\renewcommand{\SS}{\VectMode{S}}
\renewcommand{\th}{\theta}
\renewcommand{\theHfigure}{\thechapter.\arabic{figure}}



% Draft
\newcommand{\todo}[1]{\textbf{\textcolor{red}{[ToDo: #1]}}}
\newcommand{\blue}[1]{\textcolor{blue}{ #1}}
\renewcommand{\todo}[1]{}
\newcommand{\removed}[1]{}

\newcommand{\dims}{d}
\makeindex

\newcommand{\myauthor}[3]{\begin{tabular}[t]{c}#1\\#2\\\url{#3}\end{tabular}}


\title{Optimal Transport\\for Machine Learners}

\author{%
\begin{tabular}{c}
	Gabriel Peyr{\'e} \\ CNRS and ENS, PSL Universit\'e \\
	 \url{gabriel.peyre@ens.fr}
\end{tabular}
}


\date{\today}

\makeatletter
\renewcommand{\maketitle}{%
\begin{titlepage}
\thispagestyle{empty}
\vspace*{.08\textheight}
\noindent{\color{otblue}\rule{\textwidth}{1.1pt}}\par
\vspace{1.4cm}
{\centering\fontsize{30}{36}\selectfont\bfseries\color{otblue}\@title\par}
\vspace{1.2cm}
{\centering\large\@author\par}
\vfill
\begin{center}
{\color{otred}\rule{.38\textwidth}{2pt}}\par
\vspace{.65cm}
{\large\@date\par}
\end{center}
\end{titlepage}
}
\makeatother

%%

\begin{document}

\hypersetup{pageanchor=false}
\pagenumbering{gobble}
\maketitle
\cleardoublepage
\hypersetup{pageanchor=true}

\frontmatter

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
Modern machine learning repeatedly manipulates probability measures: empirical datasets, generated samples, latent distributions, class-conditional laws, particle systems, weights of wide networks and attention patterns. Optimal transport is useful in this setting because it compares such objects by asking how mass should move. It therefore combines a statistically meaningful notion of discrepancy with a geometry of interpolation, dual certificates and variational dynamics. This makes OT a common language for losses, generative modeling, domain adaptation, robust learning, barycenters, gradient flows and mean-field descriptions of learning algorithms.
\index{conditional law}% strictindex
\index{dual!certificate}% strictindex
\index{probability measure}% strictindex
\index{gradient!flow}% strictindex
\index{generative model}% strictindex

This book presents the main OT techniques with these machine-learning uses in mind. It starts from finite assignment and the Monge map viewpoint, passes to Kantorovich couplings and dual potentials, and then explains the algorithmic ideas that make transport usable: linear programming, semi-discrete cells, Sinkhorn scaling and low-dimensional projections. The same objects are then reused as a geometry of measures, giving Wasserstein distances, barycenters, gradient flows, dynamic formulations and Gaussian/Bures formulas. The final chapters emphasize the variants most relevant to modern ML: divergences and adversarial losses, entropic and unbalanced relaxations, robust or spectral ground geometries, Gromov and quantum extensions, and transport-based views of generative models, mean-field networks and attention dynamics. The goal is to keep the mathematics explicit while exposing the computational and geometric intuitions needed to turn OT into a working toolbox for machine learners.
\index{Sinkhorn!scaling}% strictindex
\index{adversarial!loss}% strictindex
\index{Monge!problem}% strictindex
\index{dual!potential}% strictindex
\index{linear programming}% strictindex
\index{Wasserstein!distance}% strictindex
\index{gradient!flow}% strictindex
\index{generative model}% strictindex

\noindent All material for this book, including the code used to reproduce the figures, is available at \href{http://github.com/gpeyre/ot4ml}{\texttt{gpeyre/ot4ml}}. Most computational figures were produced with the Python Optimal Transport (POT) library~\cite{flamary2021pot}. The author warmly thanks the POT team and contributors for their important and sustained effort in making reliable optimal-transport algorithms available to the community.

\chapter*{Guide to the Literature and Scope}
\addcontentsline{toc}{chapter}{Guide to the Literature and Scope}

Several books already cover optimal transport from complementary viewpoints. The two-volume monograph of Rachev and R\"uschendorf~\cite{rachev1998mass,rachev1998mass2} gives a broad probabilistic treatment of mass transportation and its applications. Villani's books~\cite{Villani03,Villani09} are the standard references for the modern mathematical theory, from Kantorovich duality to curvature, concentration and geometric analysis. Santambrogio's text~\cite{SantambrogioBook} offers a concise applied-mathematics route through the same foundations, with a strong emphasis on PDEs and variational arguments. Ambrosio, Gigli and Savar\'e~\cite{ambrosio2006gradient} develop the metric-space theory of gradient flows that underlies the dynamical part of the subject.
\index{Kantorovich!duality}% strictindex
\index{gradient!flow}% strictindex

On the computational side, Peyr\'e and Cuturi~\cite{peyre2019computational} provide the reference account of numerical OT, entropic regularization and applications in data sciences. Galichon's book~\cite{galichon2016optimal} explains the economic and matching-theoretic viewpoint, while the statistical theory of OT is developed in the recent lecture notes of Chewi, Niles-Weed and Rigollet~\cite{weed2025statistical}. Recent surveys complement these books by emphasizing scalable algorithms and machine-learning applications~\cite{khamis2024scalable,montesuma2023recent}, as well as the role of OT in imaging and graphics~\cite{bonneel2023survey}. These references remain the natural places to find exhaustive proofs, historical details and specialized variants.
\index{entropic!regularization}% strictindex

The aim here is different and more selective. The book keeps the core mathematics explicit, but organizes it around the questions that repeatedly arise in machine learning: how to compare singular empirical measures, how to compute differentiable transport losses, how regularization changes optimization and statistics, how dual potentials become discriminators, and how transport geometry produces flows of particles, neurons and tokens. The intended contribution is therefore not a replacement for the references above, but a compact bridge between rigorous OT and the geometric intuitions needed to use it in modern ML.
\index{neuron}% strictindex
\index{dual!potential}% strictindex
\index{empirical!measure}% strictindex

\tableofcontents

\mainmatter

\input{sections/matching}
\input{sections/monge}
\input{sections/kantorovich}
\input{sections/dual}
\input{sections/semidiscr-w1}
\input{sections/dual-norms}
\input{sections/sinkhorn}
\input{sections/sinkhorn-advanced}
\input{sections/generalized-wasserstein}
\input{sections/generalized-ot-problems}
\input{sections/beyond-comparing-measures}
\input{sections/dynamic-ot}
\input{sections/wasserstein-gradient-flows}
\input{sections/transportation-models}

\chapter*{Conclusion}
\addcontentsline{toc}{chapter}{Conclusion}

Optimal transport is useful because it keeps several viewpoints active at once. It is a linear program over couplings, a duality theory for potentials, a geometry on probability measures, a source of PDEs and gradient flows, and a computational toolbox built around linear programming, Sinkhorn scaling and low-dimensional projections. These viewpoints reinforce each other: Brenier maps explain geodesics, geodesic convexity explains convergence of flows, entropic regularization turns transport into scalable differentiable losses, and dual norms or sliced variants reveal what is gained and lost when OT is simplified.
\index{Sinkhorn!scaling}% strictindex
\index{Brenier!map}% strictindex
\index{dual!norm}% strictindex
\index{linear programming}% strictindex
\index{entropic!regularization}% strictindex
\index{probability measure}% strictindex
\index{convexity!geodesic}% strictindex
\index{gradient!flow}% strictindex

For machine learning, this interplay is especially stimulating. Modern generative modeling, diffusion and flow matching, inverse problems, robust optimization, and even continuous-depth views of transformers all ask for ways to move, compare or learn distributions in high dimension. These applications do not merely consume existing OT theory; they create difficult mathematical questions because empirical measures are singular, dimensions are large, models are parametrized and non-convex, and computational approximations are inseparable from statistical error. The strength of OT is precisely that it gives a common language for these tensions, while still leaving enough open structure for new mathematics to be needed.
\index{empirical!measure}% strictindex
\index{flow!matching}% strictindex
\index{generative model}% strictindex
\index{transformer}% strictindex

\phantomsection
\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

This work was supported by the European Research Council (ERC project WOLF) and by the French government under the management of Agence Nationale de la Recherche as part of the ``France 2030'' program, reference ANR-23-IACL-0008 (PRAIRIE-PSAI).

% \nocite{*}

\bibliographystyle{plain}
\bibliography{all}

\clearpage
\appendix
\input{sections/notation-table}

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Index}
\printindex

\end{document}
