\documentclass[10pt,a4paper,twoside]{article}

\pdfoutput=1

\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage[a4paper,inner=28mm,outer=24mm,top=26mm,bottom=28mm,headheight=20pt,headsep=7mm]{geometry}
\usepackage{newpxtext}
\usepackage[scaled=.92]{helvet}
\usepackage{microtype}
\usepackage{fancyhdr}
\usepackage[explicit]{titlesec}
\usepackage{longtable}
\usepackage{needspace}

\usepackage{mystyle}
\let\otmltransp\transp
\let\transp\undefined
\usepackage{newpxmath}
\setcounter{localmathalphabets}{0}
\let\newpxmathtransp\transp
\let\transp\otmltransp
\usepackage[scaled=.95]{inconsolata}
\usepackage{wrapfig}
\usepackage{placeins}
\usepackage{notations_ot}
\usepackage{tikz}

\definecolor{otblue}{HTML}{1F4E79}
\definecolor{otred}{HTML}{A43D3D}
\definecolor{otgray}{HTML}{5F6670}
\definecolor{otink}{HTML}{162332}
\definecolor{otpale}{HTML}{EAF2F8}
\definecolor{otline}{HTML}{B7C8D8}
\definecolor{otaccent}{HTML}{C94A4A}
\definecolor{otlink}{HTML}{1B4D74}

\newcommand{\otsectiontag}[1]{%
	\tikz[baseline=(otsectiontag.base)]%
	\node[
		anchor=base,
		rounded corners=2pt,
		fill=otpale,
		draw=otline,
		line width=.35pt,
		inner xsep=7pt,
		inner ysep=2.5pt
	] (otsectiontag) {\sffamily\bfseries\footnotesize\color{otblue}#1};%
}
\newcommand{\otsectionhead}[2]{%
	\Needspace{18\baselineskip}%
	\noindent\makebox[\linewidth][l]{%
		\otsectiontag{SECTION #1}%
		\hfill{\color{otline}\rule{.52\linewidth}{.35pt}}%
	}\par
	\vspace{.65ex}%
	{\sffamily\bfseries\fontsize{22}{25}\selectfont\color{otblue}\raggedright #2\par}%
	\vspace{.25ex}%
	\noindent{\color{otaccent}\rule{.16\linewidth}{1.4pt}\color{otline}\rule{\dimexpr.84\linewidth\relax}{.35pt}}%
}
\newcommand{\otsectionheadplain}[1]{%
	\Needspace{9\baselineskip}%
	\noindent{\color{otline}\rule{\linewidth}{.35pt}}\par
	\vspace{.55ex}%
	{\sffamily\bfseries\fontsize{21}{24}\selectfont\color{otblue}\raggedright #1\par}%
	\vspace{.25ex}%
	\noindent{\color{otaccent}\rule{.13\linewidth}{1.4pt}\color{otline}\rule{\dimexpr.87\linewidth\relax}{.35pt}}%
}
\newcommand{\otsubsectiontag}[1]{%
	\tikz[baseline=(otsubsectiontag.base)]%
	\node[
		anchor=base,
		rounded corners=2pt,
		fill=otpale,
		draw=otline,
		line width=.25pt,
		inner xsep=5pt,
		inner ysep=1.5pt
	] (otsubsectiontag) {\sffamily\bfseries\small\color{otblue}#1};%
}

\titleformat{\section}[display]
	{\normalfont}
	{}
	{0pt}
	{\otsectionhead{\thesection}{#1}}
\titleformat{name=\section,numberless}[display]
	{\normalfont}
	{}
	{0pt}
	{\otsectionheadplain{#1}}
\titleformat{\subsection}[hang]
	{\normalfont\sffamily\Large\bfseries\color{otblue}}
	{\otsubsectiontag{\thesubsection}}
	{.7em}
	{#1}
	[\vspace{-.35ex}{\color{otline}\titlerule[.25pt]}]
\titleformat{\subsubsection}[hang]
	{\normalfont\sffamily\large\bfseries\color{otgray}}
	{\color{otaccent}\rule{1pt}{1.05em}\hspace{.45em}\thesubsubsection}
	{.55em}
	{#1}
\titleformat{\paragraph}[runin]
	{\normalfont\sffamily\bfseries\color{otblue}}
	{}
	{0pt}
	{#1}
\titlespacing*{\section}{0pt}{.8ex plus .3ex minus .2ex}{2.0ex plus .2ex}
\titlespacing*{\subsection}{0pt}{2.1ex plus .5ex minus .2ex}{.85ex plus .15ex}
\titlespacing*{\subsubsection}{0pt}{1.45ex plus .35ex minus .15ex}{.55ex plus .1ex}
\titlespacing*{\paragraph}{0pt}{1.1ex plus .25ex minus .1ex}{.55em}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\small\thepage}
\fancyhead[RE]{\small\nouppercase{\leftmark}}
\fancyhead[LO]{\small\nouppercase{\rightmark}}
\renewcommand{\headrulewidth}{.35pt}
\renewcommand{\headrule}{\hbox to\headwidth{\color{otline}\leaders\hrule height \headrulewidth\hfill}}
\fancypagestyle{plain}{%
	\fancyhf{}%
	\fancyhead[LE,RO]{\small\thepage}%
	\renewcommand{\headrulewidth}{0pt}%
}
\renewcommand{\sectionmark}[1]{\markboth{\thesection\quad #1}{}}
\renewcommand{\subsectionmark}[1]{\markright{\thesubsection\quad #1}}
\setcounter{tocdepth}{2}
\setcounter{secnumdepth}{3}

\usepackage[
	bookmarks,
	bookmarksdepth=2,
	colorlinks=true,
	linkcolor=otlink,
	citecolor=red,
	urlcolor=otlink,
	pdftitle={PDE4ML: PDEs for Machine Learning},
	pdfauthor={Gabriel Peyre},
	pdfsubject={Optimal-transport-oriented review of PDE tools for machine learning},
	pdfkeywords={optimal transport, PDE, machine learning, Wasserstein gradient flow, diffusion models, flow matching}
]{hyperref}

\renewcommand{\a}{\VectMode{a}}
\renewcommand{\b}{\VectMode{b}}
\renewcommand{\c}{c}
\renewcommand{\d}{\ins{d}}
\renewcommand{\H}{H}
\renewcommand{\P}{\VectMode{P}}
\renewcommand{\S}{\VectMode{S}}
\renewcommand{\SS}{\VectMode{S}}
\renewcommand{\th}{\theta}
\renewcommand{\theHfigure}{\thesection.\arabic{figure}}

\newcommand{\todo}[1]{}
\newcommand{\removed}[1]{}
\providecommand{\index}[1]{}
\newcommand{\dims}{d}

\title{PDE4ML\\[-.05em]{\Large PDEs for Machine Learning}}

\author{%
\begin{tabular}{c}
	Gabriel Peyr{\'e} \\ CNRS and ENS, PSL Universit\'e \\
	\url{gabriel.peyre@ens.fr}
\end{tabular}
}

\date{\today}

\makeatletter
\renewcommand{\maketitle}{%
\begin{titlepage}
\thispagestyle{empty}
\vspace*{.08\textheight}
\noindent{\color{otblue}\rule{\textwidth}{1.1pt}}\par
\vspace{1.4cm}
{\centering\fontsize{30}{36}\selectfont\bfseries\color{otblue}\@title\par}
\vspace{1.2cm}
{\centering\large\@author\par}
\vfill
\begin{center}
{\color{otred}\rule{.38\textwidth}{2pt}}\par
\vspace{.65cm}
{\large\@date\par}
\end{center}
\end{titlepage}
}
\makeatother

\begin{document}

\hypersetup{pageanchor=false}
\pagenumbering{gobble}
\maketitle
\clearpage
\hypersetup{pageanchor=true}
\pagenumbering{roman}

\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}
Machine learning increasingly relies on evolutions of probability measures. Training dynamics of wide neural networks, interacting particles, diffusion models, flow matching, attention layers and generative maps can all be described through transport, continuity equations, variational flows or stochastic PDE limits. This review develops an optimal-transport-oriented account of these PDE tools. The emphasis is not on PDE theory in isolation, but on the mechanisms by which PDEs organize learning algorithms: mass conservation, least-action principles, Wasserstein gradients, Fokker--Planck equations, McKean--Vlasov limits and transport-based generative dynamics.

The document is designed as a standalone PDE/ML review extracted and reorganized from the larger OT4ML project. It first gives a compact primer on Monge and Kantorovich optimal transport, Brenier maps and Wasserstein metric geometry. It then develops the material most directly relevant to PDE methods in machine learning: dynamic optimal transport, Wasserstein gradient flows, mean-field training dynamics, diffusion and flow-matching models, transportation views of transformers, and Gaussian closures. The goal is to expose a coherent mathematical language in which ML algorithms become evolutions on spaces of measures.

\noindent The source code for the full OT4ML project, including figure-generation notebooks, is available at \href{https://github.com/gpeyre/ot4ml}{\texttt{gpeyre/ot4ml}}. Most computational figures were produced with the Python Optimal Transport (POT) library~\cite{flamary2021pot}.

\section*{Guide to the Literature and Scope}
\addcontentsline{toc}{section}{Guide to the Literature and Scope}
This review uses optimal transport as the entry point to PDE methods for machine learning. The mathematical foundations of OT are covered in Villani's monographs~\cite{Villani03,Villani09}, Santambrogio's applied text~\cite{SantambrogioBook}, the probabilistic treatment of Rachev and R\"uschendorf~\cite{rachev1998mass,rachev1998mass2}, and the computational account of Peyr\'e and Cuturi~\cite{peyre2019computational}. Gradient flows in metric spaces are developed systematically by Ambrosio, Gigli and Savar\'e~\cite{ambrosio2006gradient}, while the dynamic Benamou--Brenier viewpoint originates in~\cite{benamou2000computational}.

\paragraph{Reference map.}
The review repeatedly returns to a small set of pillars. Static OT and Wasserstein geometry follow the classical line from Monge and Kantorovich to Brenier, McCann and Villani~\cite{Monge1781,Kantorovich42,Brenier91,mccann1997convexity,Villani09}. Dynamic OT and gradient flows use the Benamou--Brenier formula, the JKO scheme and Otto's formal Riemannian calculus~\cite{benamou2000computational,JDB-JKO,otto2001geometry,ambrosio2006gradient}. Machine-learning dynamics enter through mean-field neural-network limits and particle descriptions~\cite{ChizatBach2018OverparameterizedOT,MeiMontanariNguyen2018MeanFieldNN,RotskoffVandenEijnden2018Trainability}. Generative modelling is organized around score-based diffusion, probability-flow ODEs, flow matching, rectified flows and stochastic interpolants~\cite{Song2021ScoreSDE,Lipman2022FlowMatching,Liu2023RectifiedFlow,Albergo2025StochasticInterpolants}.

For machine learning, the relevant PDE themes include Fokker--Planck and McKean--Vlasov limits of particle systems, JKO schemes, entropy and interaction-energy flows, mean-field neural-network training, diffusion models, score-based generative modeling, flow matching and stochastic interpolants. The corresponding references are recalled locally in the text, so that each citation is tied to the mathematical mechanism being discussed.

The scope is deliberately selective. The primer only includes the OT material required later: push-forwards, Monge maps, Kantorovich couplings, Brenier's theorem, Wasserstein distances and geodesics. Algorithmic OT, duality, Sinkhorn convergence and generalized OT are treated in the full OT4ML manuscript; here they are invoked only when they support PDE or ML dynamics.

\paragraph{Reading map.}
The four sections move from static OT notation to dynamic least-action principles, then to Wasserstein gradient flows and particle limits, and finally to transportation dynamics for generative models and continuous-depth architectures.

\clearpage
\tableofcontents

\clearpage
\pagenumbering{arabic}

\input{sections/primer-ot}
\input{sections/dynamic-ot}
\input{sections/wasserstein-gradient-flows}
\input{sections/transportation-models}

\section*{Conclusion}
\addcontentsline{toc}{section}{Conclusion}
The central message of this review is that many machine-learning methods can be read as PDEs on probability measures. Optimal transport supplies the geometry behind this statement: it turns a loss into a metric gradient flow, a coupling into a geodesic, a score into a Fokker--Planck drift, and a generative model into a transported law. This language is valuable because it separates modelling choices from representation choices. A finite particle algorithm, a neural parametrization and a continuum PDE may then be understood as different approximations of the same measure-valued evolution.

The same viewpoint also clarifies the limits of current theory. Empirical measures are singular, neural parametrizations are non-convex, high-dimensional transport is statistically expensive, and generative flows often use interpolants that are transportation dynamics without being Wasserstein geodesics or gradient flows. These tensions are not defects of the PDE viewpoint; they identify where mathematical analysis is still needed. For machine learning, PDEs are therefore both an explanatory tool and a source of new problems.

\begingroup
\sloppy
\hfuzz=1pt
\setlength{\emergencystretch}{2em}
\bibliographystyle{plain}
\bibliography{all}
\endgroup

\end{document}
