% !TEX root = ../CourseOT.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Extensions}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Unbalanced OT}
\label{sec-unbalanced}


%%%
\paragraph{Relaxed formulation}

We consider the following ``unbalanced'' generalization of OT for $(\al,\be) \in \Mm_+^1(\Xx) \times \Mm_+^1(\Yy)$
\eql{\label{eq-unbalanced-primal}
	\text{UW}_c(\al,\be)= \uinf{\pi \in \Mm_+^1(\Xx \times \Yy)} \int_{\Xx \times \Yy} c(x,y) \d\pi(x,y)
		+ \Dd_{\psi_1}(\pi_1|\al) + \Dd_{\psi_2}(\pi_2|\be)
}
for entropy function $\psi_1,\psi_2$.
%
This corresponds to somehow relaxing the exact conservation of mass $(\pi_1,\pi_2)=(\al,\be)$ using a ``soft'' penalty function using divergence functionals.
%
The following proposition states the dual problem.

\begin{prop}One has equality between~\eqref{eq-unbalanced-primal} and
	\eq{
		\text{\upshape UW}_c(\al,\be)=\usup{f \oplus g \leq c} - \Dd_{\psi_1}^*(-f|\al) - \Dd_{\psi_2}^*(-g|\be).
	}
\end{prop}
\begin{proof}
One gets a dual by using the formula~\eqref{eq-legendre} for the dual of a divergence
\begin{align*}
	\uinf{\pi \in \Mm_+^1(\Xx \times \Yy)} \usup{(f,g) \in \Cc(\Xx) \times \Cc(\Yy)}
		\dotp{c}{\pi} + \dotp{-f}{\pi_1} + \dotp{-g}{\pi_1} - \Dd_{\psi_1}^*(-f|\al) - \Dd_{\psi_2}^*(-g|\be). 
\end{align*}
Exchanging the min and the max gives
\begin{align*}
	&\usup{(f,g) \in \Cc(\Xx) \times \Cc(\Yy)}
	- \Dd_{\psi_1}^*(-f|\al) - \Dd_{\psi_2}^*(-g|\be)	+ 
		\uinf{\pi \in \Mm_+^1(\Xx \times \Yy)} 
		\dotp{c-f \oplus g}{\pi} + \dotp{-f}{\pi_1} + \dotp{-g}{\pi_1} \\
	&= 	\usup{f \oplus g \leq c}
	- \Dd_{\psi_1}^*(-f|\al) - \Dd_{\psi_2}^*(-g|\be).
\end{align*}
\end{proof}


%%%
\paragraph{Reverse formulation}

In order to study the metric property of this formulation, we follow Liero, Mielke and Savare, which consider a series of equivalent formulations. 
%
Assuming for simplicity that $(\pi_1,\pi_2)$ have densities with respect to $(\al,\be)$ and vice-versa, one can write the following factorization formula
\begin{align*}
	& \int_{\Xx \times \Yy} c(x,y) \d\pi(x,y)
		+ \Dd_{\psi_1}(\pi_1|\al) + \Dd_{\psi_2}(\pi_2|\be) \\
		=&
	\int_{\Xx \times \Yy} \pa{ c(x,y)  + 
		\psi_1\pa{ \frac{\d\pi_1}{\d\al}(x) } \frac{\d\al}{\d\pi_1}(x) + 
		\psi_2\pa{ \frac{\d\pi_2}{\d\be}(y) } \frac{\d\be}{\d\pi_2}(y)
	} \d\pi(x,y).
\end{align*}
More rigorously, we introduce the following functional
\eq{
	\foralls (c,r,s) \in \RR^3, \quad
	L_c(r,s) \eqdef c + r\psi_1(1/r) + s\psi_2(1/s), 
}
where we implicitly assume that $r\psi_1(1/r) = (\psi_1')^\infty$ when $r=0$, 
one has 
\eq{
	\text{UW}_c(\al,\be) = \uinf{\pi \in \Mm_+^1(\Xx \times \Yy)}
	 \int L_{c(x,y)}( F(x),G(y) ) \d\pi(x,y)
	 + \psi_1(0) \pi_1^\bot(\Xx)
	 + \psi_2(0) \pi_2^\bot(\Yy)
}
where we denoted $F=\frac{\d\pi_1}{\d\al}$, $G=\frac{\d\pi_2}{\d\be}$ and the associated Lebesgue decomposition reads
\eq{
	\pi_1 = F  \frac{\d\pi_1}{\d\al} + \pi_1^\bot
	\qandq
	\pi_2 = G  \frac{\d\pi_2}{\d\be} + \pi_2^\bot.
} 

%%%
\paragraph{Homogeneous formulation}

The second formulation is obtained by introducing the perspective transform of $L_c$
\eq{
	H_c(r,s) \eqdef \uinf{\th>0} \th L_c(r/\th,s/\th).
}
This function $H_c$ is positively 1-homogeneous.
%
The homogeneous formulation is then 
\eql{\label{eq-homogeneous}
	\text{HW}_c(\al,\be) = \uinf{\pi \in \Mm_+^1(\Xx \times \Yy)}
	 \int H_{c(x,y)}( F(x),G(y) ) \d\pi(x,y)
	 + \psi_1(0) \pi_1^\bot(\Xx)
	 + \psi_2(0) \pi_2^\bot(\Yy).
}
One automatically has $\text{HW} \leq \text{UW}$ since $H_c \leq L_c$ (by taking $\th=1$). The more remarkable and non trivial property is that in fact 
\eq{
	\text{HW} = \text{UW}. 
}	

%%%
\paragraph{Conic lifting}

We now suppose $\Xx=\Yy$ and $\psi_1=\psi_2=\psi$.
%
The last formulation is obtained by lifting the above optimization problem on the cone space $\mathfrak{C}[\Xx] \eqdef (\Xx \times \RR_+) / \sim$   associated to $\Xx$ (and similarly for $\Yy$). Here, the equivalence relation $\sim$ ``glue'' together all the points $(x,0)$ which are often called the ``apex'' of the cone. For some exponent $p \geq 1$, we define the following functional 
\eq{
	\De((x,r),(y,s)) \eqdef H_{c(x,y)}(r^p,s^p)^{1/p}.
}
The interesting case is when $c$ and $p$ are chosen so that $\De$ defines a distance on $\mathfrak{C}[\Xx]$. Here are three examples:
\begin{itemize}
	\item $\Dd_\psi=\KL$, $p=2$ and $c(x,y)=-\log \cos^2(d(x,y) \wedge \frac{\pi}{2})$ in which case $\De$ is the so-called cone metric associated to $d$ and reads
		\eq{
			\De((x,r),(y,s))^2 = r^2+s^2-2rs \cos( d(x,y) \wedge \frac{\pi}{2} ).
		}
		This corresponds to the so-called ``Hellinger-Kantorovitch''--`Wasserstein-Fisher-Rao'' setup.
	\item $\Dd_\psi=\KL$, $p=2$ and $c(x,y)=d(x,y)^2$, so that 
		\eq{
			\De((x,r),(y,s))^2 = r^2+s^2-2rs e^{-d(x,y)/2}.
		}
		This corresponds to the so-called ``Gaussian Hellinger'' setup.
	\item $\Dd_\psi=\TV$, $p=1$ and $c(x,y)=d(x,y)$, so that 
		\eq{
			\De((x,r),(y,s)) = r+s-(r \wedge s)(2-d(x,y))_+.
		}
		This corresponds to the so-called ``partial transport'' setup.
\end{itemize}
The homogeneous formulation~\eqref{eq-homogeneous} is then relaxed as an OT-type problem on the cone
\eq{
	\text{CW}(\al,\be) = \uinf{ \ga \in \Mm_+(\mathfrak{C}[\Xx]^2) }
		\enscond{
			\int \De((x,r),(y,s))^p \d\ga((x,r),(y,s))
		}{
			\int r^p \d \ga_1(\cdot,r) = \al, \:
			\int s^p \d \ga_2(\cdot,s) = \be
		}.
}
The following theorem summarize the key results.

\begin{thm}
	One has $\text{\upshape UW}=\text{\upshape HW}=\text{\upshape CW}$. If $\De$ is a distance, so is {\upshape CW}$^{1/p}$. 
\end{thm}


%%%
\paragraph{Entropic regularization}


A generic regularization of this problem reads
\eq{
	\uinf{\pi \in \Mm_+^1(\Xx \times \Yy)} \int_{\Xx \times \Yy} c(x,y) \d\pi(x,y)
		+ \Dd_{\psi_1}(\pi_1|\al) + \Dd_{\psi_2}(\pi_2|\be) + \epsilon \Dd_\phi(\pi|\al \otimes \be)
}
and its dual reads 
\eq{
	\usup{f,g}
	- \Dd_{\psi_1}^*(-f|\al) - \Dd_{\psi_2}^*(-g|\be) + \epsilon \Dd_\phi^*\pa{ \frac{c-f\oplus g}{\epsilon} }.
}
When using $\Dd_\phi=\KL$, the primal dual relation reads $\pi=e^{\frac{f\oplus g - c}{\epsilon}}$ and when also considering $\Dd_{\psi_1}=\Dd_{\psi_2} = \tau \KL$, the dual reads
\eq{
	\usup{f,g}
		- \tau \pa{ \dotp{e^{\frac{-f}{\tau}}}{\al} + \dotp{e^{\frac{-g}{\tau}}}{\be} - 2 }
		- \epsilon \iint_{\Xx \times \Yy} e^{\frac{f\oplus g - c}{\epsilon}} \d \al \otimes \be + \epsilon.
}
Sikhorn iterations are obtained by minimizing iteratively with respect to $f$ and $g$, and reads in this special case
\begin{align*}
	f &= -\frac{\tau\epsilon}{\tau+\epsilon} \log\int_\Yy \exp\pa{ \frac{g(y) - c(\cdot,y)}{\epsilon} }  \d\be(y),  \\
	g &= -\frac{\tau\epsilon}{\tau+\epsilon} \log\int_\Xx \exp\pa{ \frac{f(x) - c(x,\cdot)}{\epsilon} }  \d\al(x). \\
\end{align*}





%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{OT Barycenters}

\paragraph{Frechet means.}

This barycenter problem~\eqref{eq-wass-discr} was originally introduced by~\cite{Carlier_wasserstein_barycenter} following earlier ideas of~\cite{carlierekelandmatching}. They proved in particular uniqueness of the barycenter for $c(x,y)=\norm{x-y}^2$ over $\X=\RR^d$, if one of the input measure has a density with respect to the Lebesgue measure (and more generally under the same hypothesis as the one guaranteeing the existence of a Monge map, see Remark~\ref{rem-exist-mongemap}).

Given a set of input measure $(\be_s)_s$ defined on some space $\X$, the barycenter problem becomes 
\eql{\label{eq-barycenter-generic}
	\umin{\al \in \Mm_+^1(\X)} \sum_{s=1}^S \la_s \MK_{\c}(\al,\be_s).
}
Problem~\eqref{eq-barycenter-generic} can be viewed as a generalization of the problem of computing barycenters of points $(x_s)_{s=1}^S \in \X^S$ to arbitrary measures. Indeed, if $\be_s=\de_{x_s}$ is a single Dirac mass, then a solution to~\eqref{eq-barycenter-generic} is $\de_{x^\star}$ where $x^\star$ is a Fr\'echet mean solving~\eqref{eq-frechet-means}.
%
Note that for $c(x,y)=\norm{x-y}^2$, the mean of the barycenter $\al^\star$ is necessarily the barycenter of the mean, \emph{i.e.} 
\eq{
	\int_\Xx x \d\al^\star(x) =  \sum_s \la_s \int_\Xx x \d\al_s(x), 
}
and the support of $\al^\star$ is located in the convex hull of the supports of the $(\al_s)_s$.
%
In the case where $\X=\RR^d$ and $c(x,y)=\norm{x-y}^2$,~\cite{Carlier_wasserstein_barycenter} shows that if one of the input measures has a density, then this barycenter is unique. 	
	
The following proposition shows that \eqref{eq-barycenter-generic} is a convex optimization problem over the infinite dimensional space of measures. 
%
It is however a priori unclear how to solve this infinite dimensional problem even if $\al$ and $\be$ are discrete measure. 
%
Section~\ref{subsec:multi-marginal} shows that barycenters $\alpha$ can be computed by solving a linear program, which a multi-marginal OT problem. This implies in particular that the barycenter of $S$ discrete measure with $n$ Diracs has at most $n^s$ Diracs. An issue is that the numerical complexity of solving this problem grows exponentially with $S$. We will show later that if one forces a priori the support of $\alpha$, one can use a fast Sinkhorn's algorithm.

\begin{proposition}
	$(\al,\be) \mapsto \MK_{\c}(\al,\be)$ is convex.
\end{proposition}	


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{1-D Case}

The quantile function of the barycenter is the barycenter of the quantile function. 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Gaussians Case}

One can show that the barycenter of Gaussian is Gaussian. In 1-D, it is obtained by averaging the mean and the covariance. In higher dimension, the covariance are solution of a non-trivial fixed point. They can also be solved by minimizing a sum of Bures distances, for which one can show the convergence of a Bures' gradient flow.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Sinkhorn for barycenters}

A key difference of problem~\eqref{eq-barycenter-generic} with respect to classical Kantorovitch formulation is that there is no simple canonical measure which can be used to localized the support of the solution (for classical OT one can use the product measure $\al \otimes \be$). 
%
Indeed, from the multimarginal formulation of the barycenter, all we know a priori is that the support of barycenters are contrained in all possible barycenters of set of $S$ point extracted from all the support. If they all have $n$ points, this makes $n^S$ possibility, which is prohebitive.

So in order to reduce this complexity and develop a Sinkhorn solver, one need to fix a priori some support for the solution, with usually a much smaller size, which introduces some discretization error because this support is likely not to be the support of the target measure.

Once this reference measure is fixed, can use entropic smoothing and approximate the solution of~\eqref{eq-wass-discr} using 
\eql{\label{eq-entropic-bary}
	\umin{\a \in \simplex_n} \sum_{s=1}^S \la_s \MKD_{\C_s}^\epsilon(\a,\b_s)
}
for some $\epsilon>0$. 
%
This is a smooth convex minimization problem, which can be tackled using gradient descent~\cite{CuturiBarycenter}. An alternative is to use descent method (typically quasi-Newton) on the semi-dual~\cite{2016-Cuturi-siims}, which is useful to integrate additional regularizations on the barycenter (e.g. to impose some smoothness).
%
A simple but effective approach, as remarked in~\cite{2015-benamou-cisc} is to rewrite~\eqref{eq-entropic-bary} as a (weighted) KL projection problem
\eql{\label{eq-bary-entropy-couplings}
	\umin{ (\P_s)_s } \enscond{ \sum_{s} \la_s \KLD( \P_s|\K_s ) }{
		\foralls s, \transp{\P_s}\ones_m = \b_s, \:
		\P_1\ones_1 =  \ldots = \P_S\ones_S
	}
}
where we denoted $\K_s \eqdef e^{-\C_s/\epsilon}$. Here, the barycenter $\a$ is implicitly encoded in the row marginals of all the couplings $\P_s \in \RR^{n \times n_s}$ as $\a = \P_1\ones_1 =  \ldots = \P_S\ones_S$.
%
As detailed in~\cite{2015-benamou-cisc}, one can generalize Sinkhorn to this problem, which also corresponds to iterative projection. This can also be seen as a special case of the generalized Sinkhorn detailed in Section~\ref{sec-generalized}.
%
The optimal couplings $(\P_s)_s$ solving~\eqref{eq-bary-entropy-couplings} are computed in scaling form as 
\eql{\label{eq-bary-opt}
	\P_s=\diag(\uD_s)\K\diag(\vD_s), 
}
and the scalings are sequentially updated as
\begin{align}\label{eq-sinkhorn-bary}
	\foralls s \in \range{1,S}, \quad \itt{\vD}_s &\eqdef \frac{\b_s}{\transp{\K}_s \it{\uD}_s}, \\
	\foralls s \in \range{1,S}, \quad  \itt{\uD}_s &\eqdef \frac{\itt{\a}}{\K_s \itt{\vD}_s}, \label{eq-sinkhorn-bary-2}\\
		\qwhereq
		\itt{\a} &\eqdef \prod_s (  \K_s \itt{\vD}_s )^{\la_s}. \label{eq-sinkhorn-bary-3}
\end{align}
An alternative way to derive these iterations is to perform alternate minimization on the variables of a dual problem, which detailed in the following proposition.

\begin{prop}
	The optimal $(\uD_s,\vD_s)$ appearing in~\eqref{eq-bary-opt} can be written as $(\uD_s,\vD_s) = ( e^{\fD_s/\epsilon},e^{\gD_s/\epsilon} )$ where $( \fD_s,\gD_s )_s$ are the solutions of the following program (whose value matches the one of \eqref{eq-entropic-bary})
	\eql{\label{eq-dual-bary-entropy}
		\umax{ ( \fD_s,\gD_s )_s } \enscond{
		\sum_s \la_s \pa{
			\dotp{\gD_s}{\b_s} - \epsilon \dotp{\K_s e^{\gD_s/\epsilon}}{e^{\fD_s/\epsilon}}
		}
		}{
			\sum_s \la_s \fD_s = 0
		}.
	}
\end{prop}

\begin{proof}
	Introducing Lagrange multipliers in~\eqref{eq-bary-entropy-couplings} leads to
	\begin{align*}
		\umin{ (\P_s)_s, \a } 
		\umax{ ( \fD_s,\gD_s )_s }
		\sum_{s} \la_s \Big(
			\epsilon \KLD( \P_s|\K_s )
			+ 
			\dotp{\a - \P_s\ones_m}{ \fD_s } \\
		\qquad\qquad	+			
			\dotp{\b_s - \transp{\P_s}\ones_m}{ \gD_s }
		\Big).
	\end{align*}
	Strong duality holds, so that one can exchange the min and the max, and gets
		\begin{align*}
		&\umax{ ( \fD_s,\gD_s )_s }
			\sum_s \la_s \pa{
				 \dotp{\gD_s}{\b_s}
		+ 
		\umin{\P_s} \epsilon \KLD( \P_s|\K_s ) - \dotp{\P_s}{\fD_s \oplus \gD_s }
		} \\
	&\qquad\qquad	+ \umin{\a}
			\dotp{ \sum_{s} \la_s \fD_s }{\a}.
	\end{align*}
	The explicit minimization on $\a$ gives the constraint $\sum_{s} \la_s \fD_s=0$ together with 
	\begin{align*}
		\umax{ ( \fD_s,\gD_s )_s }
			\sum_s \la_s 
				 \dotp{\gD_s}{\b_s}
		-
		\epsilon \KLD^*\pa{ \frac{\fD_s \oplus \gD_s}{\epsilon}|\K_s }
	\end{align*}
	where $\KLD^*(\cdot|\K_s)$ is the Legendre transform~\eqref{eq-legendre} of the function $\KLD^*(\cdot|\K_s)$.
	%
	This Legendre transform reads
	\eql{\label{eq-legendre-kl}
		\KLD^*(\VectMode{U}|\K) = \sum_{i,j} \K_{i,j} (e^{\VectMode{U}_{i,j}}-1), 
	}
	which shows the desired formula. 
	%
	To show~\eqref{eq-legendre-kl}, since this function is separable, one needs to compute
	\eq{
		\foralls (u,k) \in \RR_+^2, \quad
		\KLD^*(u|k) \eqdef \umax{r} ur -   \pa{ r \log(r/k) - r + k  }
	}
	whose optimality condition reads $u=\log(r/k)$, i.e. $r = k e^{u}$, hence the result.
\end{proof}

Minimizing~\eqref{eq-dual-bary-entropy} with respect to each $\gD_s$, while keeping all the other variable fixed, is obtained in closed form by~\eqref{eq-sinkhorn-bary}. 
%
Minimizing~\eqref{eq-dual-bary-entropy} with respect to all the $(\fD_s)_s$ requires to solve for $\a$ using~\eqref{eq-sinkhorn-bary-3} and leads to the expression~\eqref{eq-sinkhorn-bary-2}.

Figures~\ref{fig-barycenters-images} and~\ref{fig-bary-shapes} show applications to 2-D and 3-D shapes interpolation. Figure~\ref{fig-barycenters-surfaces} shows a computation of barycenters on a surface, where the ground cost is the square of the geodesic distance. For this figure, the computations are performed using the geodesic in heat approximation detailed in Remark~\ref{rem-geod-heat}. We refer to~\cite{2015-solomon-siggraph} for more details and other applications to computer graphics and imaging sciences.




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multimarginal OT}
\label{subsec:multi-marginal}

\paragraph{Definition, example and properties.}


\paragraph{Multi-marginal formulation of barycenters.}


\paragraph{Entropic regularization of multi-marginal.} 



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sliced Wasserstein}

Definition. 

Link with Radon transform.

H\"older comparison with Wasserstein.

Wasserstein (particules) flow of SW and convergence to the target.

Variation : max sliced Wasserstein, sub-space SW. 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Gromov Wasserstein}

%%%%%%%%%%%%%%%%%%%%%
\paragraph{Discrete setting} 


Optimal transport needs a ground cost $\C$ to compare histograms $(\a,\b)$, it can thus not be used if the histograms are not defined on the same underlying space, or if one cannot pre-register these spaces to define a ground cost. 
%
To address this issue, one can instead only assume a weaker assumption, namely that one has at its disposal two matrices $\distD \in \RR^{n \times n}$ and $\distD' \in \RR^{m \times m}$ that represent some relationship between the points on which the histograms are defined. A typical scenario is when these matrices are (power of) distance matrices.
%
The Gromov-Wasserstein problem reads
\eql{\label{eq-gw-def}
	\GWD( (\a,\distD), (\b,\distD') )^p \eqdef
	\umin{ \P \in \CouplingsD(\a,\b) } 
		\Ee_{\distD,\distD'}(\P) \eqdef 
		\sum_{i,j,i',j'} \De(\distD_{i,i'},\distD'_{j,j'})^q \P_{i,j}\P_{i',j'}, 
}
where $p \geq 1$ and $\De$ is a distance on $\RR$, typically $\De(u,v)=|u-v|$.
%
This is a non-convex problem, which can be recast as a Quadratic Assignment Problem (QAP)~\cite{loiola-2007} and is in full generality NP-hard to solve for arbitrary inputs. 
%
It is in fact equivalent to a graph matching problem~\cite{lyzinski-2015} for a particular cost.

One can show that $\GWD$ satisfies the triangular inequality, and in fact it defines a distance between metric spaces equipped with a probability distribution (here assumed to be discrete in definition~\eqref{eq-gw-def}) up to isometries preserving the measures, as detailed below.
%
This distance was introduced and studied in details by Memoli in~\cite{memoli-2011}. An in-depth mathematical exposition (in particular, its geodesic structure and gradient flows) is given in~\cite{SturmGW}. See also~\cite{schmitzer2013modelling} for applications in computer vision.
%
This distance is also tightly connected with the Gromov-Hausdorff distance~\cite{gromov-2001} between metric spaces, which have been used for shape matching~\cite{memoli-2007,bronstein-2010}. 


%%%%%%%%%%%%%%%%%%%%%
\paragraph{General setting} 

The general setting corresponds to computing couplings between metric measure spaces $\XX=(\X,\dist_\X,\al)$
	and $\YY=(\Y,\dist_\Y,\be)$ where $(\dist_\X,\dist_\Y)$ are distances and $(\al,\be)$ are measures on their respective spaces.
	%
	One needs to restrict the space to be complete and separable in order to be able to apply the gluing lemma and construct isometries below.
	%
	One defines 
	\begin{align}
		\label{eq-gw-generic}
		\GW( \XX,\YY )^p \eqdef 
		\umin{ \pi \in \CouplingsD(\al,\be) } 
		\int_{\X^2 \times \Y^2}
		 \De(\dist_\X(x,x'),\dist_\Y(y,y'))^p
		\d\pi(x,y)\d\pi(x',y').
	\end{align}
 
In the following, one says that $\XX$ and $\YY$ are isometric mm-spaces if there exists $\phi : \X \rightarrow \Y$ such that $\phi_{\sharp}\al=\be$ and which is a bijective isometry between the supports of $\al$ and $\be$, i.e. $\dist_\Y(\phi(x),\phi(x'))=\dist_\X(x,x')$ on the support of $\al$.
 

\begin{thm}
	$\GW$ defines a distance between metric measure spaces up to isometries
\end{thm}
\begin{proof}
	If $\GW( \XX,\YY )=0$ then if $\pi$ is an optimal plan, we have $\pi \otimes \pi$-almost everywhere that $d_\Xx(x,x')=d_\Yy(y,y')$.
	%
	We show that $\XX$ and $\YY$ are isometric by showing that they are both isometric to a third space $\ZZ \eqdef (\Zz \eqdef \X \times \Y,\dist_\Zz,\pi)$
	where 
	\eql{
		\dist_\Zz((x,y),(x',y')) \eqdef \frac{1}{2} \dist_\X(x,x') + \frac{1}{2} \dist_\Y(y,y').
	}
	We now show that $\psi : (x,y) \in \Zz \mapsto x \in \Xx$ is a bijective isometry between $\supp(\pi)$ and $\supp(\al)$.
	% 
	One has, for $(x,y) \in \supp(\pi)$, $d_\Xx(x,x')=d_\Yy(y,y')$ and thus
	\eq{
		\dist_\Xx(\psi(x,y),\psi(x',y')) = \dist_\Xx(x,x') = d_\Yy(y,y') = d_\Zz((x,y),(x',y')), 
	}
	thus $\psi$ is an isometry, hence injective. We now show surjectivity, i.e. for $x \in \supp(\al)$, we need to construct $y$ such that $(x,y) \in \supp(\pi)=\Zz$. Indeed, since $\psi_\sharp \pi = \al$, there exists a sequence $(x_k,y_k) \in \supp(\pi)$ such that $x_k \rightarrow y_k$.
	%
	Since $d_\Xx(x_k,x_{k'})=d_\Yy(y_k,y_{k'})$ the sequence $(y_k)_k$ is Cauchy, and since $\supp(\pi)$ is closed, it is complete and thus $(x_k,y_k)$ is converging to some $(x,y) \in \Zz$. 
		
	Given an optimal $\pi$ and $\xi$ between three spaces $(\Xx,\Yy,\Zz)$ with measures $(\al,\be,\ga)$, we use the gluing Lemma~\ref{lem-gluing-general} to define $\si$ with marginals $(P_{\Xx,\Yy})_\sharp \si = \pi$ and $(P_{\Yy,\Zz})_\sharp \si = \xi$. 
	We denote $\rho \eqdef (P_{\Xx,\Zz})_\sharp \si$ the composition of the two coupling, so that
	$$\begin{aligned}
		\GW( \XX,\ZZ ) & \leq \Big( \int_{\Xx \times \Zz} \De(\dist_\Xx(x,x'),\dist_\Zz(z,z'))^p \d\rho(x,z)\d\rho(x',z') \Big)^{1/p}  
			\\
		 &\leq \Big( \int_{\Xx \times \Yy \times \Zz} ( \De(\dist_\Xx(x,x'),\dist_\Yy(y,y')) + \De(\dist_\Yy(y,y'),\dist_\Zz(z,z')) )^p \d\si(x,y,z) \d\si(x',y',z')\Big)^{1/p} \\
		 &\leq \Big( \int_{\Xx \times \Yy \times \Zz} \De(\dist_\Xx(x,x'),\dist_\Yy(y,y'))^p \d\si(x,y,z) \d\si(x',y',z')\Big)^{1/p} \\
		 &\qquad + 
		 \Big( \int_{\Xx \times \Yy \times \Zz} \De(\dist_\Yy(y,y'),\dist_\Zz(z,z'))^p \d\si(x,y,z) \d\si(x',y',z')\Big)^{1/p}
		  \\
		  &= \GW( \XX,\YY ) + \GW( \YY,\ZZ ), 
	\end{aligned}$$
	where the first inequality is the sub-optimality of $\rho$, the second is the triangular inequality of $\De$, 
	and the third is the Minkowski inequality.
\end{proof}


%%%%%%%%%%%%%%%%%%%%%
\begin{rem}[Gromov-Wasserstein geodesics]
The space of metric spaces (up to isometries) endowed with this $\GW$ distance~\eqref{eq-gw-generic} has a geodesic structure. \cite{SturmGW} shows that the geodesic between  $(\X_0,\dist_{\X_0},\al_0)$ and $(\X_1,\dist_{\X_1},\al_1)$ can be chosen to be 
$t \in [0,1] \mapsto (\X_0 \times \X_1,\dist_t,\pi^\star)$ where $\pi^\star$ is a solution of~\eqref{eq-gw-generic} and for all $((x_0,x_1), (x_0',x_1')) \in (\X_0 \times \X_1)^2$, 
\eq{
	\dist_t((x_0,x_1), (x_0',x_1')) \eqdef
	(1-t)\dist_{\X_0}(x_0,x_0') + t\dist_{\X_1}(x_1,x_1').
}
This formula allows one to define and analyze gradient flows which minimize functionals involving metric spaces, see~\cite{SturmGW}. It is however difficult to handle numerically, because it involves computations over the product space $\X_0 \times \X_1$. 
%
A heuristic approach is used in~\cite{peyre2016gromov} to define geodesics and barycenters of metric measure spaces while imposing the cardinality of the involved spaces and making use of the entropic smoothing~\eqref{eq-gw-entropy} detailed below.
\end{rem}
%%%%%%%%%%%%%%%%%%%%%

\paragraph{Entropic regularization and iterative solver}

We consider the case $p=1$, $\De(u,v)=|u-v|$.
%
To approximate the computation of $\GWD$, and to help convergence of minimization schemes to better minima, one can consider the entropic regularized variant
\eql{\label{eq-gw-entropy}
	\umin{ \P \in \CouplingsD(\a,\b) } 
		\Ee_{\distD,\distD'}(\P) - \epsilon \HD(\P).
}
As proposed initially in~\cite{gold-1996,rangarajan-1999}, and later revisited in~\cite{2016-solomon-gw} for applications in graphics, one can use iteratively Sinkhorn's algorithm to progressively compute a stationary point of~\eqref{eq-gw-entropy}. 
%
Indeed, successive linearizations of the objective function lead to consider the succession of updates
\eql{\label{eq-gw-sinkh}
	\itt{\P} \eqdef \umin{ \P \in \CouplingsD(\a,\b) } \dotp{\P}{\it{\C}} - \epsilon\H(\P)
		\qwhereq
}
\eq{
		\it{\C} \eqdef \nabla \Ee_{\distD,\distD'}(\it{\P}) = -\transp{\distD'} \it{\P} \distD, 
}
which can be interpreted as a mirror-descent scheme~\cite{2016-solomon-gw}. Each update can thus be solved using Sinkhorn iterations~\eqref{eq-sinkhorn} with cost $\it{\C}$. Figure~\eqref{fig-gw} illustrates the use of this entropic Gromov-Wasserstein to compute soft maps between domains. 



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Metric learning and inverse OT}

%%%
\paragraph{Metric learning and derivatives of OT}

OT is convex with respect to the measure and concave with respect to the cost.

Bilevel non-convex optimization problem.

%%%
\paragraph{Inverse Optimal Transport}

Convex optimization problem. 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\if 0
\subsection{Quantum OT}

Static formulation.

Gurvits algorithm, Q-sinkhorn.
\fi