% !TEX root = ../OT4ML.tex



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Barycenters}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fr\'echet Mean over the Wasserstein Space}


The Wasserstein barycenter problem was originally introduced by~\cite{Carlier_wasserstein_barycenter} following earlier ideas of~\cite{carlierekelandmatching}. They proved in particular uniqueness of the barycenter for $c(x,y)=\norm{x-y}^2$ over $\X=\RR^d$, if one of the input measures has a density with respect to the Lebesgue measure (and more generally under the usual hypotheses guaranteeing the existence of a Monge map).

Given a set of input measures $(\be_s)_s$ defined on some space $\X$, the barycenter problem becomes
	\eql{\label{eq-barycenter-generic}
		\umin{\al \in \Mm_+^1(\X)} \sum_{s=1}^S \la_s \MK_{\c}(\al,\be_s).
	}
	In the case where $\X=\RR^d$ and $c(x,y)=\norm{x-y}^2$,~\cite{Carlier_wasserstein_barycenter} shows that if one of the input measures has a density, then this barycenter is unique.
	%
	Problem~\eqref{eq-barycenter-generic} can be viewed as a generalization of the problem of computing barycenters of points $(x_s)_{s=1}^S \in \X^S$ to arbitrary measures. Indeed, if $\be_s=\de_{x_s}$ is a single Dirac mass, then a solution to~\eqref{eq-barycenter-generic} is $\de_{x^\star}$ where $x^\star$ is a Fr\'echet mean of the points $(x_s)_s$.
	%
	Note that for $c(x,y)=\norm{x-y}^2$, the mean of the barycenter $\al^\star$ is necessarily the barycenter of the mean, \emph{i.e.}
	\eq{
		\int_\Xx x \d\al^\star(x) = \sum_s \la_s \int_\Xx x \d\be_s(x),
	}
	and the support of $\al^\star$ is located in the convex hull of the supports of the $(\be_s)_s$.
	%
	The consistency of the approximation of the infinite dimensional optimization~\eqref{eq-barycenter-generic} when approximating the input distribution using discrete ones (and thus solving~\eqref{eq-wass-discr} in place) is studied in~\cite{Carlier-NumericsBarycenters}.
	%
	One can also recast~\eqref{eq-barycenter-generic} as a multi-marginal OT problem; see Remark~\ref{eq-multimarg-bary}.


	Existence follows from compactness and lower semicontinuity under the standard assumptions used for the two-marginal problem. Dual formulations and Monge-type characterizations can also be derived, but their structure is more involved because the barycenter is itself an unknown measure.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{1-D Case}


	In one dimension, the barycenter has an explicit expression in terms of quantile functions. If $\c(x,y)=|x-y|^p$, then
	\eq{
		\cumul{\al^\star}^{-1}(t)
		\in
			\uargmin{x \in \RR} \sum_{s=1}^S \la_s
		\abs{x-\cumul{\be_s}^{-1}(t)}^p,
		\qquad t \in [0,1].
	}
	For $p=2$, this reduces to the weighted average of the input quantiles.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Gaussians Case}

	For $c(x,y)=\norm{x-y}^2$, the barycenter of Gaussian measures is Gaussian. Its mean is the weighted average of the input means, while its covariance is characterized by a fixed-point equation involving the Bures metric on covariance matrices.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Discrete Barycenters}

	Given input histograms $\{\b_s\}_{s=1}^S$, where $\b_s \in \simplex_{n_s}$, and weights $\la \in \simplex_S$, a Wasserstein barycenter is computed by minimizing
\eql{\label{eq-wass-discr}
	\umin{\a \in \simplex_n} \sum_{s=1}^S \la_s \MKD_{\C_s}(\a,\b_s)
}
where the cost matrices $\C_s \in \RR^{n \times n_s}$ need to be specified.
%
A typical setup is ``Eulerian'', so that all the barycenters are defined on the same grid, $n_s=n$, $\C_s=\C=\distD^p$ is set to be a distance matrix, so that one solves
\eq{
	\umin{\a \in \simplex_n} \sum_{s=1}^S \la_s \WassD_p^p(\a,\b_s).
}


The barycenter problem for histograms~\eqref{eq-wass-discr} is in fact a linear program, since one can look for the $S$ couplings $(\P_s)_s$ between each input and the barycenter itself
\eq{
	\umin{\a \in \simplex_n, (\P_s \in \RR^{n \times n_s})_s } \enscond{
		\sum_{s=1}^S \la_s \dotp{\P_s}{\C_s}
	}{
			\foralls s, \P_s \ones_{n_s}=\a, \P_s^\top \ones_{n} = \b_s
	}.
}
	Although this problem is an LP, its scale forbids the use of generic solvers for medium-scale problems. One can therefore resort to first-order methods such as subgradient descent on the dual~\cite{Carlier-NumericsBarycenters}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sinkhorn for barycenters}

	A key difference with the regularized two-marginal OT problem is that there is no longer a canonical reference measure $\al \otimes \be$, since the barycenter $\al$ is unknown.

One can use entropic smoothing and approximate the solution of~\eqref{eq-wass-discr} using
\eql{\label{eq-entropic-bary}
	\umin{\a \in \simplex_n} \sum_{s=1}^S \la_s \MKD_{\C_s}^\epsilon(\a,\b_s)
}
for some $\epsilon>0$.
%
This is a smooth convex minimization problem, which can be tackled using gradient descent~\cite{CuturiBarycenter}. An alternative is to use a descent method, typically quasi-Newton, on the semi-dual~\cite{2016-Cuturi-siims}; this is useful when adding extra regularization on the barycenter, for instance to impose smoothness.
%
A simple but effective approach, as remarked in~\cite{2015-benamou-cisc} is to rewrite~\eqref{eq-entropic-bary} as a (weighted) KL projection problem
\eql{\label{eq-bary-entropy-couplings}
	\umin{ (\P_s)_s } \enscond{ \sum_{s} \la_s \KLD( \P_s|\K_s ) }{
		\foralls s, \transp{\P_s}\ones_n = \b_s, \:
			\P_1\ones_{n_1} = \ldots = \P_S\ones_{n_S}
	}
}
where we denoted $\K_s \eqdef e^{-\C_s/\epsilon}$. Here, the barycenter $\a$ is implicitly encoded in the row marginals of all the couplings $\P_s \in \RR^{n \times n_s}$ as $\a = \P_1\ones_{n_1} = \ldots = \P_S\ones_{n_S}$.
%
As detailed in~\cite{2015-benamou-cisc}, one can generalize Sinkhorn to this problem, which also corresponds to iterative projection.
%
The optimal couplings $(\P_s)_s$ solving~\eqref{eq-bary-entropy-couplings} are computed in scaling form as
\eql{\label{eq-bary-opt}
	\P_s=\diag(\uD_s)\K_s\diag(\vD_s),
}
and the scalings are sequentially updated as
\begin{align}\label{eq-sinkhorn-bary}
	\foralls s \in \range{1,S}, \quad \itt{\vD}_s &\eqdef \frac{\b_s}{\transp{\K_s} \it{\uD}_s}, \\
	\foralls s \in \range{1,S}, \quad \itt{\uD}_s &\eqdef \frac{\itt{\a}}{\K_s \itt{\vD}_s}, \label{eq-sinkhorn-bary-2}\\
		\qwhereq
		\itt{\a} &\eqdef \prod_s ( \K_s \itt{\vD}_s )^{\la_s}. \label{eq-sinkhorn-bary-3}
\end{align}
An alternative way to derive these iterations is to perform alternate minimization on the variables of a dual problem, as detailed in the following proposition.

\begin{prop}[Dual of entropic barycenters]
	The optimal $(\uD_s,\vD_s)$ appearing in~\eqref{eq-bary-opt} can be written as $(\uD_s,\vD_s) = ( e^{\fD_s/\epsilon},e^{\gD_s/\epsilon} )$ where $( \fD_s,\gD_s )_s$ are the solutions of the following program (whose value matches the one of \eqref{eq-entropic-bary})
	\eql{\label{eq-dual-bary-entropy}
		\umax{ ( \fD_s,\gD_s )_s } \enscond{
		\sum_s \la_s \pa{
			\dotp{\gD_s}{\b_s} - \epsilon \dotp{\K_s e^{\gD_s/\epsilon}}{e^{\fD_s/\epsilon}}
		}
		}{
			\sum_s \la_s \fD_s = 0
		}.
	}
\end{prop}

\begin{proof}
	Introducing Lagrange multipliers in~\eqref{eq-bary-entropy-couplings} leads to
	\begin{align*}
		\umin{ (\P_s)_s, \a }
		\umax{ ( \fD_s,\gD_s )_s }
		\sum_{s} \la_s \Big(
			\epsilon \KLD( \P_s|\K_s )
			+
			\dotp{\a - \P_s\ones_{n_s}}{ \fD_s } \\
		\qquad\qquad	+
			\dotp{\b_s - \transp{\P_s}\ones_n}{ \gD_s }
		\Big).
	\end{align*}
	Strong duality holds, so one can exchange the minimum and the maximum and obtain
		\begin{align*}
		&\umax{ ( \fD_s,\gD_s )_s }
			\sum_s \la_s \pa{
				 \dotp{\gD_s}{\b_s}
		+
		\umin{\P_s} \epsilon \KLD( \P_s|\K_s ) - \dotp{\P_s}{\fD_s \oplus \gD_s }
		} \\
	&\qquad\qquad	+ \umin{\a}
			\dotp{ \sum_{s} \la_s \fD_s }{\a}.
	\end{align*}
	The explicit minimization on $\a$ gives the constraint $\sum_{s} \la_s \fD_s=0$ together with
	\begin{align*}
		\umax{ ( \fD_s,\gD_s )_s }
			\sum_s \la_s
				 \dotp{\gD_s}{\b_s}
		-
		\epsilon \KLD^*\pa{ \frac{\fD_s \oplus \gD_s}{\epsilon}|\K_s }
	\end{align*}
		where $\KLD^*(\cdot|\K_s)$ is the Legendre transform~\eqref{eq-legendre} of the function $\KLD(\cdot|\K_s)$.
	%
	This Legendre transform reads
	\eql{\label{eq-legendre-kl}
		\KLD^*(\VectMode{U}|\K) = \sum_{i,j} \K_{i,j} (e^{\VectMode{U}_{i,j}}-1),
	}
	which shows the desired formula.
	%
	To show~\eqref{eq-legendre-kl}, since this function is separable, one needs to compute
		\eq{
			\foralls (u,k) \in \RR \times \RR_+^*, \quad
			\KLD^*(u|k) \eqdef \umax{r \geq 0} ur - \pa{ r \log(r/k) - r + k }
		}
		whose optimality condition reads $u=\log(r/k)$, i.e. $r = k e^{u}$, hence the result. The case $k=0$ follows by lower semicontinuity and gives zero.
	\end{proof}

Minimizing~\eqref{eq-dual-bary-entropy} with respect to each $\gD_s$, while keeping all the other variable fixed, is obtained in closed form by~\eqref{eq-sinkhorn-bary}.
%
Minimizing~\eqref{eq-dual-bary-entropy} with respect to all the $(\fD_s)_s$ requires to solve for $\a$ using~\eqref{eq-sinkhorn-bary-3} and leads to the expression~\eqref{eq-sinkhorn-bary-2}.

Classical applications include 2-D image interpolation, 3-D shape interpolation, and barycenters on surfaces where the ground cost is the square of the geodesic distance. We refer to~\cite{2015-solomon-siggraph} for more details and other applications to computer graphics and imaging sciences.
