\section{Beyond Comparing Measures}
\label{sec-beyond-comparing-measures}
\subsection{Vector and Matrix-Valued Measures}
\label{sec-vector-matrix-valued-measures}
\paragraph{Positive vector-valued measures.}
\begin{defn}[Positive vector-valued measure]\label{def-positive-vector-valued-measure}
A positive $\RR_+^m$-valued measure on $\X$ is a tuple
\(\mu=(\mu^1,\ldots,\mu^m)\in\Mm_+(\X;\RR_+^m),\)
where each component $\mu^k$ is a nonnegative finite measure.
\end{defn}
To keep the notation readable, first assume that the endpoints and the curve have densities. The direct analogue of Benamou--Brenier fixes a vector density $u_t(x)\in\RR_+^m$ and a spatial flux $V_t(x)=(V_{t,1},\ldots,V_{t,d})\in(\RR^m)^d$, where $V_{t,\ell}^k$ is the momentum of channel $k$ in spatial direction $\ell$.

\begin{equation}\label{eq-vector-valued-bb}
\mathcal W_{\Phi}^2(\mu_0,\mu_1)
\eqdef
\inf_{u,V}
\int_0^1\!\int_\X
\Phi(u_t(x),V_t(x))\,\d x\,\d t
\end{equation}
\begin{equation}\label{eq-vector-valued-continuity}
\partial_t u_t+\nabla_x\cdot V_t=0,
\qquad
(\nabla_x\cdot V_t)^k=\sum_{\ell=1}^d \partial_{x_\ell} V_{t,\ell}^k.
\end{equation}
A simple quadratic family is obtained from a mobility matrix $\mathsf M(u)\in\mathbb S_+^m$, where $\mathbb S_+^m$ denotes the cone of real symmetric positive semidefinite matrices:

\(\Phi_{\mathsf M}(u,V) = \sum_{\ell=1}^d V_{\ell}^{\top}\mathsf M(u)^\dagger V_{\ell},\)
\(\mathsf M_{\mathrm{diag}}(u)=\diag(u_1,\ldots,u_m),\)
\(\mathsf M_\kappa(u)=\diag(u)+\kappa\Big(\sum_{k=1}^m u_k\Big) q q^\top\)
\begin{prop}[Diagonal positive vector Benamou--Brenier]\label{prop-diagonal-positive-vector-bb}
Assume that $\mu_0^k,\mu_1^k\in\Mm_+(\X)$ have the same mass $m_k$ for every $k$. For the diagonal mobility $\mathsf M_{\mathrm{diag}}$, the value of~\eqref{eq-vector-valued-bb} is
\[
\mathcal W_{\mathrm{diag}}^2(\mu_0,\mu_1)
=
\sum_{k:m_k>0} m_k\,\Wass_2^2\!\left(\frac{\mu_0^k}{m_k},\frac{\mu_1^k}{m_k}\right),
\]
with the convention that zero-mass channels contribute zero.
\end{prop}
\begin{proof}
For $\mathsf M_{\mathrm{diag}}(u)=\diag(u_1,\ldots,u_m)$, the action separates as \(\sum_{\ell=1}^d V_\ell^\top \mathsf M_{\mathrm{diag}}(u)^\dagger V_\ell = \sum_{k=1}^m \frac{|V^k|^2}{u^k},\)
where $V^k=(V_1^k,\ldots,V_d^k)$ is the spatial momentum of channel $k$, and the scalar perspective convention is used. The constraint~\eqref{eq-vector-valued-continuity} also separates into $\partial_t u^k+\nabla\cdot V^k=0$. The minimization therefore splits into $m$ independent scalar Benamou--Brenier problems. If $m_k=0$, nonnegativity and conservation force the whole channel to vanish. If $m_k>0$, normalizing $\rho_t^k=u_t^k/m_k$ and $p_t^k=V_t^k/m_k$ factors the channel action as $m_k\int |p_t^k|^2/\rho_t^k$, hence the scalar value is $m_k\Wass_2^2(\mu_0^k/m_k,\mu_1^k/m_k)$. Summing over the channels proves the claim.
\end{proof}
\paragraph{Positive matrix-valued measures.}
\begin{defn}[Positive matrix-valued measure]\label{def-positive-matrix-valued-measure}
Write $\mathbb S^m$ for real symmetric matrices and $\mathbb S_+^m$ for the positive semidefinite cone. A positive $\mathbb S_+^m$-valued measure is an element
\(\mathcal A\in\Mm_+(\X;\mathbb S_+^m).\)
\end{defn}
If $\mathcal A$ has density $A(x)\succeq0$, then $\tr A(x)$ is the scalar amount of mass at $x$, while, wherever $\tr A(x)>0$, the normalized matrix $A(x)/\tr A(x)$ records an internal covariance or orientation. This is the matrix analogue of the positive vector case: diagonal matrices encode nonnegative vector components, and non-diagonal matrices add a local eigenbasis.

\begin{equation}\label{eq-matrix-valued-bb}
\begin{aligned}
\mathcal W_{\mathrm{mat}}^2(\mathcal A_0,\mathcal A_1)
\eqdef
\inf_{A,P}\int_0^1\!\int_\X
\sum_{\ell=1}^d \tr\big(P_{t,\ell}^{\top} A_t^\dagger P_{t,\ell}\big)\d x\d t
\end{aligned}
\end{equation}
\begin{equation}\label{eq-matrix-valued-continuity}
\partial_t A_t+\nabla_x\cdot P_t=0,
\qquad
\nabla_x\cdot P_t=\sum_{\ell=1}^d \partial_{x_\ell}P_{t,\ell}.
\end{equation}
\begin{prop}[Diagonal matrix subproblem]\label{prop-matrix-diagonal-reduction}
Assume that the endpoints are diagonal in a fixed orthonormal basis, \(\mathcal A_i=\diag(\mu_i^1,\ldots,\mu_i^m), \qquad i=0,1,\)
and that $\mu_0^k(\X)=\mu_1^k(\X)=m_k$ for every $k$. If one restricts the admissible curves in~\eqref{eq-matrix-valued-bb} to remain diagonal in that basis,
\(A_t=\diag(u_t^1,\ldots,u_t^m), \qquad P_{t,\ell}=\diag(V_{t,\ell}^1,\ldots,V_{t,\ell}^m),\)
then the value of this restricted matrix problem is
\[
\sum_{k:m_k>0} m_k\,\Wass_2^2\!\left(\frac{\mu_0^k}{m_k},\frac{\mu_1^k}{m_k}\right),
\]
with zero contribution from zero-mass channels. Thus the commuting matrix submodel is exactly the diagonal positive vector-valued Benamou--Brenier model of Proposition~\ref{prop-diagonal-positive-vector-bb}.
\end{prop}
\begin{proof}
The continuity equation~\eqref{eq-matrix-valued-continuity} is diagonal entry by diagonal entry and gives $\partial_t u^k+\nabla\cdot V^k=0$. Moreover,
\(\sum_{\ell=1}^d \tr\big(P_{t,\ell}^{\top}A_t^\dagger P_{t,\ell}\big) = \sum_{k=1}^m \frac{|V_t^k|^2}{u_t^k}\)
with the same scalar perspective convention as before. The admissible set and the action are therefore exactly those of the diagonal vector model.
\end{proof}
\subsection{Gromov--Wasserstein}
\label{sec-gromov-wasserstein}
\paragraph{Discrete formulation.}
Optimal transport needs a ground cost $\C$ to compare histograms $(\a,\b)$, and thus cannot be used directly if the histograms are not defined on the same underlying space, or if one cannot pre-register these spaces to define a ground cost.

To address this issue, one can instead use a weaker requirement: two matrices $\distD \in \RR^{n \times n}$ and $\distD' \in \RR^{m \times m}$ are available and represent relationships between the points on which the histograms are defined. A typical scenario is when these matrices are powers of distance matrices.

\eql{\label{eq-gw-def}
\GWD( (\a,\distD), (\b,\distD') )^p \eqdef
\umin{ \P \in \CouplingsD(\a,\b) }
\Ee_{\distD,\distD'}(\P) \eqdef
\sum_{i,j,i',j'} \De(\distD_{i,i'},\distD'_{j,j'})^p \P_{i,j}\P_{i',j'},
}
where $p \geq 1$ and $\De$ is a distance on $\RR$, typically $\De(u,v)=|u-v|$.

When the matrices $\distD,\distD'$ are genuine distance matrices, the general construction below shows that $\GWD$ satisfies the triangle inequality and defines a distance between metric spaces equipped with a probability distribution, up to measure-preserving isometries.

\paragraph{General setting.}
\begin{defn}[Metric-measure space]\label{def-metric-measure-space}
A metric-measure space is a triple
\(\XX=(\X,\dist_\X,\al),\)
where $(\X,\dist_\X)$ is a metric space and $\al$ is a probability measure on $\X$.
\end{defn}
The general setting corresponds to computing couplings between metric-measure spaces $\XX=(\X,\dist_\X,\al)$ and $\YY=(\Y,\dist_\Y,\be)$, where the distance and the measure are both part of the data.

\begin{align}
\label{eq-gw-generic}
\GW( \XX,\YY )^p \eqdef
\umin{ \pi \in \Couplings(\al,\be) }
\int_{\X^2 \times \Y^2}
\De(\dist_\X(x,x'),\dist_\Y(y,y'))^p
\d\pi(x,y)\d\pi(x',y').
\end{align}
\begin{prop}[Euclidean GW is controlled by Wasserstein]\label{prop-gw-controlled-by-wasserstein}
Let $\alpha,\beta$ be probability measures on $\RR^d$, equipped with the Euclidean distance, and take $\De(u,v)=|u-v|$ in~\eqref{eq-gw-generic}. Then
\(\GW((\RR^d,\norm{\cdot},\alpha),(\RR^d,\norm{\cdot},\beta)) \leq 2\Wass_p(\alpha,\beta).\)
\end{prop}
\begin{proof}
Let $\pi$ be any coupling between $\alpha$ and $\beta$. For two independent pairs $(X,Y),(X',Y')\sim\pi$, the reverse triangle inequality gives \(\big|\norm{X-X'}-\norm{Y-Y'}\big| \leq \norm{X-Y}+\norm{X'-Y'}.\)
Taking the $L^p$ norm and using Minkowski gives a bound by $2(\int\norm{x-y}^p\d\pi)^{1/p}$. Optimizing over $\pi$ proves the claim.
\end{proof}
\begin{defn}[Isometric metric-measure spaces]\label{def-isometric-mm-spaces}
Two metric-measure spaces $\XX=(\X,\dist_\X,\al)$ and $\YY=(\Y,\dist_\Y,\be)$ are isometric if there exists a measurable map $\phi:\supp(\al)\to\supp(\be)$ such that $\phi_\sharp\al=\be$, $\phi(\supp(\al))=\supp(\be)$, and
\(\dist_\Y(\phi(x),\phi(x'))=\dist_\X(x,x')\)
for all $x,x'\in\supp(\al)$.
\end{defn}
\begin{thm}[Gromov--Wasserstein metric modulo isometries]\label{thm-gw-metric}
For compact metric-measure spaces, $p\geq1$ and $\De(u,v)=|u-v|$, $\GW$ defines a distance up to measure-preserving isometries.
\end{thm}
\begin{proof}
If $\GW(\XX,\YY)=0$ and $\pi$ is an optimal plan, then $\dist_\X(x,x')=\dist_\Y(y,y')$ holds $\pi\otimes\pi$-almost everywhere. By continuity, this equality holds on $\supp(\pi)^2$.
\(\dist_\pi((x,y),(x',y')) \eqdef \frac12\dist_\X(x,x')+\frac12\dist_\Y(y,y').\)
The first projection $\psi:(x,y)\mapsto x$ is measure-preserving. For $((x,y),(x',y'))\in\supp(\pi)^2$, \(\dist_\X(\psi(x,y),\psi(x',y')) = \dist_\X(x,x') = \dist_\Y(y,y') = \dist_\pi((x,y),(x',y')),\)
so $\psi$ is an isometry and therefore injective. To see surjectivity onto $\supp(\al)$, take $x\in\supp(\al)$. Since $\psi_\sharp\pi=\al$, there is a sequence $(x_k,y_k)\in\supp(\pi)$ with $x_k\to x$. The equality of distances on $\supp(\pi)$ makes $(y_k)_k$ Cauchy, and compactness gives a convergent subsequence with limit $(x,y)\in\supp(\pi)$. The same argument for the second projection shows that the support space is also isometric to $\YY$.

For the triangle inequality, let $\pi$ be an optimal coupling between $\XX$ and $\YY$, and $\xi$ an optimal coupling between $\YY$ and $\ZZ=(\Z,\dist_\Z,\ga)$. By the gluing lemma, take $\sigma$ on $\X\times\Y\times\Z$ whose $(\X,\Y)$ and $(\Y,\Z)$ marginals are $\pi$ and $\xi$. Let $\rho=(P_{\X,\Z})_\sharp\sigma$, and write $\bar\sigma=\sigma\otimes\sigma$ for the product law of two independent triples $(x,y,z)$ and $(x',y',z')$. Then $\rho$ is feasible between $\XX$ and $\ZZ$, and
\begin{align*}
\GW(\XX,\ZZ)
&\leq
\left(
\int
\abs{\dist_\X(x,x')-\dist_\Z(z,z')}^p
\d\bar\sigma
\right)^{1/p} \\
&\leq
\left(
\int
\abs{\dist_\X(x,x')-\dist_\Y(y,y')}^p
\d\bar\sigma
\right)^{1/p}
+
\left(
\int
\abs{\dist_\Y(y,y')-\dist_\Z(z,z')}^p
\d\bar\sigma
\right)^{1/p} \\
&=
\GW(\XX,\YY)+\GW(\YY,\ZZ),
\end{align*}
where the second inequality uses the pointwise triangle inequality followed by Minkowski's inequality. Symmetry and non-negativity are immediate.
\end{proof}
\begin{prop}[Gromov--Wasserstein geodesics]\label{prop-gw-geodesics}
Let $\XX_0=(\X_0,\dist_{\X_0},\al_0)$ and $\XX_1=(\X_1,\dist_{\X_1},\al_1)$ be compact metric-measure spaces, take $\De(u,v)=|u-v|$ in~\eqref{eq-gw-generic}, and let $\pi^\star$ be an optimal coupling. Define, on $\Z=\X_0\times\X_1$,
\(\dist_t((x_0,x_1),(x'_0,x'_1)) \eqdef (1-t)\dist_{\X_0}(x_0,x'_0)+t\dist_{\X_1}(x_1,x'_1), \qquad \XX_t=(\Z,\dist_t,\pi^\star).\)
At $t=0$ and $t=1$, and possibly in degenerate cases, one quotients $\Z$ by the zero-distance relation associated with $\dist_t$. Then $t\mapsto\XX_t$ is a constant-speed geodesic:
\(\GW(\XX_s,\XX_t)=|t-s|\GW(\XX_0,\XX_1) \qquad \forall s,t\in[0,1].\)
\end{prop}
\begin{proof}
Write $D=\GW(\XX_0,\XX_1)$. For $s<t$, couple $\XX_s$ and $\XX_t$ by the diagonal coupling induced by the identity on $\Z$ and the measure $\pi^\star$. For two independent points $z=(x_0,x_1)$ and $z'=(x'_0,x'_1)$ sampled from $\pi^\star$,
\(\dist_t(z,z')-\dist_s(z,z') = (t-s)\big(\dist_{\X_1}(x_1,x'_1)-\dist_{\X_0}(x_0,x'_0)\big).\)
Using this feasible coupling gives $\GW(\XX_s,\XX_t)\leq(t-s)D$. The same construction with the projections from $\Z$ to $\X_0$ and $\X_1$ gives $\GW(\XX_0,\XX_t)\leq tD$ and $\GW(\XX_t,\XX_1)\leq(1-t)D$. The triangle inequality for $\GW$ then yields, for $0\leq s\leq t\leq1$,
\(D \leq \GW(\XX_0,\XX_s)+\GW(\XX_s,\XX_t)+\GW(\XX_t,\XX_1) \leq sD+\GW(\XX_s,\XX_t)+(1-t)D,\)
so $\GW(\XX_s,\XX_t)\geq(t-s)D$. This proves equality.
\end{proof}
\begin{prop}[M\'emoli profile lower bound]\label{prop-memoli-gw-profile-lower-bound}
Let $\XX=(\X,\dist_\X,\alpha)$ and $\YY=(\Y,\dist_\Y,\beta)$ be compact metric-measure spaces and take $\De(u,v)=|u-v|$ in~\eqref{eq-gw-generic}, with the same exponent $p\geq1$. For each $x\in\X$ and $y\in\Y$, define the distance-profile measures on $\RR_+$ by
\(\alpha_x\eqdef(\dist_\X(x,\cdot))_\sharp\alpha, \qquad \beta_y\eqdef(\dist_\Y(y,\cdot))_\sharp\beta.\)
Let $\mathsf E_\XX=(x\mapsto\alpha_x)_\sharp\alpha$ and $\mathsf E_\YY=(y\mapsto\beta_y)_\sharp\beta$, which are probability measures on $\Pp(\RR_+)$. Then
\(\Wass_p\bigl(\mathsf E_\XX,\mathsf E_\YY\bigr) \leq \GW(\XX,\YY).\)
Here the left-hand distance is taken on the space $\Pp(\RR_+)$ of profile measures.
Its ground cost is the one-dimensional Wasserstein distance $\Wass_p$.
\end{prop}
\begin{proof}
Fix any $\pi\in\Couplings(\alpha,\beta)$. It induces a coupling $(x,y)\mapsto(\alpha_x,\beta_y)$ between $\mathsf E_\XX$ and $\mathsf E_\YY$, hence
\(\Wass_p\bigl(\mathsf E_\XX,\mathsf E_\YY\bigr)^p \leq \int_{\X\times\Y}\Wass_p(\alpha_x,\beta_y)^p\,\d\pi(x,y).\)
For fixed $(x,y)$, the map $(x',y')\mapsto(\dist_\X(x,x'),\dist_\Y(y,y'))$ pushes the same coupling $\pi$ to a coupling between $\alpha_x$ and $\beta_y$. Therefore
\(\Wass_p(\alpha_x,\beta_y)^p \leq \int_{\X\times\Y} \abs{\dist_\X(x,x')-\dist_\Y(y,y')}^p \d\pi(x',y').\)
Integrating in $(x,y)$ gives \(\Wass_p\bigl(\mathsf E_\XX,\mathsf E_\YY\bigr)^p \leq \int_{\X^2\times\Y^2} \abs{\dist_\X(x,x')-\dist_\Y(y,y')}^p \d\pi(x,y)\d\pi(x',y').\)
Taking the infimum over $\pi$ and then the $p$-th root proves the claim.
\end{proof}
\paragraph{Entropic regularization and iterative solver.}
For the common squared distortion $\De(u,v)^2=(u-v)^2$, one often computes a stationary point of the entropic relaxation

\eql{\label{eq-gw-entropy}
\umin{\P\in\CouplingsD(\a,\b)}
\Ee_{\distD,\distD'}(\P)-\epsilon\HD(\P).
}
\eql{\label{eq-gw-sinkh}
\itt{\P}
\eqdef
\umin{\P\in\CouplingsD(\a,\b)}
\dotp{\P}{\it{\C}}-\epsilon\HD(\P),
\qquad
\it{\C}
\eqdef
\distD^{\odot2}\a\,\ones_m^\top
+
\ones_n(\distD'^{\odot2}\b)^\top
-
2\distD\,\it{\P}\,\transp{\distD'}.
}
\paragraph{Adding features.}
\[
\operatorname{FGW}_{\lambda,p}((\a,\distD),(\b,\distD'))^p
\eqdef
\umin{\P\in\CouplingsD(\a,\b)}
(1-\lambda)\sum_{i,j}M_{ij}\P_{ij}
+\lambda
\sum_{i,j,i',j'}\De(\distD_{ii'},\distD'_{jj'})^p\P_{ij}\P_{i'j'}.
\]
\paragraph{Hausdorff and Gromov--Hausdorff viewpoints.}
If $A,B$ are compact subsets of a common metric space $(\Z,\dist_\Z)$, their Hausdorff distance is

\[
d_{\mathrm H}^{\Z}(A,B)
=
\max\left\{
\sup_{a\in A}\inf_{b\in B}\dist_\Z(a,b),
\sup_{b\in B}\inf_{a\in A}\dist_\Z(a,b)
\right\}.
\]
\(d_{\mathrm{GH}}(\X,\Y) = \inf_{\Z,\phi,\psi} d_{\mathrm H}^{\Z}(\phi(\X),\psi(\Y)).\)
\subsection{Quantum Optimal Transport}
\label{sec-quantum-ot}
\paragraph{Finite-dimensional states and couplings.}
\begin{defn}[Hermitian and density matrices]\label{def-hermitian-density-matrices}
Let $\mathbb H_n$ be the real vector space of $n\times n$ Hermitian matrices, \(\mathbb H_n^+ = \enscond{A\in\mathbb H_n}{A\succeq0}, \qquad \mathbb H_n^{+,1} = \enscond{A\in\mathbb H_n^+}{\tr(A)=1}.\)
Elements of $\mathbb H_n^{+,1}$ are density matrices.
\end{defn}
A joint quantum state between $\CC^n$ and $\CC^m$ is a matrix $T\in\mathbb H_{nm}^+$ acting on $\CC^n\otimes\CC^m$. Its marginals are the partial traces, defined by duality through

\begin{equation}\label{eq-qot-partial-traces}
\tr(F\,\operatorname{Tr}_B T)=\tr((F\otimes\Id_m)T),
\qquad
\tr(G\,\operatorname{Tr}_A T)=\tr((\Id_n\otimes G)T),
\end{equation}
for all $F\in\mathbb H_n$ and $G\in\mathbb H_m$. Thus $\operatorname{Tr}_B(T)\in\mathbb H_n^+$ and $\operatorname{Tr}_A(T)\in\mathbb H_m^+$ play exactly the role of the two marginals of a classical coupling.

\begin{defn}[Finite-dimensional quantum OT]\label{def-finite-dimensional-qot}
Let $A\in\mathbb H_n^{+,1}$, $B\in\mathbb H_m^{+,1}$ and let $C\in\mathbb H_{nm}$ be a Hermitian cost observable. The quantum OT value is the semidefinite program
\begin{equation}\label{eq-qot-primal}
\mathrm{QOT}_C(A,B)
\eqdef
\min_{T\in\mathbb H_{nm}^+}
\enscond{\tr(CT)}{\operatorname{Tr}_B(T)=A,\ \operatorname{Tr}_A(T)=B}.
\end{equation}
\end{defn}
\begin{prop}[Quantum Kantorovich duality]\label{prop-qot-duality}
For $A\in\mathbb H_n^{+,1}$ and $B\in\mathbb H_m^{+,1}$, the dual of~\eqref{eq-qot-primal} is
\begin{equation}\label{eq-qot-dual}
\mathrm{QOT}_C(A,B)
=
\max_{F\in\mathbb H_n,\ G\in\mathbb H_m}
\left\{
\tr(FA)+\tr(GB)
:\,
F\otimes\Id_m+\Id_n\otimes G\preceq C
\right\}.
\end{equation}
If $A$ and $B$ are positive definite, strong duality follows directly from Slater's condition; the semidefinite case follows by restriction to the supports of $A$ and $B$ or by approximation.
\end{prop}
\begin{proof}
Introduce Hermitian Lagrange multipliers $F$ and $G$ for the two marginal constraints. The Lagrangian is
\[
\tr(CT)+\tr\!\left(F(A-\operatorname{Tr}_B T)\right)
+\tr\!\left(G(B-\operatorname{Tr}_A T)\right)
=
\tr(FA)+\tr(GB)
+\tr\!\left((C-F\otimes\Id_m-\Id_n\otimes G)T\right),
\]
where~\eqref{eq-qot-partial-traces} was used in the last equality. Minimizing over $T\succeq0$ gives a finite lower bound if and only if $C-F\otimes\Id_m-\Id_n\otimes G\succeq0$, in which case the infimum in $T$ is $0$. When $A,B\succ0$, the coupling $A\otimes B$ is strictly feasible, so Slater's theorem gives equality of primal and dual values and dual attainment. The general finite-dimensional semidefinite case is obtained by approximation $A_\delta=(1-\delta)A+\delta\Id_n/n$, $B_\delta=(1-\delta)B+\delta\Id_m/m$ and by compactness, or equivalently by reducing to the supports of $A$ and $B$.
\end{proof}
\paragraph{Entropic regularization and Bregman iterations.}
\begin{defn}[von Neumann quantum entropy]\label{def-von-neumann-quantum-entropy}
For a density matrix or positive semidefinite matrix $T$, the shifted von Neumann entropy functional used here is
\[
H(T)=\tr\!\left(T(\log T-\Id)\right),
\qquad
\nabla H(T)=\log T,
\]
with the convention $0\log0=0$ on eigenvalues. This is the convex negative quantum entropy; on trace-one states it differs from the physical entropy $-\tr(T\log T)$ by a sign and an additive constant.
\end{defn}
For $\epsilon>0$ define

\begin{equation}\label{eq-qot-entropic-primal}
\mathrm{QOT}_C^\epsilon(A,B)
=
\min_{T\succeq0}
\left\{
\tr(CT)+\epsilon H(T)
:\,
\operatorname{Tr}_B(T)=A,
\operatorname{Tr}_A(T)=B
\right\}.
\end{equation}
\begin{prop}[Entropic quantum OT duality]\label{prop-qot-entropic-duality}
Assume $A\succ0$, $B\succ0$ and $\epsilon>0$. Then~\eqref{eq-qot-entropic-primal} has a unique positive minimizer. Its dual is
\begin{equation}\label{eq-qot-entropic-dual}
\mathrm{QOT}_C^\epsilon(A,B)
=
\max_{F\in\mathbb H_n,\ G\in\mathbb H_m}
\left\{
\tr(FA)+\tr(GB)
-\epsilon\,
\tr\exp\!\left(\frac{F\otimes\Id_m+\Id_n\otimes G-C}{\epsilon}\right)
\right\}.
\end{equation}
At optimality, primal and dual variables are linked by the Gibbs formula
\begin{equation}\label{eq-qot-gibbs-coupling}
T_e(F,G)
=
\exp\!\left(\frac{F\otimes\Id_m+\Id_n\otimes G-C}{\epsilon}\right),
\end{equation}
with $\operatorname{Tr}_B(T_e)=A$ and $\operatorname{Tr}_A(T_e)=B$.
\end{prop}
\begin{proof}
The feasible set is compact and nonempty, and it contains the positive definite point $A\otimes B$. The trace entropy is strictly convex on positive semidefinite matrices, hence the regularized primal has a unique minimizer. Slater's condition justifies the Lagrange dual computation. The Fenchel identity
\(\sup_{T\succeq0}\ \tr(YT)-\epsilon H(T) = \epsilon\,\tr\exp(Y/\epsilon)\)
is the matrix analogue of the scalar exponential conjugacy. Applying it to the Lagrangian of~\eqref{eq-qot-entropic-primal}, with
$Y=F\otimes\Id_m+\Id_n\otimes G-C$, gives~\eqref{eq-qot-entropic-dual}. The stationarity condition of this Fenchel identity gives~\eqref{eq-qot-gibbs-coupling}; differentiating the dual objective with respect to $F$ and $G$ yields the two marginal equations.
\end{proof}
\[
D_H(T|K)
=
\tr\!\left(T(\log T-\log K)-T+K\right).
\]
\(\mathcal M_A=\{T\succeq0:\operatorname{Tr}_B(T)=A\}, \qquad \mathcal M_B=\{T\succeq0:\operatorname{Tr}_A(T)=B\}.\)
\begin{prop}[Exact Bregman projections]\label{prop-qot-bregman-projections}
Assume $A,B\succ0$ and let $K=\exp(-C/\epsilon)$. The minimizer of~\eqref{eq-qot-entropic-primal} is equivalently the minimizer of $D_H(T|K)$ over $\mathcal M_A\cap\mathcal M_B$. Moreover, if a current positive definite matrix has Gibbs form $T_e(F,G)$, then its Bregman projection onto $\mathcal M_A$ has the form $T_e(F^+,G)$, where $F^+$ is chosen so that $\operatorname{Tr}_B T_e(F^+,G)=A$. The projection onto $\mathcal M_B$ is analogous. Thus, when each one-block marginal equation is solved exactly, alternating Bregman projections are equivalent to alternating block maximization of the dual~\eqref{eq-qot-entropic-dual}.
\end{prop}
\begin{proof}
Since $\log K=-C/\epsilon$, the identity
\(\tr(CT)+\epsilon H(T) = \epsilon D_H(T|K)-\epsilon\tr(K)\)
holds, so the primal minimizer is the constrained Bregman projection of $K$ up to an additive constant. For the projection of a positive definite matrix $S$ onto $\mathcal M_A$, the affine set contains the positive definite point $A\otimes\Id_m/m$; the entropy derivative $\log T-\log S$ is singular at the boundary, so the projection lies in the interior of the positive cone. Its Lagrangian first variation is
\(\log T-\log S-\Lambda\otimes\Id_m=0\)
for a Hermitian multiplier $\Lambda$. Hence $T=\exp(\log S+\Lambda\otimes\Id_m)$. If $S=T_e(F,G)$, this is again of the form $T_e(F+\epsilon\Lambda,G)$. The multiplier is fixed by the marginal equation $\operatorname{Tr}_B(T)=A$. The same argument applies to $\mathcal M_B$. Finally, the first-order optimality condition for maximizing~\eqref{eq-qot-entropic-dual} over one block is exactly the corresponding marginal equation, so the Bregman and block-dual views coincide.
\end{proof}
\(\operatorname{Tr}_B T_e(F,G)=A, \qquad \operatorname{Tr}_A T_e(F,G)=B\)
\paragraph{Gurvits scaling and quantum Sinkhorn.}
\begin{equation}\label{eq-qot-symmetric-scaling}
T_s(F,G)
=
\exp\!\left(\frac{Z}{2\epsilon}\right)\exp(-C/\epsilon)\exp\!\left(\frac{Z}{2\epsilon}\right)
=
(U\otimes V)K(U\otimes V),
\qquad
Z=F\otimes\Id_m+\Id_n\otimes G,
\end{equation}
where $U=\exp(F/(2\epsilon))$, $V=\exp(G/(2\epsilon))$ and $K=\exp(-C/\epsilon)$. If $[Z,C]=0$, then $T_s(F,G)=T_e(F,G)$; otherwise this is a Strang-type symmetric surrogate.

Fix a Choi convention and let $\mathcal K:\mathbb H_m\to\mathbb H_n$ be the completely positive map represented by the positive Choi matrix $K$; let $\mathcal K^\star$ be its Hilbert--Schmidt adjoint. Up to the transpose dictated by the chosen Choi convention, the marginal equations for the symmetric coupling take the operator-scaling form

\(U\,\mathcal K(V^2)\,U=A, \qquad V\,\mathcal K^\star(U^2)\,V=B.\)
\begin{equation}\label{eq-qot-gurvits-updates}
\begin{aligned}
R_V&=\mathcal K(V^2),
&
U&\leftarrow
R_V^{-1/2}\bigl(R_V^{1/2} A R_V^{1/2}\bigr)^{1/2}R_V^{-1/2},
\\
S_U&=\mathcal K^\star(U^2),
&
V&\leftarrow
S_U^{-1/2}\bigl(S_U^{1/2} B S_U^{1/2}\bigr)^{1/2}S_U^{-1/2}.
\end{aligned}
\end{equation}
These inverse square roots are well-defined when $K\succ0$ and $U,V,A,B\succ0$. This is Gurvits/operator scaling with prescribed targets; when all matrices are diagonal it reduces to classical Sinkhorn scaling, and when the targets are proportional to identities it matches the usual bistochastic operator-scaling normalization, up to the conventional trace normalization.