\section{Dual Problem}
\label{sec-dual}
\subsection{Discrete dual}
The Kantorovich problem~\eqref{eq-kanto-discr} is a linear program so that one can equivalently compute its value by solving a dual linear program.

\begin{defn}[Admissible potentials]\label{def-admissible-potentials}
For a discrete problem with marginal sizes $n,m$ and cost matrix $\C\in\RR^{n\times m}$, a pair $(\fD,\gD)\in\RR^n\times\RR^m$ is admissible if it lies below the cost:
\eql{\label{eq-feasible-potential}
\PotentialsD(\a,\b) \eqdef \enscond{
(\fD,\gD) \in \RR^n \times \RR^m
}{ \foralls (i,j) \in \range{n} \times \range{m}, \fD_i+\gD_j \leq \C_{i,j} }.
}
Equivalently, $\fD\oplus\gD\leq\C$ entrywise. The notation suppresses the dependence on $\C$, which is fixed in the surrounding problem.
\end{defn}
\begin{prop}[Discrete Kantorovich duality]\label{prop-duality-discr}
One has
\eql{\label{eq-dual}
\MKD_\C(\a,\b) =
\umax{(\fD,\gD) \in \PotentialsD(\a,\b)} \dotp{\fD}{\a} + \dotp{\gD}{\b}
}
\end{prop}
\begin{proof}

For the sake of completeness, let us derive this dual problem using Lagrangian duality. The Lagrangian associated to~\eqref{eq-kanto-discr} reads
\eql{\label{eq-mk-lagr}
\umin{\P \geq 0} \umax{ (\fD,\gD) \in \RR^n \times \RR^m }
\dotp{\C}{\P} + \dotp{\a - \P\ones_m}{\fD} + \dotp{\b - \P^\top \ones_n}{\gD}.
}
For a linear program, if the primal constraint set is non-empty, one can always exchange the min and the max and get the same value. We thus consider
\(\umax{ (\fD,\gD) \in \RR^n \times \RR^m } \dotp{\a}{\fD} + \dotp{\b}{\gD} + \umin{\P \geq 0} \dotp{\C - \fD\ones_m^\top - \ones_n \gD^\top}{\P}.\)
We conclude by remarking that
\eq{
\umin{\P \geq 0} \dotp{\Q}{\P} =
\choice{
0 \qifq \Q \geq 0\\
-\infty \quad \text{otherwise}
}
}
so that the constraint reads $\C - \fD\ones_m^\top - \ones_n \gD^\top = \C-\fD\oplus \gD \geq 0$.
\end{proof}
\eql{\label{eq-mk-pd-rel}
\Supp(\P) \subset \enscond{(i,j) \in \range{n} \times \range{m}}{ \fD_i+\gD_j=\C_{i,j} }.
}
\subsection{Auction Algorithm and Dual Prices}
\label{sec-auction-dual-ascent}
Consider the square assignment problem with costs $C_{i,j}$ and rewrite it as the profit maximization problem with $a_{i,j}=-C_{i,j}$. The auction algorithm keeps prices $p_j$ on the target points and a partial assignment.

\(v_i=\max_j (a_{i,j}-p_j),\qquad j_i\in\operatorname*{argmax}_j(a_{i,j}-p_j),\qquad w_i=\max_{j\neq j_i}(a_{i,j}-p_j).\)
\[
p_{j_i}\leftarrow p_{j_i}+v_i-w_i+\epsilon.
\]
For fixed prices $p$, eliminating the bidder utilities $u_i$ in the dual minimization gives the convex objective

\(D(p)=\sum_j p_j+\sum_i \max_j(a_{i,j}-p_j),\)
\begin{defn}[$\epsilon$-complementary slackness]
An assignment $\sigma$ and prices $p$ satisfy $\epsilon$-complementary slackness if, for every source $i$, \(a_{i,\sigma(i)}-p_{\sigma(i)} \geq \max_j(a_{i,j}-p_j)-\epsilon.\)
\end{defn}
\begin{prop}[Auction optimality certificate]\label{prop-auction-eps-cs}
If a complete assignment $\sigma$ satisfies $\epsilon$-complementary slackness, then it is $n\epsilon$-optimal for the profit maximization problem, or equivalently $n\epsilon$-optimal for the original cost minimization problem. If all costs are integers and $\epsilon<1/n$, then $\sigma$ is optimal.
\end{prop}
\begin{proof}
Let $\tau$ be any assignment. By $\epsilon$-complementary slackness, \(a_{i,\tau(i)}-p_{\tau(i)} \leq \max_j(a_{i,j}-p_j) \leq a_{i,\sigma(i)}-p_{\sigma(i)}+\epsilon.\)
Summing over $i$ cancels prices, because both $\sigma$ and $\tau$ are permutations: \(\sum_i a_{i,\tau(i)} \leq \sum_i a_{i,\sigma(i)}+n\epsilon.\)
Thus no assignment has profit more than $n\epsilon$ above that of $\sigma$. Since $a=-C$, the same statement says that the cost of $\sigma$ is at most $n\epsilon$ above the minimum cost. If the costs are integers, all assignment costs are integers; a gap strictly smaller than one therefore forces the gap to be zero.
\end{proof}
\subsection{General formulation}
\label{sec-dual-general}
\begin{prop}[Kantorovich duality]\label{prop-kantorovich-duality-general}
Assume that $\Xx$ and $\Yy$ are compact metric spaces and that $c\in\Cc(\Xx\times\Yy)$. Then
\eql{\label{eq-dual-generic}
\MK_\c(\al,\be) =
\umax{(\f,\g) \in \Potentials(\c)}
\int_\X \f(x) \d\al(x) + \int_\Y \g(y) \d\be(y),
}
where the set of admissible dual potentials is
\eql{\label{eq-dfn-pot-dual}
\Potentials(\c) \eqdef \enscond{
(\f,\g) \in \Cc(\X) \times \Cc(\Y)
}{
\forall (x,y), \f(x)+\g(y) \leq \c(x,y)
}.
}
Here, $(\f,\g)$ is a pair of continuous functions, often called ``Kantorovich potentials''.
The same formula extends under the usual lower-semicontinuity and integrability assumptions, replacing maxima by suprema when dual optimizers need not exist.
\end{prop}
\begin{proof}
Weak duality is immediate: if $f(x)+g(y)\leq c(x,y)$ and $\pi\in\Couplings(\alpha,\beta)$, then
\(\int f\d\alpha+\int g\d\beta = \int (f(x)+g(y))\d\pi(x,y) \leq \int c\d\pi.\)
Taking the supremum over admissible potentials and the infimum over couplings gives ``$\leq$''.

For the reverse inequality, view the primal problem as a linear program over the locally convex space of Radon measures, paired with $\Cc(\X\times\Y)$. The affine map $\pi\mapsto(\pi_1,\pi_2)$ is continuous for the weak topology, the feasible set is non-empty because it contains $\alpha\otimes\beta$, and the cost is continuous and bounded on compact sets. Since the set of probability measures on the compact product is weakly compact, the set of attainable cost-marginal triples is closed after adding the epigraph variable below. The separating-hyperplane theorem applied to the convex set of attainable triples
\(\Big\{\big(\pi_1,\pi_2,\int c\d\pi+r\big): \pi\geq0,\ r\geq0\Big\}\)
gives a continuous affine separator, hence functions $(f,g)$ and a scalar multiplier which can be normalized so that $f\oplus g\leq c$. The separating inequality then states that the supremum over such potentials is at least the primal value. This proves equality. The same argument is the infinite-dimensional analogue of the finite linear-programming proof in Proposition~\ref{prop-duality-discr}.
\end{proof}
\eql{\label{eq-mk-pd-rel-cont}
\Supp(\pi) \subset \enscond{(x,y) \in \X \times \Y}{ \f(x)+\g(y)=\c(x,y) }.
}
For the one-dimensional quadratic cost, the continuous potentials can be read from the monotone map $T=F_\be^{-1}\circ F_\al$: on the active graph, $f'(x)=2(x-T(x))$ and $g=f^c$.

\subsection{\texorpdfstring{$c$-transforms}{c-transforms}}
\label{sec-c-transfo}
\paragraph{\texorpdfstring{Best-response potentials and the $c$-transform.}{Best-response potentials and the c-transform.}}
\(\usup{\g \in \Cc(\Yy)} \enscond{ \int g \d \be }{ \foralls (x,y), \g(y) \leq c(x,y) - \f(x) }.\)
\(\foralls y \in \Yy, \quad \g(y) \leq \f^\c(y)\)
\begin{defn}[$c$-transform]\label{def-c-transform}
For a function $f:\Xx\to\RR\cup\{-\infty\}$, its $c$-transform is
\eql{\label{eq-c-transform}
\foralls y \in \Y, \quad
\f^\c(y) \eqdef \uinf{x \in \X} \c(x,y) - \f(x).
}
For a function $g:\Yy\to\RR\cup\{-\infty\}$, the $\bar c$-transform associated with $\bar c(y,x)=c(x,y)$ is \(\foralls x \in \X, \quad \g^{\bar\c}(x) \eqdef \uinf{y \in \Y} \c(x,y) - \g(y).\)
\end{defn}
\begin{prop}[$c$-transforms solve the semi-relaxed problems]\label{prop-c-transform-semi-relaxed}
For fixed $f$, the maximizers of the dual objective over all $g$ such that $f\oplus g\leq c$ are exactly the functions satisfying $g=f^c$ $\be$-almost everywhere. Equivalently, $f^c$ gives the value of the one-marginal primal problem
\(\inf_{\pi:\,\pi_2=\be} \int c(x,y)\d\pi(x,y)-\int f(x)\d\pi_1(x) = \int f^c(y)\d\be(y).\)
Symmetrically, for fixed $g$, the maximizers over $f$ are the functions satisfying $f=g^{\bar c}$ $\al$-almost everywhere.
\end{prop}
\begin{proof}
The constraint $f(x)+g(y)\leq c(x,y)$ for all $x$ is equivalent, for each fixed $y$, to
\(g(y)\leq \inf_x c(x,y)-f(x)=f^c(y).\)
Since $\be$ is nonnegative, the largest possible value of $\int g\d\be$ is obtained by saturating this pointwise upper bound on the support of $\be$. The proof for $f=g^{\bar c}$ is identical after exchanging the two marginals.

For the primal formula, disintegrate any feasible $\pi$ as $\pi(\d x,\d y)=\pi_y(\d x)\be(\d y)$. Then
\[
\int c\d\pi-\int f\d\pi_1
=
\int\left(\int (c(x,y)-f(x))\d\pi_y(x)\right)\d\be(y)
\geq
\int f^c(y)\d\be(y).
\]
If minimizers admit a measurable selection, equality is obtained by choosing $\pi_y$ supported on minimizers of $x\mapsto c(x,y)-f(x)$. Otherwise one uses approximate measurable selections and lets the approximation error vanish.
\end{proof}
The map $(f,g)\mapsto(g^{\bar c},f^c)$ replaces dual potentials by better ones, in the sense that it preserves feasibility and improves the dual objective. Functions of the form $f^c$ and $g^{\bar c}$ are called $c$-concave and $\bar c$-concave functions.

\begin{prop}[Lipschitz stability of $c$-transforms]\label{prop-c-transform-lipschitz}
If $c$ is $L$-Lipschitz with respect to its second variable, uniformly in the first one and for the metric $d_\Y$ on $\Y$, then $f^c$ is $L$-Lipschitz.
\end{prop}
\begin{proof}
For each $x$, set $F_x(y)=c(x,y)-f(x)$ and $F(y)=f^c(y)=\inf_x F_x(y)$. Since all the functions $F_x$ are $L$-Lipschitz,
\[
|F(y)-F(y')|
=
\left|\inf_x F_x(y)-\inf_x F_x(y')\right|
\leq
\sup_x |F_x(y)-F_x(y')|
\leq
L d_\Y(y,y').
\]
\end{proof}
\paragraph{Euclidean case.}
The Euclidean quadratic cost is the model case where $c$-transforms become ordinary convex conjugates after removing the quadratic terms. This is the algebraic bridge between Kantorovich duality and Brenier maps.

\(\int \norm{x-y}^2\d\pi(x,y) = \int\norm{x}^2\d\al(x)+\int\norm{y}^2\d\be(y) -2\int \dotp{x}{y}\d\pi(x,y).\)
For $c(x,y)=-\dotp{x}{y}$, one has

\(f^c(y) = \inf_x -\dotp{x}{y}-f(x) = -(-f)^*(y), \qquad h^*(y)\eqdef\sup_x \dotp{x}{y}-h(x).\)
Thus $c$-concave functions are negatives of convex functions. In the one-dimensional bilinear model case, the hard double $c$-transform is therefore an operation of taking concave envelopes.

\paragraph{The failure of alternate optimization.}
\begin{prop}[Algebra of $c$-transforms]
\[
\textnormal{(i) } f \leq f' \Rightarrow f^{c}\geq f'^{c},
\qquad
\textnormal{(ii) } f^{\c\bar{\c}} \geq f,
\qquad
\textnormal{(iii) } g^{\bar{\c}\c} \geq g,
\qquad
\textnormal{(iv) } \f^{\c\bar{\c}\c}=\f^{\c}.
\]
\end{prop}
\begin{proof} The first inequality (i) follows from the definition of $\c$-transforms (because of the $-$ sign). To prove (ii), expanding the definition of $\f^{\c\bar{\c}}$ we have
\eq{
\left(\f^{\c\bar{\c}}\right)(x)= \min_{y} \c(x,y)-\f^{\c}(y)
= \min_{y} \c(x,y) - \min_{x'}( \c(x',y) - \f(x') ).}
Now, since $-\min_{x'}\big(\c(x',y)-\f(x')\big) \geq -(\c(x,y)-\f(x))$, we recover
\((\f^{\c\bar{\c}})(x) \geq \min_{y} \c(x,y) - \c(x,y)+\f(x) = \f(x).\)
The relation $\g^{\bar{\c}\c} \geq \g$ is obtained in the same way.

Now, to prove (iv), we first apply (ii) and then (i) with $f'=f^{c\bar c}$ to have $f^c \geq f^{c \bar c c}$.
Then we apply (iii) to $g=f^c$ to obtain $f^c \leq f^{c\bar c c}$.

\end{proof}
\eql{\label{eq-iter-c-trans}
(f,g) \mapsto (f,f^c) \mapsto (f^{c\bar c},f^c) \mapsto (f^{c\bar c},f^{c\bar c c}) = (f^{c\bar c},f^c) \ldots
}
For the bilinear cost $c(x,y)=-xy$ on a compact interval, the $c$-concave functions are ordinary concave functions and $f^{c\bar c}$ is the smallest concave majorant of $f$. In that model case, a hard transform removes non-concave oscillations in one closure step rather than producing a gradual ascent.