\section{Divergences and Dual Norms}
\label{sec-divergences-dual-norms}
\subsection{Dual norms (Integral Probability Metrics)}
\label{sec-dual-norms}
\eql{\label{eq-dual-norm-cont}
\norm{\al}_B \eqdef
\usup{\f} \enscond{ \int_{\X} \f(x) \d\al(x) }{ \f \in B}.
}
\begin{example}[Total variation]
\(B = \enscond{f \in \Cc(\X)}{\norm{f}_\infty \leq 1}.\)
\end{example}
\begin{example}[$\Wass_1$ norm]
$\Wass_1$ as defined in~\eqref{eq-w1-cont}, is a special case of dual norm~\eqref{eq-dual-norm-cont}, using

\(B = \enscond{f}{\Lip(f) \leq 1}\)
\end{example}
\begin{example}[Flat norm and Dudley metric]
If the set $B$ is bounded and separates measures, then $\norm{\cdot}_B$ is a norm on the whole space $\Mm(\Xx)$ of finite measures.

This is not the case of $\Wass_1$, which is only defined for $\al$ such that $\int_\X \d\al=0$ (otherwise $\norm{\al}_B=+\infty$).

\eql{\label{eq-set-flatnorm}
B=\enscond{f}{\Lip(f) \leq 1 \qandq \norm{\f}_\infty \leq 1}.
}
\eq{\label{eq-set-dudley}
B=\enscond{f}{\norm{\nabla \f}_\infty + \norm{\f}_\infty \leq 1}.
}
\end{example}
\begin{prop}[Metrization by dual norms]\label{prop-metrization}
Assume $\Xx$ is compact, $B=-B$, and that all measures below are probability measures.
(i) If $\Cc(\Xx) \subset \overline{\Span(B)}$ (i.e. if the span of $B$ is dense in continuous functions for the sup-norm $\norm{\cdot}_\infty$), then $\norm{\al_k-\al}_B \rightarrow 0$ implies $\al_k \rightharpoonup \al$.
(ii) If $B \subset \Cc(\Xx)$ is compact for $\norm{\cdot}_\infty$, then $\al_k \rightharpoonup \al$ implies $\norm{\al_k-\al}_B \rightarrow 0$.
\end{prop}
\begin{proof}
(i) If $\norm{\al_k-\al}_B \rightarrow 0$, then by duality and the symmetry of $B$, for any $f \in B$, \(\abs{\dotp{f}{\al_k-\al}} \leq \norm{\al_k-\al}_B,\)
so $\dotp{f}{\al_k} \rightarrow \dotp{f}{\al}$. By linearity, this convergence extends to $\Span(B)$.

Let $u \in \Cc(\Xx)$ and choose $h \in \Span(B)$ with $\norm{u-h}_\infty \leq \eta$. Since $\alpha_k$ and $\alpha$ have mass one, \(\abs{\dotp{u}{\al_k-\al}} \leq \abs{\dotp{h}{\al_k-\al}} + 2\eta.\)
Taking the limsup in $k$ and then letting $\eta \rightarrow 0$ proves weak convergence.

(ii) We assume $\al_k \rightharpoonup \al$ and we consider a sub-sequence $\al_{n_k}$ such that
\(\norm{\al_{n_k}-\al}_B \longrightarrow \lim\sup_k \norm{\al_{k}-\al}_B\)
Since $B$ is compact, the maximum appearing in the definition of $\norm{\al_{n_k}-\al}_B$ is reached, so there exists $f_{n_k}\in B$ such that $\dotp{\al_{n_k}-\al}{f_{n_k}} = \norm{\al_{n_k}-\al}_B$. Once again, by compactness, we can extract from $(f_{n_k})_k$ a (not relabelled for simplicity) subsequence converging uniformly to some $f \in B$. One has $\norm{\al_{n_k}-\al}_{B} = \dotp{\al_{n_k}-\al}{f_{n_k}}$, and this quantity converges to 0 because
\(\dotp{\al_{n_k}-\al}{f_{n_k}} = \dotp{\al_{n_k}-\al}{f} + \dotp{\al_{n_k}}{f_{n_k} - f} - \dotp{\al}{f_{n_k} - f}\)
and these three terms go to zero because $\al_{n_k}-\al \rightharpoonup 0$ for the first one and $\norm{f_{n_k} - f}_\infty \rightarrow 0$ for the last two.
\end{proof}
\begin{cor}[Wasserstein metrizes weak convergence]\label{cor-topol-wass}
On a compact space, the Wasserstein-$p$ distance metrizes the weak convergence.
\end{cor}
\begin{proof}
For $\Wass_1$, take $B=\enscond{f}{\Lip(f) \leq 1}$. The span of $B$ contains all Lipschitz functions, and Lipschitz functions are dense in $\Cc(\Xx)$ on a compact metric space.

For the converse implication, constants do not change the pairing with $\al_k-\al$ because both measures have mass one. We may therefore normalize every $f \in B$ by imposing $f(x_0)=0$ for a fixed $x_0 \in \Xx$. The normalized class is uniformly bounded by $\mathrm{diam}(\Xx)$ and equicontinuous, hence compact in $\norm{\cdot}_\infty$ by Ascoli--Arzel\`a. Proposition~\ref{prop-metrization} then applies.

Proposition~\ref{prop-comp-wass-p} shows that $\Wass_p$ has the same topology as $\Wass_1$ on a compact space, so every $\Wass_p$ metrizes convergence in law.
\end{proof}
\paragraph{Dual RKHS Norms and Maximum Mean Discrepancies.}
\begin{defn}[Positive and conditionally positive kernels]\label{def-negativedefinitekernel}
A symmetric function $k$ defined on $\X \times \X$ is said to be positive definite if for any $n\geq0$, for any family $x_1,\dots,x_n\in\X$ the matrix $(k(x_i,x_j))_{i,j}$ is positive semidefinite, i.e. for all $r \in \RR^n$
\eql{\label{eq-dual-kern}
\sum_{i,j=1}^n r_i r_j k(x_i,x_j)\geq 0,
}
The kernel is said to be conditionally positive definite if positivity only holds in~\eqref{eq-dual-kern} for zero-mean vectors $r$, i.e. such that $\dotp{r}{\ones_n}=0$.
\end{defn}
\begin{example}[Riesz, energy and Mat\'ern-type kernels]
\(-\iint \norm{x-y}\d\xi(x)\d\xi(y)\)
\end{example}
\begin{defn}[Kernel norm and MMD]\label{def-kernel-mmd-norm}
If $k$ is positive definite, or conditionally positive definite and restricted to signed measures of total mass zero, the associated kernel norm is
\eql{\label{eq-kernel-dual}
\norm{\xi}^2_{\Krkhs} \eqdef \int_{\X \times \X} \Krkhs(x,y) \d \xi(x) \d \xi(y).
}
For two probability measures, $\norm{\alpha-\beta}_{\Krkhs}$ is the maximum mean discrepancy associated with $k$.
\end{defn}
\(\norm{\al}^2_{\Krkhs} = \EE_{X,X'}( \Krkhs(X,X') ).\)
\begin{prop}[Kernel norm as an RKHS dual norm]
Let $\RKHS$ be the RKHS with reproducing kernel $\Krkhs$, and assume the kernel mean embedding
\(m_\xi \eqdef \int \Krkhs(x,\cdot)\,\d\xi(x)\)
is well-defined for the signed measure $\xi$. Then
\(\norm{\xi}_{\Krkhs} = \usup{\norm{h}_{\RKHS}\leq1}\int h(x)\,\d\xi(x),\)
so $\norm{\cdot}_{\Krkhs}$ is the dual norm in the sense of~\eqref{eq-dual-norm-cont} associated with the RKHS unit ball.
\end{prop}
\begin{proof}
By the reproducing property,
\[
\int h(x)\,\d\xi(x)
=
\left\langle h,\int \Krkhs(x,\cdot)\,\d\xi(x)\right\rangle_{\RKHS}
=
\langle h,m_\xi\rangle_{\RKHS}.
\]
Cauchy--Schwarz gives \(\sup_{\norm{h}_{\RKHS}\leq1}\int h\,\d\xi=\norm{m_\xi}_{\RKHS}.\)
Finally, \(\norm{m_\xi}_{\RKHS}^2 = \iint \Krkhs(x,y)\,\d\xi(x)\,\d\xi(y),\)
which is exactly~\eqref{eq-kernel-dual}.
\end{proof}
\begin{rem}[Universal kernels]
\end{rem}
In the special case where $\al$ is a discrete measure, one thus has the simple expression

\eq{
\norm{\al}_{\Krkhs}^2 = \sum_{i=1}^n \sum_{i'=1}^n \a_i\a_{i'} \KrkhsD_{i,i'} = \dotp{\KrkhsD\a}{\a}
\qwhereq
\KrkhsD_{i,i'} \eqdef \Krkhs(x_i,x_{i'}).
}
\eql{\label{eq-mmd-discr}
\norm{\al-\be}_{\Krkhs}^2 =
\sum_{i,i'} \a_i \a_{i'} \Krkhs(x_i,x_{i'})	+
\sum_{j,j'} \b_j \b_{j'} \Krkhs(y_j,y_{j'}) - 2
\sum_{i,j} \a_i \b_j \Krkhs(x_i,y_j).
}
\subsection[phi-divergences]{$\phi$-divergences}
\label{sec-phi-div}
\begin{defn}[Entropy function]
\label{def_entropy}
A function $\phi: \RR \to \RR \cup \{\infty\}$ is an entropy function if it is lower semicontinuous, convex, $\dom \phi\subset [0,\infty[$, and satisfies the feasibility condition $\dom \phi \cap (0,+\infty) \neq \emptyset$. The speed of growth of $\phi$ at $\infty$ is described by
\eq{
\phi'_\infty = \lim_{x\rightarrow +\infty} \phi(x)/x \in \RR \cup \{\infty\}.
}
\end{defn}
If $\phi'_\infty = \infty$, then $\phi$ grows faster than any linear function and $\phi$ is said to be \emph{superlinear}.

\begin{defn}[$\phi$-Divergences]
\label{def_divergence}
Let $\phi$ be an entropy function.
For $\al,\be \in \Mm(\X)$, let $\frac{\d \al}{\d \be} \be + \al^{\perp}$ be the Lebesgue decomposition\footnote{The Lebesgue decomposition theorem asserts that, given $\be$, $\al$ admits a unique decomposition as the sum of two measures $\al^{\mathrm{ac}} + \al^\perp$ such that $\al^{\mathrm{ac}}$ is absolutely continuous with respect to $\be$ and $\al^\perp$ and $\be$ are singular.} of $\al$ with respect to $\be$. The divergence $\Divergm_\phi$ is defined by
\eql{\label{eq-phi-div}
\Divergm_\phi (\al|\be) \eqdef \int_\X \phi\left(\frac{\d \al}{\d \be} \right) \d \be
+ \phi'_\infty \al^{\perp}(\X)
}
if $\al,\be$ are nonnegative and $\infty$ otherwise.
\end{defn}
\eql{\label{eq-div-disc-meas}
\al=\sum_i \a_i \de_{x_i}
\qandq \be=\sum_i \b_i \de_{x_i}
}
\eql{\label{eq-discr-diverg}
\DivergmD_\phi(\a|\b) = \sum_{i \in \Supp(\b)} \phi\pa{ \frac{\a_i}{\b_i} } \b_i + \phi'_\infty \sum_{i \notin \Supp(\b)} \a_i,
}
where $\Supp(\b) \eqdef \enscond{i \in \range{n}}{ \b_i \neq 0 }$.

\begin{proposition}[Basic properties of $\phi$-divergences]
If $\phi$ is an entropy function, then $\Divergm_\phi$ is jointly $1$-homogeneous, convex and weak-* lower semicontinuous in $(\al,\be)$.
\end{proposition}
\begin{proof}
One defines the associated perspective function
\eq{
\foralls (u,v) \in (\RR_+)^2, \quad
\psi(u,v) = \choice{
\phi(u/v) v \qifq v \neq 0 \\
u \phi'_\infty \qifq v=0
}
}
The joint $1$-homogeneity follows from the definition of this perspective.
\(\DivergmD_\phi(\a|\b) = \sum_{i} \psi(\a_i,\b_i),\)
and it is enough to show that $\psi$ is convex on $(\RR_+)^2$. We first prove this on $\RR_+ \times \RR_+^*$; the extension to $v=0$ follows by lower semicontinuity of the recession value $u\phi'_\infty$.

Indeed, for any $\la \in [0,1]$, $\tau=1-\la$
\begin{align*}
\phi\pa{ \frac{\tau u_1 + \la u_2}{\tau v_1 + \la v_2} }
(\tau v_1 + \la v_2)
=
\phi\pa{
\frac{\tau v_1 }{\tau v_1 + \la v_2} \frac{u_1}{v_1}+
\frac{\la v_2 }{\tau v_1 + \la v_2} \frac{u_2}{v_2}
}
(\tau v_1 + \la v_2)
\leq
\tau \phi\pa{\frac{u_1}{v_1}} v_1
+
\la \phi\pa{\frac{u_2}{v_2}} v_2.
\end{align*}
In the general measure case, weak-* lower semicontinuity is the standard lower-semicontinuity theorem for convex integral functionals with recession extension; in the discrete case it is immediate from the lower semicontinuity of $\psi$.

\end{proof}
\begin{proposition}[Non-negativity of $\phi$-divergences]\label{phi-div-positive}
Assume that $\phi$ is normalized by $\phi(1)=0$. For probability distributions $(\al,\be) \in \Mm_+^1(\Xx)$, one has $\Divergm_\phi(\al|\be) \geq 0$. If $\phi$ is strictly convex, then one has $\Divergm_\phi(\al|\be)=0$ if and only if $\al=\be$.
\end{proposition}
\begin{proof}
Let $m=\al+\be$ and write $a=\frac{\d\al}{\d m}$ and $b=\frac{\d\be}{\d m}$. Using the perspective function $\psi$ from the previous proof, \(\Divergm_\phi(\al|\be)=\int \psi(a,b)\d m.\)
Since $\al$ and $\be$ are probabilities, $\int a\d m=\int b\d m=1$. Jensen's inequality and $\psi(1,1)=\phi(1)=0$ give
\(\Divergm_\phi(\al|\be) \geq \psi\pa{\int a\d m,\int b\d m}=0.\)
If $\phi$ is strictly convex, equality in Jensen forces $a=b$ $m$-almost everywhere, hence $\al=\be$. In the general non-probability case, if $\phi \geq 0$ then the divergence is positive by construction.
\end{proof}
\begin{example}[Kullback--Leibler divergence]
\eql{\label{eq-shannon-entropy}
\phi_{\KL}(s)= \begin{cases}
s\log(s)-s+1 & \textnormal{for } s>0, \\
1 & \textnormal{for } s=0, \\
+\infty & \textnormal{otherwise.}
\end{cases}
}
\end{example}
\begin{example}[Total variation]\label{exmp-tv}
\eql{\label{eq-tv-entropy}
\phi_{\TV}(s)= \begin{cases}
|s-1| & \textnormal{for } s\geq0, \\
+\infty & \textnormal{otherwise.}
\end{cases}
}
It actually defines a norm on the full space of measures $\Mm(\X)$ where

\eql{\label{eq-defn-tv}
\TV(\al|\be) = \norm{\al-\be}_{\TV},
\qwhereq
\norm{\al}_{\TV} = |\al|(\X) = \int_\X \d|\al|(x).
}
If $\al$ has a density $\density{\al}$ on $\X=\RR^\dim$, then the TV norm is the $L^1$ norm on functions, $\norm{\al}_{\TV} = \int_\X |\density{\al}(x)| \d x = \norm{\density{\al}}_{L^1}$.

If $\al$ is discrete as in~\eqref{eq-div-disc-meas}, then the TV norm is the $\ell^1$ norm of vectors in $\RR^n$, $\norm{\al}_{\TV}=\sum_i |\a_i| = \norm{\a}_{\ell^1}$.

\end{example}
\begin{rem}[Strong vs. weak topology]
\(\Wass_1(\al,\be) \leq R \norm{\al-\be}_{\TV}\)
\end{rem}
\paragraph{Main families of $\phi$-divergences.}
\(\phi_\gamma(s)=\frac{s^\gamma-\gamma s+\gamma-1}{\gamma(\gamma-1)} \qquad(\gamma\neq0,1)\)
\[
\JS(\alpha,\beta)^2
=
\frac12\KL\!\left(\alpha\middle|\frac{\alpha+\beta}{2}\right)
+
\frac12\KL\!\left(\beta\middle|\frac{\alpha+\beta}{2}\right),
\]
\begin{rem}[$\phi$-divergences versus Bregman divergences]
\end{rem}
\begin{proposition}[Dual expression]
A $\phi$-divergence can be expressed using the Legendre transform
\(\phi^{*,\geq 0}(s) \eqdef \usup{t \in \RR^+} st - \phi(t)\)
(notice that we restrict the function to the positive real)
of $\phi$ as
\eql{\label{eq-dual-div}
\Divergm_\phi(\al|\be) = \usup{f: \X \rightarrow \RR} \int_\X f(x) \d\al(x) - \int_\X \phi^{*,\geq 0}(f(x)) \d\be(x).
}
which equivalently reads that the Legendre transform of $\Divergm_\phi(\cdot|\be)$ reads
\eql{\label{eq-legendre}
\foralls f \in \Cc(\Xx), \quad
\Divergm_\phi^*(f|\be) = \int_\X \phi^{*,\geq 0}(f(x)) \d\be(x).
}
\end{proposition}
\begin{proof}
We first consider the superlinear case $\phi'_\infty=+\infty$, so that $\Dd_\phi(\al|\be)=+\infty$ if $\al$ does not have a density $\rho \geq 0$ with respect to $\be$, $\d\al=\rho \d\be$. Thus the Legendre-Fenchel transform of $\Dd_\phi(\cdot|\be)$
reads
\begin{align*}
\Dd_\phi^*(f|\be) &= \usup{\rho \geq 0} \int_\Xx f(x) \rho(x)\d\be(x) - \int_\Xx \phi(\rho(x)) \d\be(x)\\
&= \int_\Xx \usup{\rho(x) \geq 0} \pa{ f(x) \rho(x) -\phi(\rho(x)) }\d\be(x)
= \int_\Xx \phi^{*,\geq 0}(f(x)) \d\be(x).
\end{align*}
Fenchel--Moreau then gives the displayed dual expression. For a general entropy, the same argument is applied to the perspective with its recession term; the singular part is exactly encoded by the effective domain of $\phi^{*,\geq0}$.
\end{proof}
\subsection{GANs via Duality}
\eq{
\umin{\th} \norm{\al_\th - \be}_B
= \umin{\th}
\usup{\f \in B} \int_{\X} \f(x) \d(\al_\th-\be)(x)
= \umin{\th}
\usup{\f \in B} \int_{\Zz} \f( g_\th(z)) \d\zeta - \frac{1}{m} \sum_j f(y_j).
}
\eq{
\umin{\th} \Divergm_\phi(\al_\th|\be)
= \umin{\th} \usup{\f} \int_\Xx f(x) \d\al_\th(x) - \Divergm_\phi^*(f|\be)
= \umin{\th} \usup{\f} \int_\Xx f(g_\th(z)) \d\zeta(z) -
\frac{1}{m}\sum_j \phi^*(f(y_j)).
}