\section{Dynamic Optimal Transport}
\label{sec-dynamic-optimal-transport}
\label{sec-wasserstein-flows}
\subsection{Evolutions over the Space of Measures}
\paragraph{Lagrangian and Eulerian descriptions.}
\begin{equation}
\frac{\d x(t)}{\d t} = v_t(x(t)), \label{eq:lagrangian-advection}
\end{equation}
\begin{equation}
\frac{\partial \alpha_t}{\partial t} + \diverg(v_t \alpha_t) = 0. \label{eq:eulerian-advection}
\end{equation}
\begin{equation}\label{eq:eulerian-advection-weak}
\int_0^1\!\int_{\RR^d}
\left(\partial_t\varphi(t,x)+\dotp{v_t(x)}{\nabla_x\varphi(t,x)}\right)
\d\alpha_t(x)\d t
=
0.
\end{equation}
This equation is obtained from~\eqref{eq:eulerian-advection} by integration by parts. Hence, for smooth positive densities, the classical and weak formulations are equivalent; the weak formulation is useful because it still makes sense for discrete measures whose particles evolve according to~\eqref{eq:lagrangian-advection}.

\begin{prop}[Lagrangian flows solve the continuity equation]\label{prop-lagrangian-flow-continuity}
Consider a smooth flow $T_t:\RR^d\to\RR^d$ and define $\alpha_t=(T_t)_\sharp\alpha_0$. Define the Eulerian velocity field by \(v_t(T_t(y))=\partial_t T_t(y).\)
Then $(\alpha_t,v_t)$ solves the continuity equation in the weak sense~\eqref{eq:eulerian-advection-weak}.
In particular, if $\alpha_0=\frac1n\sum_i\delta_{x_i(0)}$ is empirical, then $\alpha_t=\frac1n\sum_i\delta_{x_i(t)}$ is empirical as well, with particle velocities $\dot x_i(t)=v_t(x_i(t))$.
\end{prop}
\begin{proof}
Let $\varphi(t,x)$ be a smooth test function vanishing at $t=0$ and $t=1$. Since $\alpha_t=(T_t)_\sharp\alpha_0$, \(\frac{\d}{\d t}\int \varphi(t,x)\d\alpha_t(x) = \frac{\d}{\d t}\int \varphi(t,T_t(y))\d\alpha_0(y).\)
The chain rule gives
\[
\frac{\d}{\d t}\int \varphi(t,T_t(y))\d\alpha_0(y)
=
\int
\left(
\partial_t\varphi(t,T_t(y))
+\dotp{\nabla_x\varphi(t,T_t(y))}{\partial_t T_t(y)}
\right)
\d\alpha_0(y).
\]
Using the definition of $v_t$ and the push-forward relation, this equals
\[
\int
\left(
\partial_t\varphi(t,x)+\dotp{\nabla_x\varphi(t,x)}{v_t(x)}
\right)
\d\alpha_t(x).
\]
Integrating in time and using the boundary values of $\varphi$ gives~\eqref{eq:eulerian-advection-weak}.
\end{proof}
\paragraph{From measure evolutions to vector fields.}
For a given evolution $(\alpha_t)_t$, there are typically infinitely many velocity fields $v_t$ satisfying

\begin{equation}\label{eq:inverse-flow}
\partial_t\alpha_t+\diverg(\alpha_t v_t)=0.
\end{equation}
\(\Hh_\alpha=\enscond{v}{\diverg(\alpha v)=0}.\)
\paragraph{Dacorogna--Moser inversion.}
\begin{equation}\label{eq:dacorogna-moser}
v_t=-\frac{1}{\alpha_t}\nabla\Delta^{-1}(\partial_t\alpha_t),
\end{equation}
\paragraph{Least-square inversion and gradient structure.}
\begin{equation}\label{eq:least-square-field}
\umin{v}
\frac12\int_0^1\!\int_{\RR^d}\norm{v_t(x)}^2\,\d\alpha_t(x)\d t
\quad
\text{subject to}
\quad
\partial_t\alpha_t+\diverg(\alpha_t v_t)=0.
\end{equation}
\begin{prop}[Least-square velocities are gradients]\label{prop-least-square-gradient-velocity}
Assume that $\alpha_t=\rho_t\,\d x$ is a smooth positive density curve and that boundary terms vanish. The minimizer of~\eqref{eq:least-square-field}, if it exists, is a gradient field
\(v_t=\nabla\phi_t,\)
where $\phi_t$, unique up to an additive constant, solves the weighted Poisson equation
\begin{equation}\label{eq:least-square-field-explicit}
-\diverg(\rho_t\nabla\phi_t)=\partial_t\rho_t,
\qquad
v_t=-\nabla\Delta_{\alpha_t}^{-1}(\partial_t\alpha_t),
\quad
\Delta_{\alpha_t}\phi=\diverg(\alpha_t\nabla\phi).
\end{equation}
\end{prop}
\begin{proof}
Introduce a Lagrange multiplier $\phi_t$ for the continuity equation. The constrained problem has the formal saddle formulation
\[
\umin{v}\umax{\phi}
\int_0^1
\left[
\frac12\int_{\RR^d}\norm{v_t(x)}^2\,\d\alpha_t(x)
+\int_{\RR^d}\phi_t(x)\big(\diverg(\alpha_t v_t)(x)+\partial_t\alpha_t(x)\big)\d x
\right]\d t.
\]
Integrating by parts in the divergence term gives, for each $t$,
\[
\int \left(\frac12\norm{v_t}^2-\dotp{\nabla\phi_t}{v_t}\right)\d\alpha_t
+
\int \phi_t\,\partial_t\alpha_t.
\]
The pointwise minimizer in $v_t$ is therefore $v_t=\nabla\phi_t$. Substituting this into the constraint $\partial_t\rho_t+\diverg(\rho_t v_t)=0$ gives the weighted Poisson equation in~\eqref{eq:least-square-field-explicit}. The inverse notation is just a shorthand for solving this equation on zero-mean right-hand sides, modulo additive constants.
\end{proof}
\subsection{Benamou--Brenier dynamic formulation of OT}
\label{sec-benamou-brenier-dynamic}
\begin{thm}[Benamou--Brenier]\label{thm-benamou-brenier}
For probability measures $\alpha_0,\alpha_1\in\Pp_2(\RR^d)$,
\begin{equation}\label{eq:benamou-brenier}
\Wass_2^2(\alpha_0,\alpha_1)
=
\inf_{(\alpha_t,v_t)}
\int_0^1\!\int_{\RR^d}\norm{v_t(x)}^2\,\d\alpha_t(x)\d t,
\end{equation}
where the infimum is over $(\alpha_t,v_t)$ solving $\partial_t\alpha_t+\nabla\!\cdot(\alpha_t v_t)=0$ with $\alpha_{t=0}=\alpha_0$ and $\alpha_{t=1}=\alpha_1$. If $\alpha_0$ has a density and $T$ is the optimal Monge map $T_\sharp\alpha_0=\alpha_1$, the minimizer is
\begin{equation}\label{eq:static-to-dynamic}
\alpha_t=((1-t)\Id+tT)_\sharp\alpha_0,
\qquad
v_t\big((1-t)x+tT(x)\big)=T(x)-x.
\end{equation}
\end{thm}
\begin{proof}
For the inequality ``dynamic $\leq$ static'', assume first that a Monge map $T$ exists and define $(\alpha_t,v_t)$ by~\eqref{eq:static-to-dynamic}. Since the Lagrangian velocity $T(x)-x$ is independent of $t$,
\(\int_0^1\!\int\norm{v_t}^2\,\d\alpha_t\d t = \int\norm{T(x)-x}^2\,\d\alpha_0(x),\)
so the dynamic cost is no larger than the static Monge cost. Without a Monge map, the same construction is made with an optimal coupling $\pi$: sample $(X,Y)\sim\pi$ and move along the straight path $\gamma_{X,Y}(t)=(1-t)X+tY$. This path measure has action $\int\norm{x-y}^2\d\pi(x,y)$; projecting path velocities onto their conditional mean at time $t$ gives an admissible Eulerian velocity with no larger action, so the dynamic value is no larger than the Kantorovich value.

Conversely, for a smooth deterministic path, take the flow $T_t$ defined by $\dot T_t=v_t\circ T_t$ and $T_0=\Id$. Then $\alpha_t=(T_t)_\sharp\alpha_0$ and $(T_1)_\sharp\alpha_0=\alpha_1$. Jensen's inequality gives
\(\norm{T_1(x)-x}^2 \leq \int_0^1\norm{v_t(T_t(x))}^2\d t.\)
After integration with respect to $\alpha_0$, the Monge cost is bounded above by the dynamic action. For general finite-energy solutions of the continuity equation, the superposition principle lifts the curve to a probability measure on absolutely continuous paths; applying Jensen's inequality pathwise gives a coupling of the endpoints whose quadratic cost is no larger than the action. Thus the Kantorovich value is bounded above by the dynamic value.
\end{proof}
\subsubsection{Convex moment-based reformulation}
Although~\eqref{eq:benamou-brenier} is not jointly convex in $(\alpha_t,v_t)$, it becomes convex after replacing velocities by the momentum measure $m_t=v_t\alpha_t$ and using the perspective action. In the absolutely continuous case $\alpha_t=\rho_t\,\d x$ and $m_t(x)=\rho_t(x)v_t(x)$, this reads

\begin{equation}\label{eq:benamou-brenier-convex}
\Wass_2^2(\alpha_0,\alpha_1)
=
\inf_{\substack{\partial_t\rho_t+\diverg m_t=0\\
\rho_{t=0}\d x=\alpha_0,\ \rho_{t=1}\d x=\alpha_1}}
\int_0^1\!\int_{\RR^d}
\frac{\norm{m_t(x)}^2}{\rho_t(x)}\,\d x\,\d t,
\end{equation}
\paragraph{Dual Hamilton--Jacobi formulation.}
The momentum formulation also has a useful dual. It turns the least-action problem into a Hamilton--Jacobi subsolution inequality for a scalar potential, with equality on the part of space-time actually visited by the optimal curve.

\begin{prop}[Dual Benamou--Brenier problem]\label{prop-benamou-brenier-dual}
Assume, for simplicity, that the densities are smooth, compactly supported, and that boundary terms vanish. Then the convex dynamic value~\eqref{eq:benamou-brenier-convex} has the dual formulation
\begin{equation}\label{eq:benamou-brenier-dual}
\Wass_2^2(\alpha_0,\alpha_1)
=
\sup_{\phi}
\enscond{
\int_{\RR^d}\phi_1\,\d\alpha_1
-
\int_{\RR^d}\phi_0\,\d\alpha_0
}{
\partial_t\phi_t+\frac14\norm{\nabla\phi_t}^2\leq 0
},
\end{equation}
where the supremum is over smooth test potentials $\phi:[0,1]\times\RR^d\to\RR$. In the general metric-measure statement, the same formula is understood after relaxation: $\phi$ is a Hamilton--Jacobi subsolution, for instance in the viscosity sense, and finite-dimensional discretizations follow directly from Fenchel--Rockafellar duality.
If $(\rho,m)$ and $\phi$ are smooth primal and dual optimizers, then
\begin{equation}\label{eq:bb-primal-dual-relation}
m_t=\frac{\rho_t}{2}\nabla\phi_t,
\qquad
\partial_t\phi_t+\frac14\norm{\nabla\phi_t}^2=0
\quad\text{on }\{\rho_t>0\}.
\end{equation}
Equivalently, the optimal Eulerian velocity is $v_t=m_t/\rho_t=\nabla\phi_t/2$.
\end{prop}
\begin{proof}
Let $(\rho,m)$ satisfy the continuity equation and let $\phi$ be smooth. Multiplying the constraint by $\phi$ and integrating by parts gives
\[
\int_{\RR^d}\phi_1\,\d\alpha_1
-
\int_{\RR^d}\phi_0\,\d\alpha_0
=
\int_0^1\!\int_{\RR^d}
\left(\rho_t\,\partial_t\phi_t+\dotp{m_t}{\nabla\phi_t}\right)\d x\d t.
\]
If $\partial_t\phi_t+\norm{\nabla\phi_t}^2/4\leq0$, Young's inequality yields, pointwise, \(\rho\,\partial_t\phi+\dotp{m}{\nabla\phi} \leq -\frac{\rho}{4}\norm{\nabla\phi}^2+\dotp{m}{\nabla\phi} \leq \frac{\norm m^2}{\rho},\)
with the same perspective convention when $\rho=0$. Hence the dual objective of every feasible $\phi$ is no larger than the action of every feasible primal pair.

Conversely, introduce $\phi$ as a Lagrange multiplier for $\partial_t\rho+\diverg m=0$. After integration by parts, and up to the fixed endpoint contribution in~\eqref{eq:benamou-brenier-dual}, the pointwise minimization over $(\rho,m)$ contains
\(\frac{\norm m^2}{\rho}-\rho\,\partial_t\phi-\dotp{m}{\nabla\phi}.\)
For fixed $\rho>0$, its minimum over $m$ is attained at $m=\rho\nabla\phi/2$ and equals $-\rho(\partial_t\phi+\norm{\nabla\phi}^2/4)$. The infimum over $\rho\geq0$ is finite exactly when the Hamilton--Jacobi inequality holds, in which case the pointwise infimum is $0$. Fenchel--Rockafellar duality then gives no duality gap. Equality in the two pointwise inequalities gives~\eqref{eq:bb-primal-dual-relation}.
\end{proof}
\(\frac{\d}{\d t}\phi_t(\gamma(t)) = \partial_t\phi_t(\gamma(t))+\dotp{\nabla\phi_t(\gamma(t))}{\dot\gamma(t)} \leq \norm{\dot\gamma(t)}^2,\)
where the last inequality uses $\partial_t\phi+\norm{\nabla\phi}^2/4\leq0$ and Young's inequality. Integrating and minimizing over curves gives

\(\phi_1(y)-\phi_0(x)\leq\norm{x-y}^2.\)
Thus $(\phi_0,\phi_1)$ is a feasible static Kantorovich dual pair for the quadratic cost. At optimality the inequality is saturated on the endpoint pairs connected by the primal characteristics, which is the Eulerian version of complementary slackness.

\paragraph{Proximal splitting.}
\[
J(\rho,m)=
\begin{cases}
\norm{m}^2/\rho, & \rho>0,\\
0, & (\rho,m)=(0,0),\\
+\infty, & \text{otherwise}.
\end{cases}
\]
\[
\mathcal F(U)=\int_0^1\!\int_{\RR^d}J(\rho_t(x),m_t(x))\,\d x\,\d t,
\qquad
\mathcal C=\enscond{(\rho,m)}{\partial_t\rho+\diverg m=0,\ \rho_0\d x=\alpha_0,\ \rho_1\d x=\alpha_1},
\]
\(\min_U\; \mathcal F(U)+\mathcal G(U).\)
\[
\prox_{\tau\mathcal F}(\bar U)
=
\argmin_U
\frac12\norm{U-\bar U}_{\mathsf H}^2+\tau\mathcal F(U),
\qquad
\prox_{\tau\mathcal G}(\bar U)
=
\argmin_{U\in\mathcal C}
\frac12\norm{U-\bar U}_{\mathsf H}^2.
\]
For the standard product metric, the first map is local in $(t,x)$: it is the proximal operator of the convex perspective $J$. The second map is the orthogonal projection onto the affine set defined by the divergence equation and endpoint constraints. Douglas--Rachford then alternates these two simple operations:

\[
\begin{aligned}
U^{k+1} &= \prox_{\tau\mathcal F}(Z^k),\\
\widetilde U^{k+1} &= \prox_{\tau\mathcal G}(2U^{k+1}-Z^k),\\
Z^{k+1} &= Z^k+\widetilde U^{k+1}-U^{k+1}.
\end{aligned}
\]
\paragraph{Path-space formulation.}\label{rem-bb-path-space}
Let $\Ss=C([0,1];\RR^d)$ be the space of continuous paths endowed with the uniform topology. For $t\in[0,1]$ define the evaluation map

\(P_t:\Ss\to\RR^d, \qquad P_t(\gamma)=\gamma(t).\)
\[
\Wass_2^2(\alpha_0,\alpha_1)
=
\inf_{M\in\Pp(\Ss)}
\enscond{
\int_{\Ss}\!\int_0^1\norm{\dot\gamma(t)}^2\d t\,\d M(\gamma)
}{
(P_0)_\sharp M=\alpha_0,\ (P_1)_\sharp M=\alpha_1
}.
\]
If $\alpha_0$ has a density, the minimizer $M^*$ is unique. Its time marginals reproduce the optimal curve: $\alpha_t=(P_t)_\sharp M^*$ for all $t$. Furthermore, for a.e. $t$, denoting $Q_t(\gamma)=\dot\gamma(t)$ on absolutely continuous paths, the conditional law of the velocity is deterministic:

\((P_t,Q_t)_\sharp M^*(\d x,\d q) = \alpha_t(\d x)\delta_{v_t^*(x)}(\d q),\)
where $v_t^*$ is the optimal velocity field in the Benamou--Brenier formulation. Hence $M^*$ concentrates on straight-line geodesics and, for a.e. $t$, assigns exactly one direction at $\alpha_t$-a.e. spatial point.

\subsection{Extensions of the dynamic formulation}
\label{sec-bb-extensions}
\paragraph{Convex action densities.}\label{rem-generalized-bb}
For a convex admissible density range $I$ and an extended-valued convex l.s.c. action $A:I\times\RR^d\to[0,+\infty]$, define
\[
\mathcal A_A(\rho_0,\rho_1)
=
\inf_{\substack{\partial_t\rho_t+\nabla\!\cdot m_t=0\\
\rho_{t=0}=\rho_0,\ \rho_{t=1}=\rho_1}}
\int_0^1\!\int A(\rho_t,m_t)\,\d x\,\d t .
\]
If $A$ is even, $q$-homogeneous in $m$, and $A(\rho,m)=0$ only for $m=0$, then $\mathcal A_A^{1/q}$ is an extended distance on finite-action components of the equal-mass class; otherwise it should only be viewed as an action principle.

\begin{example}[Wasserstein-\texorpdfstring{$p$}{p} action]
\[
\Wass_p^p(\alpha_0,\alpha_1)
=
\inf_{\substack{\partial_t\alpha_t+\nabla\!\cdot(\alpha_t v_t)=0\\
\alpha_{t=0}=\alpha_0,\ \alpha_{t=1}=\alpha_1}}
\int_0^1\!\int_{\RR^d} |v_t(x)|^p\,\d\alpha_t(x)\,\d t.
\]
When $\alpha_t=\rho_t\,\d x$ and $m_t=\rho_t v_t$, this becomes the convex perspective action

\(\int_0^1\!\int_{\RR^d} \frac{|m_t(x)|^p}{\rho_t(x)^{p-1}}\,\d x\,\d t, \qquad \partial_t\rho_t+\nabla\!\cdot m_t=0,\)
with the perspective convention at $\rho=0$. Its $p$-th root is exactly $\Wass_p$.
\end{example}

\begin{example}[Concave-mobility action]
A second example changes the mobility: $|m|^2/\rho$ is replaced by $|m|^2/\theta(\rho)$ for a concave $\theta\geq0$,
\[
\mathsf W_\theta^2(\rho_0\d x,\rho_1\d x)
=
\inf_{\partial_t\rho_t+\nabla\!\cdot m_t=0}
\int_0^1\!\int \frac{|m_t|^2}{\theta(\rho_t)}\,\d x\,\d t .
\]
Concavity makes the integrand convex; $\theta(\rho)=\rho$ gives $\Wass_2$, while $\theta(\rho)=\rho^\gamma$ or $\rho(1-\rho/M)$ models nonlinear or volume-filling mobility. For $F(\rho)=\int U(\rho)+\int V\rho$, the formal gradient flow is $\partial_t\rho=\nabla\cdot(\theta(\rho)(U''(\rho)\nabla\rho+\nabla V))$: $\theta(\rho)=\rho$, $U(\rho)=\rho^m/(m-1)$ yields $\Delta(\rho^m)$, while entropy with $\theta(\rho)=\rho^\gamma$, $0<\gamma<1$, yields $(1/\gamma)\Delta(\rho^\gamma)$. The finite-state version, where the mobility is dictated by a reversible Markov chain, is developed in Section~\ref{sec-discrete-wasserstein-markov}.
\end{example}

\paragraph{Dynamic unbalanced OT.}
\label{sec-unbalanced-ot}
\[
\partial_t\rho_t+\nabla\!\cdot m_t=s_t,
\qquad
\int_0^1\!\int
\left(\frac{|m_t|^2}{\rho_t}+\kappa^2\frac{s_t^2}{\rho_t}\right)\d x\,\d t,
\]
\[
\mathcal A_\kappa(\rho,m,s)
\eqdef
\int
\left(
\frac{\norm{\dot m}^2}{\dot\rho}
+
\kappa^2\frac{\dot s^2}{\dot\rho}
\right)\d\lambda,
\qquad
(\dot\rho,\dot m,\dot s)
=
\left(\frac{\d\rho}{\d\lambda},\frac{\d m}{\d\lambda},\frac{\d s}{\d\lambda}\right),
\]
where $\lambda$ dominates $\rho$ and the total variations of $m$ and $s$, and the value is independent of this choice. The convention is $0/0=0$ and $a/0=+\infty$ for $a>0$, so finite action forces both the flux and source to be absolutely continuous with respect to the transported mass.

\begin{prop}[Static/dynamic equivalence for unbalanced OT]\label{prop-static-dynamic-unbalanced}
Fix the action above and let $\CW_\kappa$ be the cone value of Theorem~\ref{thm-cone-unbalanced-ot} with the cone metric normalized to the same growth scale $\kappa$. For nonnegative finite measures $\alpha_0,\alpha_1$ on $\RR^d$, the dynamic value
\begin{equation}\label{eq-dynamic-unbalanced-ot}
\WFR_\kappa^2(\alpha_0,\alpha_1)
\eqdef
\inf_{\substack{\partial_t\rho_t+\nabla\cdot m_t=s_t\\
\rho_0=\alpha_0,\ \rho_1=\alpha_1}}
\int_0^1
\mathcal A_\kappa(\rho_t,m_t,s_t)\,\d t
\end{equation}
equals the static cone formulation $\CW_\kappa(\alpha_0,\alpha_1)$. Hence the static unbalanced problem and the balance-equation least-action problem define the same geodesic distance.
\end{prop}
\begin{proof}
The cone construction turns variation of mass into radial motion and spatial transport into angular motion on $\mathfrak C[\RR^d]$. Applying the Benamou--Brenier theorem on the cone to the lifted endpoint measures gives a dynamic least-action problem on $\mathfrak C[\RR^d]$ whose static value is the cone value $\CW_\kappa$ of Theorem~\ref{thm-cone-unbalanced-ot}. This is the standard static/dynamic identification for the Hellinger--Kantorovich and Wasserstein--Fisher--Rao metrics.

Projecting a cone curve back to the base space with the weight $r^2$ produces a measure curve $\rho_t$, a spatial flux $m_t$ and a source term $s_t$ satisfying the balance equation. With the matching normalization of the cone metric, the cone kinetic energy decomposes exactly into the perspective action $\mathcal A_\kappa$ in~\eqref{eq-dynamic-unbalanced-ot}. Conversely, any finite-action triple $(\rho_t,m_t,s_t)$ can be lifted to a cone curve whose radial velocity realizes the growth term and whose spatial velocity realizes the transport term, with the same action after relaxation. The two infima are therefore equal; lower semicontinuity gives the general finite-measure statement from the smooth positive case.
\end{proof}
\subsection{Discrete Wasserstein Geometries on Markov Chains}
\label{sec-discrete-wasserstein-markov}
\[
\Sigma_n\eqdef\left\{p\in\RR_+^n:\sum_i p_i=1\right\}.
\]
\(\operatorname{Ent}_\pi(\rho)\eqdef\sum_i\pi_i\rho_i\log\rho_i.\)
\[
\theta(a,b)\eqdef
\begin{cases}
\displaystyle\frac{a-b}{\log a-\log b}, & a\neq b,\\[.4em]
a, & a=b,
\end{cases}
\]
\begin{equation}
\label{eq-discrete-markov-onsager}
(\mathcal K_\rho\psi)_i
\eqdef
\sum_j K_{ij}\theta(\rho_i,\rho_j)(\psi_i-\psi_j).
\end{equation}
\begin{equation}
\label{eq-discrete-markov-action}
\mathcal A(\rho,\psi)
\eqdef
\frac12\sum_{i,j}\pi_iK_{ij}\theta(\rho_i,\rho_j)(\psi_i-\psi_j)^2.
\end{equation}
\begin{equation}
\label{eq-discrete-markov-distance}
\mathcal W_K^2(\rho^0,\rho^1)
\eqdef
\inf_{\rho_t,\psi_t}
\int_0^1\mathcal A(\rho_t,\psi_t)\,\d t,
\qquad
\dot\rho_t+\mathcal K_{\rho_t}\psi_t=0,
\end{equation}
\begin{equation}
\label{eq-discrete-markov-jko}
\rho^{k+1}\in\argmin_\rho
\frac{1}{2\tau}\mathcal W_K^2(\rho,\rho^k)+\operatorname{Ent}_\pi(\rho).
\end{equation}
\(\frac{\rho^{k+1}-\rho^k}{\tau} =-\mathcal K_{\rho^k}\log\rho^k+O(\tau) =K\rho^k+O(\tau),\)
\begin{prop}[Entropy gradient flow of a reversible Markov chain]
\label{prop-discrete-markov-entropy-gradient}
Let \(K\) be reversible with invariant law \(\pi\). The gradient flow of \(\operatorname{Ent}_\pi\) for the metric \(\mathcal W_K\) is the forward equation of the Markov chain,
\begin{equation}
\label{eq-discrete-markov-gradient-flow}
\dot\rho_i(t)=\sum_jK_{ij}\bigl(\rho_j(t)-\rho_i(t)\bigr).
\end{equation}
Equivalently, for the masses \(p_i(t)=\pi_i\rho_i(t)\), this is \(\dot p_i(t)=\sum_j(p_j(t)K_{ji}-p_i(t)K_{ij})\).
\end{prop}
\begin{proof}
The first variation of the entropy, with respect to the weighted pairing \(\sum_i\pi_i\xi_i\varphi_i\), is \(\log\rho_i+1\). Constants do not contribute to \(\mathcal K_\rho\), hence the metric gradient-flow equation is
\(\dot\rho=-\mathcal K_\rho\log\rho.\)
Using the identity
\(\theta(a,b)(\log a-\log b)=a-b,\)
one obtains, componentwise, \(\dot\rho_i =-\sum_jK_{ij}\theta(\rho_i,\rho_j)(\log\rho_i-\log\rho_j) = \sum_jK_{ij}(\rho_j-\rho_i),\)
which is the density form of the Kolmogorov forward equation. Multiplying by \(\pi_i\) and using detailed balance gives the equation for the probability masses.
\end{proof}
\paragraph{Closed forms in dimensions two and three.}
\begin{equation}
\label{eq-two-state-markov-distance}
\mathcal W_K(p,q)
=
\left|\int_s^r \frac{\d u}{\sqrt{\theta(u,1-u)}}\right|,
\qquad
0<r,s<1.
\end{equation}
\(a_{ij}(p)\eqdef\frac12\theta(p_i,p_j), \qquad 1\leq i<j\leq3.\)
\begin{equation}
\label{eq-three-state-markov-norm}
\|u\|_p^2
=
\min_{q_{12},q_{13},q_{23}}
\left\{
\frac{q_{12}^2}{a_{12}}+
\frac{q_{13}^2}{a_{13}}+
\frac{q_{23}^2}{a_{23}}
\right\},
\end{equation}
\(u_1+q_{12}+q_{13}=0, \qquad u_2-q_{12}+q_{23}=0, \qquad u_3-q_{13}-q_{23}=0.\)
\(q_{12}^*=\frac{u_2/a_{23}-u_1/a_{13}}{D}, \qquad q_{13}^*=-u_1-q_{12}^*, \qquad q_{23}^*=q_{12}^*-u_2,\)
\begin{equation}
\label{eq-three-state-markov-distance}
\mathcal W_K^2(p^0,p^1)
=
\inf_{p_t\in\operatorname{int}(\Sigma_3)}
\int_0^1\|\dot p_t\|_{p_t}^2\,\d t,
\qquad
p_0=p^0,
\quad p_1=p^1.
\end{equation}
