\section{Generalized OT Problems}
\label{sec-generalized-ot-problems}
\subsection{OT Barycenters}
\label{sec-barycenters}
\paragraph{Fr\'echet means.}
For discrete input histograms $\{\b_s\}_{s=1}^S$, with $\b_s \in \simplex_{n_s}$, and weights $\la \in \simplex_S$, a Wasserstein barycenter can be computed by minimizing

\eql{\label{eq-wass-discr}
\umin{\a \in \simplex_n} \sum_{s=1}^S \la_s \MKD_{\C_s}(\a,\b_s),
}
where the cost matrices $\C_s \in \RR^{n \times n_s}$ are prescribed.

Given a set of input measures $(\be_s)_s$ defined on some space $\X$, the barycenter problem becomes

\eql{\label{eq-barycenter-generic}
\umin{\al \in \Mm_+^1(\X)} \sum_{s=1}^S \la_s \MK_{\c}(\al,\be_s).
}
\begin{prop}[Convexity of the OT cost]\label{prop-barycenter-ot-cost-convexity}
The map $(\al,\be)\mapsto\MK_\c(\al,\be)$ is convex.
\end{prop}
\begin{proof}
Let $(\al_0,\be_0)$ and $(\al_1,\be_1)$ be two pairs of probability measures and let $t\in[0,1]$. For $\eta>0$, choose couplings $\pi_i\in\Couplings(\al_i,\be_i)$ such that
\(\int c\d\pi_i\leq\MK_\c(\al_i,\be_i)+\eta \qquad (i=0,1).\)
Then $\pi_t=(1-t)\pi_0+t\pi_1$ is a coupling between $(1-t)\al_0+t\al_1$ and $(1-t)\be_0+t\be_1$. Hence
\(\MK_\c((1-t)\al_0+t\al_1,(1-t)\be_0+t\be_1) \leq (1-t)\MK_\c(\al_0,\be_0)+t\MK_\c(\al_1,\be_1)+\eta.\)
Letting $\eta\to0$ gives the claim.
\end{proof}
\paragraph{One-dimensional case.}
\begin{prop}[Quantile barycenters on the line]\label{prop-quantile-barycenters}
For $\X=\RR$ and $c(x,y)=|x-y|^2$, the quantile function of a Wasserstein barycenter is the weighted average of the input quantile functions:
\(\cumul{\al^\star}^{-1}(r) = \sum_{s=1}^S\la_s\cumul{\be_s}^{-1}(r), \qquad r\in[0,1].\)
\end{prop}
\begin{proof}
The one-dimensional formula~\eqref{eq-wass-cumul} gives \(\sum_s\la_s\Wass_2^2(\al,\be_s) = \int_0^1 \sum_s\la_s \abs{\cumul{\al}^{-1}(r)-\cumul{\be_s}^{-1}(r)}^2 \d r.\)
The minimization decouples pointwise in $r$. For each fixed $r$, the minimizer of $z\mapsto\sum_s\la_s|z-\cumul{\be_s}^{-1}(r)|^2$ is the weighted average $\sum_s\la_s\cumul{\be_s}^{-1}(r)$. This function is nondecreasing because it is a positive weighted sum of nondecreasing quantile functions, hence it is a valid quantile function.
\end{proof}
\paragraph{Gaussian case.}
\paragraph{Sinkhorn for barycenters.}
\eql{\label{eq-entropic-bary}
\umin{\a\in\simplex_n}
\sum_{s=1}^S\la_s\MKD_{\C_s}^\epsilon(\a,\b_s)
}
\eql{\label{eq-bary-entropy-couplings}
\umin{(\P_s)_s}
\enscond{
\epsilon\sum_s\la_s\KLD(\P_s|\K_s)
}{
\foralls s,\ \transp{\P_s}\ones_n=\b_s,
\quad
\P_1\ones_{n_1}=\ldots=\P_S\ones_{n_S}
},
}
where $\K_s\eqdef e^{-\C_s/\epsilon}$. The barycenter $\a$ is implicitly encoded in the common row marginal

\(\a=\P_1\ones_{n_1}=\ldots=\P_S\ones_{n_S}.\)
\eql{\label{eq-bary-opt}
\P_s=\diag(\uD_s)\K_s\diag(\vD_s),
}
\begin{align}
\foralls s\in\range{1,S},\qquad
\itt{\vD}_s
&\eqdef
\frac{\b_s}{\transp{\K_s}\it{\uD}_s},
\label{eq-sinkhorn-bary}
\\
\foralls s\in\range{1,S},\qquad
\itt{\uD}_s
&\eqdef
\frac{\itt{\a}}{\K_s\itt{\vD}_s},
\label{eq-sinkhorn-bary-2}
\\
\qwhereq
\itt{\a}
&\eqdef
\prod_s(\K_s\itt{\vD}_s)^{\la_s}.
\label{eq-sinkhorn-bary-3}
\end{align}
\begin{prop}[Dual of entropic barycenters]\label{prop-dual-entropic-barycenters}
The optimal scalings in~\eqref{eq-bary-opt} can be written as $(\uD_s,\vD_s)=(e^{\fD_s/\epsilon},e^{\gD_s/\epsilon})$, where $(\fD_s,\gD_s)_s$ solve the dual problem
\eql{\label{eq-dual-bary-entropy}
\umax{(\fD_s,\gD_s)_s}
\enscond{
\sum_s\la_s
\left(
\dotp{\gD_s}{\b_s}
-\epsilon\dotp{\K_s e^{\gD_s/\epsilon}}{e^{\fD_s/\epsilon}}
\right)
}{
\sum_s\la_s\fD_s=0
}.
}
\end{prop}
\begin{proof}
Introduce Lagrange multipliers in~\eqref{eq-bary-entropy-couplings}:
\begin{align*}
\umin{(\P_s)_s,\a}
\umax{(\fD_s,\gD_s)_s}
\sum_s\la_s\Big(
\epsilon\KLD(\P_s|\K_s)
+\dotp{\a-\P_s\ones_{n_s}}{\fD_s}
+\dotp{\b_s-\transp{\P_s}\ones_n}{\gD_s}
\Big).
\end{align*}
Strong duality holds, so one can exchange the minimum and maximum. The minimization with respect to $\a$ gives the constraint $\sum_s\la_s\fD_s=0$, and the minimization with respect to $\P_s$ gives the Legendre transform of $\KLD(\cdot|\K_s)$:
\[
\umax{(\fD_s,\gD_s)_s}
\sum_s\la_s
\left[
\dotp{\gD_s}{\b_s}
-\epsilon
\KLD^*\left(\frac{\fD_s\oplus\gD_s}{\epsilon}\middle|\K_s\right)
\right],
\qquad
\sum_s\la_s\fD_s=0.
\]
The separable conjugate is
\eql{\label{eq-legendre-kl-bary}
\KLD^*(\VectMode{U}|\K)
=
\sum_{i,j}\K_{i,j}\big(e^{\VectMode{U}_{i,j}}-1\big),
}
because for $k>0$, \(\sup_{r\geq0} ur-\big(r\log(r/k)-r+k\big) = k(e^u-1),\)
and the case $k=0$ follows by lower semicontinuity. Dropping constants independent of $(\fD_s,\gD_s)_s$ gives~\eqref{eq-dual-bary-entropy}. The coordinate maximization in $\gD_s$ gives~\eqref{eq-sinkhorn-bary}; the block maximization in all $(\fD_s)_s$ gives the common marginal~\eqref{eq-sinkhorn-bary-3} and then~\eqref{eq-sinkhorn-bary-2}.
\end{proof}
\subsection{Multimarginal OT}
\label{sec-multimarginal-ot}
\label{subsec:multi-marginal}
\paragraph{Definition and basic structure.}
The multi-marginal formulation replaces a coupling between two measures by a joint distribution with several prescribed marginals. Given measures $(\al_s)_{s=1}^S$ on spaces $(\X_s)_{s=1}^S$ and a cost $c:\X_1\times\cdots\times\X_S\to\RR$, the problem reads

\(\inf_{\pi\in\Couplings(\al_1,\ldots,\al_S)} \int_{\X_1\times\cdots\times\X_S} c(x_1,\ldots,x_S)\d\pi(x_1,\ldots,x_S),\)
where $\Couplings(\al_1,\ldots,\al_S)$ is the set of probability measures whose $s$-th marginal is $\al_s$. This is still a linear program in the discrete setting, but its ambient tensor has size $\prod_s n_s$.

\paragraph{Monge structure and splitting-set twist.}
\begin{defn}[Twist on splitting sets]\label{def-twist-splitting-sets}
Fix $x_1\in\X_1$. A set $M\subset\X_2\times\cdots\times\X_S$ is a $c$-splitting set at $x_1$ if there exist functions $u_s:\X_s\to\RR\cup\{-\infty\}$, for $s=2,\ldots,S$, such that
\(\sum_{s=2}^S u_s(x_s)\leq c(x_1,x_2,\ldots,x_S)\)
for all $(x_2,\ldots,x_S)$, with equality on $M$. Assume $c$ is differentiable in $x_1$. The cost is twisted on splitting sets if, for every $x_1$ and every $c$-splitting set $M$ at $x_1$, the map
\((x_2,\ldots,x_S) \longmapsto \nabla_{x_1}c(x_1,x_2,\ldots,x_S)\)
is injective on $M$.
\end{defn}
\begin{prop}[Multi-marginal Monge structure]\label{prop-multimarginal-monge-structure}
Assume that $\X_s\subset\RR^d$, that $c$ is continuous and differentiable with respect to $x_1$, and that $c$ is twisted on splitting sets. Suppose that Kantorovich dual maximizers $(\varphi_s)_{s=1}^S$ exist and that $\varphi_1$ is differentiable $\al_1$-a.e. If $\al_1$ is absolutely continuous, then every optimal plan $\pi^\star\in\Couplings(\al_1,\ldots,\al_S)$ is concentrated on the graph of maps
\(\pi^\star=(\Id,\T_2,\ldots,\T_S)_\sharp\al_1, \qquad (\T_s)_\sharp\al_1=\al_s.\)
In particular, under these hypotheses the optimizer is unique.
\end{prop}
\begin{proof}
Let $(\varphi_s)_{s=1}^S$ be optimal dual potentials, so that $\sum_s\varphi_s(x_s)\leq c(x_1,\ldots,x_S)$ with equality on $\supp(\pi^\star)$. Fix a point $x_1$ where $\varphi_1$ is differentiable and consider the fiber
\(M(x_1)=\{(x_2,\ldots,x_S):(x_1,x_2,\ldots,x_S)\in\supp(\pi^\star)\}.\)
Indeed, set $u_2=\varphi_2+\varphi_1(x_1)$ and $u_s=\varphi_s$ for $s\geq3$. Dual feasibility gives $\sum_{s=2}^S u_s(x_s)\leq c(x_1,x_2,\ldots,x_S)$, and equality holds on $M(x_1)$. If $(x_2,\ldots,x_S)\in M(x_1)$, the function
\(z\longmapsto c(z,x_2,\ldots,x_S)-\sum_{s=2}^S\varphi_s(x_s)\)
touches $\varphi_1$ from above at $z=x_1$. Differentiating at this contact point gives \(\nabla\varphi_1(x_1)=\nabla_{x_1}c(x_1,x_2,\ldots,x_S).\)
All points in the fiber therefore have the same value of $\nabla_{x_1}c$. Twist on splitting sets makes the fiber a singleton for $\al_1$-a.e. $x_1$. Disintegrating $\pi^\star$ with respect to its first marginal gives Dirac conditional measures, hence measurable maps $(\T_2,\ldots,\T_S)$. If $\pi^1$ and $\pi^2$ are two optimal plans, their average is also optimal. The conditional measure of this average over $x_1$ is the average of the two Dirac conditionals, and it must again be a Dirac mass by the preceding argument. Hence the two Dirac masses coincide for $\al_1$-a.e. $x_1$, proving uniqueness.
\end{proof}
\paragraph{Coulomb cost and density-functional theory.}
\(c_{\mathrm{Coul}}(x_1,\ldots,x_N) \eqdef \sum_{1\leq i<j\leq N}\frac{1}{\norm{x_i-x_j}},\)
\[
V_{\mathrm{ee}}^{\mathrm{SCE}}[\rho]
\eqdef
\inf_{\pi\in\Couplings(\mu,\ldots,\mu)}
\int_{(\RR^3)^N}
c_{\mathrm{Coul}}(x_1,\ldots,x_N)
\d\pi(x_1,\ldots,x_N).
\]
\(\pi=(\Id,\T_2,\ldots,\T_N)_\sharp\mu, \qquad (\T_i)_\sharp\mu=\mu,\)
\begin{prop}[Cyclic co-motion plans]\label{prop-cyclic-comotion-plans}
Let $\mu\in\Pp(\RR^3)$ and let $\T:\RR^3\to\RR^3$ be measurable with $\T_\sharp\mu=\mu$ and $\T^N=\Id$ $\mu$-a.e. Set $\T^0=\Id$ and
\(\pi_\T=(\T^0,\T^1,\ldots,\T^{N-1})_\sharp\mu.\)
Then $\pi_\T\in\Couplings(\mu,\ldots,\mu)$. If $R_\sigma(x_1,\ldots,x_N)=(x_{\sigma(1)},\ldots,x_{\sigma(N)})$ for $\sigma\in\Perm(N)$, then
\(\bar\pi_\T\eqdef\frac1{N!}\sum_{\sigma\in\Perm(N)}(R_\sigma)_\sharp\pi_\T\)
is a symmetric admissible plan and
\[
\int c_{\mathrm{Coul}}\d\bar\pi_\T
=
\int c_{\mathrm{Coul}}\d\pi_\T
=
\int_{\RR^3}
\sum_{0\leq i<j\leq N-1}
\frac{1}{\norm{\T^i(x)-\T^j(x)}}\d\mu(x).
\]
\end{prop}
\begin{proof}
Since $\T_\sharp\mu=\mu$, all iterates $\T^i$ preserve $\mu$, so every marginal of $\pi_\T$ is $\mu$. The plan $\bar\pi_\T$ is an average of coordinate permutations of $\pi_\T$, hence has the same marginals and is invariant under coordinate permutations. The Coulomb cost is symmetric in its arguments, so its integral is unchanged by each $R_\sigma$. The last identity follows by evaluating $c_{\mathrm{Coul}}$ on the graph $(x,\T(x),\ldots,\T^{N-1}(x))$.
\end{proof}
Proposition~\ref{prop-multimarginal-monge-structure} explains the general mechanism that can force graph solutions, while Proposition~\ref{prop-cyclic-comotion-plans} records the additional equal-marginal symmetry used by co-motion maps. Thus the DFT problem is both a central application and a warning that multi-marginal OT is richer than a naive deterministic matching problem.

\paragraph{Multi-marginal formulation of barycenters.}
\(c_{\mathrm{bar}}(x_1,\ldots,x_S) = \min_{x\in\RR^d} \sum_{s=1}^S\la_s\norm{x-x_s}^2.\)
\begin{prop}[Multi-marginal formula for quadratic barycenters]\label{prop-multimarginal-barycenter}
Let $\be_1,\ldots,\be_S\in\Pp_2(\RR^d)$ and $\la\in\simplex_S$. Define
\(B(x_1,\ldots,x_S)=\sum_{s=1}^S\la_s x_s, \qquad c_{\mathrm{bar}}(x_1,\ldots,x_S)=\min_x\sum_s\la_s\norm{x-x_s}^2.\)
If $\pi^\star$ solves the multi-marginal OT problem with marginals $(\be_s)_s$ and cost $c_{\mathrm{bar}}$, then $\al^\star=B_\sharp\pi^\star$ is a Wasserstein barycenter. Conversely, every barycenter is obtained this way from an optimal multi-marginal plan.
\end{prop}
\begin{proof}
For any candidate barycenter $\al$ and couplings $\pi_s\in\Couplings(\al,\be_s)$, glue the couplings along their common $\al$ marginal to obtain a joint law of $(X,Y_1,\ldots,Y_S)$. Conditioning on $(Y_s)_s$ and minimizing over $X$ gives
\(\sum_s\la_s\EE\norm{X-Y_s}^2 \geq \EE\min_x\sum_s\la_s\norm{x-Y_s}^2 = \EE c_{\mathrm{bar}}(Y_1,\ldots,Y_S).\)
Taking the infimum over the couplings gives that the barycenter value is at least the multi-marginal value. Conversely, from an optimal multi-marginal plan $\pi^\star$, set $X=B(Y_1,\ldots,Y_S)$. The couplings between $X$ and each $Y_s$ are feasible for the barycenter problem and attain exactly the multi-marginal cost, proving equality and the formula.
If $\al^\star$ is any barycenter, choose optimal couplings between $\al^\star$ and each $\be_s$ and glue them along the common $\al^\star$ marginal. Since the barycenter and multi-marginal values are equal, the conditional minimization inequality above must be an equality. Thus $X=B(Y_1,\ldots,Y_S)$ almost surely for the induced optimal multi-marginal plan, and $\al^\star=B_\sharp\pi^\star$.
\end{proof}
\begin{cor}[Gaussian and discrete barycenters]\label{cor-gaussian-discrete-barycenters}
Quadratic Wasserstein barycenters of Gaussian measures are Gaussian. If the input measures are discrete, then there exists a barycenter supported on the set of weighted averages $\sum_s\la_s x_{s,i_s}$ of one support point from each input; in particular, if the $s$-th input has $n_s$ atoms, a barycenter exists with at most $\prod_s n_s$ atoms.
\end{cor}
\begin{proof}
Let the input Gaussians have means $\mean_s$ and covariances $\cov_s$. For any candidate barycenter $\al$ with mean $\mean$ and covariance $\cov$, Gelbrich's inequality, proved later in Theorem~\ref{thm-gelbrich-projection}, gives
\(\Wass_2^2(\al,\be_s) \geq \norm{\mean-\mean_s}^2+\Bb(\cov,\cov_s)^2,\)
with equality for the Gaussian law with mean $\mean$ and covariance $\cov$. Therefore the barycenter objective is bounded below by a function depending only on $(\mean,\cov)$, and this lower bound is attained by the Gaussian measure with the minimizing mean and covariance. Hence at least one barycenter is Gaussian, and uniqueness in the usual nondegenerate setting gives the Gaussian barycenter mentioned above. For discrete inputs, any multi-marginal optimizer is supported on the finite product of the input supports, and $B$ maps this product to at most $\prod_s n_s$ points.
\end{proof}
\paragraph{Entropic regularization of multi-marginal OT.}
\(\inf_{\pi\in\Couplings(\al_1,\ldots,\al_S)} \int c\d\pi+\epsilon\KL(\pi|\al_1\otimes\cdots\otimes\al_S).\)
\[
\d\pi^\star(x_1,\ldots,x_S)
=
\exp\!\left(\frac{\sum_s f_s(x_s)-c(x_1,\ldots,x_S)}{\epsilon}\right)
\prod_s\d\al_s(x_s),
\]
\subsection{Low-Rank Optimal Transport}
\label{sec-low-rank-ot}
\begin{defn}[Low-rank factored couplings]\label{def-low-rank-couplings}
Let $\a\in\simplex_n$, $\b\in\simplex_m$ and let $r\geq1$. A rank-$r$ factored coupling is a triple $(Q,R,g)$ such that
\(Q\in\RR_+^{n\times r},\qquad R\in\RR_+^{m\times r}, \qquad g\in\simplex_r,\)
with \(Q\ones_r=\a, \qquad R\ones_r=\b, \qquad \transp Q\ones_n=\transp R\ones_m=g.\)
It induces the coupling
\begin{equation}\label{eq-low-rank-coupling-factor}
P(Q,R,g)
=
Q\diag(g)^{-1}\transp R,
\qquad
P_{i,j}=\sum_{k=1}^r \frac{Q_{i,k}R_{j,k}}{g_k},
\end{equation}
where columns with $g_k=0$ are discarded before applying the formula.
\end{defn}
\begin{prop}[Factored couplings and nonnegative rank]\label{prop-low-rank-factorization}
For every feasible triple $(Q,R,g)$ in Definition~\ref{def-low-rank-couplings}, the matrix $P(Q,R,g)$ belongs to $\CouplingsD(\a,\b)$ and has nonnegative rank at most $r$. Conversely, every coupling $P\in\CouplingsD(\a,\b)$ with nonnegative rank at most $r$ admits a representation of the form~\eqref{eq-low-rank-coupling-factor}, after possibly deleting zero latent components.
\end{prop}
\begin{proof}
The marginal constraints follow by direct summation: \(P\ones_m=Q\diag(g)^{-1}\transp R\ones_m =Q\diag(g)^{-1}g=Q\ones_r=\a,\)
and similarly $\transp P\ones_n=\b$. Since $P$ is a sum of $r$ nonnegative rank-one matrices $Q_{:,k}R_{:,k}^\top/g_k$, its nonnegative rank is at most $r$.

Conversely, suppose $P=UV^\top$ with $U\in\RR_+^{n\times r}$ and $V\in\RR_+^{m\times r}$. Set $q_k=\sum_i U_{i,k}$ and $s_k=\sum_j V_{j,k}$. Components with $q_ks_k=0$ do not contribute and can be removed. Define $g_k=q_ks_k$, $Q_{i,k}=U_{i,k}s_k$ and $R_{j,k}=V_{j,k}q_k$. Since $P$ has total mass one, $g\in\simplex_r$. Moreover $Q\ones_r=P\ones_m=\a$, $R\ones_r=\transp P\ones_n=\b$, and both column marginals are equal to $g$. Finally,
\(\sum_k\frac{Q_{i,k}R_{j,k}}{g_k} = \sum_k\frac{U_{i,k}s_kV_{j,k}q_k}{q_ks_k} =P_{i,j}.\)
\end{proof}
For a cost matrix $C\in\RR^{n\times m}$, the low-rank constrained OT value is therefore

\(\min_{(Q,R,g)} \sum_{k=1}^r \frac{\transp{Q_{:,k}} C R_{:,k}}{g_k} \quad\text{subject to Definition~\ref{def-low-rank-couplings}.}\)
\begin{equation}\label{eq-low-rank-entropic-ot}
\min_{(Q,R,g)}
\sum_{k=1}^r \frac{\transp{Q_{:,k}} C R_{:,k}}{g_k}
+
\epsilon\KLD(Q|\a\otimes g)
+
\epsilon\KLD(R|\b\otimes g),
\end{equation}
where the constraints are those of Definition~\ref{def-low-rank-couplings}.

\subsection{Capacity-Constrained Optimal Transport}
\label{sec-capacity-constrained-ot}
Let $\alpha\in\Mm_+^1(\Xx)$, $\beta\in\Mm_+^1(\Yy)$ and let $\kappa:\Xx\times\Yy\to[0,+\infty]$ be a measurable capacity. The capacity-constrained transport value is

\begin{equation}\label{eq-capacity-constrained-ot}
\MK_c^\kappa(\alpha,\beta)
\eqdef
\inf_{\pi\in\Couplings(\alpha,\beta)}
\enscond{
\int_{\Xx\times\Yy} c(x,y)\d\pi(x,y)
}{
\pi\ll\alpha\otimes\beta,\quad
\frac{\d\pi}{\d(\alpha\otimes\beta)}(x,y)\leq\kappa(x,y)
}.
\end{equation}
The product coupling $\alpha\otimes\beta$ is feasible whenever $\kappa\geq1$. Thus the constraint is not meant to promote the product plan; rather, it prevents the optimizer from using any pair more than the prescribed density ratio.

For discrete measures $\alpha=\sum_i a_i\de_{x_i}$ and $\beta=\sum_j b_j\de_{y_j}$, a capacity is an upper matrix $U\in\RR_+^{n\times m}$. The density-ratio discretization of~\eqref{eq-capacity-constrained-ot} is $U_{i,j}=\kappa_{i,j}a_i b_j$, and the finite-dimensional problem is the linear program

\begin{equation}\label{eq-discrete-capacity-constrained-ot}
\min_{P\in\CouplingsD(\a,\b)}
\dotp{C}{P}
\quad\text{subject to}\quad
0\leq P_{i,j}\leq U_{i,j}\quad\forall(i,j).
\end{equation}
\begin{equation}\label{eq-entropic-capacity-constrained-ot}
\min_{P\in\CouplingsD(\a,\b),\,0\leq P\leq U}
\dotp{C}{P}
+
\epsilon\KLD(P|\a\otimes\b).
\end{equation}
\subsection{Metric learning and inverse OT}
\subsubsection{Differentiating OT losses}
\begin{prop}[First variations of unregularized OT]\label{prop-ot-first-variations-unregularized}
Let $\alpha\in\Pp(\Xx)$ and $\beta\in\Pp(\Yy)$, where $\Xx,\Yy$ are compact metric spaces, and let $c\in\Cc(\Xx\times\Yy)$. Define
\(\mathcal V_c(\alpha,\beta) \eqdef \inf_{\pi\in\Couplings(\alpha,\beta)} \int_{\Xx\times\Yy}c(x,y)\d\pi(x,y).\)
Let $\mathcal O_c(\alpha,\beta)$ be the set of optimal couplings and let $\mathcal D_c(\alpha,\beta)$ be the set of optimal dual potentials $(f,g)$ with $f(x)+g(y)\leq c(x,y)$. If $\chi$ is a signed measure with $\chi(\Xx)=0$ and $\alpha_t=\alpha+t\chi$ is a probability measure for $0\leq t\leq t_0$, then
\[
\left.\frac{\d}{\d t}\right|_{t=0^+}
\mathcal V_c(\alpha_t,\beta)
=
\sup_{(f,g)\in\mathcal D_c(\alpha,\beta)}
\int_{\Xx} f(x)\d\chi(x).
\]
If $h\in\Cc(\Xx\times\Yy)$ and $c_t=c+th$, then
\[
\left.\frac{\d}{\d t}\right|_{t=0^+}
\mathcal V_{c_t}(\alpha,\beta)
=
\inf_{\pi\in\mathcal O_c(\alpha,\beta)}
\int_{\Xx\times\Yy} h(x,y)\d\pi(x,y).
\]
In particular, if the normalized optimal potential $f^\star$ and the optimal plan $\pi^\star$ are unique, then
\(\frac{\delta \mathcal V_c}{\delta\alpha}=f^\star, \qquad \frac{\delta \mathcal V_c}{\delta c}=\pi^\star.\)
The second identity means that the first variation with respect to the function $c$ is represented by the optimal measure $\pi^\star$ on $\Xx\times\Yy$.
\end{prop}
\begin{proof}
Kantorovich duality writes
\(\mathcal V_c(\alpha,\beta) = \sup_{f\oplus g\leq c} \int f\d\alpha+\int g\d\beta.\)
This is a supremum of affine functions of $\alpha$, so Danskin's theorem gives the one-sided directional derivative as the supremum over active maximizers, namely the optimal dual potentials. The condition $\chi(\Xx)=0$ makes the formula independent of the additive gauge $(f,g)\mapsto(f+\lambda,g-\lambda)$.

For the cost variable, \(\mathcal V_{c_t}(\alpha,\beta) = \inf_{\pi\in\Couplings(\alpha,\beta)} \int c\d\pi+t\int h\d\pi\)
is an infimum of affine functions of $t$. Danskin's theorem for a minimum gives the right directional derivative as the infimum of $\int h\d\pi$ over the active minimizers. If the active dual potential or coupling is unique, the corresponding directional derivative is linear in the perturbation, which is the displayed first variation.
\end{proof}
\(f^\star\in\partial_{\a}\MKD_{\C}(\a,\b), \qquad \P^\star\in\partial_{\C}^{\mathrm{sup}}\MKD_{\C}(\a,\b).\)
\begin{prop}[First variations of entropic OT]\label{prop-ot-first-variations-entropic}
Let $\epsilon>0$ and define the KL-normalized entropic value
\(\mathcal V_{c,\epsilon}(\alpha,\beta) \eqdef \inf_{\pi\in\Couplings(\alpha,\beta)} \int c\d\pi+\epsilon\KL(\pi|\alpha\otimes\beta).\)
Assume that the optimal entropic potentials $(f_\epsilon,g_\epsilon)$ exist and are normalized so that the following density has marginals $\alpha$ and $\beta$:
\(\d\pi_\epsilon(x,y) = \exp\pa{\frac{f_\epsilon(x)+g_\epsilon(y)-c(x,y)}{\epsilon}} \d\alpha(x)\d\beta(y)\)
This defines the unique optimal coupling. For the same perturbations $\alpha_t=\alpha+t\chi$ and $c_t=c+th$ as in Proposition~\ref{prop-ot-first-variations-unregularized},
\[
\left.\frac{\d}{\d t}\right|_{t=0^+}
\mathcal V_{c,\epsilon}(\alpha_t,\beta)
=
\int f_\epsilon\d\chi,
\qquad
\left.\frac{\d}{\d t}\right|_{t=0^+}
\mathcal V_{c_t,\epsilon}(\alpha,\beta)
=
\int h\d\pi_\epsilon.
\]
Equivalently, \(\frac{\delta \mathcal V_{c,\epsilon}}{\delta\alpha}=f_\epsilon, \qquad \frac{\delta \mathcal V_{c,\epsilon}}{\delta c}=\pi_\epsilon.\)
In finite dimension, for positive histograms, these are ordinary derivatives on the relative interior of the simplices.
\end{prop}
\begin{proof}
The cost derivative follows directly from the primal envelope theorem, because the entropic optimizer is unique. For the measure derivative, use the dual formula~\eqref{eq-dual-sinkhorn-objective}. At an optimal pair, the soft-transform equations imply the row normalization
\(\int_{\Yy} \exp\pa{\frac{f_\epsilon(x)+g_\epsilon(y)-c(x,y)}{\epsilon}} \d\beta(y) =1 \qquad\text{for all }x.\)
Differentiating the dual objective with respect to $\alpha$ at fixed optimal potentials gives \(\int f_\epsilon\d\chi - \epsilon \int_{\Xx\times\Yy} \pa{ e^{(f_\epsilon(x)+g_\epsilon(y)-c(x,y))/\epsilon}-1 } \d\chi(x)\d\beta(y).\)
The second term vanishes by the previous normalization identity, leaving $\int f_\epsilon\d\chi$. The gauge ambiguity of $(f_\epsilon,g_\epsilon)$ again disappears because $\chi(\Xx)=0$.
\end{proof}
For a finite-dimensional parametrization $c_\theta$ or $C_\theta$, the entropic formula gives the backpropagation rule

\(\partial_{\theta}\mathcal V_{c_\theta,\epsilon} = \int \partial_\theta c_\theta(x,y)\d\pi_\epsilon(x,y),\)
where $\pi_\epsilon$ is the entropic optimizer. For the unregularized value $\mathcal V_{c_\theta}$, uniqueness of the optimal plan $\pi^\star$ gives $\partial_{\theta}\mathcal V_{c_\theta}=\int \partial_\theta c_\theta\d\pi^\star$.

\(\MKD_\C^\epsilon(\a,\b) = \mathcal V_{\C,\epsilon}(\a,\b) -\epsilon\HD(\a)-\epsilon\HD(\b),\)
\paragraph{Inverse Optimal Transport.}
\(\inf_{\pi\in\Couplings(\alpha,\beta)}\int c(x,y)\d\pi(x,y).\)
A useful statistical methodology is to measure the suboptimality of the observed plan through a Fenchel--Young loss. Write the score as $s=-c$ and define the convex regularized prediction value

\(G_\epsilon(s)= \sup_{\pi\in\Couplings(\alpha,\beta)} \int s\d\pi-\epsilon\KL(\pi|\alpha\otimes\beta).\)
\(\mathcal L_\epsilon(c;\widehat\pi) = G_\epsilon(-c)+G_\epsilon^*(\widehat\pi)+\int c\d\widehat\pi\)
is nonnegative by Fenchel's inequality and vanishes exactly when $\widehat\pi\in\partial G_\epsilon(-c)$, i.e. when $\widehat\pi$ satisfies the regularized optimality conditions for $c$. Entropic regularization is important here because it makes the forward map smoother and provides gradients with respect to $c$, at the price of a bias that must be controlled statistically.

In the discrete unregularized case, this loss reduces to the optimality gap of the observed coupling. For $\widehat P\in\CouplingsD(\a,\b)$ and a cost matrix $C$, denote it by

\(\mathcal L_{\mathrm{iOT}}(C;\widehat P) = \dotp{C}{\widehat P} - \umin{P\in\CouplingsD(\a,\b)} \dotp{C}{P}.\)
\(C_\theta=\sum_{r=1}^R \theta_r C^{(r)}, \qquad \theta\in\Theta,\)
where $\Theta$ is convex and the matrices $C^{(r)}$ encode features, graph distances or a Mahalanobis parameterization.

\(c_A(x,y)=\dotp{Ax}{y}, \qquad A\in\RR^{d\times d}.\)
For empirical measures $\al=\frac1n\sum_i\de_{x_i}$ and $\be=\frac1n\sum_j\de_{y_j}$, this gives the cost matrix

\(C(A)_{i,j}=\dotp{Ax_i}{y_j},\)
For a fixed matrix $A$, the forward prediction is the optimal face

\(\mathcal P_A\eqdef \uargmin{P\in\CouplingsD(\ones_n/n,\ones_n/n)} \dotp{C(A)}{P}.\)
\(\norm{x-y}^2=\norm{x}^2+\norm{y}^2-2\dotp{x}{y},\)
\(A_t=-\diag(1+t,1-t), \qquad -1\leq t\leq 1,\)
\[
\mathcal L_{\mathrm{iOT}}(C(A_t);\widehat P)
=
\dotp{C(A_t)}{\widehat P}
-
\umin{P\in\CouplingsD(\ones_n/n,\ones_n/n)}
\dotp{C(A_t)}{P},
\qquad
C(A_t)_{i,j}=\dotp{A_t x_i}{y_j}.
\]
\begin{prop}[Convex dual-gap formulation of inverse OT]\label{prop-inverse-ot-convex}
Let $\widehat P\in\CouplingsD(\a,\b)$ be an observed coupling and let $C_\theta$ depend affinely on $\theta\in\Theta$, where $\Theta$ is convex. The condition that $\widehat P$ is optimal for the cost $C_\theta$ is equivalent to the existence of dual potentials $(f,g)$ such that
\(f_i+g_j\leq (C_\theta)_{i,j} \qandq \sum_{i,j}\widehat P_{i,j}\big((C_\theta)_{i,j}-f_i-g_j\big)=0.\)
\eql{\label{eq-inverse-ot-convex}
\umin{\theta\in\Theta,f,g}
R(\theta)+\lambda
\sum_{i,j}\widehat P_{i,j}\big((C_\theta)_{i,j}-f_i-g_j\big)
\quad\text{subject to}\quad
f_i+g_j\leq (C_\theta)_{i,j}\quad\forall i,j.
}
\end{prop}
\begin{proof}
For a fixed cost $C_\theta$, Kantorovich duality gives \(\min_{P\in\CouplingsD(\a,\b)}\dotp{C_\theta}{P} = \max_{f_i+g_j\leq (C_\theta)_{i,j}} \dotp{f}{\a}+\dotp{g}{\b}.\)
Since $\widehat P$ has marginals $(\a,\b)$, every dual feasible pair satisfies \(\dotp{C_\theta}{\widehat P}-\dotp{f}{\a}-\dotp{g}{\b} = \sum_{i,j}\widehat P_{i,j}\big((C_\theta)_{i,j}-f_i-g_j\big) \geq0.\)
It vanishes if and only if $\widehat P$ reaches the dual value and is therefore optimal. If $C_\theta$ is affine and $\Theta$ and $R$ are convex, the constraints and objective in~\eqref{eq-inverse-ot-convex} are convex, proving the relaxation claim.
\end{proof}
\(\widehat P_{i,j}\approx a_i b_j \exp\pa{\frac{f_i+g_j-(C_\theta)_{i,j}}{\epsilon}},\)
\subsection{Weak Optimal Transport}
\label{sec-weak-ot}
\paragraph{Barycentric projection of a coupling.}
\begin{defn}[Barycentric projection of a coupling]\label{def-barycentric-projection}
Let $\alpha,\beta\in\Pp_1(\RR^d)$ and let $\pi\in\Couplings(\alpha,\beta)$. Disintegrate $\pi$ with respect to its first marginal as
$\pi(\d x,\d y)=\pi_x(\d y)\alpha(\d x)$. Since $\beta$ has finite first moment, the conditional mean is finite for $\alpha$-a.e. $x$, and the barycentric projection of $\pi$ is the map
\begin{equation}\label{eq-barycentric-projection}
\bar T_\pi(x)
\eqdef
\int_{\RR^d}y\d\pi_x(y),
\qquad
\bar\beta_\pi
\eqdef
(\bar T_\pi)_\sharp\alpha.
\end{equation}
\end{defn}
\begin{prop}[Barycentric projection of a quadratic optimal plan]\label{prop-barycentric-projection-optimal}
Let $\pi\in\Couplings(\alpha,\beta)$ be optimal for the quadratic cost $\norm{x-y}^2$ between $\alpha,\beta\in\Pp_2(\RR^d)$, and define $\bar T_\pi$ and $\bar\beta_\pi$ by~\eqref{eq-barycentric-projection}. Then $(\Id,\bar T_\pi)_\sharp\alpha$ is an optimal coupling between $\alpha$ and $\bar\beta_\pi$. Equivalently, $\bar T_\pi$ is a quadratic optimal transport map from $\alpha$ to the projected target $\bar\beta_\pi$.
\end{prop}
\begin{proof}
By Theorem~\ref{thm:opt_ccm}, $\pi$ is concentrated on a $c$-cyclically monotone set $\Gamma$ for $c(x,y)=\norm{x-y}^2$. For the quadratic cost, and since it is enough to check cyclic permutations, this means that every finite cycle $(x_i,y_i)_{i=1}^m\subset\Gamma$ satisfies
\(\sum_{i=1}^m \dotp{x_i}{y_i} \geq \sum_{i=1}^m \dotp{x_i}{y_{i+1}}, \qquad y_{m+1}=y_1.\)
After changing the disintegration on an $\alpha$-negligible set, $\pi_x$ is supported on the section
\(\Gamma_x=\enscond{y}{(x,y)\in\Gamma}\)
for $\alpha$-a.e. $x$.
Choose $x_1,\ldots,x_m$ in this full-measure set and independently sample $Y_i\sim\pi_{x_i}$. Applying the cyclic inequality to $(x_i,Y_i)$ and taking expectations gives
\(\sum_{i=1}^m \dotp{x_i}{\bar T_\pi(x_i)} \geq \sum_{i=1}^m \dotp{x_i}{\bar T_\pi(x_{i+1})}.\)
Thus $(\Id,\bar T_\pi)_\sharp\alpha$ is concentrated on a cyclically monotone graph. By the cyclic-monotonicity characterization of quadratic optimality, this plan is optimal between its two marginals, namely $\alpha$ and $\bar\beta_\pi$.
\end{proof}
\begin{equation}\label{eq-weak-ot}
\WOT_C(\alpha,\beta)
\eqdef
\inf_{\pi\in\Couplings(\alpha,\beta)}
\int C(x,\pi_x)\d\alpha(x).
\end{equation}
\begin{prop}[Weak Kantorovich duality]\label{prop-weak-ot-duality}
Assume that $\X,\Y$ are compact metric spaces and that $C(x,\nu)$ is lower semicontinuous, bounded from below and convex in $\nu$, with the standard qualification assumptions ensuring Fenchel--Rockafellar duality. For $g\in\Cc(\Y)$ define the weak $C$-transform
\[
g^C(x)
\eqdef
\inf_{\nu\in\Pp(\Y)}
\left\{
C(x,\nu)-\int g(y)\d\nu(y)
\right\}.
\]
Then
\[
\WOT_C(\alpha,\beta)
=
\sup_{g\in\Cc(\Y)}
\left\{
\int g^C(x)\d\alpha(x)+\int g(y)\d\beta(y)
\right\}.
\]
When $C(x,\nu)=\int c(x,y)\d\nu(y)$, this reduces to the usual Kantorovich dual with $g^C(x)=\inf_y(c(x,y)-g(y))$.
\end{prop}
\begin{proof}
For any coupling $\pi$ and any $g\in\Cc(\Y)$, the definition of $g^C$ gives \(C(x,\pi_x)\geq g^C(x)+\int g(y)\d\pi_x(y).\)
After integration with respect to $\alpha$, the second term becomes $\int g\d\beta$ because the second marginal of $\pi$ is $\beta$. This proves weak duality.

For the reverse inequality, consider the convex minimization over probability kernels $x\mapsto\pi_x$ with the affine constraint $\int \pi_x\d\alpha(x)=\beta$. Fenchel--Rockafellar duality gives a continuous Lagrange multiplier $g$ for this marginal constraint. Minimizing the Lagrangian over each conditional law gives exactly the pointwise term $g^C(x)$, while the multiplier contributes $\int g\d\beta$. The compactness, lower semicontinuity, convexity and qualification assumptions ensure no duality gap. This is the weak-cost analogue of Proposition~\ref{prop-duality-discr}.
\end{proof}
\begin{prop}[Barycentric weak transport is weaker than $\Wass_2$]\label{prop-barycentric-weak-ot}
Let $\alpha,\beta\in\Pp_2(\RR^d)$ and define
\(C_{\mathrm{bar}}(x,\nu) = \norm{x-\int y\d\nu(y)}^2.\)
Equivalently, for a coupling $\pi$, the integrand is $\norm{x-\bar T_\pi(x)}^2$. Then
\(\mathcal{W}_{C_{\mathrm{bar}}}(\alpha,\beta) \leq \Wass_2^2(\alpha,\beta).\)
\end{prop}
\begin{proof}
Let $\pi$ be any coupling and disintegrate it as $\pi_x\alpha$. By Jensen's inequality, \(\norm{x-\bar T_\pi(x)}^2 \leq \int\norm{x-y}^2\d\pi_x(y).\)
Integrating in $x$ gives $\int C_{\mathrm{bar}}(x,\pi_x)\d\alpha(x)\leq\int\norm{x-y}^2\d\pi(x,y)$. Taking the infimum over $\pi$ proves the claim.
\end{proof}
The barycentric cost is the canonical example to keep in mind: admissibility still constrains the full conditional laws to have second marginal $\beta$, but the objective only charges the displacement from $x$ to $\bar T_\pi(x)$ and ignores the conditional variance around this barycenter.

\subsubsection{Martingale Optimal Transport}
\label{sec-martingale-ot}
\begin{defn}[Martingale couplings and martingale OT]\label{def-martingale-coupling}
Let $\alpha,\beta\in\Pp_1(\RR^d)$. A coupling $\pi\in\Couplings(\alpha,\beta)$ is a martingale coupling if, for the disintegration $\pi(\d x,\d y)=\pi_x(\d y)\alpha(\d x)$,
\begin{equation}\label{eq-martingale-coupling}
\int_{\RR^d}y\d\pi_x(y)=x
\qquad\text{for $\alpha$-a.e. }x.
\end{equation}
Equivalently, $\bar T_\pi=\Id$ in~\eqref{eq-barycentric-projection}. The set of such couplings is denoted
\(\Couplings_{\mathrm{mart}}(\alpha,\beta) \eqdef \enscond{\pi\in\Couplings(\alpha,\beta)}{\bar T_\pi=\Id\quad\alpha\text{-a.e.}}.\)
\end{defn}
For a cost $c$, the martingale OT problem minimizes $\int c\d\pi$ over $\Couplings_{\mathrm{mart}}(\alpha,\beta)$, with value $+\infty$ when the admissible set is empty.

The terminology comes from probability: if $(X,Y)\sim\pi$, then~\eqref{eq-martingale-coupling} is exactly $\EE[Y|X]=X$. Hence martingale OT is a Kantorovich problem with the usual two marginal constraints plus a barycentric constraint on the conditional laws.

\((\Id,\bar T_\pi)_\sharp\alpha\)
\paragraph{Stochastic orders.}
\[
\alpha\preceq_{\mathrm{st}}\beta
\quad\Longleftrightarrow\quad
\int \varphi\,\d\alpha\leq\int \varphi\,\d\beta
\quad\text{for every increasing }\varphi.
\]
\[
\alpha\preceq_{\mathrm{cx}}\beta
\quad\Longleftrightarrow\quad
\int \varphi\,\d\alpha\leq\int \varphi\,\d\beta
\quad\text{for every convex }\varphi\text{ with finite integrals}.
\]
\(\Gaussian(m,\Sigma_0)\preceq_{\mathrm{cx}}\Gaussian(m,\Sigma_1) \quad\Longleftrightarrow\quad \Sigma_1-\Sigma_0\succeq0.\)
Indeed, if $\Sigma_1-\Sigma_0\succeq0$, then $\Gaussian(m,\Sigma_1)$ is obtained from $\Gaussian(m,\Sigma_0)$ by adding independent centered Gaussian noise; the converse follows by testing the convex quadratic functions $x\mapsto\dotp{u}{x}^2$. Strassen's martingale theorem states that $\alpha\preceq_{\mathrm{cx}}\beta$ if and only if $\Couplings_{\mathrm{mart}}(\alpha,\beta)$ is nonempty.