\section{Generalized Wasserstein Distances}
\label{sec-extensions}
\label{sec-generalized-wasserstein-distances}
\subsection{Unbalanced OT}
\label{sec-unbalanced}
\paragraph{Relaxed formulation.}
For nonnegative measures $(\al,\be) \in \Mm_+(\Xx) \times \Mm_+(\Yy)$, a generic relaxed formulation is

\eql{\label{eq-unbalanced-primal}
\UW_c(\al,\be)= \uinf{\pi \in \Mm_+(\Xx \times \Yy)} \int_{\Xx \times \Yy} c(x,y) \d\pi(x,y)
+ \Dd_{\psi_1}(\pi_1|\al) + \Dd_{\psi_2}(\pi_2|\be),
}
where $\psi_1,\psi_2$ are convex entropy functions. Exact conservation $(\pi_1,\pi_2)=(\al,\be)$ is replaced by a cost for changing the marginals.

\(\UW_{c,\tau}(\alpha,\beta) = \inf_{\pi\geq0} \int c\d\pi + \tau\Dd_{\bar\psi_1}(\pi_1|\alpha) + \tau\Dd_{\bar\psi_2}(\pi_2|\beta).\)
These figures should be read as pictures of a single relaxed plan $\pi$: the same nonnegative measure determines both the transported coupling and its two relaxed marginals. The small-$\tau$ result below formalizes the opposite regime, where transport becomes negligible compared with local mass variation.

\begin{prop}[Small-transport-scale limit for marginal penalties]\label{prop-unbalanced-small-scale-limit}
Assume that $\alpha,\beta$ are finite measures on a compact metric space $\Xx$, that $c$ is continuous, $c\geq0$, and $c(x,y)=0$ if and only if $x=y$. Assume also that the marginal divergences are nonnegative, weak-* lower semicontinuous, and have weak-* compact sublevel sets on $\Mm_+(\Xx)$. Then
\[
\lim_{\tau\downarrow0}\frac1\tau\UW_{c,\tau}(\alpha,\beta)
=
\inf_{\rho\in\Mm_+(\Xx)}
\Dd_{\bar\psi_1}(\rho|\alpha)
+
\Dd_{\bar\psi_2}(\rho|\beta).
\]
The right-hand side is the infimal gluing divergence obtained by matching the two measures through a common zero-transport marginal $\rho$. In the dominated case, if $\alpha=a\lambda$, $\beta=b\lambda$, and the minimizing common marginal is absolutely continuous, $\rho=r\lambda$, this divergence decouples pointwise as
\[
\int \mathfrak m_{\bar\psi_1,\bar\psi_2}(a(x),b(x))\d\lambda(x),
\qquad
\mathfrak m_{\bar\psi_1,\bar\psi_2}(a,b)
\eqdef
\inf_{r\geq0} a\,\bar\psi_1(r/a)+b\,\bar\psi_2(r/b),
\]
with the usual recession conventions when $a=0$ or $b=0$. For superlinear entropies, and in particular for KL, finite energy forces this dominated form. Thus, when $\bar\psi_1=\bar\psi_2$ is the KL entropy,
\(\inf_{\rho\in\Mm_+(\Xx)} \KL(\rho|\alpha)+\KL(\rho|\beta) = \int (\sqrt{a}-\sqrt{b})^2\d\lambda.\)
Thus the KL marginal relaxation contains the squared Hellinger distance as its local mass-variation limit.
\end{prop}
\begin{proof}
For the upper bound, restrict to diagonal plans $\pi=(\Id,\Id)_\sharp\rho$, whose transport cost is zero and whose two marginals are both $\rho$.

For the lower bound, let $\tau_n\downarrow0$ and let $\pi_n$ be almost minimizing plans with bounded scaled values $\tau_n^{-1}\UW_{c,\tau_n}(\alpha,\beta)$. Since the divergences are nonnegative, $\int c\d\pi_n=O(\tau_n)$, hence $\int c\d\pi_n\to0$. The bounded scaled values also put the two marginals in compact divergence sublevel sets. Since a coupling has the same total mass as each marginal, the couplings are tight on $\Xx\times\Xx$. Up to subsequences, $\pi_n\rightharpoonup\pi_0$. Lower semicontinuity of the transport cost yields $\int c\d\pi_0=0$, so $\pi_0$ is concentrated on the diagonal. Its two marginals are therefore equal to a common measure $\rho$. Lower semicontinuity of the marginal divergences gives
\(\liminf_n\frac1{\tau_n}\UW_{c,\tau_n}(\alpha,\beta) \geq \Dd_{\bar\psi_1}(\rho|\alpha)+\Dd_{\bar\psi_2}(\rho|\beta),\)
and optimizing over $\rho$ gives the lower bound.

In the dominated case, writing $\rho=r\lambda$ gives \(\Dd_{\bar\psi_1}(\rho|\alpha)+\Dd_{\bar\psi_2}(\rho|\beta) = \int a\,\bar\psi_1(r/a)+b\,\bar\psi_2(r/b) \d\lambda,\)
so the minimization over $\rho$ decouples into the scalar envelope $\mathfrak m_{\bar\psi_1,\bar\psi_2}$. For KL, no singular part is admissible when $\alpha$ and $\beta$ are dominated by $\lambda$. The pointwise objective is $r\log(r/a)-r+a+r\log(r/b)-r+b$. Its optimality condition is $\log(r/a)+\log(r/b)=0$, hence $r=\sqrt{ab}$, and the minimum is $a+b-2\sqrt{ab}=(\sqrt a-\sqrt b)^2$.
\end{proof}
\begin{prop}[Dual of unbalanced optimal transport]\label{prop-dual-unbalanced-ot}
Under the usual Fenchel--Rockafellar qualification assumptions, one has equality between~\eqref{eq-unbalanced-primal} and
\(\UW_c(\al,\be) = \usup{f\oplus g\leq c} - \Dd_{\psi_1}^*(-f|\al) - \Dd_{\psi_2}^*(-g|\be).\)
\end{prop}
\begin{proof}
Use the variational formula~\eqref{eq-legendre} for the dual of a divergence and introduce the marginal variables through continuous potentials:
\(\inf_{\pi\geq0}\sup_{f,g} \int c\d\pi + \int -f\d\pi_1 + \int -g\d\pi_2 - \Dd_{\psi_1}^*(-f|\al) - \Dd_{\psi_2}^*(-g|\be).\)
Exchanging the infimum and the supremum gives \(\sup_{f,g} - \Dd_{\psi_1}^*(-f|\al) - \Dd_{\psi_2}^*(-g|\be) + \inf_{\pi\geq0}\int\big(c-(f\oplus g)\big)\d\pi.\)
The last infimum is $0$ when $f\oplus g\leq c$ and $-\infty$ otherwise, which gives the displayed dual.
\end{proof}
\paragraph{Partial optimal transport.}
\(0\leq m\leq \min\{\al(\Xx),\be(\Yy)\},\)
\[
\operatorname{POT}_m(\al,\be)
\eqdef
\inf_{\substack{\pi\in\Mm_+(\Xx\times\Yy)\\
\pi_1\leq\al,\ \pi_2\leq\be\\
\pi(\Xx\times\Yy)=m}}
\int c\,\d\pi.
\]
Thus only a submeasure of $\al$ is transported onto a submeasure of $\be$; the remaining mass is left unmatched. The corresponding Lagrangian form is obtained by adding total-variation penalties.

\(\inf_{\pi\in\Mm_+(\Xx\times\Yy)} \int c\,\d\pi + \lambda\norm{\al-\pi_1}_{\TV} + \lambda\norm{\be-\pi_2}_{\TV}.\)
Assume $c\geq0$. Allowing transported marginals larger than the available marginals does not improve the value: excess transported mass can be trimmed, which decreases the transport cost and cannot increase the sum of the two total-variation penalties. Hence an optimal plan may be chosen with $\pi_1\leq\al$ and $\pi_2\leq\be$. If $m=\pi(\Xx\times\Yy)$, the penalty then reduces to

\(\lambda\big(\al(\Xx)-m\big)+\lambda\big(\be(\Yy)-m\big),\)
\begin{prop}[TV penalization selects fixed-mass partial OT]\label{prop-tv-partial-ot-lagrange}
Assume that $c\geq0$ and set $A=\al(\Xx)$, $B=\be(\Yy)$, $M=\min(A,B)$. For $\lambda\geq0$, define the TV-penalized partial value
\[
\operatorname{POT}^{\TV}_\lambda(\al,\be)
\eqdef
\inf_{\substack{\pi\in\Mm_+(\Xx\times\Yy)\\ \pi_1\leq\al,\ \pi_2\leq\be}}
\left\{
\int c\,\d\pi
+
\lambda\big(A+B-2\pi(\Xx\times\Yy)\big)
\right\}.
\]
Then
\[
\operatorname{POT}^{\TV}_\lambda(\al,\be)
=
\lambda(A+B)
+
\inf_{0\leq m\leq M}
\left\{
\operatorname{POT}_m(\al,\be)-2\lambda m
\right\}.
\]
If $\pi_\lambda$ minimizes $\operatorname{POT}^{\TV}_\lambda$ and $m_\lambda=\pi_\lambda(\Xx\times\Yy)$, then $m_\lambda$ minimizes $m\mapsto\operatorname{POT}_m(\al,\be)-2\lambda m$ and $\pi_\lambda$ minimizes $\operatorname{POT}_{m_\lambda}(\al,\be)$. Conversely, if $2\lambda$ is a subgradient at $m$ of the convex function $m\mapsto\operatorname{POT}_m(\al,\be)$ extended by $+\infty$ outside $[0,M]$, then every minimizer of $\operatorname{POT}_m(\al,\be)$ is a minimizer of $\operatorname{POT}^{\TV}_\lambda(\al,\be)$. In particular, every prescribed mass $m\in[0,M]$ is selected by at least one multiplier $\lambda\geq0$, possibly together with a whole interval of masses when the value function has a flat exposed face.
\end{prop}
\begin{proof}
Write $V(m)=\operatorname{POT}_m(\al,\be)$. The function $V$ is convex on $[0,M]$: if $\pi_0,\pi_1$ are admissible competitors of masses $m_0,m_1$, then $(1-t)\pi_0+t\pi_1$ is admissible for mass $(1-t)m_0+tm_1$ and has the corresponding convex-combination cost. It is also nondecreasing because any competitor of mass $m_1>m_0$ can be rescaled by $m_0/m_1$ to give a competitor of mass $m_0$ with no larger cost.

For any admissible $\pi$ in the definition of $\operatorname{POT}^{\TV}_\lambda$, let $m=\pi(\Xx\times\Yy)$. The plan is admissible for $V(m)$, hence its objective is at least $V(m)+\lambda(A+B-2m)$. Taking the infimum over $\pi$ gives the lower bound by the displayed right-hand side. Conversely, for each $m$ and each minimizing sequence for $V(m)$, the same plans are admissible in $\operatorname{POT}^{\TV}_\lambda$ and have objective tending to $V(m)+\lambda(A+B-2m)$. Taking the infimum over $m$ gives the reverse inequality.

The optimizer statement follows from the same inequalities. If $\pi_\lambda$ minimizes the penalized problem with mass $m_\lambda$, then equality in the lower bound forces both $V(m_\lambda)-2\lambda m_\lambda$ to be minimal over $[0,M]$ and $\pi_\lambda$ to be optimal at fixed mass $m_\lambda$. Conversely, $2\lambda\in\partial V(m)$ means
\(V(m')-2\lambda m'\geq V(m)-2\lambda m \quad\text{for all }m'\in[0,M].\)
Thus any minimizer of $V(m)$ attains the penalized infimum. Since a finite convex function on an interval has a nonempty subdifferential at every point after adding the normal cone of the constraint interval, such a $\lambda$ exists for every $m\in[0,M]$; monotonicity of $V$ ensures that one may choose $\lambda\geq0$.
\end{proof}
\begin{prop}[Metric case of fixed-mass partial OT]\label{prop-partial-ot-metric-slice}
Let $(\Xx,d)$ be a metric space, let $p\geq1$, set $c=d^p$, and fix $m>0$. On the class of nonnegative measures with total mass exactly $m$ and finite $p$th moment,
\(\mathcal M_m^p(\Xx)\eqdef \enscond{\mu\in\Mm_+(\Xx)}{\mu(\Xx)=m,\ \int d(x_0,x)^p\d\mu(x)<+\infty}\)
for one, hence any, base point $x_0$, the quantity
\(d_m(\al,\be)\eqdef \operatorname{POT}_m(\al,\be)^{1/p}\)
is a distance. More precisely, if $\bar\al=\al/m$ and $\bar\be=\be/m$, then
\(d_m(\al,\be)=m^{1/p}\Wass_p(\bar\al,\bar\be).\)
If instead $\operatorname{POT}_m$ is considered on a larger class of measures with mass at least $m$, then it is not a distance in general for $m>0$: different measures can share a common submeasure of mass $m$, giving zero partial cost.
\end{prop}
\begin{proof}
If $\al(\Xx)=\be(\Xx)=m$ and $\pi$ is admissible in $\operatorname{POT}_m(\al,\be)$, then $\pi_1\leq\al$ and both measures have mass $m$, hence $\pi_1=\al$. Similarly $\pi_2=\be$. Thus the admissible plans are exactly the ordinary couplings between $\al$ and $\be$. Dividing such a coupling by $m$ gives a probability coupling between $\bar\al$ and $\bar\be$, and conversely multiplying any coupling of $\bar\al,\bar\be$ by $m$ gives a coupling of $\al,\be$. Therefore
\(\operatorname{POT}_m(\al,\be) = m\Wass_p(\bar\al,\bar\be)^p.\)
The metric properties of $d_m$ follow from those of $\Wass_p$ in Proposition~\ref{prop-metric-measure}.

For the last claim, take three distinct points $x,y,z$ in a metric space and a number $r>0$. The measures $\al=m\de_x+r\de_y$ and $\be=m\de_x+r\de_z$ are different but the partial plan $m\de_{(x,x)}$ is admissible and has zero cost. Hence $\operatorname{POT}_m(\al,\be)=0$ although $\al\neq\be$.
\end{proof}
Thus the TV-penalized formulation is the Lagrangian envelope of constrained partial OT: increasing $\lambda$ rewards transporting more mass, while small $\lambda$ makes deletion and creation cheaper. As the prescribed transported mass decreases, the plan keeps only the least costly facing portions of the two marginals.

\paragraph{Reverse and homogeneous formulations.}
\begin{align*}
&\int c(x,y)\d\pi(x,y)
+\Dd_{\psi_1}(\pi_1|\al)+\Dd_{\psi_2}(\pi_2|\be) \\
&\qquad =
\int
\left(
c(x,y)
+
\psi_1\!\left(\frac{\d\pi_1}{\d\al}(x)\right)\frac{\d\al}{\d\pi_1}(x)
+
\psi_2\!\left(\frac{\d\pi_2}{\d\be}(y)\right)\frac{\d\be}{\d\pi_2}(y)
\right)
\d\pi(x,y).
\end{align*}
\eql{\label{eq-unbalanced-reverse-local-cost}
L_c(r,s) \eqdef c + r\psi_1(1/r)+s\psi_2(1/s),
}
with the usual recession convention at $r=0$ or $s=0$. If $\al=F\pi_1+\al^\perp$ and $\be=G\pi_2+\be^\perp$ are the Lebesgue decompositions of the reference marginals with respect to the transported marginals, then the reverse formulation reads

\(\UW_c(\al,\be) = \inf_{\pi\geq0} \int L_{c(x,y)}(F(x),G(y))\d\pi(x,y) + \psi_1(0)\al^\perp(\Xx) + \psi_2(0)\be^\perp(\Yy).\)
\eql{\label{eq-unbalanced-homogeneous-local-cost}
H_c(r,s) \eqdef \inf_{\theta>0}\theta L_c(r/\theta,s/\theta),
}
\eql{\label{eq-homogeneous}
\HW_c(\al,\be)
=
\inf_{\pi\geq0}
\int H_{c(x,y)}(F(x),G(y))\d\pi(x,y)
+
\psi_1(0)\al^\perp(\Xx)
+
\psi_2(0)\be^\perp(\Yy).
}
\begin{prop}[Homogenization does not change the unbalanced cost]\label{prop-homogeneous-unbalanced}
One has $\HW_c(\al,\be)=\UW_c(\al,\be)$.
\end{prop}
\begin{proof}
The inequality $\HW\leq\UW$ follows from $H_c\leq L_c$ by taking $\theta=1$. Conversely, take a feasible measure $\pi$ in the homogeneous formulation. By definition of the perspective transform, for every $(x,y)$ and every $\eta>0$ there exists a scale $\theta(x,y)>0$ such that
\(H_{c(x,y)}(F(x),G(y))+\eta \geq \theta(x,y)L_{c(x,y)}\big(F(x)/\theta(x,y),G(y)/\theta(x,y)\big).\)
Replacing $\pi$ by the rescaled measure $\tilde\pi=\theta\pi$ and the densities by $F/\theta$ and $G/\theta$ gives an admissible competitor for the reverse formulation with cost no larger than the homogeneous cost plus $\eta\pi(\Xx\times\Yy)$. Letting $\eta\to0$ yields $\UW\leq\HW$. The singular terms are unchanged because the same rescaling is performed before taking the Lebesgue decomposition of the marginals.
\end{proof}
\paragraph{Conic lifting.}
Assume now that $\Xx=\Yy$ and $\psi_1=\psi_2=\psi$. The last formulation lifts the problem to the cone space $\mathfrak C[\Xx]\eqdef(\Xx\times\RR_+)/\sim$, where all points $(x,0)$ are identified at the apex. For an exponent $p\geq1$, define

\(\De((x,r),(y,s))\eqdef H_{c(x,y)}(r^p,s^p)^{1/p}.\)
\begin{itemize}
\item $\Dd_\psi=\KL$, $p=2$, and $c(x,y)=-\log\cos^2(d(x,y)\wedge\pi/2)$ give the Hellinger--Kantorovich or Wasserstein--Fisher--Rao cone metric
\(\De((x,r),(y,s))^2=r^2+s^2-2rs\cos(d(x,y)\wedge\pi/2).\)
\item $\Dd_\psi=\KL$, $p=2$, and $c(x,y)=d(x,y)^2$ give the Gaussian Hellinger cone metric
\(\De((x,r),(y,s))^2=r^2+s^2-2rs e^{-d(x,y)^2/2}.\)
\item $\Dd_\psi=\TV$, $p=1$, and $c(x,y)=d(x,y)$ give the partial-transport cone cost
\(\De((x,r),(y,s))=r+s-(r\wedge s)(2-d(x,y))_+.\)
\end{itemize}
\[
\CW(\al,\be)
=
\inf_{\gamma\in\Mm_+(\mathfrak C[\Xx]^2)}
\left\{
\int \De((x,r),(y,s))^p\d\gamma
\;; \;
\int r^p\d\gamma_1(\cdot,r)=\al,
\int s^p\d\gamma_2(\cdot,s)=\be
\right\}.
\]
\begin{thm}[Cone formulation of unbalanced OT]\label{thm-cone-unbalanced-ot}
One has $\UW=\HW=\CW$. If $\De$ is a distance, then $\CW^{1/p}$ is a distance between nonnegative measures.
\end{thm}
\begin{proof}
The equality $\UW=\HW$ is Proposition~\ref{prop-homogeneous-unbalanced}. To prove $\HW=\CW$, disintegrate an admissible cone coupling $\gamma$ with respect to its spatial variables $(x,y)$ and radii $(r,s)$. The cone marginal constraints say precisely that the spatial marginals are recovered after weighting by $r^p$ and $s^p$. Since $\De((x,r),(y,s))^p=H_{c(x,y)}(r^p,s^p)$, integrating the cone cost gives the homogeneous objective. Conversely, any homogeneous competitor can be lifted to the cone by placing, over each $(x,y)$, radii whose $p$th powers are the two density factors appearing in $H_c$.

If $\De$ is a distance on the cone, then $\CW^{1/p}$ is the usual $p$-Wasserstein distance between lifted measures under the linear cone-marginal constraints. Symmetry and the triangle inequality follow from the corresponding Wasserstein properties and the gluing lemma on the cone. If the distance is zero, an optimal cone coupling is concentrated on the diagonal of the cone, so the weighted projections agree and therefore $\alpha=\beta$.
\end{proof}
\paragraph{Entropic KL relaxation.}
\(\inf_{\pi\in\Mm_+(\Xx\times\Yy)} \int c\d\pi + \Dd_{\psi_1}(\pi_1|\al) + \Dd_{\psi_2}(\pi_2|\be) + \epsilon\Dd_\phi(\pi|\al\otimes\be).\)
\[
\sup_{f,g}
-
\Dd_{\psi_1}^*(-f|\al)
-
\Dd_{\psi_2}^*(-g|\be)
-
\epsilon\Dd_\phi^*\left(\frac{f\oplus g-c}{\epsilon}\middle|\al\otimes\be\right).
\]
For $\Dd_\phi=\KL$, the primal-dual relation is $\d\pi=e^{(f\oplus g-c)/\epsilon}\d\al\d\be$. If, in addition, $\Dd_{\psi_1}=\Dd_{\psi_2}=\tau\KL$, the dual specializes to

\[
\sup_{f,g}
-\tau\int(e^{-f/\tau}-1)\d\al
-\tau\int(e^{-g/\tau}-1)\d\be
-\epsilon\iint
\left(e^{(f\oplus g-c)/\epsilon}-1\right)
\d\al\d\be,
\]
\begin{align*}
f(x)
&=
-
\frac{\tau\epsilon}{\tau+\epsilon}
\log\int_{\Yy}
\exp\left(\frac{g(y)-c(x,y)}{\epsilon}\right)\d\be(y),\\
g(y)
&=
-
\frac{\tau\epsilon}{\tau+\epsilon}
\log\int_{\Xx}
\exp\left(\frac{f(x)-c(x,y)}{\epsilon}\right)\d\al(x).
\end{align*}
\[
u_i\leftarrow \left(\frac{\a_i}{(K v)_i}\right)^\rho,
\qquad
v_j\leftarrow \left(\frac{\b_j}{(\transp{K}u)_j}\right)^\rho,
\qquad
\P=\diag(u)K\diag(v).
\]
\subsection{Sliced Wasserstein Distances}
\label{sec-sliced-wasserstein}
\begin{defn}[Sliced Wasserstein distance]\label{def-sliced-wasserstein}
Let $\sigma$ be the uniform probability measure on the sphere $\Sphere^{d-1}$. The sliced $p$-Wasserstein distance is
\eql{\label{eq-sliced-wasserstein}
\SW_p(\al,\be)^p
\eqdef
\int_{\Sphere^{d-1}}
\Wass_p\big((P_\theta)_\sharp\al,(P_\theta)_\sharp\be\big)^p
\d\sigma(\theta).
}
\end{defn}
\begin{prop}[Metric properties of sliced Wasserstein]\label{prop-sliced-wasserstein-metric}
For $p\geq1$, $\SW_p$ is a distance on $\Pp_p(\RR^d)$. Moreover, $\SW_p$ metrizes weak convergence together with convergence of the $p$th moment. Finally, \(\SW_p(\alpha,\beta)\leq \Wass_p(\alpha,\beta),\)
and, for $p=2$ with the uniform probability measure on the sphere, \(\SW_2(\alpha,\beta)^2\leq \frac1d\Wass_2(\alpha,\beta)^2.\)
\end{prop}
\begin{proof}
Non-negativity and symmetry follow from the one-dimensional Wasserstein distance. For the triangle inequality, apply the triangle inequality of $\Wass_p$ for each direction $\theta$ and then Minkowski's inequality in $L^p(\Sphere^{d-1})$.

If $\SW_p(\alpha,\beta)=0$, then $(P_\theta)_\sharp\alpha=(P_\theta)_\sharp\beta$ for almost every direction. By continuity of characteristic functions this holds for all directions, and the Cram\'er--Wold theorem implies $\alpha=\beta$. This proves separation.

The bound $\SW_p\leq\Wass_p$ follows because $P_\theta$ is $1$-Lipschitz, so
\(\Wass_p((P_\theta)_\sharp\alpha,(P_\theta)_\sharp\beta) \leq \Wass_p(\alpha,\beta)\)
for every $\theta$. For $p=2$, using any coupling $\pi$ between $\alpha$ and $\beta$, \(\int_{\Sphere^{d-1}}\int |\dotp{x-y}{\theta}|^2\d\pi(x,y)\d\sigma(\theta) = \frac1d\int\norm{x-y}^2\d\pi(x,y).\)
Optimizing over $\pi$ gives the sharper inequality. The weak-convergence statement follows from the same Cram\'er--Wold mechanism plus the moment condition: convergence in $\SW_p$ gives convergence of almost all one-dimensional projections and tightness of the $p$th moments; conversely, weak convergence with $p$th-moment convergence implies convergence of projected $\Wass_p$ distances and dominated convergence on the sphere.
\end{proof}
\begin{defn}[Max-sliced Wasserstein]\label{def-sliced-variants}
The max-sliced distance replaces the average over directions by the most discriminating one: \(\MaxSW_p(\alpha,\beta) \eqdef \sup_{\theta\in\Sphere^{d-1}} \Wass_p((P_\theta)_\sharp\alpha,(P_\theta)_\sharp\beta).\)
\end{defn}
\paragraph{Subspace-sliced variants.}
\begin{defn}[Subspace-sliced Wasserstein]\label{def-subspace-sliced-wasserstein}
Fix $1\leq k\leq d$. Subspace-sliced variants replace one-dimensional lines by $k$-dimensional orthogonal projections. If $U\in\RR^{d\times k}$ satisfies $\transp{U}U=\Id_k$, then
\(\SW_{p,k}(\alpha,\beta)^p \eqdef \int \Wass_p((\transp{U})_\sharp\alpha,(\transp{U})_\sharp\beta)^p\d U,\)
where $\d U$ denotes the normalized invariant measure on the Stiefel manifold
$\mathrm{St}(d,k)=\{U\in\RR^{d\times k}\;:\;\transp{U}U=\Id_k\}$, and
\(\MaxSW_{p,k}(\alpha,\beta) \eqdef \sup_{\transp{U}U=\Id_k} \Wass_p((\transp{U})_\sharp\alpha,(\transp{U})_\sharp\beta).\)
The case $k=1$ recovers ordinary sliced and max-sliced Wasserstein, while $k=d$ recovers the original Wasserstein distance.
\end{defn}
\begin{prop}[Basic bounds for sliced variants]\label{prop-sliced-variant-bounds}
Let $p\geq1$ and let $\alpha,\beta\in\Pp_p(\RR^d)$. With normalized spherical and Stiefel measures, \(\SW_p(\alpha,\beta) \leq \MaxSW_p(\alpha,\beta) \leq \Wass_p(\alpha,\beta).\)
For $k$-dimensional subspace projections, \(\SW_{p,k}(\alpha,\beta) \leq \MaxSW_{p,k}(\alpha,\beta) \leq \Wass_p(\alpha,\beta).\)
\end{prop}
\begin{proof}
The first inequality in each line follows because an $L^p$ average over a probability space is bounded by the corresponding supremum. The second inequality follows because orthogonal projections are $1$-Lipschitz: pushing any admissible coupling between $\alpha$ and $\beta$ through a projection gives an admissible coupling for the projected measures with no larger transport cost. Optimizing over couplings and then averaging or maximizing over the projection gives the result.
\end{proof}
\paragraph{Min-sliced lifted transport plans.}
\(\pi_\theta = \frac1n\sum_{i=1}^n\delta_{(x_i,y_{\sigma_\theta(i)})}\)
\(\operatorname{MSWGG}_2(\alpha,\beta)^2 \eqdef \min_{\theta\in\Sphere^{d-1}} \int\norm{x-y}^2\d\pi_\theta(x,y).\)
\(\Wass_2^2(\alpha,\beta) \leq \int\norm{x-y}^2\d\pi_\theta(x,y), \qquad \Wass_2^2(\alpha,\beta) \leq \operatorname{MSWGG}_2(\alpha,\beta)^2.\)
\subsection{Quotient Wasserstein and Wasserstein--Procrustes}
\label{sec-quotient-wasserstein-procrustes}
\begin{defn}[Quotient Wasserstein distance]\label{def-quotient-wasserstein}
Let a group $\mathcal G$ act on a metric space $(\Xx,d)$ by isometries, and assume that the action preserves finite $p$th moments. Write $g_\sharp\alpha$ for the push-forward of a measure $\alpha$ by $g\in\mathcal G$, and $[\alpha]\eqdef\{g_\sharp\alpha\;:\;g\in\mathcal G\}$ for its orbit. The quotient $p$-Wasserstein distance between the orbits of $\alpha,\beta\in\Pp_p(\Xx)$ is
\[
\Wass_{p,\mathcal G}([\alpha],[\beta])
\eqdef
\inf_{g,h\in\mathcal G}\Wass_p(g_\sharp\alpha,h_\sharp\beta)
=
\inf_{g\in\mathcal G}\Wass_p(\alpha,g_\sharp\beta).
\]
\end{defn}
\begin{prop}[Metric property on the quotient]\label{prop-quotient-wasserstein-metric}
If $\mathcal G$ acts by isometries and preserves $\Pp_p(\Xx)$, then $\Wass_{p,\mathcal G}$ is a pseudo-distance on $\Pp_p(\Xx)$. If, in addition, the infimum is attained between any two orbits and all orbits are closed, then it is a distance on the orbit space $\Pp_p(\Xx)/\mathcal G$. This applies in particular to continuous actions of compact groups.
\end{prop}
\begin{proof}
Isometry of the action gives $\Wass_p(g_\sharp\mu,g_\sharp\nu)=\Wass_p(\mu,\nu)$, which proves the second formula in Definition~\ref{def-quotient-wasserstein}. Non-negativity and symmetry are inherited from $\Wass_p$. For the triangle inequality, choose $g_1,g_2\in\mathcal G$. Then
\[
\Wass_p(\alpha,(g_1)_\sharp\beta)
+
\Wass_p(\beta,(g_2)_\sharp\gamma)
=
\Wass_p(\alpha,(g_1)_\sharp\beta)
+
\Wass_p((g_1)_\sharp\beta,(g_1g_2)_\sharp\gamma).
\]
Applying the triangle inequality for $\Wass_p$ and then taking infima over $g_1,g_2$ gives the result. If the infimum is attained and the quotient distance is zero, there exist $g,h\in\mathcal G$ such that $\Wass_p(g_\sharp\alpha,h_\sharp\beta)=0$, hence $g_\sharp\alpha=h_\sharp\beta$. Closed orbits then imply $[\alpha]=[\beta]$. For compact groups, continuity of the action and compactness give the required attainment.
\end{proof}
\paragraph{Rigid motions and Wasserstein--Procrustes.}
\(\inf_{R\in\mathrm O(d),\,t\in\RR^d,\,\pi\in\Couplings(\alpha,\beta)} \int\norm{Rx+t-y}^2\d\pi(x,y).\)
This is an extrinsic counterpart of the Gromov--Wasserstein viewpoint developed in Section~\ref{sec-gromov-wasserstein}. Procrustes alignment searches only over ambient rigid motions, whereas GW is invariant under all measure-preserving isometries of the metric-measure spaces.

\(\GW((\RR^d,\norm{\cdot},\alpha),(\RR^d,\norm{\cdot},\beta)) = \GW((\RR^d,\norm{\cdot},g_\sharp\alpha),(\RR^d,\norm{\cdot},h_\sharp\beta)).\)
\(\GW((\RR^d,\norm{\cdot},g_\sharp\alpha),(\RR^d,\norm{\cdot},h_\sharp\beta)) \leq 2\,\Wass_p(g_\sharp\alpha,h_\sharp\beta).\)
\begin{equation}\label{eq-gw-procrustes-upper}
\GW((\RR^d,\norm{\cdot},\alpha),(\RR^d,\norm{\cdot},\beta))
\leq
2\,\Wass_{p,\mathrm E(d)}([\alpha],[\beta]),
\end{equation}
for the same exponent $p$ and $\De(u,v)=|u-v|$. Thus a good Wasserstein--Procrustes registration forces small GW distortion.

\begin{equation}\label{eq-wasserstein-procrustes-coupling-update}
\P^{(k)}
\in
\argmin_{\P\in\CouplingsD(\a,\b)}
\sum_{i,j}\P_{ij}\norm{R^{(k)}x_i+t^{(k)}-y_j}^2.
\end{equation}
\begin{equation}\label{eq-wasserstein-procrustes-rigid-update}
(R^{(k+1)},t^{(k+1)})
\in
\argmin_{R\in\mathrm O(d),\,t\in\RR^d}
\sum_{i,j}\P^{(k)}_{ij}\norm{Rx_i+t-y_j}^2,
\end{equation}
\begin{prop}[Rigid update for a fixed coupling]\label{prop-wasserstein-procrustes-rigid-update}
Let $\alpha=\sum_i a_i\delta_{x_i}$ and $\beta=\sum_j b_j\delta_{y_j}$, and fix $\P\in\CouplingsD(\a,\b)$. Define
\(\bar x=\sum_i a_i x_i,\qquad \bar y=\sum_j b_j y_j,\qquad M_\P=\sum_{i,j} \P_{ij}(y_j-\bar y)(x_i-\bar x)^\top.\)
If $M_\P=U\Sigma V^\top$ is a singular value decomposition, then the minimizers over the full orthogonal group of
\(\min_{R\in\mathrm O(d),\,t\in\RR^d} \sum_{i,j}\P_{ij}\norm{Rx_i+t-y_j}^2\)
are given by $R^\star=UV^\top$ and $t^\star=\bar y-R^\star\bar x$, up to the usual non-uniqueness when $M_\P$ is rank deficient. If one imposes $R\in\mathrm{SO}(d)$, set
\(D=\diag(1,\ldots,1,\det(UV^\top)), \qquad R^\star=UDV^\top, \qquad t^\star=\bar y-R^\star\bar x.\)
\end{prop}
\begin{proof}
For fixed $R$, differentiating with respect to $t$ gives $t=\bar y-R\bar x$. Substituting this value centers the variables and leaves
\(\sum_{i,j}\P_{ij}\norm{R(x_i-\bar x)-(y_j-\bar y)}^2.\)
The two quadratic terms independent of $R$ are fixed, so the problem is equivalent to maximizing
\(\sum_{i,j}\P_{ij}\dotp{R(x_i-\bar x)}{y_j-\bar y} = \tr(R^\top M_\P).\)
Von Neumann's trace inequality gives $\tr(R^\top U\Sigma V^\top)\leq\tr(\Sigma)$ for $R\in\mathrm O(d)$, with equality for $R=UV^\top$. Under the constraint $R\in\mathrm{SO}(d)$, the same argument applies with the additional determinant constraint on $U^\top R V$, which gives the displayed diagonal correction.
\end{proof}
\subsection{Vector Quantiles and Linear Optimal Transport}
\label{sec-linear-ot}
\label{sec-vector-quantiles-linearized-transport}
\paragraph{Vector quantiles.}
Assume that $\rho$ is absolutely continuous. For a target law $\mu$ with finite second moment, its vector quantile relative to $\rho$ is the Brenier map

\(\T_\mu=\nabla\phi_\mu, \qquad (\T_\mu)_\sharp\rho=\mu,\)
\(\min_{\T_\sharp\rho=\mu} \int \norm{x-\T(x)}^2\d\rho(x).\)
\paragraph{Linearized Wasserstein coordinates.}
\begin{equation}\label{eq-lot-embedding}
\alpha \mapsto T_\alpha-\Id\in L^2(\rho;\RR^d),
\qquad
\LOT_\rho(\alpha,\beta)=\norm{T_\alpha-T_\beta}_{L^2(\rho)}.
\end{equation}
If one of the two targets equals the reference, the linearized distance is exact: for instance, $\LOT_\rho(\rho,\alpha)=\norm{T_\alpha-\Id}_{L^2(\rho)}=\Wass_2(\rho,\alpha)$. For a family $(\alpha_s)_s$ with weights $(\lambda_s)_s$, the linearized barycenter is obtained by averaging maps,

\(\bar T=\sum_s\lambda_s T_{\alpha_s}, \qquad \bar\alpha_{\LOT}=\bar T_\sharp\rho.\)
This is exact in one dimension, where quantile functions linearize $\Wass_2$, and it is especially useful when many barycenters with changing weights must be evaluated quickly.

\paragraph{Principal components in linear OT coordinates.}
\(z_i \eqdef T_{\alpha_i}-\Id, \qquad \bar z \eqdef \frac1N\sum_{i=1}^N z_i.\)
\(\mathcal C_{\LOT} h \eqdef \frac1N\sum_{i=1}^N \dotp{z_i-\bar z}{h}_{L^2(\rho)} (z_i-\bar z),\)
\(e_k=\frac{1}{\sqrt{N\lambda_k}}\sum_{i=1}^N v_i^{(k)}(z_i-\bar z)\)
\(a_{i,k} \eqdef \dotp{z_i-\bar z}{e_k}_{L^2(\rho)}.\)
\(T_a(x) \eqdef x+\bar z(x)+\sum_{k=1}^m a_k e_k(x), \qquad \alpha_a \eqdef (T_a)_\sharp\rho.\)
\begin{prop}[Local stability of linear OT]\label{prop-linear-ot-stability}
Assume that the measures are supported on a fixed convex compact set, with densities bounded above and below, and that the Brenier maps from $\rho$ are regular. Then, for $\alpha,\beta$ in a sufficiently small regular neighborhood of $\rho$,
\(\Wass_2(\alpha,\beta)\leq \LOT_\rho(\alpha,\beta) \quad\text{and}\quad \LOT_\rho(\alpha,\beta)\leq C\Wass_2(\alpha,\beta)^\eta\)
for constants $C>0$ and $\eta\in(0,1]$ depending on regularity.
\end{prop}
\begin{proof}
The first inequality is immediate: $(T_\alpha,T_\beta)_\sharp\rho$ is a feasible coupling between $\alpha$ and $\beta$. The reverse local estimate is a standard stability statement for the Monge--Amp\`ere equation under the stated regularity assumptions: changes in the target measure control changes in the Brenier potential in H\"older norms, hence control $T_\alpha-T_\beta$ in $L^2(\rho)$. In simple one-dimensional settings, quantile functions make this exact with $\eta=1$.
In several dimensions one should not read the statement as a global Lipschitz estimate in $\Wass_2$. Quantitative stability results for semi-discrete and Monge--Amp\`ere maps give H\"older exponents depending on the dimension, density bounds, support geometry and regularity; see for instance the estimates of M\'erigot, Delalande and Chazal. Under stronger smooth perturbations of uniformly convex smooth densities, elliptic regularity can give Lipschitz dependence in stronger function norms, but converting those controls to Wasserstein perturbations generally loses powers.
\end{proof}
\subsection{Spectral and Robust Wasserstein Distances}
\label{sec-spectral-subspace-wasserstein}
\begin{defn}[Monotone spectral gauge]\label{def-monotone-spectral-gauge}
A monotone spectral gauge on positive semidefinite matrices is a convex, positively $1$-homogeneous map $\gamma:\mathbb S_+^d\to\RR_+$ such that $\gamma(M)=0$ only for $M=0$, $\gamma(QM\transp{Q})=\gamma(M)$ for every orthogonal matrix $Q$, and
\(0\preceq M\preceq N \quad\Longrightarrow\quad \gamma(M)\leq\gamma(N).\)
\end{defn}
\begin{defn}[Spectral Wasserstein distance]\label{def-spectral-wasserstein}
Let $\gamma$ be a monotone spectral gauge. For a coupling $\pi\in\Couplings(\alpha,\beta)$, define its displacement covariance
\(M_\pi\eqdef\int (x-y)(x-y)^\top\d\pi(x,y).\)
The spectral Wasserstein distance associated with $\gamma$ is
\eql{\label{eq-spectral-wasserstein}
\Wass_\gamma(\alpha,\beta)^2
\eqdef
\inf_{\pi\in\Couplings(\alpha,\beta)}\gamma(M_\pi).
}
\end{defn}
The special case $\gamma(M)=\tr(M)$ gives the usual quadratic Wasserstein distance $\Wass_2$. The spectral gauge $\gamma(M)=\lambda_{\max}(M)$ instead measures the worst transported variance direction.

\eql{\label{eq-quadratic-projected-cost}
\Wass_{2,A}(\alpha,\beta)^2
\eqdef
\inf_{\pi\in\Couplings(\alpha,\beta)}
\int (x-y)^\top A(x-y)\d\pi(x,y)
=
\Wass_2((A^{1/2})_\sharp\alpha,(A^{1/2})_\sharp\beta)^2.
}
\eql{\label{eq-spectral-polar-set}
\mathcal B_\gamma
\eqdef
\enscond{A\succeq0}{\tr(AM)\leq\gamma(M)\quad\text{for all }M\succeq0},
}
\begin{prop}[Robust representation and metric equivalence]\label{prop-spectral-wasserstein-robust}
Assume, for simplicity, that the measures are compactly supported and that $\gamma$ is closed and finite on the positive semidefinite cone. Then
\(\Wass_\gamma(\alpha,\beta)^2 = \sup_{A\in\mathcal B_\gamma} \Wass_{2,A}(\alpha,\beta)^2.\)
If there exist constants $0<a\leq b<+\infty$ such that $a\Id\in\mathcal B_\gamma$ and $\mathcal B_\gamma\subset\enscond{A}{0\preceq A\preceq b\Id}$, equivalently
\(a\tr(M)\leq\gamma(M)\leq b\tr(M) \qquad (M\succeq0),\)
then
\(\sqrt a\,\Wass_2(\alpha,\beta) \leq \Wass_\gamma(\alpha,\beta) \leq \sqrt b\,\Wass_2(\alpha,\beta).\)
In particular, $\Wass_\gamma$ is a distance. When $\gamma$ is the restriction of a norm to the positive semidefinite cone, these bounds hold automatically in finite dimension, so $\Wass_\gamma$ is equivalent to $\Wass_2$ on measures with finite second moments.
\end{prop}
\begin{proof}
Using the polar representation of $\gamma$, \(\Wass_\gamma(\alpha,\beta)^2 = \inf_{\pi\in\Couplings(\alpha,\beta)} \sup_{A\in\mathcal B_\gamma}\tr(AM_\pi).\)
The coupling set is convex and compact for weak convergence under compact support. Since $\gamma$ is a finite gauge on a finite-dimensional cone and vanishes only at the origin, it is equivalent to the trace norm on the slice $\tr(M)=1$, so $\mathcal B_\gamma$ is convex and compact. The map $(\pi,A)\mapsto\tr(AM_\pi)$ is affine in each variable and continuous. Sion's minimax theorem gives
\[
\inf_\pi\sup_{A\in\mathcal B_\gamma}\tr(AM_\pi)
=
\sup_{A\in\mathcal B_\gamma}\inf_\pi\tr(AM_\pi)
=
\sup_{A\in\mathcal B_\gamma}\Wass_{2,A}(\alpha,\beta)^2.
\]
For fixed $A\succeq0$, $\Wass_{2,A}$ is the Wasserstein pseudodistance associated with the seminorm $x\mapsto\norm{A^{1/2}x}$. Since all terms are nonnegative, the robust identity also gives
\(\Wass_\gamma(\alpha,\beta) = \sup_{A\in\mathcal B_\gamma}\Wass_{2,A}(\alpha,\beta).\)
A supremum of pseudodistances is symmetric and satisfies the triangle inequality.

If $a\Id\in\mathcal B_\gamma$ and $A\preceq b\Id$ for all $A\in\mathcal B_\gamma$, then
\(a\Wass_2(\alpha,\beta)^2 = \Wass_{2,a\Id}(\alpha,\beta)^2 \leq \Wass_\gamma(\alpha,\beta)^2 \leq b\Wass_2(\alpha,\beta)^2,\)
which proves definiteness and equivalence with $\Wass_2$. The equivalence between these operator bounds and $a\tr(M)\leq\gamma(M)\leq b\tr(M)$ follows directly from the polar formula. In finite dimension, any norm restricted to the positive semidefinite cone is equivalent to the trace norm on that cone. The finite-second-moment case follows by truncation when these norm-equivalence bounds hold.
\end{proof}
\begin{defn}[Subspace robust Wasserstein]\label{def-subspace-robust-wasserstein}
For $1\leq k\leq d$, the Paty--Cuturi subspace robust Wasserstein distance is
\[
\SRW_{2,k}(\alpha,\beta)
\eqdef
\sup_{\transp{U}U=\Id_k}
\Wass_2((\transp{U})_\sharp\alpha,(\transp{U})_\sharp\beta)
=
\sup_{P^2=P=\transp{P},\ \tr(P)=k}\Wass_{2,P}(\alpha,\beta).
\]
\end{defn}
\(\gamma_k(M)=\sum_{\ell=1}^k\lambda_\ell(M),\)
\(\mathcal B_{\gamma_k}=\enscond{A}{0\preceq A\preceq\Id,\ \tr(A)\leq k}.\)
Thus $k=d$ gives $\gamma_d(M)=\tr(M)$ and recovers $\Wass_2$. The convex hull of rank-$k$ projectors is

\(\enscond{A}{0\preceq A\preceq\Id,\ \tr(A)=k},\)
and, since $M\succeq0$, the associated support function is the same Ky Fan gauge. Thus $\Wass_{\gamma_k}$ is the convexified spectral counterpart of $\SRW_{2,k}$, while $\SRW_{2,k}$ keeps the original non-convex rank constraint.