\section{Optimal Matching between Point Clouds}
\label{sec-matching}
\subsection{Monge Problem for Discrete Points}
\label{sec-monge-pbm}
\paragraph{Matching Problem.}
Given a cost matrix $(\C_{i,j})_{i \in \range{n}, j \in \range{m}}$ and assuming $n=m$, the optimal assignment problem aims to find a bijection $\sigma$ within the set $\Perm(n)$ of permutations of $n$ elements that solves

\eql{\label{eq-optimal-assignment}
\umin{\sigma \in \Perm(n)} \frac{1}{n}\sum_{i=1}^n \C_{i,\sigma(i)}.
}
\paragraph{1-D Case.}
\begin{proposition}[Monotone matching on the line]\label{prop-matching-1d-monotone}
Assume that the points $(x_i)_i$ and $(y_j)_j$ are pairwise distinct. If the cost is of the form $\C_{i,j}=h(x_i-y_j)$, where $h: \RR \rightarrow \RR^+$ is strictly convex (for example, $\C_{i,j}=|x_i-y_j|^p$ for $p > 1$), then any optimal $\sigma$ defines a strictly increasing map $x_i \mapsto y_{\sigma(i)}$ (and thus is unique), i.e.,
\(\forall (i,i'), \quad (x_i-x_{i'})(y_{\sigma(i)}-y_{\sigma(i')}) > 0.\)
\end{proposition}
\begin{proof}
Indeed, if this property is violated, i.e., there exists $(i,i')$ such that $(x_i-x_{i'})(y_{\sigma(i)}-y_{\sigma(i')}) < 0$, then one can define a permutation $\tilde{\sigma}$ by swapping the match, i.e., $\tilde{\sigma}(i)=\sigma(i')$ and $\tilde{\sigma}(i')=\sigma(i)$, yielding a strictly better cost, as proved in the following fact.

Let $h:\RR\to\RR$ be strictly convex and let $x < x'$ and $y < y'$. Then
\(h(x-y)+h(x'-y') < h(x-y')+h(x'-y).\)
We set the gap $d:=y'-y > 0$ and define for every $s\in\RR$
\(D(s):=\frac{h(s)-h(s-d)}{d} \quad\text{and}\quad \Delta:=h(x-y')+h(x'-y)-h(x-y)-h(x'-y') = d ( D(x'-y)-D(x-y) ).\)
Because $h$ is strictly convex, $D$ is strictly increasing.
Since $x-y < x'-y$, monotonicity yields $D(x-y) < D(x'-y)$, that is $\Delta > 0$.
\end{proof}
\(x_{\sigma_X(1)} \leq x_{\sigma_X(2)} \leq \ldots \qandq y_{\sigma_Y(1)} \leq y_{\sigma_Y(2)} \leq \ldots\)
Note that if $h$ is strictly convex, then all optimal assignments are increasing, and if the points are all distinct, this increasing map is unique. If $h$ is not strictly convex, for instance $c(x,y)=|x-y|$, non-increasing optimal assignments can also exist.
\begin{example}[Book-shifting for $\Wass_1$]\label{ex-book-shifting-w1}
Let
\(\al=\frac12\mathds{1}_{[0,2]}\d x\) and \(\be=\frac12\mathds{1}_{[1,3]}\d y\).
Both \(T_{\rm tr}(x)=x+1\) and
\[
T_{\rm book}(x)=
\begin{cases}
x+2, & 0\leq x\leq1,\\
x, & 1<x\leq2,
\end{cases}
\]
push $\al$ to $\be$. For any coupling $\pi\in\Couplings(\al,\be)$,
\(\int |y-x|\d\pi(x,y)\geq\int (y-x)\d\pi(x,y)=1,\)
and equality is attained by every coupling supported on $\{y\geq x\}$. The two displayed maps are therefore $\Wass_1$-optimal. The second map leaves the overlap $[1,2]$ fixed and shifts the remaining mass like a book. For $p>1$, strict convexity selects the unique quantile coupling \(x\mapsto x+1\): its $p$-cost is $1$, while \(T_{\rm book}\) has $p$-cost \(2^{p-1}>1\).
\end{example}

\paragraph{Optimal transport on the circle.}\label{rem-circle-ot-cut}
\(d_{\mathbb S^1}(x,y):=\min_{k\in\ZZ}|x-y+k|, \qquad c_p(x,y):=d_{\mathbb S^1}(x,y)^p,\qquad p>1.\)
\begin{proposition}[Discrete circle transport by a cut]\label{prop-circle-ot-cut}
Let $x_1,\ldots,x_n$ and $y_1,\ldots,y_n$ be two families of distinct points on $\mathbb S^1$, with equal weights. Let $x_{(1)},\ldots,x_{(n)}$ and $y_{(1)},\ldots,y_{(n)}$ denote any fixed cyclic orderings, with the convention $y_{(k+n)}=y_{(k)}$. For the cost $c_p$, $p>1$, an optimal assignment is one of the cyclic shifts
\(x_{(k)}\longmapsto y_{(k+s)},\qquad k\in\range{n},\quad s\in\{0,\ldots,n-1\},\)
and is found by minimizing
\[
\sum_{k=1}^n d_{\mathbb S^1}\!\left(x_{(k)},y_{(k+s)}\right)^p
\]
over the $n$ possible shifts. Equivalently, for an optimal shift one can choose a cut $\theta\in\mathbb S^1\setminus(\{x_i\}_i\cup\{y_j\}_j)$ so that, after lifting all points to $(\theta,\theta+1)$ and sorting them, the optimal matching is the equal-rank monotone matching on this interval.
\end{proposition}
\begin{proof}
Call two matched pairs cyclically inverted if the circular order of their source endpoints is opposite to the circular order of their target endpoints. Among optimal assignments, choose one with the smallest number of such inversions. The elementary exchange step is the circular analogue of the line argument in Proposition~\ref{prop-matching-1d-monotone}: if two matched pairs are inverted, then cutting the circle in a gap which does not meet the four endpoints and choosing integer lifts realizes the four geodesic distances involved in the exchange as ordinary distances between two ordered source lifts and two oppositely ordered target lifts. The one-dimensional Monge inequality for the strictly convex function $r\mapsto |r|^p$ then shows that swapping the two targets cannot increase the cost, and decreases it unless the four endpoints are in a degenerate tie configuration.

Thus an optimal assignment can be chosen with no cyclic inversion. A bijection between two finite cyclically ordered sets with no cyclic inversion is a rotation of the order, hence a cyclic shift. This shift specifies how the two cyclic orderings should be opened; after this cut, the rotation becomes an ordinary linear order and the matching is the equal-rank monotone assignment on the unfolded interval. Conversely, each cut gives one such cyclic shift, so minimizing over the finitely many shifts gives an optimal discrete circle assignment. Repeated points or ties are obtained by the same argument after an arbitrarily small perturbation and a limiting passage. This is the discrete fast circle-Monge construction.
\end{proof}
\paragraph{Rational weights.}
\begin{proposition}[Rational weights as duplicated uniform matching]\label{prop-rational-weights-duplicated-matching}
Let
\(\mu=\sum_{i=1}^n \frac{k_i}{N}\delta_{x_i}, \qquad \nu=\sum_{j=1}^m \frac{\ell_j}{N}\delta_{y_j}, \qquad \sum_i k_i=\sum_j\ell_j=N,\)
with $k_i,\ell_j\in\NN$. The discrete Kantorovich problem between $(\mu,\nu)$ is equivalent to the uniform assignment problem obtained by replacing each $x_i$ by $k_i$ identical copies and each $y_j$ by $\ell_j$ identical copies. More precisely, after multiplying transport masses by $N$, optimal couplings correspond to optimal integer count matrices $(n_{ij})$ with row sums $k_i$ and column sums $\ell_j$, and these count matrices are exactly the collapsed form of assignments between the duplicated clouds.
\end{proposition}
\begin{proof}
Any assignment between the duplicated source and target clouds defines integers $n_{ij}$ counting how many copied particles of type $x_i$ are matched to copied particles of type $y_j$. These counts satisfy $\sum_j n_{ij}=k_i$ and $\sum_i n_{ij}=\ell_j$, and the associated coupling $P_{ij}=n_{ij}/N$ has marginals $k_i/N$ and $\ell_j/N$. The assignment cost is
\(\frac1N\sum_{i,j} n_{ij}c(x_i,y_j) = \sum_{i,j}P_{ij}c(x_i,y_j).\)
Conversely, any nonnegative integer count matrix with those row and column sums can be realized by matching the corresponding duplicated particles. Finally, the transportation constraint matrix is totally unimodular, so the linear problem with integer supplies and demands has an optimal integer count matrix. Thus the optimum of the rational-weight Kantorovich problem is the same as the optimum of the duplicated uniform assignment problem.
\end{proof}
\paragraph{2D case.}
\begin{proposition}[Non-crossing optimal matchings]
In dimension 2, for $c(x,y) = \norm{x-y}$, if $\sigma$ is an optimal assignment, then segments $[x_i,y_{\sigma(i)}]$ cannot cross.
\end{proposition}
\begin{proof}
If two segments $[x_i,y_{\sigma(i)}]$ and $[x_j,y_{\sigma(j)}]$ cross at an interior point $z$, then the triangle inequality gives
\(\norm{x_i-y_{\sigma(j)}}+\norm{x_j-y_{\sigma(i)}} < \norm{x_i-y_{\sigma(i)}}+\norm{x_j-y_{\sigma(j)}}.\)
The assignment obtained by swapping $(\sigma(i),\sigma(j))$ therefore has a strictly smaller cost, which contradicts optimality.
\end{proof}
\(C_n=\frac{1}{n+1}\binom{2n}{n}\sim \frac{4^n}{\sqrt{\pi}n^{3/2}}.\)
\subsection{Matching Algorithms}
\paragraph{Hungarian primal-dual method.}
\(u_i+v_j\leq C_{i,j}\qquad\forall i,j.\)
\[
\delta=\min_{i\in S,\ j\notin T}\bigl(C_{i,j}-u_i-v_j\bigr),
\qquad
u_i\leftarrow u_i+\delta\ (i\in S),\qquad
v_j\leftarrow v_j-\delta\ (j\in T).
\]
\begin{prop}[Correctness and complexity of the Hungarian primal-dual method]\label{prop-hungarian-correct}
Assume the Hungarian method terminates with a perfect matching $\sigma$ contained in the equality graph
\(E(u,v)=\enscond{(i,j)}{u_i+v_j=C_{i,j}},\)
where $(u,v)$ is dual feasible, i.e. $u_i+v_j\leq C_{i,j}$ for all $(i,j)$. Then $\sigma$ is an optimal assignment. Moreover, the usual Hungarian updates terminate after finitely many augmentations.
With maintained slacks, the method uses $O(n^3)$ arithmetic operations.
\end{prop}
\begin{proof}
For any permutation $\tau$, dual feasibility gives \(\sum_i C_{i,\tau(i)} \geq \sum_i (u_i+v_{\tau(i)}) = \sum_i u_i+\sum_j v_j.\)
This is the weak duality lower bound. If $\sigma$ is contained in the equality graph, then
\(\sum_i C_{i,\sigma(i)} = \sum_i u_i+\sum_j v_j,\)
so the primal cost of $\sigma$ reaches the dual lower bound and is optimal.

At each successful augmentation, the matching cardinality increases by one, so there are at most $n$ augmentations. During one augmentation phase, the algorithm grows an alternating tree in the equality graph. If no augmenting path is available, the dual update uses the smallest slack of an edge leaving the current tree. For edges inside the tree, adding $\delta$ to source labels and subtracting $\delta$ from target labels preserves tightness; for edges from $S$ to $T^c$, the definition of $\delta$ preserves feasibility and makes at least one new edge tight; all other inequalities are unchanged or become looser. Thus the reachable sets strictly grow between two failed augmentation attempts, and they can grow at most $n$ times within one phase. If the current slacks $\min_{i\in S}(C_{i,j}-u_i-v_j)$ are updated when a source enters $S$, each tree expansion costs $O(n)$. A phase therefore costs $O(n^2)$, and the $n$ phases give $O(n^3)$ operations. Hence the method reaches a perfect optimal matching.
\end{proof}
