\section{Monge Problem between Measures}
\label{sec-monge}
\subsection{Measures}
\label{sec-measures}
\paragraph{Histograms.}
\begin{defn}[Probability simplex]\label{def-probability-simplex}
The probability simplex of length $n$ is \(\simplex_n \eqdef \enscond{\a \in \RR_+^n}{ \sum_{i=1}^n \a_i = 1 }.\)
Its elements are also called probability vectors or histograms.
\end{defn}
\paragraph{Discrete measure, empirical measure.}
\begin{defn}[Discrete measure]\label{def-discrete-measure}
A discrete measure with weights $\a$ and locations $x_1,\dots,x_n\in\X$ is
\eql{\label{eq-discr-meas}
\al = \sum_{i=1}^n \a_i \de_{x_i},
}
where $\de_x$ is the Dirac mass at position $x$. It is a probability measure if $\a\in\simplex_n$, and a positive measure if all weights $\a_i$ are nonnegative.
\end{defn}
\paragraph{General measures.}
A Dirac measure $\de_x$ is then defined as $\de_x(A)=1$ if $x \in A$ and $0$ otherwise, and this extends by linearity for discrete measures of the form~\eqref{eq-discr-meas} as

\(\al(A) = \sum_{x_i \in A} \a_i\)
We denote $\Mm_+(\X)$ the subset of all positive measures on $\X$, i.e. $\al(A) \geq 0$ (and $\al(\X)<+\infty$ for the measure to be finite). The set of probability measures is denoted $\Mm_+^1(\X)$, which means that any $\al \in \Mm_+^1(\X)$ is positive, and that $\al(\X)=1$.

\paragraph{Polish metric spaces.}
\begin{defn}[Polish metric space]\label{def-polish-metric-space}
A metric space $(\X,d)$ is Polish if it is complete and separable: every Cauchy sequence converges in $\X$, and $\X$ contains a countable dense subset. More generally, a topological space is called Polish if its topology can be induced by some complete separable metric.
\end{defn}
\(\Ss=C([0,1];\RR^d),\qquad d_\infty(\gamma,\eta) \eqdef \sup_{t\in[0,1]}\norm{\gamma(t)-\eta(t)}\)
\begin{defn}[Support of a measure]\label{def:support}
The support $\supp(\al)$ of a Borel measure $\al$ on a metric space $\Xx$ is the smallest closed set of full $\al$-mass. Equivalently, $x\in\supp(\al)$ if every open ball centered at $x$ has positive $\al$-mass.
\end{defn}
\paragraph{Radon measures.}
Using Lebesgue integration, a Borel measure can be used to compute the integral of measurable functions (i.e. such that level sets $\enscond{x}{f(x) < t}$ are Borel sets), and we denote this pairing as

\(\dotp{f}{\al} \eqdef \int f(x) \d\al(x).\)
\(\int_\X f(x) \d\al(x) = \sum_{i=1}^n \a_i f(x_i).\)
\paragraph{Relative densities.}
\begin{defn}[Relative density]\label{def-relative-density}
If $\al$ is absolutely continuous with respect to a reference measure $\lambda$, its relative density is the Radon--Nikodym derivative
\(\density{\al}\eqdef\frac{\d\al}{\d\lambda}, \qquad \d\al(x)=\density{\al}(x)\d\lambda(x).\)
Equivalently, for all $h\in\Cc(\Xx)$, \(\int_{\Xx} h(x) \d\al(x) = \int_{\Xx} h(x) \density{\al}(x) \d\lambda(x).\)
\end{defn}
\paragraph{Total variation norm.}
\begin{defn}[Total variation]\label{defn-total-variation}
For a finite signed Radon measure $\al$ on a compact space $\Xx$, \(\norm{\al}_{\TV} \eqdef \usup{f \in \Cc(\Xx)} \enscond{\dotp{f}{\al}}{\norm{f}_\infty \leq 1}.\)
\end{defn}
\(|\al|(A) \eqdef \usup{A=\cup_i B_i} \sum_i |\al(B_i)|,\)
where the supremum is over finite or countable measurable partitions of $A$. Thus, if $\al=\sum_i \a_i \de_{x_i}$ with distinct atoms, $|\al|=\sum_i |\a_i|\de_{x_i}$; if $\d\al(x)=\rho(x)\d\lambda(x)$, then $\d|\al|(x)=|\rho(x)|\d\lambda(x)$.

\begin{prop}[Dual and measure definitions of total variation]\label{prop-tv-dual-measure}
For a finite signed Radon measure $\al$ on a compact space $\Xx$, \(\norm{\al}_{\TV}=|\al|(\Xx).\)
\end{prop}
\begin{proof}
The inequality $\norm{\al}_{\TV}\leq |\al|(\Xx)$ follows from
\(\abs{\int f\,\d\al}\leq \int |f|\,\d|\al|\leq \norm{f}_\infty |\al|(\Xx).\)
For the reverse inequality, write the Jordan decomposition $\al=\al^+-\al^-$, so that $|\al|=\al^++\al^-$. The measurable sign $s=\frac{\d\al}{\d|\al|}$ takes values in $\{-1,1\}$ outside a $|\al|$-null set and satisfies $\d\al=s\,\d|\al|$. By regularity of Radon measures on compact spaces, $s$ can be approximated in $L^1(|\al|)$ by continuous functions $f_k$ with $\norm{f_k}_\infty\leq1$. Hence $\int f_k\,\d\al\to\int s\,\d\al=|\al|(\Xx)$, which proves the equality. The final statement is the Riesz--Markov--Kakutani representation theorem with this norm.
\end{proof}
For two absolutely continuous measures $\d\al=\rho_\al\d\lambda$ and $\d\be=\rho_\be\d\lambda$, this gives the concrete formula

\(\norm{\al-\be}_{\TV} = \int_\Xx |\rho_\al(x)-\rho_\be(x)|\,\d\lambda(x).\)
For discrete measures on the union support \(z_k\), write
\(\al=\sum_{i=1}^n \tilde a_i\delta_{x_i}, \qquad \be=\sum_{j=1}^m \tilde b_j\delta_{y_j},\)
\(a_k\eqdef\sum_{i:x_i=z_k}\tilde a_i, \qquad b_k\eqdef\sum_{j:y_j=z_k}\tilde b_j.\)
Then $\al=\sum_{k=1}^r a_k\delta_{z_k}$ and $\be=\sum_{k=1}^r b_k\delta_{z_k}$. This convention merges masses located at the same point before comparing the two measures.

\(\norm{\al-\be}_{\TV} = \sum_{k=1}^r |a_k-b_k|.\)
\paragraph{Probabilistic interpretation.}
Let $(\Omega,\Ff,\PP)$ be an abstract probability space. If \(X:\Omega\to\X\) is measurable, its law is \(\al=X_\sharp\PP\), i.e.
\(\al(A)=\PP(\enscond{\omega\in\Omega}{X(\omega)\in A}) \qquad\text{for Borel sets } A\subset \X.\)
\(\int_\X f(x)\,\d\al(x)=\mathbb{E}[f(X)].\)
\subsection{Push Forward}
\label{sec-push-forward}
For a continuous map $\T: \X \rightarrow \Y$, the push-forward operator is $\T_\sharp: \Mm(\X) \rightarrow \Mm(\Y)$.

For a Dirac mass, $\T_\sharp \de_x=\de_{\T(x)}$, and the formula extends to arbitrary measures by linearity.

For discrete measures~\eqref{eq-discr-meas}, the pushforward operation consists simply in moving the positions of all the points in the support of the measure

\(\T_{\sharp} \al \eqdef \sum_i \a_i \de_{\T(x_i)}.\)
\begin{defn}[Push-forward]\label{defn-pushfwd}
For $\T: \X \rightarrow \Y$, the push forward measure $\be = \T_\sharp \al \in \Mm(\Y)$ of some $\al \in \Mm(\X)$ satisfies
\eql{\label{eq-push-fwd}
\foralls h \in \Cc(\Y), \quad \int_\Y h(y) \d \be(y) = \int_\X h(\T(x)) \d\al(x).
}
Equivalently, for any measurable set $B \subset \Y$, one has
\eql{\label{eq-equiv-pushfwd}
\be(B) = \al( \enscond{x \in \X}{\T(x) \in B} ).
}
Note that $\T_\sharp$ preserves positivity and total mass, so that if $\al \in \Mm_+^1(\X)$ then $\T_\sharp \al \in \Mm_+^1(\Y)$.
\end{defn}
\begin{prop}[Push-forward formula for densities]\label{prop-push-forward-densities}
Let $\alpha$ and $\beta$ have densities $\density{\al}$ and $\density{\be}$ on open subsets of $\RR^\dims$. Assume that $\T$ is a $C^1$ diffeomorphism and that $\beta=\T_\sharp\alpha$. Then, for all $x$,
\eql{\label{eq-pfwd-density}
\density{\al}(x) = |\det(\T'(x))| \density{\be}(\T(x)).
}
Equivalently, for $y=\T(x)$, \(\rho_\beta(y)=\rho_\alpha(\T^{-1}y)\, \frac{1}{|\det(\T'(\T^{-1}y))|}.\)
\end{prop}
\begin{proof}
Explicitly doing the change of variable $y=\T(x)$, so that $\d y = |\det(\T'(x))| \d x$ in formula~\eqref{eq-push-fwd} for measures with densities $(\density{\al},\density{\be})$ on $\RR^\dims$, one has for all $h \in \Cc(\Yy)$
\begin{align*}
\int_\Yy h(y)\rho_\be(y) \d y
&= \int_\Xx h(\T(x)) \rho_\al(x) \d x \\
&= \int_\Yy h(y) \rho_\al(\T^{-1}y) \frac{\d y}{|\det(\T'(\T^{-1} y))|},
\end{align*}
which shows that
\(\rho_\be(y) = \rho_\al(\T^{-1}y) \frac{1}{|\det(\T'(\T^{-1} y))|}.\)
Since $\T$ is a diffeomorphism, one obtains equivalently
\(\density{\al}(x) = |\det(\T'(x))| \density{\be}(\T(x))\)
where $\T'(x) \in \RR^{\dims \times \dims}$ is the Jacobian matrix of $\T$ (the matrix formed by taking the gradient of each coordinate of $\T$).

This implies, denoting $y=\T(x)$
\(|\det(\T'(x))| = \frac{ \density{\al}(x) }{ \density{\be}(y) }.\)
\end{proof}
\subsection{Monge's Formulation}
\label{sec-monge-formulation}
\label{sec-continuous-monge}
\paragraph{Monge problem.}
Given $\al\in\Mm_+^1(\Xx)$, $\be\in\Mm_+^1(\Yy)$ and a cost $c:\Xx\times\Yy\to\RR_+$, the Monge problem is

\eql{\label{eq-monge-continuous}
\Monge_c(\al,\be)
\eqdef
\inf_{\T:\Xx\to\Yy}
\enscond{\int_\Xx c(x,\T(x))\d\al(x)}{\T_\sharp\al=\be}.
}
\begin{prop}[Empirical Monge maps and matchings]\label{prop-empirical-monge-matching}
Assume that the source atoms $x_1,\ldots,x_n$ are distinct and that
\(\al=\frac1n\sum_{i=1}^n\de_{x_i}, \qquad \be=\frac1n\sum_{j=1}^n\de_{y_j}.\)
If $\T_\sharp\al=\be$, then for each distinct target value $z$ in the support of $\be$, exactly $n\be(\{z\})$ source atoms are mapped to $z$. In particular, if the $y_j$ are distinct, then there exists a permutation $\sigma\in\Perm(n)$ such that $\T(x_i)=y_{\sigma(i)}$ for all $i$. Conversely, every such assignment of source atoms to target atoms with the correct masses defines a feasible Monge map on the support of $\al$, and in the distinct-target case
\(\int_\Xx c(x,\T(x))\d\al(x)=\frac1n\sum_{i=1}^n c(x_i,y_{\sigma(i)}).\)
If source locations are repeated, they should first be merged into atoms with larger masses; such atoms cannot be split by a Monge map.
\end{prop}
\begin{proof}
Since $\T_\sharp\al=\be$, all images $\T(x_i)$ must belong to the support of $\be$; otherwise the push-forward would give positive mass to a point outside that support. For any target atom $z$,
\(\be(\{z\})=\al(\T^{-1}(\{z\}))=\frac1n\#\enscond{i}{\T(x_i)=z}.\)
This proves the counting statement. If all target atoms have mass $1/n$, each target receives exactly one source atom, which is a permutation. The converse and the cost identity follow by direct substitution.
\end{proof}
\begin{prop}[Existence of transport maps from atomless sources]\label{prop-existence-transport-map-atomless}
Let $\al$ and $\be$ be Borel probability measures on Polish spaces, and assume that $\al$ is atomless. Then there exists a measurable map $\T$ such that $\T_\sharp\al=\be$.
\end{prop}
\begin{proof}
A standard measure-isomorphism theorem identifies the atomless probability space generated by $\al$ with Lebesgue measure on $[0,1]$, modulo null sets. It is therefore enough to construct a map from $[0,1]$ to the target law $\be$. Choose a Borel isomorphism $i$ from the support of $\be$ onto a Borel subset of $[0,1]$, set $\nu=i_\sharp\be$, and use the generalized inverse of the cumulative distribution function of $\nu$. This map sends Lebesgue measure on $[0,1]$ to $\nu$ and takes values in $i(\supp\be)$ almost surely. Composing with $i^{-1}$ and then with the source isomorphism gives a measurable transport map from $\al$ to $\be$.
\end{proof}
\paragraph{Monge distance.}
When $\Xx=\Yy$ and $c(x,y)=d(x,y)^p$ for a metric $d$, set

\(\Ee_\al(\T) \eqdef \int_\Xx d(x,\T(x))^p\d\al(x).\)
\eql{\label{eq-monge-distance}
\tilde\Wass_p(\al,\be)^p
\eqdef
\inf_{\T:\Xx\to\Xx}
\enscond{\Ee_\al(\T)}{\T_\sharp\al=\be}.
}
If the constraint set is empty, then $\tilde\Wass_p(\al,\be)=+\infty$.

\begin{prop}[Directed Monge distance]\label{prop-directed-monge-distance}
Assume that $\Xx=\Yy$ is a metric space. The quantity $\tilde\Wass_p$ is nonnegative, vanishes only on the diagonal, and satisfies the triangle inequality. It is therefore a directed extended distance: it need not be symmetric and may take the value $+\infty$.
\end{prop}
\begin{proof}
Nonnegativity is immediate. If $\tilde\Wass_p(\al,\be)=0$, choose feasible maps $\T_k$ with $\int d(x,\T_k(x))^p\d\al(x)\to0$. For every bounded $1$-Lipschitz function $h$,
\[
\left|\int h\d\be-\int h\d\al\right|
=
\left|\int h(\T_k(x))-h(x)\d\al(x)\right|
\leq
\left(\int d(x,\T_k(x))^p\d\al(x)\right)^{1/p}
\longrightarrow0.
\]
Since bounded Lipschitz functions separate probability measures on metric spaces, $\al=\be$.

If $\tilde\Wass_p(\al,\be)=+\infty$ while both $\tilde\Wass_p(\al,\ga)$ and $\tilde\Wass_p(\ga,\be)$ were finite, there would be maps $S_\sharp\al=\ga$ and $T_\sharp\ga=\be$, hence $(T\circ S)_\sharp\al=\be$, a contradiction. Thus the inequality is automatic whenever the left-hand side is infinite. In the finite case, fix $\epsilon>0$ and choose $\epsilon$-minimizers $S_\sharp\al=\ga$ and $T_\sharp\ga=\be$ such that
\(\Ee_\al(S)^{1/p}\leq\tilde\Wass_p(\al,\ga)+\epsilon, \qquad \Ee_\ga(T)^{1/p}\leq\tilde\Wass_p(\ga,\be)+\epsilon.\)
The composed map is feasible from $\al$ to $\be$, and Minkowski's inequality gives
\[
\begin{aligned}
\tilde\Wass_p(\al,\be)
&\leq
\left(\int d(x,T(S(x)))^p\d\al(x)\right)^{1/p} \\
&\leq
\left(\int d(x,S(x))^p\d\al(x)\right)^{1/p}
+
\left(\int d(S(x),T(S(x)))^p\d\al(x)\right)^{1/p} \\
&=
\Ee_\al(S)^{1/p}+\Ee_\ga(T)^{1/p}
\leq
\tilde\Wass_p(\al,\ga)+\tilde\Wass_p(\ga,\be)+2\epsilon.
\end{aligned}
\]
Letting $\epsilon\to0$ gives the result.
\end{proof}
\subsection{Existence and Uniqueness of the Monge Map}
\label{sec-monge-existence-uniqueness}
\paragraph{Brenier's theorem.}
\begin{thm}[Brenier]\label{thm-brenier}
Let $\al,\be\in\Mm_+^1(\RR^d)$ have finite second moments, and assume that $\al$ is absolutely continuous with respect to Lebesgue measure. For the quadratic cost $c(x,y)=\norm{x-y}^2$, there exists a convex function $\phi:\RR^d\to\RR\cup\{+\infty\}$ such that
\(\T=\nabla\phi, \qquad \T_\sharp\al=\be,\)
and $\T$ is the unique optimal Monge map $\al$-almost everywhere. The optimal Kantorovich plan is $(\Id,\T)_\sharp\al$.
\end{thm}
\begin{proof}
Kantorovich duality for the quadratic cost gives optimal potentials $(f,g)$ with equality $f(x)+g(y)=\norm{x-y}^2$ on the support of any optimal plan. After subtracting the harmless quadratic terms, this equality is equivalent to the Fenchel equality $\phi(x)+\phi^*(y)=\dotp{x}{y}$ for a convex function $\phi$. Hence the support of every optimal plan lies in the graph of the subdifferential $\partial\phi$. Since $\al$ has a density and convex functions are differentiable Lebesgue-almost everywhere, $\partial\phi(x)$ is a singleton for $\al$-almost every $x$. The plan is therefore concentrated on the graph of $\T=\nabla\phi$, which proves that the relaxed optimizer is induced by a Monge map. Any two optimal plans are concentrated on the same single-valued graph $\al$-almost everywhere, which gives uniqueness of the map. This is the standard route behind Brenier's polar factorization theorem; related existence and uniqueness results for more general strictly convex costs are developed for instance in.
\end{proof}
\(\dotp{\nabla\phi(x)-\nabla\phi(x')}{x-x'}\geq0.\)
\begin{defn}[Measures not charging hypersurfaces]\label{defn-not-charging-hypersurfaces}
A Borel measure $\al$ on $\RR^d$ does not charge hypersurfaces if $\al(S)=0$ for every countably $(d-1)$-rectifiable set $S$, i.e. every set that can be covered, up to an $\mathcal H^{d-1}$-null set, by countably many $C^1$ hypersurfaces.
\end{defn}
\paragraph{Radial measures.}
Radial symmetry gives a useful higher-dimensional case where the Brenier map reduces to a one-dimensional monotone rearrangement. A measure $\mu$ on $\RR^d$ is radial if $Q_\sharp\mu=\mu$ for every orthogonal map $Q$.

\begin{prop}[Optimal transport between radial measures]\label{prop-radial-transport}
Let $\al,\be\in\Mm_+^1(\RR^d)$ be radial probability measures with finite second moments, and assume that $\al$ is absolutely continuous with respect to Lebesgue measure. Define their radial distribution functions
\(F_\al(r)\eqdef \al(\{x:\norm{x}\leq r\}), \qquad F_\be(r)\eqdef \be(\{y:\norm{y}\leq r\}), \qquad r\geq0,\)
and let $F_\be^{-1}(u)\eqdef\inf\{r\geq0:F_\be(r)\geq u\}$ be the generalized inverse. Set
\(\tau(r)\eqdef F_\be^{-1}(F_\al(r)).\)
Then the quadratic Brenier map from $\al$ to $\be$ is the radial map
\(\T(0)=0, \qquad \T(x)=\frac{\tau(\norm{x})}{\norm{x}}x \quad\text{for }x\neq0.\)
Moreover, if $\mu_\al=(x\mapsto\norm{x})_\sharp\al$ is the law of the source radius, then
\(\Wass_2^2(\al,\be) = \int_0^\infty |r-\tau(r)|^2\,\d\mu_\al(r).\)
\end{prop}
\begin{proof}
Let $\mu_\al=(\norm{\cdot})_\sharp\al$ and $\mu_\be=(\norm{\cdot})_\sharp\be$ be the radial laws. Since $\al$ is absolutely continuous, $\mu_\al$ has no atoms. The map $\tau=F_\be^{-1}\circ F_\al$ is therefore the monotone one-dimensional transport from $\mu_\al$ to $\mu_\be$. If $X\sim\al$ and $R=\norm{X}$, then $\tau(R)$ has law $\mu_\be$. Moreover, by radiality, the conditional direction $X/\norm{X}$ given $R=r$ is uniform on the sphere for $\mu_\al$-a.e. $r>0$. Thus $\T$ changes only the radius, from $R$ to $\tau(R)$, and keeps the uniform angular distribution. It follows that $\T_\sharp\al$ is radial with radial law $\mu_\be$, hence $\T_\sharp\al=\be$.

The function $\tau$ is nondecreasing and nonnegative. Define $\psi(r)\eqdef\int_0^r\tau(s)\d s$ on $\RR_+$. Then $\psi$ is convex and nondecreasing, so $\phi(x)\eqdef\psi(\norm{x})$ is convex on $\RR^d$. Since $\al$ is absolutely continuous, it does not charge the origin nor the radii where the monotone function $\tau$ is discontinuous. Thus
\(\nabla\phi(x)=\frac{\tau(\norm{x})}{\norm{x}}x =\T(x) \qquad\text{for }\al\text{-a.e. }x.\)
The map $\T$ is therefore a gradient of a convex potential and pushes $\al$ to $\be$. Brenier's theorem identifies it as the unique quadratic optimal Monge map. The displayed cost formula follows by substituting $\T(x)$ in $\int\norm{x-\T(x)}^2\d\al(x)$.
\end{proof}
If $\al$ is not absolutely continuous, the same radial idea is still useful but has to be interpreted at the level of couplings of the radial variables: atoms on spheres may have to be split, so a Monge map need not exist.

\paragraph{Polar factorization.}
Brenier's theorem does more than solve one transport problem: it provides a canonical way to extract the ``monotone part'' of an arbitrary map. Suppose one starts from a square-integrable deformation $u:\Omega\to\RR^d$, for instance a velocity snapshot, an image deformation or a generative map.

\begin{prop}[Polar factorization]\label{prop-polar-factorization}
Let $\Omega\subset\RR^d$ be endowed with normalized Lebesgue measure $\lambda$, and let $u\in L^2(\Omega;\RR^d)$. Assume that the law $\nu=u_\sharp\lambda$ has finite second moment. Then there exist a measure-preserving map $s:\Omega\to\Omega$ and a convex function $\phi$ such that
\(u=\nabla\phi\circ s \qquad \lambda\text{-a.e.}\)
The Brenier factor $\nabla\phi$ is unique $\lambda$-almost everywhere.
\end{prop}
\begin{proof}
By Brenier's theorem there is a unique gradient of a convex function $\T=\nabla\phi$ transporting $\lambda$ to $\nu$. The maps $u$ and $\T$ have the same image law. The rearrangement theorem for non-atomic probability spaces gives a measure-preserving map $s$ such that $u=\T\circ s$; equivalently, $s$ chooses, with the correct conditional probabilities, preimages of $u(x)$ under $\T$. Uniqueness of the Brenier factor follows from Theorem~\ref{thm-brenier}.
\end{proof}
The name is meant to echo the usual polar decomposition of matrices. This analogy becomes literal for linear maps under the Gaussian reference measure.

\(A = S O, \qquad O=S^\dagger A,\)
\paragraph{Displacement interpolation.}
\label{sec-monge-interpolation}
\begin{defn}[Monge and McCann displacement interpolation]\label{def-monge-mccann-interpolation}
If $\T_\sharp\al=\be$, the Monge interpolation generated by $\T$ is the curve
\(\T_t(x)\eqdef(1-t)x+t\T(x), \qquad \al_t\eqdef(\T_t)_\sharp\al, \qquad t\in[0,1].\)
For the quadratic cost, when $\T$ is the optimal Brenier map, this curve is called McCann's displacement interpolation.
\end{defn}
\begin{prop}[Directed Monge displacement geodesics]\label{prop-monge-displacement-geodesic}
Let $p\geq1$, let $\al,\be\in\Mm_+^1(\RR^d)$ have finite $p$-th moments, and let $\T$ be an optimal map for $\tilde\Wass_p(\al,\be)$ in~\eqref{eq-monge-distance}. Set $\al_t=(\T_t)_\sharp\al$ with $\T_t=(1-t)\Id+t\T$. Assume that, for every $t<1$, $\T_t$ is one-to-one on a full $\al$-measure Borel set, so that $\T_t^{-1}$ is defined $\al_t$-almost everywhere. Then, for $0\leq s\leq t\leq1$,
\(\tilde\Wass_p(\al_s,\al_t)=(t-s)\tilde\Wass_p(\al,\be).\)
Thus $t\mapsto\al_t$ is an oriented constant-speed geodesic for the directed Monge distance. In particular, for $p=2$, this applies to the Brenier map under the hypotheses of Theorem~\ref{thm-brenier}.
\end{prop}
\begin{proof}
The case $s=t$ is trivial, so assume $s<t$. Since $s<1$, the inverse $\T_s^{-1}$ is defined $\al_s$-almost everywhere along the transported particles. Define
\(S_{s,t}\eqdef\T_t\circ\T_s^{-1} \qquad \al_s\text{-a.e.}\)
Then $(S_{s,t})_\sharp\al_s=\al_t$, and, using the optimality of $\T$,
\[
\begin{aligned}
\tilde\Wass_p(\al_s,\al_t)^p
&\leq
\int \norm{S_{s,t}(z)-z}^p\d\al_s(z) \\
&=
\int \norm{\T_t(x)-\T_s(x)}^p\d\al(x)
=
(t-s)^p\tilde\Wass_p(\al,\be)^p.
\end{aligned}
\]
The same particle construction gives $\tilde\Wass_p(\al,\al_s)\leq s\tilde\Wass_p(\al,\be)$. If $t<1$, the map $\T\circ\T_t^{-1}$ sends $\al_t$ to $\be$ and gives $\tilde\Wass_p(\al_t,\be)\leq(1-t)\tilde\Wass_p(\al,\be)$; if $t=1$, this latter distance is zero. Using the triangle inequality from Proposition~\ref{prop-directed-monge-distance},
\[
\tilde\Wass_p(\al,\be)
\leq
\tilde\Wass_p(\al,\al_s)+\tilde\Wass_p(\al_s,\al_t)+\tilde\Wass_p(\al_t,\be)
\leq
s\tilde\Wass_p(\al,\be)+\tilde\Wass_p(\al_s,\al_t)+(1-t)\tilde\Wass_p(\al,\be).
\]
Hence $\tilde\Wass_p(\al_s,\al_t)\geq(t-s)\tilde\Wass_p(\al,\be)$, which proves equality. For a Brenier map $\T=\nabla\phi$, the map $\T_t$ is the gradient of
\(x\mapsto (1-t)\frac{\norm{x}^2}{2}+t\phi(x),\)
which is $(1-t)$-strongly convex for every $t<1$. On the full-measure set where $\phi$ is differentiable, this gives \(\dotp{\T_t(x)-\T_t(y)}{x-y}\geq(1-t)\norm{x-y}^2,\)
so $\T_t$ is injective there.
\end{proof}
\paragraph{Regularity and the Monge--Amp\`ere equation.}
\begin{prop}[Caffarelli regularity]\label{prop-caffarelli-regularity}
Let $\Omega,\Lambda\subset\RR^d$ be bounded uniformly convex domains with $C^2$ boundaries. Let $\al=\rho(x)\d x$ be supported on $\Omega$ and $\be=\eta(y)\d y$ be supported on $\Lambda$, with $0<m\leq\rho,\eta\leq M<+\infty$. If $\rho,\eta\in C^\alpha$ for some $\alpha\in(0,1)$, then the Brenier potential $\phi$ transporting $\al$ to $\be$ is strictly convex and satisfies $\phi\in C^{2,\alpha}_{\mathrm{loc}}(\Omega)$; in particular $\nabla\phi$ is locally H\"older. Under the corresponding boundary compatibility and smoothness assumptions, the regularity holds up to the boundary.
\end{prop}
\begin{proof}
The potential solves the Monge--Amp\`ere equation
\(\det(\nabla^2\phi(x)) = \frac{\rho(x)}{\eta(\nabla\phi(x))}\)
in the Alexandrov sense, with second boundary condition $\nabla\phi(\Omega)=\Lambda$. The density bounds and convexity of the domains give strict convexity and localization of sections. Caffarelli's interior theory then yields $C^{2,\alpha}_{\mathrm{loc}}$ estimates for $\phi$; the boundary statement follows from the boundary regularity theory under uniform convexity and compatibility assumptions.
\end{proof}
For smooth densities, the change-of-variables formula~\eqref{eq-pfwd-density} gives the Monge--Amp\`ere equation

\eql{\label{eq-monge-ampere}
\det(\nabla^2\phi(x))\density{\be}(\nabla\phi(x))=\density{\al}(x).
}
\begin{prop}[Linearization of the Monge--Amp\`ere equation]\label{prop-linearized-monge-ampere}
Let $\rho_\epsilon=\rho_0+\epsilon r+o(\epsilon)$ be a smooth perturbation of a positive reference density $\rho_0$ on a smooth bounded domain, with $\int r=0$. If $\T_\epsilon(x)=x+\epsilon\nabla u(x)+o(\epsilon)$ transports $\rho_0$ to $\rho_\epsilon$, then, to first order,
\(-\nabla\cdot(\rho_0\nabla u)=r.\)
In particular, when $\rho_0$ is constant, the linearized equation is $-\Delta u=r/\rho_0$.
\end{prop}
\begin{proof}
The change-of-variables equation for $\T_\epsilon$ is \(\rho_0(x) = \rho_\epsilon(\T_\epsilon(x))\det(\nabla \T_\epsilon(x)).\)
Expanding $\rho_\epsilon(x+\epsilon\nabla u)=\rho_0(x)+\epsilon r(x)+\epsilon\dotp{\nabla\rho_0}{\nabla u}+o(\epsilon)$ and
\(\det(I+\epsilon\nabla^2u)=1+\epsilon\Delta u+o(\epsilon)\)
gives
\[
\rho_0
=
\rho_0+
\epsilon\bigl(r+\dotp{\nabla\rho_0}{\nabla u}+\rho_0\Delta u\bigr)+o(\epsilon)
=
\rho_0+
\epsilon\bigl(r+\nabla\cdot(\rho_0\nabla u)\bigr)+o(\epsilon).
\]
The first-order term must vanish.
\end{proof}
\subsection{Beyond the Quadratic Euclidean Cost}
\label{sec-beyond-quadratic-euclidean-cost}
\paragraph{\texorpdfstring{$W_p$ costs.}{Wp costs}.}
\(\nabla f(x)=\nabla_x c(x,\T(x)).\)
\(\T(x)=x-\norm{\nabla f(x)}^{q-2}\nabla f(x), \qquad \frac1p+\frac1q=1.\)
\paragraph{Squared geodesic distance.}
\(\T(x)=\exp_x(-\nabla\phi(x)),\)
\paragraph{Twist condition.}
\begin{defn}[Twist condition]\label{def-twist-condition}
Let $X,Y\subset\RR^d$ and let $c\in C^1(X\times Y)$. The cost $c$ satisfies the twist condition from $X$ to $Y$ if, for every fixed $x\in X$, the map
\(y\in Y \longmapsto \nabla_x c(x,y)\in\RR^d\)
is injective. The dual twist condition is the analogous injectivity of $x\mapsto\nabla_y c(x,y)$ for every fixed $y$.
\end{defn}
\begin{prop}[Twist prevents splitting]\label{prop-twist-prevents-splitting}
Assume that $c\in C^1(X\times Y)$ satisfies the twist condition from $X$ to $Y$. Suppose that optimal pairs for the cost $c$ admit a source potential $f$ with the following first-order certificate: $f$ is differentiable on a full $\al$-measure subset of $X$, and whenever $x$ belongs to this differentiability set and $y$ can be paired optimally with $x$, one has $\nabla f(x)=\nabla_xc(x,y)$. Then, for $\al$-almost every source point $x$, there is at most one target point $y$ that can be paired optimally with $x$.
\end{prop}
\begin{proof}
Let $(f,g)$ be optimal Kantorovich potentials and let $\pi$ be an optimal plan. On $\supp(\pi)$ one has $f(x)+g(y)=c(x,y)$, while dual feasibility gives $f(z)+g(y)\leq c(z,y)$ for all $z$. Thus, for each contact pair $(x,y)$, the function $z\mapsto c(z,y)-g(y)$ touches $f$ from above at $x$. At a point where $f$ is differentiable, this gives
\(\nabla f(x)=\nabla_x c(x,y).\)
If the cost is twisted, this equation determines at most one $y$. Hence no optimal plan can split the mass of such an $x$ between several target points. Since differentiability holds on a full $\al$-measure set, the plan is concentrated on the graph of a measurable map there.
\end{proof}
The quadratic cost satisfies twist since $\nabla_x\norm{x-y}^2=2(x-y)$; the bilinear cost $c(x,y)=-\dotp{x}{y}$ satisfies twist since $\nabla_x c(x,y)=-y$; more generally $c(x,y)=h(x-y)$ is twisted when $h$ is smooth strictly convex and $\nabla h$ is injective. On a Riemannian manifold, the squared geodesic cost is twisted locally away from the cut locus.

\paragraph{Ma--Trudinger--Wang curvature.}
\begin{defn}[Weak MTW condition]\label{def-mtw-condition}
Assume that $c\in C^4(X\times Y)$, that $X,Y\subset\RR^d$, that $c$ is twisted, and that the mixed Hessian $\nabla^2_{xy}c(x,y)$ is invertible. For $(x,p)$ in the image of the change of variables $(x,y)\mapsto(x,\nabla_xc(x,y))$, let $Y(x,p)$ be defined by
\(\nabla_x c(x,Y(x,p))=p,\)
and set
\(A_{ij}(x,p)\eqdef \partial^2_{x_i x_j}c(x,Y(x,p)).\)
The cost satisfies the weak MTW condition if
\(\sum_{i,j,k,l}\partial^2_{p_kp_l} A_{ij}(x,p)\,\xi_i\xi_j\eta_k\eta_l\geq0 \qquad\text{whenever }\dotp{\xi}{\eta}=0.\)
A strong MTW condition asks for a positive lower bound, proportional to $\norm{\xi}^2\norm{\eta}^2$, on the same orthogonal directions.
\end{defn}
\begin{prop}[MTW condition and regularity]\label{prop-mtw-regularity-implication}
Assume that $c$ is smooth, twisted and non-degenerate, that the source and target domains satisfy the usual $c$-convexity assumptions, and that the source and target densities are positive and bounded above and below. Under the weak MTW condition, optimal maps are continuous for smooth positive data. Under the corresponding strong MTW condition and higher smoothness of the data and domains, one obtains higher regularity estimates for the optimal map.
\end{prop}
\begin{proof}
This is the regularity theorem of Ma--Trudinger--Wang and Loeper, stated here in its structural form. The $c$-convex potential solves a generated Jacobian equation. Twist and non-degeneracy turn the potential into a single-valued map, while the weak MTW inequality controls the convexity of contact sets of $c$-convex functions. This prevents the contact set from tearing apart and yields continuity. Strong MTW gives a quantitative curvature lower bound; combined with smooth densities, boundary geometry and elliptic estimates for the generated Jacobian equation, it yields the usual higher regularity. Loeper also proved that nonnegative MTW curvature is essentially necessary for continuity for arbitrary smooth positive data.
\end{proof}
\subsection{One-Dimensional Transport and Quantiles}
\label{sec-1d-transport-quantiles}
\begin{defn}[Cumulative and quantile functions]\label{def-cdf-quantile}
For $\al\in\Mm_+^1(\RR)$, its cumulative distribution function is
\eql{\label{eq-cumul-defn}
\cumul{\al}(x)\eqdef\al((-\infty,x]).
}
Its generalized inverse, or quantile function, is
\eql{\label{eq-OT-map-1d}
\cumul{\al}^{-1}(r)
\eqdef
\inf\enscond{x\in\RR}{\cumul{\al}(x)\geq r},
\qquad r\in(0,1).
}
\end{defn}
\begin{prop}[Quantile push-forward]\label{prop-quantile-pushforward}
One has $(\cumul{\al}^{-1})_\sharp\mathrm{Leb}_{[0,1]}=\al$. If $\al$ has no atoms, then $(\cumul{\al})_\sharp\al=\mathrm{Leb}_{[0,1]}$.
\end{prop}
\begin{proof}
For simplicity, assume first that $\al$ has a strictly positive density, so that $\cumul{\al}$ is strictly increasing and continuous. Denote $\ga\eqdef(\cumul{\al}^{-1})_\sharp\mathrm{Leb}_{[0,1]}$. It is enough to prove that $\cumul{\ga}=\cumul{\al}$. For every $x$,
\[
\begin{aligned}
\cumul{\ga}(x)
&=
\int_0^1 \mathbf 1_{(-\infty,x]}(\cumul{\al}^{-1}(z))\d z
=
\int_0^1 \mathbf 1_{[0,\cumul{\al}(x)]}(z)\d z
=
\cumul{\al}(x),
\end{aligned}
\]
where we used $\cumul{\al}^{-1}(z)\leq x$ if and only if $z\leq\cumul{\al}(x)$. General measures follow from the same argument with generalized inverses and right-continuity of the cumulative distribution function. If $\al$ has no atoms, the probability integral transform gives $(\cumul{\al})_\sharp\al=\mathrm{Leb}_{[0,1]}$.
\end{proof}
If $\al$ has no atoms, the map

\eql{\label{eq-1d-monge-map}
\T=\cumul{\be}^{-1}\circ\cumul{\al}
}
satisfies $\T_\sharp\al=\be$.

\begin{prop}[Monotone rearrangement on the line]\label{prop-1d-quantile-map}
Let $\al,\be\in\Mm_+^1(\RR)$ have finite $p$-th moments, with $p\geq1$. The coupling
\(\pi^\star=(\cumul{\al}^{-1},\cumul{\be}^{-1})_\sharp\mathrm{Leb}_{[0,1]}\)
minimizes $\int |x-y|^p\d\pi(x,y)$ among couplings. If $\al$ has no atoms, it is induced by the monotone Monge map~\eqref{eq-1d-monge-map}.
\end{prop}
\begin{proof}
The displayed measure is a coupling by Proposition~\ref{prop-quantile-pushforward}. Its support is monotone: for $s<t$, both quantile functions satisfy $\cumul{\al}^{-1}(s)\leq\cumul{\al}^{-1}(t)$ and $\cumul{\be}^{-1}(s)\leq\cumul{\be}^{-1}(t)$. If a coupling had two crossed pairs $x<x'$ and $y>y'$ with positive mass, exchanging the targets decreases the cost for strictly convex powers and does not increase it for $p=1$, by the two-point inequality used in Proposition~\ref{prop-matching-1d-monotone}. Eliminating crossings yields a monotone optimal coupling, and the monotone coupling with prescribed marginals is exactly the quantile coupling. If $\al$ has no atoms, $(\cumul{\al})_\sharp\al=\mathrm{Leb}_{[0,1]}$, so the coupling is generated by~\eqref{eq-1d-monge-map}.
\end{proof}
For discrete measures, one cannot directly apply the map formula when the source has atoms, but if the measures are uniform on the same number of Dirac masses, then it is exactly the sorting formula of Proposition~\ref{prop-matching-1d-monotone}.

\begin{prop}[One-dimensional Wasserstein formulas]\label{prop-wass-quantile-1d}
Let $\al,\be\in\Mm_+^1(\RR)$ have finite $p$-th moments. For every $p\geq1$,
\eql{\label{eq-wass-cumul}
\Wass_p(\al,\be)^p
=
\int_0^1\abs{\cumul{\al}^{-1}(r)-\cumul{\be}^{-1}(r)}^p\d r
=
\norm{\cumul{\al}^{-1}-\cumul{\be}^{-1}}_{L^p([0,1])}^p.
}
For $p=1$, this is equivalently
\begin{align}\label{eq-w1-1d}
\Wass_1(\al,\be)
&=
\norm{\cumul{\al}-\cumul{\be}}_{L^1(\RR)}
=
\int_\RR \abs{\cumul{\al}(x)-\cumul{\be}(x)}\d x \\
&=
\int_\RR \abs{\int_{-\infty}^x\d(\al-\be)}\d x.
\end{align}
\end{prop}
\begin{proof}
The first formula follows from Proposition~\ref{prop-1d-quantile-map}: the optimal coupling is obtained by taking the same quantile level $r$ for both measures. For $p=1$, use the layer-cake identity. If $q_\al$ and $q_\be$ are the two quantile functions, then
\(\int_0^1 |q_\al(r)-q_\be(r)|\d r = \int_\RR \lambda\bigl(\{r:q_\al(r)\leq x<q_\be(r)\}\cup\{r:q_\be(r)\leq x<q_\al(r)\}\bigr)\d x.\)
The measure of the displayed set is exactly $|\cumul{\al}(x)-\cumul{\be}(x)|$ for almost every $x$.
\end{proof}
Formula~\eqref{eq-wass-cumul} means that the map $\al\mapsto\cumul{\al}^{-1}$ embeds one-dimensional Wasserstein geometry isometrically into a linear $L^p$ space. For $p=2$, the Wasserstein distance on probability measures over the real line is therefore Hilbertian.

\(\cumul{\al_t}^{-1}(r) = (1-t)\cumul{\al}^{-1}(r)+t\cumul{\be}^{-1}(r), \qquad r\in(0,1).\)
\paragraph{Triangular rearrangements.}
\begin{prop}[Knothe--Rosenblatt triangular rearrangement]\label{prop-knothe-rosenblatt}
Let $\al,\be\in\Mm_+^1(\RR^d)$. Assume, for simplicity, that the first marginal of $\al$ and the one-dimensional conditional laws of $\al$ used below are atomless, and that regular conditional distributions are fixed. There is a triangular map
\(\T(x_1,\ldots,x_d) = (\T_1(x_1),\T_2(x_1,x_2),\ldots,\T_d(x_1,\ldots,x_d))\)
such that $\T_\sharp\al=\be$ and, for each $k$, the function $x_k\mapsto\T_k(x_1,\ldots,x_k)$ is nondecreasing for $\al$-almost every value of $(x_1,\ldots,x_{k-1})$.
\end{prop}
\begin{proof}
The construction is recursive. For $k=1$, let $\T_1$ be the monotone rearrangement between the first marginal of $\al$ and the first marginal of $\be$. Suppose that $\T_1,\ldots,\T_{k-1}$ have already been constructed. Write $x_{<k}=(x_1,\ldots,x_{k-1})$ and $\T_{<k}=(\T_1,\ldots,\T_{k-1})$. By induction, $(\T_{<k})_\sharp\al_{<k}=\be_{<k}$, where $\al_{<k}$ and $\be_{<k}$ are the first $(k-1)$-coordinate marginals. Let $\al^k_{x_{<k}}$ and $\be^k_{y_{<k}}$ be regular conditional laws of the $k$-th coordinate given the previous coordinates. Define $\T_k(x_{<k},\cdot)$ as the one-dimensional monotone rearrangement from $\al^k_{x_{<k}}$ to $\be^k_{\T_{<k}(x_{<k})}$.

The map $\T_k(x_{<k},\cdot)$ is nondecreasing by the one-dimensional rearrangement theorem. The chain rule for disintegrations then shows that, after step $k$, the first $k$ coordinates of $\T_\sharp\al$ have the same law as the first $k$ coordinates of $\be$. At $k=d$ this gives $\T_\sharp\al=\be$. Target atoms are handled by generalized quantiles. If a source conditional law has atoms that must be split, the same recursive construction defines a triangular Markov kernel rather than a deterministic map.
\end{proof}
\begin{prop}[Anisotropic Brenier maps converge to Knothe--Rosenblatt]\label{prop-knothe-limit-anisotropic-brenier}
Let $\alpha,\beta$ be compactly supported probability measures on $\RR^d$ with densities bounded above and below on their rectangular supports, and assume that the conditional laws entering Proposition~\ref{prop-knothe-rosenblatt} are atomless. For $\epsilon>0$, set
\(c_\epsilon(x,y) \eqdef \sum_{k=1}^d \epsilon^{k-1}\abs{x_k-y_k}^2.\)
Let $T_\epsilon$ be the Monge map from $\alpha$ to $\beta$ for the cost $c_\epsilon$, and let $T_{\mathrm{KR}}$ be the triangular Knothe--Rosenblatt rearrangement with the coordinate order used above. Then
\(T_\epsilon \longrightarrow T_{\mathrm{KR}} \qquad\text{in } L^2(\alpha;\RR^d) \qquad\text{as } \epsilon\to0.\)
\end{prop}
\begin{proof}
Let $\pi_\epsilon=(\Id,T_\epsilon)_\sharp\alpha$. Since the supports are compact, a subsequence converges weakly to some coupling $\pi$ between $\alpha$ and $\beta$. The optimality of $\pi_\epsilon$ for
\(F_\epsilon(\gamma) = \sum_{k=1}^d \epsilon^{k-1} \int\abs{x_k-y_k}^2\d\gamma(x,y)\)
first implies, by letting $\epsilon\to0$, that $\pi$ minimizes the one-dimensional quadratic cost in the first coordinate among all couplings. Since the first marginal of $\alpha$ is atomless, the one-dimensional minimizer is the monotone rearrangement, so $y_1=T_1(x_1)$ under $\pi$.

Now restrict attention to couplings satisfying this first-coordinate constraint. Subtract the common minimal value of the first-coordinate cost, divide the optimality inequality by $\epsilon$, and let $\epsilon\to0$. The limit coupling must minimize the second-coordinate quadratic cost among all couplings that already realize the first monotone rearrangement. Disintegrating with respect to $(x_1,y_1)$ reduces this constrained problem to the one-dimensional monotone rearrangement between the conditional laws of $x_2$ and $y_2$. Hence $y_2=T_2(x_1,x_2)$ under $\pi$.

Repeating the same subtraction-and-rescaling argument gives, for every $k$, the conditional monotone rearrangement $y_k=T_k(x_1,\ldots,x_k)$. Thus every weak limit of $(\pi_\epsilon)_\epsilon$ is concentrated on the graph of $T_{\mathrm{KR}}$. This graph coupling is unique, so the whole family $\pi_\epsilon$ converges weakly to $(\Id,T_{\mathrm{KR}})_\sharp\alpha$.

Finally, let $X\sim\alpha$. The graph couplings are the laws of $(X,T_\epsilon(X))$, and they converge weakly to the law of $(X,T_{\mathrm{KR}}(X))$. To turn this into convergence of maps, fix $\zeta>0$. By Lusin's theorem, there is a compact set $K$ with $\alpha(K)>1-\zeta$ on which $T_{\mathrm{KR}}$ is continuous. On $K$, the set
\(\enscond{(x,y)}{x\in K,\ \norm{y-T_{\mathrm{KR}}(x)}\geq\delta}\)
is closed and has zero mass under the limiting graph coupling. Portmanteau's theorem gives
\[
\limsup_{\epsilon\to0}
\alpha\!\left(\enscond{x}{x\in K,\ \norm{T_\epsilon(x)-T_{\mathrm{KR}}(x)}\geq\delta}\right)
=0.
\]
Adding the complement of $K$ and letting $\zeta\to0$ proves convergence in $\alpha$-probability. Since all maps take values in a common compact set, this convergence is uniformly integrable and hence holds in $L^2(\alpha)$.
\end{proof}
\subsection{Gaussian Measures and the Bures Metric}
\label{sec-gaussian-bures}
\paragraph{One-dimensional Gaussians.}
Let $\al=\Gaussian(m_\al,\sigma_\al^2)$ and $\be=\Gaussian(m_\be,\sigma_\be^2)$ be two nondegenerate Gaussians on $\RR$. Then

\(\T(x)=\frac{\sigma_\be}{\sigma_\al}(x-m_\al)+m_\be\)
satisfies $\T_\sharp\al=\be$. It is the derivative of the convex function

\(\phi(x)=\frac{\sigma_\be}{2\sigma_\al}(x-m_\al)^2+m_\be x,\)
\[
\Wass_2(\al,\be)^2
=
\int_\RR\left(\frac{\sigma_\be}{\sigma_\al}(x-m_\al)+m_\be-x\right)^2\d\al(x)
=
(m_\al-m_\be)^2+(\sigma_\al-\sigma_\be)^2.
\]
The formula extends by continuity to Dirac masses, although the affine Monge map itself only pushes a Dirac source to another Dirac. Thus the OT geometry of one-dimensional Gaussians is the Euclidean geometry of the half-plane $(m,\sigma)\in\RR\times\RR_+$.

\paragraph{Multivariate Gaussians.}
\eql{\label{eq-gauss-pf}
\al=\Gaussian(\mean_\al,\cov_\al),
\qquad
\be=\Gaussian(\mean_\be,\cov_\be),
\qquad
\T(x)=\mean_\be+A(x-\mean_\al),
}
\begin{prop}[Affine push-forward of Gaussians]\label{prop-gaussian-affine-push-forward}
One has $\T_\sharp\al=\be$ if and only if
\eql{\label{eq-gauss-covariance-pushforward}
A\cov_\al A^\top=\cov_\be.
}
\end{prop}
\begin{proof}
An affine function maps a Gaussian to a Gaussian, so the law of $\T(X)$ is determined by its mean and covariance. If $X\sim\al$ and $Y=\T(X)$, then
\(\EE(Y)=\mean_\be+A\EE(X-\mean_\al)=\mean_\be,\)
and
\(\EE((Y-\mean_\be)(Y-\mean_\be)^\top) = A\EE((X-\mean_\al)(X-\mean_\al)^\top)A^\top = A\cov_\al A^\top.\)
Thus $A\cov_\al A^\top=\cov_\be$ is necessary and sufficient for $Y$ to have the same mean and covariance as $\be$.
\end{proof}
\begin{defn}[Bures metric]\label{def-bures-metric}
For positive semidefinite covariance matrices $\Sigma$ and $\Lambda$, the Bures metric is
\eql{\label{eq-bure-defn}
\Bb(\Sigma,\Lambda)^2
\eqdef
\tr\pa{\Sigma+\Lambda-2(\Sigma^{1/2}\Lambda\Sigma^{1/2})^{1/2}}.
}
\end{defn}
\begin{prop}[Gaussian $\Wass_2$ formula and Bures covariance term]\label{prop-gaussian-w2-bures}
Assume that $\cov_\al$ and $\cov_\be$ are positive definite. The unique symmetric positive-definite solution of
\(A\cov_\al A=\cov_\be\)
is
\eql{\label{eq-bures-map}
A=
\cov_\al^{-1/2}
\pa{\cov_\al^{1/2}\cov_\be\cov_\al^{1/2}}^{1/2}
\cov_\al^{-1/2}.
}
The affine map $\T(x)=\mean_\be+A(x-\mean_\al)$ is the optimal quadratic-cost transport from $\Gaussian(\mean_\al,\cov_\al)$ to $\Gaussian(\mean_\be,\cov_\be)$, and
\eql{\label{eq-dist-gauss}
\Wass_2(\al,\be)^2
=
\norm{\mean_\al-\mean_\be}^2+\Bb(\cov_\al,\cov_\be)^2,
}
where $\Bb$ is the Bures metric of Definition~\ref{def-bures-metric}.
\end{prop}
\begin{proof}
Multiplying $A\cov_\al A=\cov_\be$ on the left and right by $\cov_\al^{1/2}$ gives \((\cov_\al^{1/2}A\cov_\al^{1/2})^2 = \cov_\al^{1/2}\cov_\be\cov_\al^{1/2}.\)
Since $A$ is symmetric positive, $\cov_\al^{1/2}A\cov_\al^{1/2}$ is symmetric positive and is therefore the unique positive square root of the right-hand side. Conversely, the matrix in~\eqref{eq-bures-map} is symmetric positive and satisfies the covariance equation.

By Proposition~\ref{prop-gaussian-affine-push-forward}, this affine map pushes $\al$ to $\be$. It is the gradient of a convex quadratic potential, so Brenier's theorem implies optimality. If $X\sim\al$, then
\begin{align*}
\EE\norm{X-\T(X)}^2
&=
\norm{\mean_\al-\mean_\be}^2
+
\EE\norm{(I-A)(X-\mean_\al)}^2 \\
&=
\norm{\mean_\al-\mean_\be}^2
+
\tr((I-A)\cov_\al(I-A)^\top) \\
&=
\norm{\mean_\al-\mean_\be}^2
+
\tr(\cov_\al)+\tr(A\cov_\al A)-2\tr(A\cov_\al) \\
&=
\norm{\mean_\al-\mean_\be}^2
+
\tr(\cov_\al)+\tr(\cov_\be)
-2\tr((\cov_\al^{1/2}\cov_\be\cov_\al^{1/2})^{1/2}),
\end{align*}
which is the desired expression. The formula for singular covariance matrices follows by adding $\eta I$ and letting $\eta\downarrow0$.
\end{proof}
The same formula can be read without first restricting to Gaussian measures. Let $\Phi_2$ keep only raw second moments.

\begin{prop}[Bures metric as a second-moment quotient]\label{prop-bures-second-moment-lift}
For $\mu\in\Pp_2(\RR^d)$, set its raw second-moment matrix to be
\(\Phi_2(\mu)\eqdef \int_{\RR^d} uu^\top \d\mu(u)\in\mathbb S_+^d.\)
For $\Sigma,\Lambda\in\mathbb S_+^d$, define
\(D_{\Phi_2}(\Sigma,\Lambda)^2 \eqdef \inf_{\Phi_2(\mu)=\Sigma,\ \Phi_2(\nu)=\Lambda} \Wass_2(\mu,\nu)^2.\)
Then
\(D_{\Phi_2}(\Sigma,\Lambda)=\Bb(\Sigma,\Lambda).\)
In other words, the Bures metric is the Wasserstein quotient distance induced on positive semidefinite matrices by identifying all probability measures with the same raw second-moment matrix.
\end{prop}
\begin{proof}
Assume first that $\Sigma$ and $\Lambda$ are positive definite. Fix two admissible laws $\mu,\nu$ and an arbitrary coupling $\pi$ between them, and write
\(K\eqdef \int_{\RR^d\times\RR^d}uv^\top \d\pi(u,v).\)
Then
\(\int\norm{u-v}^2\d\pi(u,v) = \tr(\Sigma)+\tr(\Lambda)-2\tr(K).\)
The block second-moment matrix
\[
\int
\begin{pmatrix}u\\ v\end{pmatrix}
\begin{pmatrix}u\\ v\end{pmatrix}^{\!\top}
\d\pi(u,v)
=
\begin{pmatrix}
\Sigma & K\\
K^\top & \Lambda
\end{pmatrix}
\]
is positive semidefinite. The Schur complement gives \(R\eqdef\Sigma^{-1/2}K\Lambda^{-1/2}, \qquad RR^\top\preceq I,\)
so $K=\Sigma^{1/2}R\Lambda^{1/2}$ with $\norm{R}_{\mathrm{op}}\leq1$. By duality between the nuclear and operator norms,
\[
\tr(K)
\leq
\sup_{\norm{R}_{\mathrm{op}}\leq1}\tr(\Sigma^{1/2}R\Lambda^{1/2})
=
\norm{\Lambda^{1/2}\Sigma^{1/2}}_*
=
\tr\pa{(\Sigma^{1/2}\Lambda\Sigma^{1/2})^{1/2}}.
\]
For singular matrices, fix $\eta>0$. Adding $\eta I$ to the two diagonal blocks preserves block positivity, so the same Schur-complement argument gives
\(\tr(K) \leq \tr\pa{((\Sigma+\eta I)^{1/2}(\Lambda+\eta I)(\Sigma+\eta I)^{1/2})^{1/2}}.\)
Letting $\eta\downarrow0$ and using continuity of the matrix square root gives the same trace bound with $\Sigma,\Lambda$. Hence every admissible coupling has cost at least $\Bb(\Sigma,\Lambda)^2$.

Conversely, choose the centered Gaussian laws $\Gaussian(0,\Sigma)$ and $\Gaussian(0,\Lambda)$. They satisfy the two second-moment constraints, and Proposition~\ref{prop-gaussian-w2-bures} gives
\(\Wass_2(\Gaussian(0,\Sigma),\Gaussian(0,\Lambda))^2 = \Bb(\Sigma,\Lambda)^2,\)
again by continuity in the singular case. This admissible pair attains the lower bound, proving the claim.
\end{proof}
\[
\Sigma=\begin{pmatrix} a & c \\ c & b \end{pmatrix},
\qquad
t=\frac{a+b}{\sqrt2},\qquad
u=\frac{a-b}{\sqrt2},\qquad
v=\sqrt2\,c,
\]
\(a=\frac{t+u}{\sqrt2},\qquad b=\frac{t-u}{\sqrt2},\qquad c=\frac{v}{\sqrt2}.\)
\(t^2-u^2-v^2=2(ab-c^2)=2\det(\Sigma).\)
\(t\geq \sqrt{u^2+v^2}.\)
Thus the cone of $2\times2$ covariance matrices is the Lorentz, or ice-cream, cone. The Bures distance is not the ambient Euclidean distance in these coordinates: on the positive definite interior it is the geodesic distance of a smooth nonlinear geometry, whose geodesics are the covariance parts of Gaussian optimal transport rather than the straight chords inherited from $\RR^3$.

\begin{prop}[Metric and convexity properties of the Bures term]\label{prop-bures-metric-convex}
The function $\Bb$ is a distance on positive semidefinite covariance matrices. Moreover, $\Bb^2$ is jointly convex: for $t\in[0,1]$,
\(\Bb^2((1-t)\Sigma_0+t\Sigma_1,(1-t)\Lambda_0+t\Lambda_1) \leq (1-t)\Bb^2(\Sigma_0,\Lambda_0)+t\Bb^2(\Sigma_1,\Lambda_1).\)
\end{prop}
\begin{proof}
The key identity is the Procrustes representation
\(\Bb^2(\Sigma,\Lambda) = \min_{Q^\top Q=I}\norm{\Sigma^{1/2}-\Lambda^{1/2}Q}_F^2.\)
Indeed, expanding the square gives $\tr\Sigma+\tr\Lambda-2\max_{Q^\top Q=I}\tr(\Sigma^{1/2}Q^\top\Lambda^{1/2})$, and the orthogonal Procrustes formula identifies the maximum with $\tr((\Sigma^{1/2}\Lambda\Sigma^{1/2})^{1/2})$. Symmetry, positivity and separation follow immediately from this representation. The triangle inequality follows by choosing two almost optimal orthogonal matrices $Q_1,Q_2$ and writing
\(\norm{\Sigma^{1/2}-\Gamma^{1/2}Q_2Q_1}_F \leq \norm{\Sigma^{1/2}-\Lambda^{1/2}Q_1}_F + \norm{\Lambda^{1/2}-\Gamma^{1/2}Q_2}_F.\)
Letting the two choices become optimal proves the metric property.

For convexity, use the equivalent factor formulation
\(\Bb^2(\Sigma,\Lambda) = \min_{UU^\top=\Sigma,\thinspace VV^\top=\Lambda}\norm{U-V}_F^2.\)
Choose nearly optimal factors $(U_0,V_0)$ and $(U_1,V_1)$ for the two pairs, and define block factors
\(U_t=[\sqrt{1-t}\,U_0,\sqrt t\,U_1], \qquad V_t=[\sqrt{1-t}\,V_0,\sqrt t\,V_1].\)
Then $U_tU_t^\top=(1-t)\Sigma_0+t\Sigma_1$ and $V_tV_t^\top=(1-t)\Lambda_0+t\Lambda_1$, while
\(\norm{U_t-V_t}_F^2 = (1-t)\norm{U_0-V_0}_F^2+t\norm{U_1-V_1}_F^2.\)
Taking the infimum over the initial factors proves joint convexity.
\end{proof}