\section{Generative Models via Transportation}
\label{sec-generative-models-transportation}
\subsection{Generative Models via Flow Matching}
\paragraph{Stochastic interpolant.}
The word ``stochastic'' can hide two different levels of randomness. We first use the simpler one: after drawing a latent variable $U\sim\pi$, the path $t\mapsto P_t(U)$ is deterministic and differentiable.

We assume first that $\alpha_t$ is obtained by pushing a latent distribution $\pi \in \Pp(\RR^{d'})$ through a time-dependent map $P_t: \RR^{d'} \to \RR^d$; the latent dimension $d'$ may be larger than the data dimension $d$, i.e.

\begin{equation}\label{eq:interp-coupling}
\forall t \in [0,1], \quad \alpha_t:= (P_t)_\sharp \pi.
\end{equation}
If $\pi = \alpha \otimes \beta$ and $\alpha = \frac{1}{n} \sum_i \delta_{x_i}$, $\beta = \frac{1}{m} \sum_j \delta_{y_j}$, then $\alpha_t$ consists of $n \times m$ Dirac masses

\(\alpha_t = \frac{1}{nm} \sum_{i,j} \delta_{P_t(x_i,y_j)}.\)
If $\pi = (\Id, T)_\sharp \alpha$ is a Brenier-type coupling, then $\alpha_t = ((1-t)\Id + tT)_\sharp \alpha$ is the so-called McCann OT interpolation.

\paragraph{Flow matching formula.}
\begin{prop}[Flow matching vector field]\label{prop-flow-matching-vector-field}
For each fixed $t$, assume $\partial_tP_t\in L^2(\pi;\RR^d)$. Consider the flow-matching problem over measurable fields $v_t:\RR^d\to\RR^d$
\begin{equation}
\min_{v_t} \int_{\RR^{d'}} \norm{v_t(P_t(u)) - [\partial_t P_t](u)}^2 \, \d\pi(u). \label{eq:flow-matching}
\end{equation}
Its minimizer is characterized $\alpha_t$-almost everywhere by the conditional expectation
\begin{equation}\label{eq:flow-match-conditional}
v_t(z) = \EE_{u \sim \pi} \big( [\partial_t P_t](u) \, \big| \, z = P_t(u) \big).
\end{equation}
Then the pair $(\alpha_t,v_t)$ satisfies the continuity equation~\eqref{eq:eulerian-advection}.
\end{prop}
\begin{proof}
We first recall the two equivalent ways of writing the interpolated measure. Formally, one may write
\(\alpha_t(z)=\int_{\RR^{d'}}\delta(z-P_t(u))\,\d\pi(u),\)
while the rigorous meaning is that, for every smooth test function $\varphi$,
\begin{equation}\label{eq:flow-matching-pushforward-test}
\int_{\RR^d}\varphi(z)\,\d\alpha_t(z)
=
\int_{\RR^{d'}}\varphi(P_t(u))\,\d\pi(u).
\end{equation}
The minimizer in~\eqref{eq:flow-matching} is the orthogonal projection in $L^2(\pi;\RR^d)$ of the latent velocity $\partial_tP_t(u)$ onto the closed subspace of functions that depend on $u$ only through $P_t(u)$. This projection is the conditional expectation~\eqref{eq:flow-match-conditional}. Formally, this can be read as
\(v_t(z)=\frac{1}{\alpha_t(z)} \int_{\RR^{d'}}\delta(z-P_t(u))[\partial_tP_t](u)\,\d\pi(u),\)
and rigorously it means that, for every smooth test vector field $m$,
\begin{equation}\label{eq:v_t}
\int \dotp{m(z)}{v_t(z)} \, \d\alpha_t(z)
=
\int \dotp{m(P_t(u))}{[\partial_t P_t](u)} \, \d\pi(u).
\end{equation}

The weak form of
$\partial_t\alpha_t+\diverg(\alpha_t v_t)=0$ is that, for every smooth scalar test function $\varphi$,
\begin{equation}\label{eq:flow-matching-weak-target}
\frac{\d}{\d t}\int\varphi(z)\,\d\alpha_t(z)
-
\int\dotp{v_t(z)}{\nabla\varphi(z)}\,\d\alpha_t(z)
=0.
\end{equation}
Using~\eqref{eq:flow-matching-pushforward-test} and differentiating under the integral sign gives
\begin{equation}\label{eq:flow-matching-test-derivative}
\frac{\d}{\d t}\int \varphi(z)\d\alpha_t(z)
=
\int \dotp{\nabla\varphi(P_t(u))}{[\partial_t P_t](u)}\d\pi(u).
\end{equation}
On the other hand, applying~\eqref{eq:v_t} with $m=\nabla\varphi$ gives
\begin{equation}\label{eq:flow-matching-velocity-test}
\int\dotp{v_t(z)}{\nabla\varphi(z)}\,\d\alpha_t(z)
=
\int \dotp{\nabla\varphi(P_t(u))}{[\partial_t P_t](u)}\d\pi(u).
\end{equation}
Comparing~\eqref{eq:flow-matching-test-derivative} and~\eqref{eq:flow-matching-velocity-test} yields~\eqref{eq:flow-matching-weak-target}, which is the desired continuity equation.
\end{proof}
The conditional expectation in~\eqref{eq:flow-match-conditional} has a simple measure-theoretic meaning. Let $\alpha_t=(P_t)_\sharp\pi$ and define the vector-valued measure $m_t$ on $\RR^d$ by

\(\int_{\RR^d}\dotp{\psi(z)}{\d m_t(z)} \eqdef \int_{\RR^{d'}}\dotp{\psi(P_t(u))}{[\partial_tP_t](u)}\d\pi(u)\)
\(\d m_t(z)=v_t(z)\d\alpha_t(z), \qquad v_t=\frac{\d m_t}{\d\alpha_t}.\)
In the language of Lebesgue decomposition, the flux measure $m_t$ has only an absolutely continuous part with respect to $\alpha_t$ and no singular part; the conditional expectation is precisely this density. Equivalently, disintegrating $\pi$ with respect to the map $P_t$ gives $\pi(\d u)=\pi_{t,z}(\d u)\alpha_t(\d z)$, where $\pi_{t,z}$ is supported on the fiber $\{u:\,P_t(u)=z\}$, and

\(v_t(z)=\int_{\{P_t(u)=z\}}[\partial_tP_t](u)\d\pi_{t,z}(u).\)
Thus the solution of \eqref{eq:flow-matching} is the conditional expectation of the velocities $\partial_t P_t$: intuitively, $v_t(z)$ is the average velocity of all trajectories passing through $z$.

For the exact field $v_t$, integrating the ODE $\dot{x}=v_t(x)$ defines a transport map $T_t$. If $v_t$ is regular enough, or more generally if the continuity equation has a unique solution for this velocity, then $(T_t)_\sharp\alpha_0=\alpha_t$.

\paragraph{Connection with diffusion models.}
In the special case where $P_t(x,y)=(1-t)x+ty$ is a linear interpolation and $\pi = \alpha \otimes \beta$, the curve $\alpha_t$ is a convolution of rescaled versions of $\alpha_0$ and $\alpha_1$. The flow-matching problem~\eqref{eq:flow-matching} becomes

\(\min_{(v_t)_t} \int_{\RR^{d} \times \RR^d} \norm{v_t( (1-t)x+t y ) - (y-x) }^2 \, \d\alpha_0(x) \d\alpha_1(y).\)
\begin{proposition}[Tweedie identity]\label{prop:Tweedie}
Let $W$ be a random vector in $\RR^{d}$ with density $\beta$.
For $\sigma>0$, observe
\(Z \;=\; W + \sigma\,\varepsilon, \quad\text{where } \varepsilon \sim \Gaussian(0,I_{d}) \text{ is independent of } W.\)
Denote by \(\beta_\sigma \;=\; \beta * \Gaussian\bigl(0,\sigma^{2}I_{d}\bigr)\)
the density of $Z$.
Then
\(\EE\bigl[\,W \mid Z=z\bigr] \;=\; z \;+\;\sigma^{2}\,\nabla \log \beta_\sigma(z) \qquad\text{for all } z \in \RR^{d}.\)
\end{proposition}
\begin{proof}
Bayes' rule gives the conditional density
$
p_{W|Z}(w\mid z)
= \dfrac{\beta(w)\,\varphi_\sigma(z-w)}{\beta_\sigma(z)}
$
with $\varphi_\sigma$ the $\Gaussian(0,\sigma^{2}I_{d})$ density.
Hence
\(\EE[W\mid Z=z] = \frac{1}{\beta_\sigma(z)} \int_{\RR^{d}} w\, \beta(w)\,\varphi_\sigma(z-w)\,\d w.\)
Differentiating the Gaussian convolution under the integral sign and using
$
\nabla_z\varphi_\sigma(z-w)
= -\sigma^{-2}(z-w)\,\varphi_\sigma(z-w)
$
yields
\(\nabla_z\beta_\sigma(z) = \int \beta(w)\,\nabla_z\varphi_\sigma(z-w)\,\d w = -\sigma^{-2}\Bigl(z-\EE[W\mid Z=z]\Bigr)\,\beta_\sigma(z).\)
Rearranging finishes the proof.
\end{proof}
\begin{proposition}[Gaussian-endpoint flow-matching field]\label{prop:flow}
Let $X\sim\alpha$ and $Y\sim\Gaussian(0,I_{d})$ be independent.
For $t\in(0,1)$ set
\(Z_t \;=\; (1-t)\,X + t\,Y, \qquad \alpha_t =\text{Law}(Z_t).\)
The regression minimizer $v^\star:\RR^d\times(0,1)\to\RR^d$ of
\(\min_{v}\;\int_{0}^{1}\! \iint_{\RR^{d}\times\RR^{d}} \bigl|y-x-v\bigl((1-t)x+t y,t\bigr)\bigr|^{2}\, \d\alpha(x)\,\d\Gaussian(y)\,\d t\)
is \(v^\star(x,t) = -\frac{1}{1-t}\,x \;-\; \frac{t}{1-t}\,\nabla\log\alpha_t(x) \qquad (x\in\RR^{d},\;t\in(0,1)).\)
In particular, for each $t\in(0,1)$ this field is a gradient field,
\[
v^\star(\cdot,t)=-\nabla
\left(
\frac{\norm{\cdot}^2}{2(1-t)}
+\frac{t}{1-t}\log\alpha_t
\right).
\]
\end{proposition}
\begin{proof}
Fix $t\in(0,1)$ and write $W=(1-t)X$, $\sigma=t$, so that
$Z_t = W + \sigma\,Y$ matches the setting of Proposition~\ref{prop:Tweedie}.
Conditional expectations satisfy
$
v^\star(z,t)
= \EE[Y-X\mid Z_t=z]
= \frac{1}{t}\,\EE[Z_t-W\mid Z_t=z]
-\,\frac{1}{1-t}\,\EE[W\mid Z_t=z].
$
Applying Proposition~\ref{prop:Tweedie} to $\EE[W\mid Z_t=z]$ and
noting $\EE[Y\mid Z_t=z]
= -\,t\,\nabla\log\alpha_t(z)$
gives the claimed formula.
\end{proof}
\paragraph{When is the induced map optimal?}
Integrating the learned velocity gives a deterministic map from $\alpha_0$ to $\alpha_1$, but this map is not automatically the Brenier optimal map. It is optimal only in special cases where the accumulated flow remains the gradient of a convex potential.

\begin{prop}[Gaussian flow matching and optimality]\label{prop-gaussian-flow-matching-optimality}
Let $\Sigma_0,\Sigma_1\succ0$ and let $X_0\sim\Gaussian(0,\Sigma_0)$ and $X_1\sim\Gaussian(0,\Sigma_1)$ be independent. Consider the linear flow-matching interpolation
\(Z_t=(1-t)X_0+tX_1, \qquad \alpha_t=\operatorname{Law}(Z_t)=\Gaussian(0,\Sigma_t),\)
where
\begin{equation}\label{eq-gaussian-product-fm-covariance}
\Sigma_t=(1-t)^2\Sigma_0+t^2\Sigma_1.
\end{equation}
Then the exact flow-matching velocity is affine, $v_t(z)=A_tz$, with
\begin{equation}\label{eq-gaussian-product-fm-velocity}
A_t=\bigl(t\Sigma_1-(1-t)\Sigma_0\bigr)\Sigma_t^{-1}.
\end{equation}
The induced flow map $T_t^{\rm FM}$ from $\alpha_0$ to $\alpha_t$ is
\begin{equation}\label{eq-gaussian-product-fm-map}
T_t^{\rm FM}
=
\Sigma_0^{1/2}
\Bigl((1-t)^2\Id+t^2\Sigma_0^{-1/2}\Sigma_1\Sigma_0^{-1/2}\Bigr)^{1/2}
\Sigma_0^{-1/2}.
\end{equation}
In particular,
\begin{equation}\label{eq-gaussian-product-fm-terminal-map}
T_1^{\rm FM}
=
\Sigma_0^{1/2}
\bigl(\Sigma_0^{-1/2}\Sigma_1\Sigma_0^{-1/2}\bigr)^{1/2}
\Sigma_0^{-1/2}.
\end{equation}
This terminal map coincides with the quadratic optimal transport map
\begin{equation}\label{eq-gaussian-brenier-map-comparison}
T^{\rm OT}
=
\Sigma_0^{-1/2}
\bigl(\Sigma_0^{1/2}\Sigma_1\Sigma_0^{1/2}\bigr)^{1/2}
\Sigma_0^{-1/2}
\end{equation}
if and only if $\Sigma_0\Sigma_1=\Sigma_1\Sigma_0$.
\end{prop}
\begin{proof}
The conditional-expectation formula gives \(v_t(z)=\EE[X_1-X_0\mid Z_t=z].\)
Since all variables are jointly Gaussian, this conditional expectation is linear and
\(v_t(z) = \operatorname{Cov}(X_1-X_0,Z_t)\operatorname{Cov}(Z_t)^{-1}z = \bigl(t\Sigma_1-(1-t)\Sigma_0\bigr)\Sigma_t^{-1}z,\)
which proves~\eqref{eq-gaussian-product-fm-velocity}. To solve the characteristic equation, whiten the source by setting
\(C=\Sigma_0^{-1/2}\Sigma_1\Sigma_0^{-1/2}, \qquad \widetilde Z_t=\Sigma_0^{-1/2}Z_t.\)
In these coordinates the source covariance is $\Id$ and
\(\widetilde\Sigma_t=(1-t)^2\Id+t^2C.\)
Because $\Id$ and $C$ commute, the affine flow map in whitened coordinates is simply $\widetilde T_t=\widetilde\Sigma_t^{1/2}$. Indeed,
\(\frac{\d}{\d t}\widetilde\Sigma_t^{1/2} = \bigl(tC-(1-t)\Id\bigr)\widetilde\Sigma_t^{-1/2},\)
which is exactly the equation $\dot{\widetilde T}_t=\widetilde A_t\widetilde T_t$ with $\widetilde T_0=\Id$. Returning to the original coordinates gives~\eqref{eq-gaussian-product-fm-map}, and $t=1$ gives~\eqref{eq-gaussian-product-fm-terminal-map}.

Both $T_1^{\rm FM}$ and $T^{\rm OT}$ push $\Gaussian(0,\Sigma_0)$ to $\Gaussian(0,\Sigma_1)$. The Brenier map between nondegenerate Gaussians is the unique symmetric positive definite linear map with this property. Hence $T_1^{\rm FM}=T^{\rm OT}$ if and only if $T_1^{\rm FM}$ is symmetric positive definite. The map $T_1^{\rm FM}$ is similar to $C^{1/2}$, so if it is symmetric then it is automatically positive definite. Since $C^{1/2}$ is symmetric positive definite,
\((T_1^{\rm FM})^\top = \Sigma_0^{-1/2}C^{1/2}\Sigma_0^{1/2}.\)
Thus symmetry of $T_1^{\rm FM}$ is equivalent to
$\Sigma_0 C^{1/2}=C^{1/2}\Sigma_0$, hence to $\Sigma_0 C=C\Sigma_0$ by functional calculus. Multiplying this identity on the left and right by $\Sigma_0^{1/2}$ gives $\Sigma_0\Sigma_1=\Sigma_1\Sigma_0$. Conversely, if $\Sigma_0$ and $\Sigma_1$ commute, they are orthogonally co-diagonalizable, and both~\eqref{eq-gaussian-product-fm-terminal-map} and~\eqref{eq-gaussian-brenier-map-comparison} reduce in that basis to the diagonal map with entries $\sqrt{\lambda_{1,k}/\lambda_{0,k}}$. This proves the equivalence.
\end{proof}
\paragraph{Variations on the interpolant.}
\(v_t(z)=\lambda'(t)\,v^{\rm lin}_{\lambda(t)}(z), \qquad v^{\rm lin}_{r}(z)=\EE[Y-X\mid (1-r)X+rY=z].\)
\(Z_t=a_tX+b_tY,\qquad Y\sim\Gaussian(0,\sigma^2\Id),\)
then both the mixture centers and the component variances are changed. Writing $p_t$ for the density of $Z_t$ and $s_t=\nabla\log p_t$, Tweedie's formula gives, away from times where $a_t=0$,

\[
v_t(z)=a'_t\,\EE[X\mid Z_t=z]+b'_t\,\EE[Y\mid Z_t=z]
=\frac{a'_t}{a_t}z+
\left(\frac{a'_tb_t^2}{a_t}-b'_tb_t\right)\sigma^2s_t(z).
\]
For the linear bridge, $a_t=1-t$ and $b_t=t$, this recovers the formula above. For the variance-preserving Ornstein--Uhlenbeck noising used in diffusion models,

\(a_\tau=e^{-\tau},\qquad b_\tau=\sqrt{1-e^{-2\tau}},\)
one obtains the forward probability-flow velocity $v_\tau(z)=-z-\sigma^2\nabla\log p_\tau(z)$. Sampling follows the reverse field $z+\sigma^2\nabla\log p_\tau(z)$ as $\tau$ decreases.

\(a_t=(1-t)(1-2t), \qquad b_t=t,\)
\subsection{One-Step Generative Models}
\paragraph{Training a one-step flow.}
Let $\zeta$ be a simple latent distribution and let $\alpha_\theta=(G_\theta)_\sharp\zeta$ be the model distribution. Assume that the target data distribution is $\beta$. A Wasserstein-flow construction chooses a discrepancy

\(\mathcal E_\beta(\alpha),\)
\begin{equation}\label{eq-one-step-wgf}
\partial_t\mu_t+\diverg(\mu_t w_t)=0,
\qquad
w_t(x)=-\nabla\delta_\alpha \mathcal E_\beta(\mu_t)(x).
\end{equation}
\begin{equation}\label{eq-one-step-l2-fit}
\min_\eta \int_0^1\!\int
\norm{U_\eta(t,x)-w_t(x)}^2
\,\d\mu_t(x)\,\d t.
\end{equation}
\[
\alpha_{\theta}^{+}
=
(\Id+\tau U_\eta)_\sharp \alpha_\theta,
\qquad\text{or equivalently}\qquad
G_\theta^{+}(z)=G_\theta(z)+\tau U_\eta(G_\theta(z)).
\]
\paragraph{Self-corrected drifting fields.}
\begin{equation}\label{eq-normalized-kernel-drift}
B_\epsilon[\nu](x)
\eqdef
\frac{\int (y-x)K_\epsilon(x,y)\,\d\nu(y)}
{\int K_\epsilon(x,y)\,\d\nu(y)}.
\end{equation}
For the Gaussian kernel $K_\epsilon(x,y)=\exp(-\norm{x-y}^2/(2\epsilon))$, this normalized field is a score of a smoothed density:

\begin{equation}\label{eq-normalized-kernel-score}
B_\epsilon[\nu](x)
=
\epsilon\nabla\log\!\left(\int K_\epsilon(x,y)\,\d\nu(y)\right).
\end{equation}
\begin{equation}\label{eq-cross-minus-self-drift}
u_t(x)=B_\epsilon[\beta](x)-B_\epsilon[\mu_t](x)
=
\epsilon\nabla\log
\frac{\int K_\epsilon(x,y)\,\d\beta(y)}
{\int K_\epsilon(x,y)\,\d\mu_t(y)}.
\end{equation}
The first term pulls samples toward data, while the second term corrects self-attraction and prevents all particles from collapsing onto the same high-density region. For a fixed reference measure, \(B_\epsilon[\nu]\) is precisely the Gaussian mean-shift displacement in~\eqref{eq-l2-attention-mean-shift}: it moves $x$ toward the local kernel barycenter of $\nu$.

\begin{prop}[Drifting as a time-dependent Wasserstein gradient]\label{prop-drifting-semi-relaxed-gradient}
Let $\mu_t$ be a smooth curve of positive densities and let $u_t=\nabla\phi_t$ be a smooth time-dependent gradient field. Define the semi-relaxed functional
\begin{equation}\label{eq-semi-relaxed-drift-functional}
\mathcal R_t(\alpha|\mu_t)
\eqdef
-\int \phi_t(x)\,\d\alpha(x)
+\int \phi_t(x)\,\d\mu_t(x).
\end{equation}
Here $\mu_t$ and $\phi_t$ are frozen when taking the first variation with respect to the first argument $\alpha$. Then the continuity equation
\(\partial_t\mu_t+\diverg(\mu_t u_t)=0\)
is the formal Wasserstein gradient descent of the time-dependent functional $\alpha\mapsto\mathcal R_t(\alpha|\mu_t)$.
\end{prop}
\begin{proof}
Since $\mu_t$ and $\phi_t$ are fixed in the variation with respect to $\alpha$, the first variation is \(\delta_\alpha \mathcal R_t(\alpha|\mu_t)(x)=-\phi_t(x).\)
By Proposition~\ref{prop-formal-wass-gradient}, \(\Wgrad \mathcal R_t(\alpha|\mu_t) = \nabla\delta_\alpha \mathcal R_t(\alpha|\mu_t) = -\nabla\phi_t = -u_t.\)
The Wasserstein gradient-descent velocity is the negative of this gradient, namely $u_t$. Substituting this velocity in the continuity equation gives the claimed flow.
\end{proof}
\subsection{Moment Measures}
\label{sec-moment-measures}
\begin{defn}[Moment measure of a convex function]\label{def-moment-measure}
Let $u:\RR^d\to\RR\cup\{+\infty\}$ be a proper lower-semicontinuous convex function such that
\(Z_u\eqdef \int_{\RR^d} e^{-u(x)}\,\d x <+\infty.\)
The normalized log-concave measure associated with $u$ is \(\rho_u \eqdef Z_u^{-1} e^{-u(x)}\,\d x.\)
Whenever $\nabla u$ is defined $\rho_u$-almost everywhere, the moment measure of $u$ is \(\mathfrak M(u)\eqdef (\nabla u)_\sharp \rho_u.\)
\end{defn}
\(\int y\,\d\mathfrak M(u)(y) = Z_u^{-1}\int \nabla u(x)e^{-u(x)}\,\d x = -Z_u^{-1}\int \nabla(e^{-u(x)})\,\d x =0.\)
Thus moment measures are necessarily centered. The non-smooth theory uses essentially continuous convex functions, meaning convex functions whose restriction to the closure of their effective domain is continuous except possibly on a boundary set of zero $\mathcal H^{d-1}$ measure.

\begin{thm}[Cordero--Erausquin--Klartag]\label{thm-moment-measure-characterization}
Let $\mu\in\Pp_1(\RR^d)$ be a probability measure. There exists an essentially continuous convex function $u$ such that $\mu=\mathfrak M(u)$ if and only if
\(\int y\,\d\mu(y)=0 \qquad\text{and}\qquad \supp(\mu)\text{ is not contained in a hyperplane.}\)
Under these conditions, $u$ is unique up to translations of the argument and additive constants.
\end{thm}
\paragraph{Optimal-transport variational formulation.}
\begin{equation}\label{eq-moment-max-correlation}
\mathcal C_\mu(\rho)
\eqdef
\sup_{\pi\in\Couplings(\rho,\mu)}
\int_{\RR^d\times\RR^d} x\cdot y\,\d\pi(x,y).
\end{equation}
\begin{equation}\label{eq-moment-max-correlation-dual}
\mathcal C_\mu(\rho)
=
\inf_{v\ \mathrm{convex}}
\left\{
\int v(x)\,\d\rho(x)+\int v^*(y)\,\d\mu(y)
\right\},
\end{equation}
where $v^*$ is the Legendre transform. If $\rho,\mu\in\Pp_2(\RR^d)$, then

\begin{equation}\label{eq-moment-correlation-w2}
\mathcal C_\mu(\rho)
=
\frac12\int \norm{x}^2\,\d\rho(x)
+
\frac12\int \norm{y}^2\,\d\mu(y)
-
\frac12\Wass_2^2(\rho,\mu).
\end{equation}
\begin{equation}\label{eq-moment-measure-variational}
\min_{\rho\in\Pp_1(\RR^d)}
\left\{
\mathcal H(\rho)+\mathcal C_\mu(\rho)
\right\},
\qquad
\mathcal H(r\,\d x)\eqdef \int r(x)\log r(x)\,\d x,
\end{equation}
\begin{prop}[Variational characterization of moment measures]\label{prop-moment-hidden-convexity}
Let $\mu\in\Pp_1(\RR^d)$ be centered and not supported on a hyperplane. Then the minimization problem~\eqref{eq-moment-measure-variational} admits a solution, unique up to translations. Every minimizer is a log-concave density of the form
\(\rho=Z_u^{-1}e^{-u}\,\d x,\)
where $u$ is convex, essentially continuous, and satisfies the moment-measure equation
\(\mu=(\nabla u)_\sharp \rho.\)
Conversely, any essentially continuous convex $u$ satisfying this equation yields a global minimizer. In the finite-second-moment case $\mu\in\Pp_2(\RR^d)$, the objective $\rho\mapsto\mathcal H(\rho)+\mathcal C_\mu(\rho)$ is displacement convex along $\Wass_2$ geodesics on $\Pp_2(\RR^d)$ whenever the terms are finite. For general $\mu\in\Pp_1(\RR^d)$, the same statement is obtained by approximation and lower semicontinuity.
\end{prop}
\begin{proof}
The existence proof has two ingredients. First, since $\mu$ is centered, translating $\rho$ does not change either $\mathcal H(\rho)$ or $\mathcal C_\mu(\rho)$, so one can center a minimizing sequence. Second, the assumption that $\mu$ is not supported on a hyperplane gives a coercive lower bound of the form $\mathcal C_\mu(\rho)\geq c_\mu\int\norm{x}\,\d\rho(x)$ for centered absolutely continuous $\rho$, with $c_\mu>0$. Together with the lower semicontinuity estimates for entropy and maximal correlation, this yields a minimizer.

Let $\rho=r\,\d x$ be a minimizer and let $u$ be a convex optimizer in the dual formula~\eqref{eq-moment-max-correlation-dual}. Keeping $u$ fixed and varying $\rho$ in~\eqref{eq-moment-measure-variational} gives the Euler equation
\(\log r(x)+1+u(x)=\mathrm{constant} \qquad\text{on }\{r>0\},\)
so $\rho=Z_u^{-1}e^{-u}\d x$. The optimality condition for the scalar-product transport problem says that an optimal coupling is supported on $\{(x,y):y\in\partial u(x)\}$. Since $\rho$ is absolutely continuous and $u$ is convex, $\partial u(x)=\{\nabla u(x)\}$ for $\rho$-almost every $x$, hence $\mu=(\nabla u)_\sharp\rho$.

Conversely, assume $\rho=Z_u^{-1}e^{-u}\d x$ and $\mu=(\nabla u)_\sharp\rho$. Let $\nu$ be a smooth compactly supported competitor, and let $T$ be the Brenier map from $\rho$ to $\nu$. Along the geodesic $\rho_t=((1-t)\Id+tT)_\sharp\rho$, the right derivative of the entropy at $t=0$ is
\(\frac{\d}{\d t}\mathcal H(\rho_t)\Big|_{t=0^+} = -\int \dotp{T(x)-x}{\nabla u(x)}\,\d\rho(x),\)
where the identity follows by differentiating the Jacobian formula and integrating by parts. Santambrogio's derivative estimate for the maximal-correlation functional gives
\(\frac{\d}{\d t}\mathcal C_\mu(\rho_t)\Big|_{t=0^+} \geq \int \dotp{T(x)-x}{\nabla u(x)}\,\d\rho(x),\)
because $(\Id,\nabla u)_\sharp\rho$ is optimal for the scalar-product problem. The first-order terms cancel, hence the one-sided derivative of $\mathcal H+\mathcal C_\mu$ at $\rho$ in every such direction is nonnegative. Displacement convexity then implies global minimality, and approximation removes the smooth compact-support restriction. Strict displacement convexity of entropy gives uniqueness, except for translations; translations do not change $\mathcal C_\mu$ because $\mu$ is centered.

The entropy term is displacement convex by McCann's theorem, recalled in Proposition~\ref{prop-basic-geodesic-convexity}. If $\mu$ has finite second moment, identity~\eqref{eq-moment-correlation-w2} writes $\mathcal C_\mu$ as the sum of the $1$-convex moment term $\rho\mapsto\frac12\int\norm{x}^2\,\d\rho$ and the $(-1)$-convex term $\rho\mapsto-\frac12\Wass_2^2(\rho,\mu)$, hence $\mathcal C_\mu$ is displacement convex. For merely finite first moment $\mu$, the same convexity and the variational characterization follow by the approximation and semicontinuity arguments of Santambrogio.
\end{proof}
\paragraph{Conjugate moment measures for generation.}
\begin{equation}\label{eq-conjugate-moment-measure}
\beta
=
(\nabla w^*)_\sharp
\left(Z_w^{-1}e^{-w(z)}\,\d z\right).
\end{equation}
\subsection{Evolution in Depth of Transformers}
\paragraph{Attention as a context-dependent velocity.}
After tokenization, embedding, and positional encoding, each input (from a set of tokens) is represented as a point cloud $(x_i)_{i=1}^n$ of $n$ points in the space of vectorized tokens. An attention layer with skip connection and rescaling by $1/T$ (where $T$ is the depth) defines a transformation of the tokens:

\(x_i \mapsto x_i + \frac{1}{T} \sum_j \frac{e^{\langle Q x_i, K x_j \rangle} V x_j}{\sum_{\ell} e^{\langle Q x_i, K x_\ell \rangle}},\)
where $\theta = (K, Q, V)$ are the parameters of the attention layer, represented by three matrices.

\paragraph{Token measure evolution.}
To handle an arbitrary number of tokens, we define $\alpha = \frac{1}{n} \sum_i \delta_{x_i}$ as the empirical measure of tokens and rewrite the transformer mapping as:

\(x_i \mapsto x_i + \frac{1}{T} \Gamma_\theta[\alpha](x_i),\)
\(\Gamma_\theta[\alpha](x):= \frac{\int e^{\langle Q x, K y \rangle} V y \, \d \alpha(y)} {\int e^{\langle Q x, K z \rangle} \, \d \alpha(z)}.\)
\(\alpha_{t+\tau} = (\Id + \tau \Gamma_{\theta_t}[\alpha_t])_\sharp \alpha_t.\)
\(\partial_t \alpha_t + \diverg(\alpha_t \Gamma_{\theta_t}[\alpha_t]) = 0.\)
\paragraph{\texorpdfstring{$L^2$ attention and mean shift.}{L^2 attention and mean shift.}}
\[
K_\epsilon(x,y)=\exp(-\norm{x-y}^2/(2\epsilon)),
\qquad
\rho_\epsilon[\alpha](x)=\int K_\epsilon(x,y)\d\alpha(y),
\qquad
m_\epsilon[\alpha](x)
=
\frac{\int yK_\epsilon(x,y)\d\alpha(y)}
{\rho_\epsilon[\alpha](x)}.
\]
\begin{equation}\label{eq-l2-attention-mean-shift}
M_\epsilon[\alpha](x)
\eqdef
m_\epsilon[\alpha](x)-x
=
\frac{\int (y-x)K_\epsilon(x,y)\d\alpha(y)}
{\rho_\epsilon[\alpha](x)}
=
\epsilon\nabla\log\rho_\epsilon[\alpha](x)
\end{equation}
\(x_i^{k+1} = (1-\tau)x_i^k+\tau m_\epsilon[\alpha_k](x_i^k) = x_i^k+\tau M_\epsilon[\alpha_k](x_i^k)\)
\(\partial_t\alpha_t+\diverg\bigl(\alpha_tM_\epsilon[\alpha_t]\bigr)=0.\)
\paragraph{Gradient structure and limitations.}
When the token space has dimension $d$ and the query/key space has dimension $r$, take $Q,K\in\RR^{r\times d}$ and $V\in\RR^{d\times d}$. If $V=Q^\top K$, the field $\Gamma_\theta[\alpha]$ is a gradient vector field in the token variable. Indeed, define the log-partition potential

\(\Phi_\alpha(x) = \int \exp(\dotp{Qx}{Ky})\d\alpha(y), \qquad U_\alpha(x)=\log\Phi_\alpha(x).\)
\(\nabla_x U_\alpha(x) = \frac{\int Q^\top K y\,\exp(\dotp{Qx}{Ky})\d\alpha(y)} {\int \exp(\dotp{Qx}{Kz})\d\alpha(z)} = \Gamma_\theta[\alpha](x).\)
This is an instantaneous gradient in $x$. It is not, however, the gradient of the first variation of a fixed functional of $\alpha$, because the potential $U_\alpha$ itself depends on the current measure through the same attention normalization.

\subsection{Flows over the Gaussian Manifold}
\paragraph{Gaussianity preservation.}
\begin{prop}[Affine velocities preserve Gaussianity]\label{prop-gaussian-affine-closure}
Let $\alpha_t=\Gaussian(\mean_t,\cov_t)$, with $\cov_t$ positive definite, solve the continuity equation with an affine velocity
\(v_t(x)=b_t+A_t(x-\mean_t).\)
Then $\alpha_t$ remains Gaussian and its moments solve
\(\dot \mean_t=b_t, \qquad \dot\cov_t=A_t\cov_t+\cov_t A_t^\top.\)
Conversely, any smooth Gaussian curve with positive definite covariance can be generated by such an affine velocity. If one wants the velocity to be a Wasserstein tangent gradient, one chooses the unique symmetric solution of the Lyapunov equation
\(A_t\cov_t+\cov_t A_t=\dot\cov_t.\)
\end{prop}
\begin{proof}
Let $X_t$ follow the characteristic ODE $\dot X_t=b_t+A_t(X_t-\mean_t)$. This linear ODE maps Gaussian random variables to Gaussian random variables. Taking expectation gives $\dot\mean_t=b_t$. Writing $\tilde X_t=X_t-\mean_t$, one has $\dot{\tilde X}_t=A_t\tilde X_t$, hence
\(\dot\cov_t = \frac{\d}{\d t}\EE(\tilde X_t\tilde X_t^\top) = A_t\cov_t+\cov_t A_t^\top.\)
For the converse, set $b_t=\dot\mean_t$ and choose any matrix $A_t$ satisfying the covariance equation. Since $\cov_t$ is positive definite, the Lyapunov map $A\mapsto A\cov_t+\cov_t A$ is invertible on symmetric matrices, which gives the unique symmetric choice when a gradient velocity is required. In that case $v_t$ is the gradient of the quadratic potential $x\mapsto \dotp{b_t}{x}+\dotp{A_t(x-\mean_t)}{x-\mean_t}/2$.
\end{proof}
\paragraph{Constrained evolution on the Gaussian manifold.}
\(\mathcal G=\{\Gaussian(\mean,\cov):\mean\in\RR^d,\ \cov\succ0\}\)
be the Gaussian submanifold of $\Pp_2(\RR^d)$. The Wasserstein gradient of a functional constrained to a smooth submanifold $\mathcal M\subset\Pp_2$ is defined as the Riesz representative of the differential restricted to tangent velocities of $\mathcal M$.

\(\alpha^{k+1}\in \uargmin{\alpha\in\mathcal M} \frac{1}{2\tau}\Wass_2^2(\alpha,\alpha^k)+f(\alpha).\)
For $\mathcal M=\mathcal G$, tangent velocities are affine gradient fields $v(x)=b+A(x-\mean)$ with $A=A^\top$. The constrained gradient is therefore the $L^2(\Gaussian(\mean,\cov))$ projection of the ambient Wasserstein gradient onto this finite-dimensional affine space, whenever the ambient gradient exists.

\begin{prop}[Gaussian-constrained Wasserstein gradients]\label{prop-gaussian-gradient-bullet-list}
Let $f$ be a smooth functional and assume that its restriction to nondegenerate Gaussian measures can be written as \(f(\Gaussian(\mean,\cov))=F(\mean,\cov).\)
Then the Wasserstein gradient constrained to the Gaussian family is the affine vector field
\(v_F(x) = \nabla_\mean F(\mean,\cov) + 2\nabla_\cov F(\mean,\cov)(x-\mean),\)
where $\nabla_\cov F$ denotes the symmetric matrix derivative. Equivalently, $v_F$ is the $L^2(\Gaussian(\mean,\cov))$ projection of the ambient Wasserstein gradient onto affine gradient fields, whenever the ambient gradient exists. Hence the gradient descent flow constrained to Gaussian measures satisfies
\begin{equation}\label{eq-gaussian-wgf-closure}
\dot\mean_t=-\nabla_\mean F(\mean_t,\cov_t),
\qquad
\dot\cov_t=-2\bigl(\cov_t\nabla_\cov F(\mean_t,\cov_t)+\nabla_\cov F(\mean_t,\cov_t)\cov_t\bigr),
\end{equation}
and the descent velocity is affine.
\end{prop}
\begin{proof}
Test the functional along a Gaussian tangent vector, represented by an affine gradient field
\(v(x)=b+A(x-\mean)\)
with $A$ symmetric. The induced first-order variations are $\dot\mean=b$ and $\dot\cov=A\cov+\cov A$. Therefore
\[
\d F(\mean,\cov)[b,A\cov+\cov A]
=
\dotp{\nabla_\mean F}{b}
+
\mathrm{tr}\!\left(\nabla_\cov F(A\cov+\cov A)\right).
\]
Since $A$, $\cov$ and $\nabla_\cov F$ are symmetric, the second term equals
\(2\,\mathrm{tr}(\nabla_\cov F\,A\cov) = \int \dotp{2\nabla_\cov F(x-\mean)}{A(x-\mean)}\d\Gaussian(\mean,\cov)(x).\)
Together with the mean term, this gives \(\d F(\mean,\cov)[\dot\mean,\dot\cov] = \int \dotp{v_F(x)}{v(x)}\d\Gaussian(\mean,\cov)(x)\)
for all affine gradient fields $v$. This identifies the constrained Wasserstein gradient in the induced $L^2(\alpha)$ metric, or equivalently the projection of the ambient gradient when it exists. Substituting the descent velocity $-v_F$ in Proposition~\ref{prop-gaussian-affine-closure} gives~\eqref{eq-gaussian-wgf-closure}.
\end{proof}
\paragraph{Gaussian-preserving gradient flows.}
\begin{prop}[Centered Gaussian covariance catalogue]\label{prop-centered-gaussian-covariance-catalogue}
Let $\gamma=\Gaussian(0,\Id)$ and let $\mu_t=\Gaussian(0,C_t)$ with $C_t\succ0$. For the normalizations displayed below, the Wasserstein descent constrained to the centered Gaussian manifold satisfies $\dot C_t=h(C_t)$, with
\[
\begin{array}{rcl}
\KL(\mu|\gamma) &:& h(C)=2(\Id-C), \\[.35em]
\frac12\,\mathcal I(\mu|\gamma) &:& h(C)=2(C^{-1}-C), \\[.35em]
\Wass_2^2(\mu,\gamma) &:& h(C)=4(C^{1/2}-C), \\[.35em]
\MMD_k^2(\mu,\gamma),\quad k(x,y)=\dotp{x}{y}^2
&:& h(C)=8(C-C^2), \\[.35em]
S_\epsilon(\mu,\gamma)
&:& h(C)=
4\left(C+\frac{\epsilon^2}{16}\Id\right)^{1/2}
-2\left(C^2+\frac{\epsilon^2}{16}\Id\right)^{1/2}
-2C-\frac{\epsilon}{2}\Id, \\[.75em]
\SW_2^2(\mu,\gamma)
&:& h(C)=V(C)C+CV(C),
\end{array}
\]
where $S_\epsilon$ is the debiased Sinkhorn divergence for the quadratic cost $\norm{x-y}^2$ and KL regularization strength $\epsilon$, and
\[
V(C)=
2\int_{\Sphere^{d-1}}
\left(\frac{1}{\sqrt{\theta^\top C\theta}}-1\right)
\theta\theta^\top \,\d\sigma(\theta)
\]
for the normalized spherical measure $\sigma$. Here
\[
\mathcal I(\mu|\gamma)=\int \left|\nabla\log\rho(x)+x\right|^2\rho(x)\,\d x
\qquad(\mu=\rho\,\d x).
\]
Thus the unhalved Fisher divergence has right-hand side $4(C^{-1}-C)$. Multiplying any of these energies by a constant simply rescales the corresponding right-hand side.
\end{prop}
\begin{proof}
Each row is obtained by identifying the affine descent velocity $v(x)=M_Cx$ generated by the corresponding Gaussian-constrained calculation and then applying Proposition~\ref{prop-gaussian-affine-closure}, which gives $\dot C=M_CC+CM_C^\top$. For $\KL(\cdot|\gamma)$, the Fokker--Planck velocity is $(C^{-1}-\Id)x$, hence $\dot C=2(\Id-C)$. For the Fisher row, the restriction of $\frac12\mathcal I$ to centered Gaussians is
\[
\frac12\left(\tr(C)+\tr(C^{-1})-2d\right).
\]
Using Proposition~\ref{prop-gaussian-gradient-bullet-list} gives the descent velocity $(C^{-2}-\Id)x$, hence $\dot C=2(C^{-1}-C)$. This row should be read as a Gaussian projected closure of the fourth-order Fisher flow.

For $\Wass_2^2(\cdot,\gamma)$, the Brenier map from $\Gaussian(0,C)$ to $\gamma$ is $C^{-1/2}x$, so the descent velocity for the unhalved squared distance is $2(C^{-1/2}-\Id)x$, giving $4(C^{1/2}-C)$. For the polynomial MMD row, centered Gaussians satisfy $\MMD_k^2(\mu,\gamma)=\norm{C-\Id}_{\mathrm F}^2$; the first variation is quadratic and its descent velocity is $4(\Id-C)x$, giving $8(C-C^2)$.

Gaussian Sinkhorn dual potentials are quadratic, so the velocity is again linear; differentiating the closed Gaussian formula yields the displayed spectral expression. The square roots are spectral functions of $C$, hence commute with $C$, which is why the covariance ODE closes as a matrix function of $C$ alone. For sliced Wasserstein, each one-dimensional projection is a Gaussian transport with velocity $2((\theta^\top C\theta)^{-1/2}-1)\dotp{\theta}{x}\theta$; averaging these velocities over $\Sphere^{d-1}$ gives $v(x)=V(C)x$ and thus $\dot C=V(C)C+CV(C)$.
\end{proof}
\paragraph{Non-variational Gaussian-preserving flows.}
\paragraph{Contractive Gaussian projection.}
\begin{thm}[Gelbrich theorem]\label{thm-gelbrich-projection}
For $\mu\in\Pp_2(\RR^d)$, let
\(\mathcal R\mu\eqdef \Gaussian(\mean_\mu,\cov_\mu)\)
be the Gaussian with the same mean and covariance as $\mu$. Then
\(\Wass_2^2(\mathcal R\mu,\mathcal R\nu) = \norm{\mean_\mu-\mean_\nu}^2+\Bb^2(\cov_\mu,\cov_\nu) \leq \Wass_2^2(\mu,\nu).\)
\end{thm}
\begin{proof}
Take any coupling $(X,Y)$ of $\mu$ and $\nu$, center the variables, and write $C=\EE[(X-\mean_\mu)(Y-\mean_\nu)^\top]$. In the positive definite case, positivity of the block covariance matrix implies the factorization $C=\cov_\mu^{1/2}K\cov_\nu^{1/2}$ with $\norm{K}_{\mathrm{op}}\leq1$, and therefore, by operator/nuclear norm duality,
\[
\tr C\leq \tr\left((\cov_\mu^{1/2}\cov_\nu\cov_\mu^{1/2})^{1/2}\right).
\]
The semidefinite case follows by adding $\eta\Id$ to both covariance matrices and letting $\eta\downarrow0$.
Expanding $\EE\norm{X-Y}^2$ gives the lower bound
\(\EE\norm{X-Y}^2 \geq \norm{\mean_\mu-\mean_\nu}^2+\Bb^2(\cov_\mu,\cov_\nu).\)
Taking the infimum over couplings proves the inequality, while equality for Gaussian laws is Proposition~\ref{prop-gaussian-w2-bures}.
\end{proof}
\begin{thm}[Hugo Lavenant Gaussian-preservation criterion]\label{thm-lavenant-gaussian-preserving-jko}
Let $F:\Pp_2(\RR^d)\to(-\infty,+\infty]$ satisfy
\(F(\mathcal R\mu)\leq F(\mu) \qquad\forall\mu\in\Pp_2(\RR^d),\)
with $\mathcal R$ defined in Theorem~\ref{thm-gelbrich-projection}. If $\gamma$ is Gaussian and $\nu$ minimizes the JKO step
\(\eta\mapsto F(\eta)+\frac1{2\tau}\Wass_2^2(\gamma,\eta),\)
then $\mathcal R\nu$ is also a minimizer. If the JKO minimizer is unique, it is Gaussian. Thus any unique Wasserstein gradient flow obtained as the limit of this JKO scheme preserves Gaussian initial data.
\end{thm}
\begin{proof}
For the JKO claim, $\mathcal R\gamma=\gamma$ because $\gamma$ is Gaussian. Hence, for any competitor $\eta$, \(F(\mathcal R\eta)+\frac1{2\tau}\Wass_2^2(\gamma,\mathcal R\eta) \leq F(\eta)+\frac1{2\tau}\Wass_2^2(\gamma,\eta).\)
Applying this to a minimizer $\eta=\nu$ shows that $\mathcal R\nu$ is again a minimizer. Uniqueness forces $\nu=\mathcal R\nu$.
\end{proof}