\section{Wasserstein Gradient Flows}
\label{sec-wasserstein-gradient-flows}
\subsection{Minimizing Movements and Wasserstein Gradients}
\begin{equation}
\alpha_{t+\tau}:= \arg\min_\alpha \frac{1}{2 \tau} \Wass_2(\alpha_t, \alpha)^2 + f(\alpha). \label{eq:jko-discr}
\end{equation}
\paragraph{Euclidean gradient flows.}
If we restrict \eqref{eq:jko-discr} to finite dimensions and assume $\alpha_t = \delta_{x(t)}$ and $\alpha = \delta_x$ (single Dirac measures), this matches the implicit Euler scheme:

\(x(t+\tau):= \arg\min_x \frac{1}{2 \tau} \norm{x - x(t)}^2 + h(x),\)
where $h(x) = f(\delta_x)$. Its solution is formally given by the implicit Euler formula:

\(x(t+\tau) = (\Id + \tau \nabla h)^{-1}(x(t)).\)
\(x(t+\tau) = (\Id - \tau \nabla h)(x(t)) = x(t) - \tau \nabla h(x(t)).\)
\begin{equation}
\dot{x}(t) = -\nabla h(x(t)). \label{eq:grad-flow-classical}
\end{equation}
\paragraph{Wasserstein gradient formula.}
\(f((1-\tau)\alpha + \tau \beta) = f(\alpha + \tau \rho) = f(\alpha) + \tau \int [\delta f(\alpha)](x) \d \rho(x) + o(\tau).\)
\begin{defn}[Wasserstein gradient]\label{def-wasserstein-gradient}
Assume that $f$ admits a smooth first variation $\delta f(\alpha)$. In the smooth formal calculus on $\Pp_2(\RR^d)$, the Wasserstein gradient of $f$ at $\alpha$ is the gradient vector field
\(\Wgrad f(\alpha) = \nabla_x \delta f(\alpha).\)
\end{defn}
\begin{equation}
\frac{\partial \alpha_t}{\partial t} + \diverg(-\Wgrad f(\alpha_t) \alpha_t) = 0. \label{eq:wassflow-pde}
\end{equation}
\begin{proposition}[Formal Wasserstein gradient]\label{prop-formal-wass-gradient}
Assume that $f$ admits a smooth first variation $\delta f(\alpha)$ and that $\alpha$ has a smooth positive density. For infinitesimal perturbations generated by a velocity field $v$ through $(\Id+\tau v)_\sharp\alpha$, the differential of $f$ is
\(\frac{\d}{\d\tau}_{|\tau=0} f\big((\Id+\tau v)_\sharp\alpha\big) = \int \dotp{\nabla \delta f(\alpha)(x)}{v(x)}\d\alpha(x).\)
Hence, for the Riemannian metric $\norm{v}_{L^2(\alpha)}^2=\int \norm{v}^2\d\alpha$, the Wasserstein gradient is the vector field
\(\Wgrad f(\alpha)=\nabla \delta f(\alpha).\)
\end{proposition}
\begin{proof}
The push-forward expansion gives, in the sense of distributions, \((\Id+\tau v)_\sharp\alpha = \alpha-\tau\diverg(\alpha v)+o(\tau).\)
Using the definition of the first variation, \(f((\Id+\tau v)_\sharp\alpha) = f(\alpha)-\tau\int \delta f(\alpha)\diverg(\alpha v)\d x+o(\tau).\)
An integration by parts, with either compact support or vanishing boundary flux, gives \(-\int \delta f(\alpha)\diverg(\alpha v)\d x = \int \dotp{\nabla\delta f(\alpha)}{v}\d\alpha.\)
By definition of the Riesz representative for the $L^2(\alpha)$ metric, this representative is $\nabla\delta f(\alpha)$.
\end{proof}
\paragraph{From the JKO step to the velocity field.}
\(\min_{v} \frac{1}{2 \tau} \tau^2 \norm{v}_{L^2(\alpha_t)}^2 + f( (\Id + \tau v)_\sharp \alpha_t ).\)
\((\Id + \tau v)_\sharp \alpha_t = \alpha_t - \tau \diverg( v \alpha_t ) + o(\tau)\)
\(f( (\Id + \tau v)_\sharp \alpha_t ) = f(\alpha_t) - \tau \int \delta f(\alpha_t) \diverg( v \alpha_t ) \d x + o(\tau)\)
\(= f(\alpha_t) + \tau \int \langle \nabla_x \delta f(\alpha_t)(x), v(x) \rangle \d \alpha_t(x) + o(\tau)\)
\(\min_{v} f(\alpha_t) + \tau \int \Big[ \frac{1}{2} \norm{v(x)}^2 + \langle \Wgrad f(\alpha_t)(x), v(x) \rangle \Big]\d \alpha_t(x) + o(\tau).\)
\paragraph{Discrete evolutions.}
If $f(\alpha)$ can be evaluated on discrete distributions and $\Wgrad$ is continuous in this case, the flow \eqref{eq:wassflow-pde} maintains the number of Dirac masses, $\alpha_t = \frac{1}{n} \sum_i \delta_{x_i(t)}$. The particles $X(t):= (x_i(t))_i$ evolve according to a system of coupled ODEs:

\begin{equation}
\dot{x}_i(t) = -n\nabla_{x_i} F(X(t)), \label{eq:wassflows-particles}
\end{equation}
where $F(X):= f\left(\frac{1}{n} \sum_i \delta_{x_i}\right)$ and the factor $n$ comes from the empirical Wasserstein metric $\frac1n\sum_i \norm{\dot x_i}^2$.

\paragraph{Linear Functionals.}
\paragraph{Shannon Neg-Entropy.}
\paragraph{Interaction Energies.}
\begin{equation}
f(\alpha):= \iint k(x, y) \d \alpha(x) \d \alpha(y). \label{eq:quadratic-func}
\end{equation}
For a symmetric kernel $k$:

\(\delta f(\alpha)(x) = 2 \int k(x, y) \d \alpha(y), \quad \Wgrad f(\alpha)(x) = 2 \int \nabla_x k(x, y) \d \alpha(y).\)
For $\alpha_0 = \frac{1}{n} \sum_i \delta_{x_i}$, the flow \eqref{eq:wassflow-pde} implies particles $(x_i(t))_i$ obey:

\(\dot{x}_i(t) = -\frac{2}{n} \sum_j \nabla k(x_i(t), x_j(t)).\)
If $k$ is positive definite, or more generally conditionally positive definite on signed measures of zero total mass as for the energy-distance kernel $k(x,y)=-\norm{x-y}$, and one minimizes the squared kernel discrepancy to a teacher distribution $\beta$, then

\[
\norm{\alpha-\beta}_k^2
=
\iint k\d\alpha\d\alpha
-2\int\!\left(\int k(x,y)\d\beta(y)\right)\d\alpha(x)
+\mathrm{constant}.
\]
Thus MMD-type training energies are exactly an interaction energy plus a linear potential; the teacher distribution appears through the potential $x\mapsto-2\int k(x,y)\d\beta(y)$. The corresponding empirical Wasserstein gradient flow is

\(\dot x_i(t) = -\frac{2}{n}\sum_j\nabla_x k(x_i(t),x_j(t)) + 2\int\nabla_x k(x_i(t),y)\d\beta(y).\)
\paragraph{Stochastic particles and McKean--Vlasov limits.}
\(\d X_t=b(X_t)\d t+\sqrt{2}\sigma\d B_t,\)
and the one-particle law $\alpha_t=\rho_t\d x$ directly satisfies the linear Fokker--Planck equation

\(\partial_t\rho_t=-\diverg(b\rho_t)+\sigma^2\Delta\rho_t.\)
\(\d X_i^n(t)=b(X_i^n(t),\mu_t^n)\d t+\sqrt{2}\sigma\d B_i(t), \qquad \mu_t^n=\frac1n\sum_{i=1}^n\delta_{X_i^n(t)}.\)
For finite $n$, the empirical law $\mu_t^n$ is itself random. Under suitable Lipschitz, growth and chaotic-initialization assumptions, propagation of chaos states that finitely many particles become asymptotically independent as $n\to\infty$, all with the same deterministic law $\rho_t\d x$; equivalently, the empirical measure $\mu_t^n$ converges in probability to this law.

\(\partial_t \rho_t = -\diverg\big(b(x,\rho_t)\rho_t\big) + \sigma^2\Delta\rho_t.\)
\(b(x,\rho)=-\nabla \frac{\delta \mathcal E}{\delta \rho}(x),\)
\(\mathcal E(\rho)+\sigma^2\int \rho\log\rho\,\d x.\)
\paragraph{Gradient flows under density constraints.}
\begin{equation}\label{eq-density-constrained-energy}
F(\rho\d x)=\int_\Omega h(x)\rho(x)\d x+
\iota_{\mathcal K_\kappa}(\rho\d x),
\qquad
\mathcal K_\kappa=
\enscond{\rho\d x\in\Pp_2(\Omega)}{0\leq\rho\leq\kappa\ \text{a.e.}},
\end{equation}
where $\Omega\subset\RR^d$ is a convex domain and $\mathcal K_\kappa$ is assumed nonempty; if $\Omega$ has finite volume, this requires $\kappa|\Omega|\geq1$. The indicator term has no classical first variation; it contributes the normal cone to $\mathcal K_\kappa$ in the Wasserstein first-order condition.

\begin{equation}\label{eq-density-constrained-jko}
\alpha^{k+1}
\in
\uargmin{\alpha\in\mathcal K_\kappa}
\frac{1}{2\tau}\Wass_2^2(\alpha^k,\alpha)+\int_\Omega h\,\d\alpha.
\end{equation}
\begin{proposition}[Geodesic convexity of density caps]\label{prop-density-cap-geodesic-convex}
Assume that $\Omega\subset\RR^d$ is convex and let $\kappa>0$. The set $\mathcal K_\kappa$ in~\eqref{eq-density-constrained-energy} is geodesically convex in $\Pp_2(\Omega)$: if $\alpha_0,\alpha_1\in\mathcal K_\kappa$, then the constant-speed $\Wass_2$ geodesic $(\alpha_t)_{t\in[0,1]}$ between them also belongs to $\mathcal K_\kappa$.
\end{proposition}
\begin{proof}
Assume first that the endpoints have smooth positive densities $\alpha_i=\rho_i\d x$ with $\rho_i\leq\kappa$, and let $T=\nabla\varphi$ be the Brenier map from $\alpha_0$ to $\alpha_1$. Since $\Omega$ is convex, $T_t=(1-t)\Id+tT$ maps $\Omega$ into $\Omega$. The interpolation is $\alpha_t=(T_t)_\sharp\alpha_0$ and
\(\rho_t(T_t(x))=\frac{\rho_0(x)}{\det((1-t)I+t\nabla T(x))}.\)
The concavity of $\det^{1/d}$ on positive semidefinite matrices and the identity $\rho_1(T(x))\det\nabla T(x)=\rho_0(x)$ give
\(\rho_t(T_t(x))^{-1/d} \geq (1-t)\rho_0(x)^{-1/d}+t\rho_1(T(x))^{-1/d} \geq\kappa^{-1/d}.\)
Hence $\rho_t\leq\kappa$ for all $t$. The general case follows by approximation and lower semicontinuity of the $L^\infty$ bound, as in McCann's displacement-convexity theory.
\end{proof}
\begin{equation}\label{eq-density-constrained-pressure-flow}
\partial_t\rho_t=\diverg\left(\rho_t\nabla(h+p_t)\right),
\qquad
0\leq\rho_t\leq\kappa,
\qquad
p_t\geq0,
\qquad
p_t(\kappa-\rho_t)=0.
\end{equation}
The pressure vanishes in the unsaturated region and acts only where $\rho_t=\kappa$, pushing mass away from congested zones so that the density cap remains satisfied.

\paragraph{Multi-species gradient flows.}
Several applications involve several densities living on the same physical space: chemical species, color channels, cell populations or competing phases. Let $\Omega\subset\RR^d$ be the physical domain and assume that the component masses $\mathbf m=(m_1,\ldots,m_p)$ are positive and fixed.

\begin{equation}\label{eq-multispecies-space}
\Pp_{2,\mathbf m}(\Omega;\RR_+^p)=
\enscond{\alpha=(\alpha_1,\ldots,\alpha_p)}
{\alpha_i\in\Mm_+(\Omega),\ \alpha_i(\Omega)=m_i,\ \int_\Omega\norm{x}^2\d\alpha_i(x)<+\infty}.
\end{equation}
\begin{equation}\label{eq-multispecies-product-metric}
\Wass_{2,\oplus}^2(\alpha,\eta)=
\sum_{i=1}^p m_i\,
\Wass_2^2\!\left(\frac{\alpha_i}{m_i},\frac{\eta_i}{m_i}\right).
\end{equation}
\begin{equation}\label{eq-multispecies-jko}
\alpha^{k+1}
\in
\uargmin{\alpha\in\Pp_{2,\mathbf m}(\Omega;\RR_+^p)}
\frac{1}{2\tau}\Wass_{2,\oplus}^2(\alpha^k,\alpha)+F(\alpha).
\end{equation}
For smooth densities and first variations $\phi_i=\delta F/\delta\rho_i$, this yields

\begin{equation}\label{eq-multispecies-pde}
\partial_t\rho_i=\diverg\left(\rho_i\nabla\phi_i\right),
\qquad
\phi_i=\frac{\delta F}{\delta\rho_i},
\qquad i=1,\ldots,p.
\end{equation}
\begin{equation}\label{eq-multispecies-composition-constraint}
\mathcal C_\beta=
\enscond{\alpha\in\Pp_{2,\mathbf m}(\Omega;\RR_+^p)}
{\sum_{i=1}^p \alpha_i=\beta}.
\end{equation}
\begin{equation}\label{eq-multispecies-constrained-jko}
\alpha^{k+1}
\in
\uargmin{\alpha\in\mathcal C_\beta}
\frac{1}{2\tau}\Wass_{2,\oplus}^2(\alpha^k,\alpha)+F(\alpha).
\end{equation}
\begin{equation}\label{eq-multispecies-constrained-pde}
\partial_t\rho_i=\diverg\left(\rho_i\nabla(\phi_i-\lambda_t)\right),
\qquad
\diverg(b\nabla\lambda_t)=\diverg\left(\sum_{i=1}^p \rho_i\nabla\phi_i\right).
\end{equation}
For the separable Shannon entropy $F(\alpha)=\sum_i\int_\Omega \rho_i\log\rho_i\d x$, this becomes

\begin{equation}\label{eq-multispecies-modified-heat}
\partial_t\rho_i=\Delta\rho_i-\diverg(\rho_i\nabla\lambda_t),
\qquad
\diverg(b\nabla\lambda_t)=\Delta b.
\end{equation}
When $b$ is constant, the pressure can be chosen constant and the system reduces to independent heat equations which nevertheless preserve the pointwise sum $\sum_i\rho_i=b$. This product geometry should be contrasted with the vector-valued Benamou--Brenier distances of Section~\ref{sec-vector-matrix-valued-measures}, where the metric itself can couple the components through a non-diagonal mobility.

\subsection{Unbalanced Wasserstein--Fisher--Rao Gradient Flows}
The Wasserstein gradient flows above conserve total mass, because their tangent vectors are generated by continuity equations. When the unknown measure represents an intensity, a population or a cloud of trainable particles, it is often useful to combine spatial motion with creation and destruction of mass.

\begin{defn}[WFR minimizing movement]\label{def-wfr-jko-gradient-flow}
Let $F:\Mm_+(\RR^d)\to\RR\cup\{+\infty\}$ be an energy on positive finite measures and let $\WFR_\kappa$ be the Wasserstein--Fisher--Rao distance with growth scale $\kappa>0$, defined by the dynamic balance-equation formulation~\eqref{eq-dynamic-unbalanced-ot} and its static equivalence in Proposition~\ref{prop-static-dynamic-unbalanced}. The unbalanced JKO step is
\begin{equation}\label{eq-wfr-jko}
\alpha^{k+1}
\in
\uargmin{\alpha\in\Mm_+(\RR^d)}
\frac{1}{2\tau}\WFR_\kappa^2(\alpha^k,\alpha)+F(\alpha).
\end{equation}
The continuous-time limit, when it exists, is called the $\WFR_\kappa$ gradient flow of $F$.
\end{defn}
\(\partial_t\rho_t+\diverg(\rho_t v_t)=g_t\rho_t,\)
\begin{proposition}[Formal WFR gradient-flow PDE]\label{prop-wfr-gradient-flow-pde}
Assume that $F$ has a smooth first variation $\phi_t=\delta F(\alpha_t)$ and that $\alpha_t=\rho_t\d x$ is smooth and positive. The formal $\WFR_\kappa$ gradient flow of $F$ is
\begin{equation}\label{eq-wfr-gradient-flow-pde}
\partial_t\rho_t
=
\diverg(\rho_t\nabla\phi_t)
-
\frac{1}{\kappa^2}\phi_t\rho_t.
\end{equation}
Along such a smooth solution,
\begin{equation}\label{eq-wfr-energy-dissipation}
\frac{\d}{\d t}F(\alpha_t)
=
-\int
\left(\norm{\nabla\phi_t}^2+\frac{1}{\kappa^2}\phi_t^2\right)\rho_t\,\d x
\leq 0.
\end{equation}
\end{proposition}
\begin{proof}
For an infinitesimal tangent pair $(v,g)$, the density variation is $\zeta=-\diverg(\rho v)+g\rho$. The first variation of $F$ is \(\int \phi\,\zeta = \int \dotp{\nabla\phi}{v}\rho\,\d x+\int \phi g\rho\,\d x,\)
after integration by parts. For the inner product
\(\dotp{(v,g)}{(\tilde v,\tilde g)}_\rho = \int \dotp{v}{\tilde v}\rho\,\d x+\kappa^2\int g\tilde g\rho\,\d x,\)
the Riesz representative is $(\nabla\phi,\phi/\kappa^2)$. Steepest descent therefore uses $(v,g)=(-\nabla\phi,-\phi/\kappa^2)$, which gives~\eqref{eq-wfr-gradient-flow-pde}.
Substituting this pair in the first-variation formula gives
\[
\frac{\d}{\d t}F(\alpha_t)
=
-\int
\left(\norm{\nabla\phi_t}^2+\frac{1}{\kappa^2}\phi_t^2\right)\rho_t\,\d x,
\]
which is~\eqref{eq-wfr-energy-dissipation}.
\end{proof}
\subsection{Geodesic Convexity and Convergence}
\label{sec-geodesic-convexity}
\paragraph{Geodesics and convexity.}
\(\alpha_t=((1-t)P_0+tP_1)_\sharp\pi^\star, \qquad t\in[0,1],\)
where $P_0(x,y)=x$ and $P_1(x,y)=y$. If the optimal plan is induced by a Brenier map $T$, this reduces to $((1-t)\Id+tT)_\sharp\alpha_0$.

\begin{defn}[Geodesic convexity]\label{def-geodesic-convexity}
A functional $f$ on $\Pp_2(\RR^d)$ is geodesically convex if for every $\Wass_2$ geodesic $(\alpha_t)_t$, \(f(\alpha_t)\leq (1-t)f(\alpha_0)+t f(\alpha_1).\)
It is $\lambda$-geodesically convex if the right-hand side is improved by $-\frac{\lambda}{2}t(1-t)\Wass_2^2(\alpha_0,\alpha_1)$.
\end{defn}
\begin{prop}[Basic geodesically convex energies]\phantomsection\label{prop-basic-geodesic-convexity}
\begin{enumerate}
\item If $h$ is convex, then $\alpha\mapsto\int h\d\alpha$ is geodesically convex; if $h$ is $\lambda$-strongly convex, it is $\lambda$-geodesically convex.
\item If $W(x-y)$ is convex as a function of the displacement, then $\alpha\mapsto\frac12\iint W(x-y)\d\alpha(x)\d\alpha(y)$ is geodesically convex.
\item Shannon entropy $\alpha\mapsto\int\rho\log\rho\,\d x$ is geodesically convex.
\item The relative entropy $\KL(\alpha|\gamma)$ with $\d\gamma=e^{-V}\d x/Z$ is $\lambda$-geodesically convex when $V$ is $\lambda$-strongly convex.
\end{enumerate}
\end{prop}
\begin{proof}
Along a Monge geodesic $X_t=(1-t)X_0+tX_1$, convexity of $h$ gives $h(X_t)\leq(1-t)h(X_0)+t h(X_1)$, and strong convexity gives the additional quadratic term; integrating proves the first claim. The interaction claim follows similarly by applying convexity of $W$ to pairwise differences $X_t-X_t'=(1-t)(X_0-X_0')+t(X_1-X_1')$ and integrating over two independent copies. The entropy claim is McCann's displacement convexity theorem; at the density level it follows from the concavity of the Jacobian determinant under the interpolation of optimal maps. Finally, $\KL(\alpha|\gamma)=\int\rho\log\rho\,\d x+\int V\d\alpha+\mathrm{constant}$, so it is the sum of displacement-convex entropy and a $\lambda$-geodesically convex linear potential.
\end{proof}
\paragraph{A useful non-example.}
The squared distance to a fixed measure is not geodesically convex in general. This is a sharp contrast with Hilbert spaces, where \(x\mapsto\norm{x-y}^2\) is convex along line segments.

\[
\beta=\frac12\de_{(-1/2,-1/2)}+\frac12\de_{(1/2,1/2)},
\quad
\alpha_0=\frac12\de_{(-1,0)}+\frac12\de_{(1,0)},
\quad
\alpha_1=\frac12\de_{(0,-1)}+\frac12\de_{(0,1)}.
\]
\(\alpha_{1/2}=\frac12\de_{(-1/2,1/2)}+\frac12\de_{(1/2,-1/2)}.\)
\(\Wass_2^2(\alpha_0,\beta) = \Wass_2^2(\alpha_1,\beta) = \frac12, \qquad \Wass_2^2(\alpha_{1/2},\beta) = 1,\)
which violates geodesic convexity. Thus objectives built from squared Wasserstein distances can be non-convex in the Wasserstein geometry itself.

\paragraph{Convergence of the flow.}
\begin{prop}[Energy decay for convex Wasserstein flows]\label{prop-convex-wass-flow-rate}
Assume formally that $f$ is geodesically convex, admits a smooth first variation, and has a minimizer $\alpha^\star$. Let $(\alpha_t)_t$ be a smooth solution of the Wasserstein gradient flow
\(\partial_t\alpha_t+\diverg(\alpha_t v_t)=0, \qquad v_t=-\Wgrad f(\alpha_t).\)
Then
\(\frac{\d}{\d t} f(\alpha_t) = -\int \norm{\Wgrad f(\alpha_t)(x)}^2\,\d\alpha_t(x) \leq 0.\)
If $T_t$ is the optimal map from $\alpha_t$ to $\alpha^\star$, then
\(f(\alpha_t)-f(\alpha^\star) \leq -\frac{\d}{\d t}\frac12 \Wass_2^2(\alpha_t,\alpha^\star),\)
and consequently
\(f(\alpha_t)-f(\alpha^\star) \leq \frac{\Wass_2^2(\alpha_0,\alpha^\star)}{2t}.\)
If $f$ is $\lambda$-geodesically convex with $\lambda>0$, then
\(f(\alpha_t)-f(\alpha^\star) \leq e^{-2\lambda t}\bigl(f(\alpha_0)-f(\alpha^\star)\bigr).\)
\end{prop}
\begin{proof}
The chain rule and Proposition~\ref{prop-formal-wass-gradient} give
\(\frac{\d}{\d t}f(\alpha_t) = \int \dotp{\Wgrad f(\alpha_t)(x)}{v_t(x)}\,\d\alpha_t(x) = -\int \norm{\Wgrad f(\alpha_t)(x)}^2\,\d\alpha_t(x).\)
Geodesic convexity along the geodesic
$((1-s)\Id+sT_t)_\sharp\alpha_t$ gives \(f(\alpha^\star)-f(\alpha_t) \geq \int \dotp{\Wgrad f(\alpha_t)(x)}{T_t(x)-x}\,\d\alpha_t(x).\)
Since $v_t=-\Wgrad f(\alpha_t)$, this reads \(f(\alpha_t)-f(\alpha^\star) \leq \int \dotp{v_t(x)}{T_t(x)-x}\,\d\alpha_t(x).\)
\(\frac{\d}{\d t}\frac12\Wass_2^2(\alpha_t,\alpha^\star) = \int \dotp{x-T_t(x)}{v_t(x)}\,\d\alpha_t(x),\)
which proves the differential inequality. Integrating it from $0$ to $t$ and using the monotonicity of $s\mapsto f(\alpha_s)$ gives
\[
t\bigl(f(\alpha_t)-f(\alpha^\star)\bigr)
\leq
\int_0^t \bigl(f(\alpha_s)-f(\alpha^\star)\bigr)\,\d s
\leq
\frac12\Wass_2^2(\alpha_0,\alpha^\star).
\]
If $f$ is $\lambda$-geodesically convex, the Wasserstein analogue of strong convexity gives the slope inequality
\(\int\norm{\Wgrad f(\alpha_t)}^2\d\alpha_t \geq 2\lambda\bigl(f(\alpha_t)-f(\alpha^\star)\bigr).\)
\(\frac{\d}{\d t}\bigl(f(\alpha_t)-f(\alpha^\star)\bigr) \leq -2\lambda\bigl(f(\alpha_t)-f(\alpha^\star)\bigr),\)
and Gronwall's lemma gives the exponential rate.
\end{proof}
\begin{prop}[Convex examples covered by the theory]\label{prop-convex-flow-examples}
The hypotheses of Proposition~\ref{prop-convex-wass-flow-rate} are satisfied in the following standard cases, at least at the formal smooth level used in this section.
\begin{enumerate}
\item For the linear energy $f(\alpha)=\int h\d\alpha$, geodesic convexity holds when $h$ is convex. If $h$ is $\lambda$-strongly convex, then $f$ is $\lambda$-geodesically convex and the flow enjoys the exponential rate of Proposition~\ref{prop-convex-wass-flow-rate}.
\item For the interaction energy $f(\alpha)=\frac12\iint W(x-y)\d\alpha(x)\d\alpha(y)$, geodesic convexity holds when $W$ is convex and even. This covers repulsive or attractive pairwise models whose displacement cost has no non-convex wells.
\item The Shannon entropy $f(\alpha)=\int\rho\log\rho\,\d x$ and, more generally, McCann displacement-convex internal energies generate diffusion-type Wasserstein gradient flows.
\item If $\gamma=Z^{-1}e^{-V}\d x$ and $V$ is $\lambda$-strongly convex, then the relative entropy $\KL(\alpha|\gamma)$ is $\lambda$-geodesically convex. Its flow is the Fokker--Planck equation with invariant law $\gamma$.
\end{enumerate}
\end{prop}
\begin{proof}
Let $(\alpha_t)_t$ be the McCann interpolation between $\alpha_0$ and $\alpha_1$, written with an optimal coupling as $X_t=(1-t)X_0+tX_1$. For a linear energy, Jensen's inequality gives \(h(X_t)\leq(1-t)h(X_0)+t h(X_1),\)
and the strong convexity version gives the additional term $-\frac{\lambda}{2}t(1-t)\norm{X_0-X_1}^2$. Integrating over the optimal coupling proves geodesic convexity and $\lambda$-geodesic convexity.

For interaction energies, use two independent copies of the optimal coupling. The pairwise displacement evolves as \(X_t-X_t'=(1-t)(X_0-X_0')+t(X_1-X_1').\)
Convexity of $W$ gives the convexity inequality after integration over the product coupling. Evenness of $W$ ensures that the interaction is symmetric in the two particles and matches the usual factor $1/2$ in~\eqref{eq:quadratic-func}.

The entropy claim is McCann's displacement-convexity theorem. For smooth positive densities and Brenier maps, it follows from the change-of-variables formula and the concavity of the determinant along positive matrices; the general statement is obtained by approximation. Finally,
\(\KL(\alpha|\gamma)=\int\rho\log\rho\,\d x+\int V\d\alpha+\log Z,\)
so it is the sum of the displacement-convex entropy and the $\lambda$-geodesically convex linear potential generated by $V$. Proposition~\ref{prop-convex-wass-flow-rate} then applies to all four cases.
\end{proof}
\paragraph{Convexity and curvature.}
The same language is not restricted to subsets of $\RR^d$. If $(\X,\dist,\mathfrak m)$ is a geodesic metric-measure space, $\Wass_2$ geodesics can be defined by transporting each pair of endpoints along metric geodesics, or more intrinsically by dynamical optimal plans on path space, as discussed in Section~\ref{sec-kantorovich-plan-interpolation}.

\[
\mathrm{Ent}_{\mathfrak m}(\alpha)
\eqdef
\begin{cases}
\displaystyle \int_\X \rho\log\rho\,\d\mathfrak m,
& \text{if } \alpha=\rho\,\mathfrak m,\\
+\infty,
& \text{otherwise.}
\end{cases}
\]
\begin{thm}[Ricci curvature and entropy convexity]\label{thm-ricci-entropy-convexity}
Let $(M,g)$ be a smooth compact connected Riemannian manifold without boundary, and let $\mathfrak m=\mathrm{vol}_g$. For $\lambda\in\RR$, the lower Ricci bound $\mathrm{Ric}_g\geq\lambda g$ holds if and only if $\mathrm{Ent}_{\mathfrak m}$ is $\lambda$-geodesically convex on $(\Pp_2(M),\Wass_2)$.
\end{thm}
\subsection{Training Two-Layer MLPs as Wasserstein Flows}
\label{sec-wasserstein-flows-mlp}
\(x=(u,v)\in\RR^d\times\RR^{d'},\)
where $u$ is the inner weight and $v$ is the outer vector weight. For a scalar nonlinearity $\sigma$, define the vector-valued feature

\(\psi(x,z)=v\,\sigma(\dotp{u}{z})\in\RR^{d'}.\)
\(G_X(z)=\frac1n\sum_{i=1}^n\psi(x_i,z), \qquad G_\alpha(z)=\int\psi(x,z)\d\alpha(x), \qquad \alpha=\frac1n\sum_i\delta_{x_i}.\)
Let $\rho$ be a probability distribution on data-label pairs $(z,y)\in\RR^d\times\RR^{d'}$. The population risk is

\(f(\alpha)=\int \ell(G_\alpha(z),y)\d\rho(z,y),\)
\[
\dot x_i=-n\nabla_{x_i}F(X),
\qquad
F(X)=f\!\left(\frac1n\sum_i\delta_{x_i}\right),
\]
which is the gradient flow of $F(X)=f(\alpha_X)$ for this Wasserstein particle metric, equivalently Euclidean gradient descent with the time scale multiplied by $n$. It gives a particle discretization of~\eqref{eq:wassflow-pde}.

Assume that $\ell$ is differentiable in its first variable. The first variation is

\begin{equation}\label{eq-mlp-first-variation-general}
\delta f(\alpha)(x)
=
\int
\dotp{\nabla_1\ell(G_\alpha(z),y)}{\psi(x,z)}
\d\rho(z,y),
\end{equation}
\(\Wgrad f(\alpha)(x) = \nabla_x\delta f(\alpha)(x) = \int [D_x\psi(x,z)]^\top\nabla_1\ell(G_\alpha(z),y) \d\rho(z,y).\)
For the squared Euclidean loss $\ell(s,y)=\frac12\norm{s-y}^2$, the energy is the sum of a quadratic interaction and a linear potential:

\begin{equation}\label{eq-mlp-square-loss-quadratic-linear}
f(\alpha)
=
\frac12\iint k(x,x')\d\alpha(x)\d\alpha(x')
+
\int g(x)\d\alpha(x)
+\frac12\int\norm{y}^2\d\rho(z,y),
\end{equation}
\begin{equation}\label{eq-mlp-kernel-linear-potential}
k(x,x')=\int\dotp{\psi(x,z)}{\psi(x',z)}\d\rho(z,y),
\qquad
g(x)=-\int\dotp{y}{\psi(x,z)}\d\rho(z,y).
\end{equation}
\(\delta f(\alpha)(x)=\int k(x,x')\d\alpha(x')+g(x), \qquad \Wgrad f(\alpha)(x)=\int\nabla_x k(x,x')\d\alpha(x')+\nabla_xg(x).\)
\paragraph{Classical convexity and stationarity.}
\(F(\alpha) = \frac12\iint k(x,x')\d\alpha(x)\d\alpha(x') + \int V(x)\d\alpha(x) +C,\)
\(Q((1-s)\alpha+s\beta)\leq (1-s)Q(\alpha)+sQ(\beta), \qquad Q(\alpha)=\frac12\iint k\d\alpha\d\alpha.\)
\begin{prop}[Affine convexity and stationary densities]\label{prop-classical-convex-stationary}
Let $F=Q+\int V\,\d\alpha+C$ be as above, and assume that $Q$ is classically convex. Suppose that a Wasserstein gradient flow for $F$ converges to a measure $\alpha_\infty=\rho_\infty\d x$. Assume also the standard regularity needed to pass to the limit in the first variation, and assume that the support and positivity of $\rho_\infty$ allow the stationary condition to be tested against all admissible zero-mass density perturbations. In the form needed here, assume that this stationarity yields, for every competitor $\beta$, the variational inequality
\(\int \delta F(\alpha_\infty)(x)\d(\beta-\alpha_\infty)(x)\geq0.\)
Then $\alpha_\infty$ is a global minimizer of $F$.
\end{prop}
\begin{proof}[Proof sketch]
The dissipation identity for the gradient flow gives stationarity of the limit: formally, after passing to the limit, \(\int \norm{\nabla \delta F(\alpha_\infty)}^2\d\alpha_\infty=0.\)
Without such a support and positivity assumption, this identity only controls the first variation on the region explored by the limit. The density hypothesis allows one to test against sufficiently many signed density perturbations of total mass zero. By approximation and the assumed regularity, this yields the displayed first-order variational inequality for arbitrary competitors $\beta$. Classical convexity of $F$ in the affine variable $\alpha$ then gives the usual subgradient inequality
\(F(\beta)\geq F(\alpha_\infty) + \int \delta F(\alpha_\infty)\d(\beta-\alpha_\infty) \geq F(\alpha_\infty).\)
Thus no competitor has smaller energy. For square-loss two-layer mean-field models, \eqref{eq-mlp-square-loss-quadratic-linear} is exactly of this quadratic-plus-linear form, and positive semidefiniteness of the induced kernel $k$ is the classical convexity assumption.
\end{proof}
\begin{prop}[Formal global optimality for two-homogeneous mean-field flows]\label{prop-formal-chizat-bach}
Assume that the feature is positively two-homogeneous in the neuron variable, \(\psi(\lambda x,z)=\lambda^2\psi(x,z) \qquad(\lambda>0),\)
and that $f(\alpha)=J(G_\alpha)$ with $J$ convex and differentiable as a functional of the predictor. Let $\alpha$ be a smooth stationary point of the Wasserstein flow, so that $\nabla_x\delta f(\alpha)(x)=0$ on $\supp(\alpha)$. Assume also full directional support: for every nonzero direction $\omega$, the support of $\alpha$ intersects the ray $\{\lambda\omega:\lambda>0\}$. Then $\alpha$ is a global minimizer of $f$ over the mean-field model class.
\end{prop}
\begin{proof}
Write
\[
h_\alpha(x)=\delta f(\alpha)(x)
=
\left\langle \nabla J(G_\alpha),\psi(x,\cdot)\right\rangle_\rho.
\]
By two-homogeneity of $\psi$, one has $h_\alpha(\lambda x)=\lambda^2h_\alpha(x)$. Normalize a nonzero direction $\omega$ and choose $r_\omega>0$ with $r_\omega\omega\in\supp(\alpha)$. Stationarity gives a zero radial derivative at this point:
\(0=\frac{\d}{\d r}h_\alpha(r\omega)\bigg|_{r=r_\omega} =2r_\omega h_\alpha(\omega).\)
Hence $h_\alpha(\omega)=0$ for every direction $\omega$, and by homogeneity $h_\alpha(x)=0$ for every $x$.

For any competitor $\beta$, convexity of $J$ gives \(f(\beta)-f(\alpha) \geq \int h_\alpha(x)\d(\beta-\alpha)(x)=0.\)
Thus no competitor has smaller risk. The rigorous theorem replaces the full directional support assumption by propagation and overparameterization hypotheses ensuring that a negative descent direction would be present in the support and would contradict stationarity.
\end{proof}
\subsection{Functional Inequalities via Optimal Transport}
\paragraph{Brunn--Minkowski and isoperimetry.}
\begin{prop}[Brunn--Minkowski and Euclidean isoperimetry]\label{prop-ot-brunn-minkowski-isoperimetric}
Let $A,B\subset\RR^d$ be bounded Borel sets with positive Lebesgue measure. For $t\in[0,1]$, set
\((1-t)A+tB\eqdef\{(1-t)x+ty\;:\;x\in A,\ y\in B\}.\)
Then
\(\mathrm{Vol}((1-t)A+tB)^{1/d} \geq (1-t)\mathrm{Vol}(A)^{1/d} + t\,\mathrm{Vol}(B)^{1/d}.\)
\(\mathrm{Per}(A) \geq d\,\omega_d^{1/d}\mathrm{Vol}(A)^{(d-1)/d}.\)
\end{prop}
\begin{proof}
We first give the argument for regular sets, the general Borel case following by approximation from inside and outside. Let $\alpha$ and $\beta$ be the uniform probability measures on $A$ and $B$, and let $T=\nabla\phi$ be the Brenier map with $T_\sharp\alpha=\beta$. The interpolating map $T_t=(1-t)\Id+tT$ sends $\alpha$ to a measure $\alpha_t$ supported in $(1-t)A+tB$. At points where $T$ is differentiable,
\(\det DT_t = \det((1-t)\Id+tDT).\)
Since $DT$ is symmetric positive semidefinite and $\det^{1/d}$ is concave on the positive semidefinite cone, \(\det(DT_t)^{1/d} \geq (1-t)+t\det(DT)^{1/d}.\)
The change-of-variables identity for the uniform measures gives $\det(DT)=\mathrm{Vol}(B)/\mathrm{Vol}(A)$ almost everywhere. Hence the density $\rho_t$ of $\alpha_t$ satisfies
\(\rho_t^{-1/d} \geq (1-t)\mathrm{Vol}(A)^{1/d} + t\mathrm{Vol}(B)^{1/d}.\)
This means that $\rho_t\leq c_t^{-d}$ on its support, where $c_t=(1-t)\mathrm{Vol}(A)^{1/d}+t\mathrm{Vol}(B)^{1/d}$. Since $\alpha_t$ has total mass one and is supported in $(1-t)A+tB$, this gives $\mathrm{Vol}((1-t)A+tB)\geq c_t^d$, which is the displayed Brunn--Minkowski inequality.

For isoperimetry, apply Brunn--Minkowski with $B=B_1$ and write $A+\epsilon B_1$ for the parallel set. Then
\(\mathrm{Vol}(A+\epsilon B_1)^{1/d} \geq \mathrm{Vol}(A)^{1/d}+\epsilon\omega_d^{1/d}.\)
Using $\mathrm{Vol}(A+\epsilon B_1)=\mathrm{Vol}(A)+\epsilon\,\mathrm{Per}(A)+o(\epsilon)$ and differentiating at $\epsilon=0$ yields the claimed perimeter bound.
\end{proof}
\paragraph{Pr\'ekopa--Leindler.}
\begin{prop}[Pr\'ekopa--Leindler inequality]\label{prop-ot-prekopa-leindler}
Let $u,v,w:\RR^d\to[0,+\infty)$ be integrable, and let $t\in[0,1]$. Assume that
\(w((1-t)x+ty)\geq u(x)^{1-t}v(y)^t \qquad \text{for all }x,y\in\RR^d.\)
Then
\[
\int_{\RR^d} w
\geq
\left(\int_{\RR^d}u\right)^{1-t}
\left(\int_{\RR^d}v\right)^t.
\]
\end{prop}
\begin{proof}
Set $U=\int u$ and $V=\int v$, and assume $U,V>0$. Let $\alpha=(u/U)\d x$ and $\beta=(v/V)\d x$, and let $T$ be the Brenier map from $\alpha$ to $\beta$. Put $T_t=(1-t)\Id+tT$. The change of variables gives
\(\int w \geq \int w(T_t(x))\det(DT_t(x))\d x.\)
By the assumption on $w$ and by $\det((1-t)\Id+tDT)\geq\det(DT)^t$ for positive semidefinite $DT$, \(\int w \geq \int u(x)^{1-t}v(T(x))^t\det(DT(x))^t\d x.\)
Since $T_\sharp\alpha=\beta$, the Monge--Amp\`ere identity gives \(\det(DT(x))=\frac{u(x)V}{v(T(x))U}\)
almost everywhere. Substitution yields
\[
\int w
\geq
\left(\frac{V}{U}\right)^t\int u(x)\d x
=
U^{1-t}V^t.
\]
Approximation removes the smoothness and positivity assumptions.
\end{proof}
\paragraph{Gaussian logarithmic Sobolev inequality.}
\begin{prop}[Gaussian logarithmic Sobolev inequality]\label{prop-gaussian-log-sobolev-ot}
Let $\gamma_d$ be the standard Gaussian measure on $\RR^d$. If $f$ is smooth, positive, has sufficient decay, and $\int f\d\gamma_d=1$, then
\(\mathrm{Ent}_{\gamma_d}(f) \eqdef \int f\log f\,\d\gamma_d \leq \frac12\int \frac{\norm{\nabla f}^2}{f}\,\d\gamma_d.\)
\end{prop}
\begin{proof}
Let $\mu=f\gamma_d$, and let $T=\nabla\phi$ be the Brenier map with $T_\sharp\mu=\gamma_d$. Write $T(x)=x+\theta(x)$. The Monge--Amp\`ere equation reads \(f(x)e^{-\norm{x}^2/2} = e^{-\norm{T(x)}^2/2}\det(DT(x)).\)
Thus
\(\log f(x) = -\dotp{x}{\theta(x)} -\frac12\norm{\theta(x)}^2 + \log\det(\Id+D\theta(x)).\)
Since $DT$ is positive semidefinite, $\log\det(\Id+D\theta)\leq\tr(D\theta)=\diverg\theta$. Multiplying by $f$ and integrating with respect to $\gamma_d$ gives
\(\mathrm{Ent}_{\gamma_d}(f) \leq \int f(\diverg\theta-\dotp{x}{\theta})\,\d\gamma_d -\frac12\int f\norm{\theta}^2\,\d\gamma_d.\)
Gaussian integration by parts gives \(\int f(\diverg\theta-\dotp{x}{\theta})\,\d\gamma_d = -\int\dotp{\nabla f}{\theta}\,\d\gamma_d.\)
Completing the square pointwise, \(-\dotp{\nabla f}{\theta}-\frac12 f\norm{\theta}^2 \leq \frac12\frac{\norm{\nabla f}^2}{f}.\)
This proves the inequality. The general Sobolev-class statement follows by approximation.
\end{proof}
\paragraph{Talagrand's transport-entropy inequality.}
\begin{prop}[Gaussian \(T_2\) inequality]\label{prop-gaussian-talagrand-t2}
Let $\mu=f\gamma_d$ be a probability measure with finite second moment. Then
\(\frac12\Wass_2^2(\mu,\gamma_d) \leq \KL(\mu|\gamma_d) = \int f\log f\,\d\gamma_d.\)
\end{prop}
\begin{proof}
Assume first that $f$ is smooth and positive. Let $T=\nabla\phi$ be the Brenier map with $T_\sharp\gamma_d=\mu$, and write $T(x)=x+\theta(x)$. The change of variables gives
\(f(T(x))e^{-\norm{T(x)}^2/2}\det(DT(x)) = e^{-\norm{x}^2/2}.\)
Hence
\(\log f(T(x)) = \dotp{x}{\theta(x)} + \frac12\norm{\theta(x)}^2 - \log\det(\Id+D\theta(x)).\)
Using again $\log\det(\Id+D\theta)\leq\diverg\theta$ and integrating with respect to $\gamma_d$, \(\KL(\mu|\gamma_d) \geq \frac12\int\norm{\theta}^2\,\d\gamma_d + \int(\dotp{x}{\theta}-\diverg\theta)\,\d\gamma_d.\)
The last integral vanishes by Gaussian integration by parts. Since $T$ is optimal, \(\int\norm{\theta}^2\,\d\gamma_d=\Wass_2^2(\gamma_d,\mu),\)
which proves the claim. Approximation gives the general case.
\end{proof}
\paragraph{The HWI bridge.}
\begin{prop}[Otto--Villani HWI inequality]\label{prop-hwi-inequality}
Let $\beta=\rho_\beta\d x=e^{-V}\d x/Z$ be a smooth probability measure in $\Pp_2(\RR^d)$, and assume that $\nabla^2 V\succeq \lambda \Id$ for some $\lambda\in\RR$. For an absolutely continuous probability measure $\alpha=\rho_\alpha\d x$ in $\Pp_2(\RR^d)$, define the relative Fisher information
\(\mathcal I_\beta(\alpha) \eqdef \int \norm{\nabla\log\frac{\rho_\alpha}{\rho_\beta}}^2\,\d\alpha,\)
with the convention $\mathcal I_\beta(\alpha)=+\infty$ if $\alpha$ is not absolutely continuous or if the expression is not defined. Then
\(\KL(\alpha|\beta) \leq \Wass_2(\alpha,\beta)\sqrt{\mathcal I_\beta(\alpha)} -\frac{\lambda}{2}\Wass_2^2(\alpha,\beta).\)
\end{prop}
\begin{proof}
If $\mathcal I_\beta(\alpha)=+\infty$, there is nothing to prove. Assume first that $\rho_\alpha$ is smooth and positive. Let $T$ be the Brenier map from $\alpha$ to $\beta$, set $v=T-\Id$, and let $\alpha_t=((1-t)\Id+tT)_\sharp\alpha$. By Proposition~\ref{prop-basic-geodesic-convexity}, $F(\eta)=\KL(\eta|\beta)$ is $\lambda$-geodesically convex. Thus the one-dimensional function $f(t)=F(\alpha_t)$ satisfies
\(f(1)\geq f(0)+f'(0)+\frac{\lambda}{2}\Wass_2^2(\alpha,\beta).\)
Since $f(1)=F(\beta)=0$, it remains to compute the initial derivative. The continuity equation for the geodesic gives $\partial_t\alpha_t+\nabla\cdot(\alpha_t v_t)=0$ and $v_0=v$. Hence
\(f'(0) = \int \dotp{\nabla\log\frac{\rho_\alpha}{\rho_\beta}}{v}\,\d\alpha.\)
Therefore
\(\KL(\alpha|\beta) \leq -\int \dotp{\nabla\log\frac{\rho_\alpha}{\rho_\beta}}{T-\Id}\,\d\alpha -\frac{\lambda}{2}\Wass_2^2(\alpha,\beta).\)
Cauchy's inequality and $\int\norm{T-\Id}^2\d\alpha=\Wass_2^2(\alpha,\beta)$ give the announced bound. The general case follows by the standard approximation and lower-semicontinuity of entropy, Fisher information, and $\Wass_2$.
\end{proof}
When $\lambda>0$, Young's inequality $rs-\lambda r^2/2\leq s^2/(2\lambda)$ gives the logarithmic Sobolev estimate $\KL(\alpha|\beta)\leq \mathcal I_\beta(\alpha)/(2\lambda)$. Thus HWI explains why the same curvature assumption controls both the static inequality and the exponential decay of the associated Fokker--Planck flow below.

\paragraph{Back to gradient flows.}
For the standard Gaussian $\gamma=\mathcal N(0,\Id)$, Proposition~\ref{prop-gaussian-log-sobolev-ot} gives $\lambda=1$.

\begin{prop}[Log-Sobolev inequality implies entropy decay]\label{prop-lsi-entropy-decay}
Let $\beta=e^{-V}\d x/Z$ be a smooth probability measure on $\RR^d$. Assume that $\beta$ satisfies the logarithmic Sobolev inequality with constant $\lambda>0$,
\(\KL(\alpha|\beta) \leq \frac{1}{2\lambda} \int \norm{\nabla\log\frac{\d\alpha}{\d\beta}}^2\,\d\alpha.\)
Let $\alpha_t$ solve the Wasserstein gradient flow of $F(\alpha)=\KL(\alpha|\beta)$, namely
\(\partial_t\rho_t = \nabla\cdot(\rho_t\nabla V)+\Delta\rho_t, \qquad \alpha_t=\rho_t\d x.\)
Then
\(\KL(\alpha_t|\beta) \leq e^{-2\lambda t}\KL(\alpha_0|\beta).\)
\end{prop}
\begin{proof}
The first variation of $F$ is $\log(\rho/\rho_\beta)$ up to an additive constant. Along the smooth Wasserstein gradient flow,
\(\frac{\d}{\d t}\KL(\alpha_t|\beta) = -\int \norm{\nabla\log\frac{\d\alpha_t}{\d\beta}}^2\,\d\alpha_t.\)
The logarithmic Sobolev inequality bounds the Fisher information below by $2\lambda\KL(\alpha_t|\beta)$. Hence
\(\frac{\d}{\d t}\KL(\alpha_t|\beta) \leq -2\lambda\KL(\alpha_t|\beta),\)
and Gronwall's lemma gives the result.
\end{proof}
\subsection{Second-Order Momentum Flows}
\label{sec-second-order-momentum-flows}
\paragraph{Finite-dimensional momentum.}
Let $F:\RR^{N}\to\RR$ be a smooth finite-dimensional objective. With step size $h>0$ and damping $\gamma>0$, one convenient velocity form is

\begin{equation}\label{eq-polyak-momentum}
S^{k+1}=(1-\gamma h)S^k-h\nabla F(X^k),
\qquad
X^{k+1}=X^k+hS^{k+1}.
\end{equation}
If $S^k$ approximates $\dot X(kh)$, this is the explicit Euler discretization of the first-order system

\begin{equation}\label{eq-heavy-ball-first-order-system}
\dot X(t)=S(t),
\qquad
\dot S(t)=-\gamma S(t)-\nabla F(X(t)),
\end{equation}
\begin{equation}\label{eq-heavy-ball-ode}
\ddot X(t)+\gamma\dot X(t)+\nabla F(X(t))=0.
\end{equation}
\begin{equation}\label{eq-nesterov-momentum}
Y^k=X^k+\theta_k(X^k-X^{k-1}),
\qquad
X^{k+1}=Y^k-h\nabla F(Y^k),
\end{equation}
\begin{equation}\label{eq-nesterov-ode}
\ddot X(t)+\frac{r}{t}\dot X(t)+\nabla F(X(t))=0.
\end{equation}
\paragraph{Empirical Wasserstein lift.}
\begin{equation}\label{eq-empirical-momentum-lift}
\mu_X \eqdef \frac1n\sum_{i=1}^n \delta_{x_i},
\qquad
F(X) \eqdef f(\mu_X).
\end{equation}
\(\norm{\dot X}_n^2 \eqdef \frac1n\sum_{i=1}^n\norm{\dot x_i}^2,\)
\[
\left.\frac{\d}{\d\varepsilon}\right|_{\varepsilon=0}
F(x_1,\ldots,x_i+\varepsilon\xi,\ldots,x_n)
=
\frac1n\dotp{\nabla\delta f(\mu_X)(x_i)}{\xi}.
\]
Thus, if $\nabla^{(n)}F$ denotes the gradient for the metric $\norm{\cdot}_n$, then

\begin{equation}\label{eq-empirical-lift-gradient}
(\nabla^{(n)}F(X))_i
=
n\nabla_{x_i}F(X)
=
\nabla\delta f(\mu_X)(x_i)
=
\Wgrad f(\mu_X)(x_i).
\end{equation}
\begin{equation}\label{eq-momentum-wasserstein-velocity}
v_\mu(x)\eqdef -\Wgrad f(\mu)(x)=-\nabla\delta f(\mu)(x).
\end{equation}
\begin{equation}\label{eq-wasserstein-momentum-particles}
\dot x_i(t)=s_i(t),
\qquad
\dot s_i(t)=-\gamma s_i(t)+v_{\mu_t}(x_i(t)),
\qquad
\mu_t=\frac1n\sum_{i=1}^n\delta_{x_i(t)}.
\end{equation}
\(\ddot x_i(t)+\gamma\dot x_i(t) = v_{\mu_t}(x_i(t)) = -\Wgrad f(\mu_t)(x_i(t)).\)
\begin{prop}[Discrete mechanical energy dissipation]\label{prop-second-order-energy-dissipation}
Assume that $f$ and the particle solution are smooth enough to justify differentiating the identities above along \eqref{eq-wasserstein-momentum-particles}. Define
\(\mathcal H_n(X,S) \eqdef F(X)+\frac1{2n}\sum_{i=1}^n\norm{s_i}^2, \qquad S=(s_1,\ldots,s_n).\)
Then along \eqref{eq-wasserstein-momentum-particles},
\begin{equation}\label{eq-second-order-energy-dissipation}
\frac{\d}{\d t}\mathcal H_n(X(t),S(t))
=
-\frac{\gamma}{n}\sum_{i=1}^n\norm{s_i(t)}^2
\leq 0.
\end{equation}
\end{prop}
\begin{proof}
By \eqref{eq-empirical-lift-gradient} and the definition $v_\mu=-\Wgrad f(\mu)$, one has $\nabla_{x_i}F(X)=-v_{\mu_X}(x_i)/n$. Hence
\(\frac{\d}{\d t}F(X(t)) = \sum_{i=1}^n \dotp{\nabla_{x_i}F(X(t))}{s_i(t)} = -\frac1n\sum_{i=1}^n\dotp{v_{\mu_t}(x_i(t))}{s_i(t)}.\)
On the other hand, \(\frac{\d}{\d t}\frac1{2n}\sum_i\norm{s_i(t)}^2 = \frac1n\sum_i\dotp{s_i(t)}{-\gamma s_i(t)+v_{\mu_t}(x_i(t))}.\)
The force terms cancel, leaving \eqref{eq-second-order-energy-dissipation}.
\end{proof}
\paragraph{Phase-space formulation.}
\begin{prop}[Phase-space Liouville formulation]\label{prop-second-order-liouville}
Let $v_\mu$ be smooth enough that \eqref{eq-wasserstein-momentum-particles} has a classical solution. Define the empirical phase-space measure
\(\eta_t^n \eqdef \frac1n\sum_{i=1}^n\delta_{(x_i(t),s_i(t))} \in \Pp(\RR^d_x\times\RR^d_s), \qquad \mu_t^n=(\pi_x)_\sharp\eta_t^n.\)
Then $\eta_t^n$ solves, in the sense of distributions,
\begin{equation}\label{eq-second-order-liouville-empirical}
\partial_t\eta_t^n
+
\nabla_x\cdot(s\,\eta_t^n)
+
\nabla_s\cdot\bigl((-\gamma s+v_{\mu_t^n}(x))\,\eta_t^n\bigr)
=0.
\end{equation}
If, as $n\to\infty$, the empirical phase-space laws converge to a smooth density $\eta_t(x,s)$ and the nonlinear fields converge accordingly, the limiting kinetic equation is
\begin{equation}\label{eq-second-order-liouville-density}
\partial_t\eta_t
+
\nabla_x\cdot(s\,\eta_t)
+
\nabla_s\cdot\bigl((-\gamma s+v_{\mu_t}(x))\,\eta_t\bigr)
=0,
\qquad
\mu_t=(\pi_x)_\sharp\eta_t.
\end{equation}
\end{prop}
\begin{proof}
Let $\varphi\in C_c^\infty(\RR^d_x\times\RR^d_s)$. Along the particle system,
\[
\frac{\d}{\d t}
\int \varphi\,\d\eta_t^n
=
\frac1n\sum_{i=1}^n
\left[
\dotp{\nabla_x\varphi(x_i,s_i)}{s_i}
+
\dotp{\nabla_s\varphi(x_i,s_i)}{-\gamma s_i+v_{\mu_t^n}(x_i)}
\right].
\]
Rewriting the right-hand side as an integral against $\eta_t^n$ and integrating by parts gives \eqref{eq-second-order-liouville-empirical}. Passing formally to the mean-field limit gives \eqref{eq-second-order-liouville-density}.
\end{proof}
\(k(x,y)=-\norm{x-y}, \qquad f(\mu)=\MMD_k^2(\mu,\beta) = \iint k(x,y)\,\d(\mu-\beta)(x)\d(\mu-\beta)(y).\)
\(v_\mu(x) = 2\int \frac{x-y}{\norm{x-y}}\,\d(\mu-\beta)(y).\)
\paragraph{Entropy without an empirical energy.}
For the numerical illustration, the empirical density is replaced by the Gaussian kernel estimator $\widehat\rho_h(x)=n^{-1}\sum_i K_h(x-x_i)$ with a relatively small bandwidth $h$. Its score is evaluated directly at the particle locations by the closed formula

\[
\nabla\log\widehat\rho_h(x_i)
=
\frac{\sum_j \exp(-\norm{x_i-x_j}^2/(2h^2))(x_j-x_i)}{h^2\sum_j \exp(-\norm{x_i-x_j}^2/(2h^2))},
\]
\subsection{Normalized Spectral Wasserstein Dynamics}
\label{sec-normalized-spectral-wasserstein-dynamics}
\paragraph{Spectral tangent norm.}
Let $\gamma$ be a monotone spectral gauge on $\mathbb S_+^d$, with polar set $\mathcal B_\gamma$ as in \eqref{eq-spectral-polar-set}. For a probability measure $\alpha$ and a velocity field $v\in L^2(\alpha;\RR^d)$, define the spectral tangent norm

\begin{equation}\label{eq-normalized-spectral-tangent-norm}
\mathcal N_{\gamma,\alpha}(v)^2
\eqdef
\gamma\!\left(\int v(x)v(x)^\top\d\alpha(x)\right).
\end{equation}
This is the infinitesimal analogue of the displacement covariance used in the spectral distance $\Wass_\gamma$ in Definition~\ref{def-spectral-wasserstein}. It measures the covariance of the velocity field before applying the gauge.

\begin{defn}[Spectral steepest-descent direction]\label{def-normalized-spectral-descent}
For $g\in L^2(\alpha;\RR^d)$, the spectral steepest-descent direction is any minimizer
\begin{equation}\label{eq-normalized-spectral-lmo}
\mathcal J_{\gamma}^{\alpha}(g)
\in
\uargmin{v\in L^2(\alpha;\RR^d)}
\left\{
\int \dotp{g(x)}{v(x)}\d\alpha(x)
+
\frac12\mathcal N_{\gamma,\alpha}(v)^2
\right\}.
\end{equation}
\end{defn}
The field $g$ should be read as the Wasserstein gradient $g=\Wgrad f(\alpha)=\nabla\delta f(\alpha)$. Since $\mathcal N_{\gamma,\alpha}$ is $1$-homogeneous in $v$, \eqref{eq-normalized-spectral-lmo} is the squared-norm steepest-descent step associated with the spectral tangent norm.

\begin{prop}[Polar formula for the spectral direction]\label{prop-normalized-spectral-lmo-polar}
Let
\(S_\alpha(g)\eqdef\int g(x)g(x)^\top\d\alpha(x).\)
Assume that the inverse-trace problem admits a positive definite minimizer
\[
A_\alpha^\star
\in
\uargmin{A\in\mathcal B_\gamma\cap\mathbb S_{++}^d}
\tr\!\left(A^{-1}S_\alpha(g)\right).
\]
Then
\begin{equation}\label{eq-normalized-spectral-lmo-explicit}
\mathcal J_{\gamma}^{\alpha}(g)(x)=-(A_\alpha^\star)^{-1}g(x)
\end{equation}
is a spectral steepest-descent direction. If the minimizer lies on the boundary of $\mathcal B_\gamma$, the same formula is understood by a limiting argument along positive definite minimizers, or equivalently by replacing the inverse by the corresponding inverse on the range of $S_\alpha(g)$.
\end{prop}
\begin{proof}
Using the polar formula $\gamma(M)=\sup_{A\in\mathcal B_\gamma}\tr(AM)$, the objective in \eqref{eq-normalized-spectral-lmo} can be written as
\[
\sup_{A\in\mathcal B_\gamma}
\left\{
\int \dotp{g}{v}\d\alpha
+
\frac12\int v^\top A v\d\alpha
\right\}.
\]
For a fixed $A\succ0$, the quadratic lower bound
\(\int \dotp{g}{v}\d\alpha+\frac12\int v^\top A v\d\alpha\)
is minimized by $v_A=-A^{-1}g$, with value $-\frac12\tr(A^{-1}S_\alpha(g))$. Let $v_\star=-(A_\alpha^\star)^{-1}g$ and
\(M_\star=\int v_\star(x)v_\star(x)^\top\d\alpha(x) = (A_\alpha^\star)^{-1}S_\alpha(g)(A_\alpha^\star)^{-1}.\)
Since $A_\alpha^\star$ minimizes $A\mapsto\tr(A^{-1}S_\alpha(g))$ over the convex polar set, its first-order optimality condition gives
\(\tr(AM_\star)\leq \tr(A_\alpha^\star M_\star) \qquad \text{for all } A\in\mathcal B_\gamma.\)
Thus $A_\alpha^\star$ is active in the polar representation of $\gamma(M_\star)$, and
\(\gamma(M_\star)=\tr(A_\alpha^\star M_\star) = \tr((A_\alpha^\star)^{-1}S_\alpha(g)).\)
For any $v$, the polar formula gives \(\int \dotp{g}{v}\d\alpha+\frac12\mathcal N_{\gamma,\alpha}(v)^2 \geq \int \dotp{g}{v}\d\alpha+\frac12\int v^\top A_\alpha^\star v\d\alpha,\)
and the right-hand side is minimized at $v_\star$. The activity identity above shows that equality holds at $v_\star$, hence $v_\star$ minimizes \eqref{eq-normalized-spectral-lmo}. The boundary case follows by approximation with positive definite matrices in the polar set.
\end{proof}
For the trace gauge, $\mathcal B_\gamma=\enscond{A}{0\preceq A\preceq\Id}$ and the minimizer is $A_\alpha^\star=\Id$. For the operator gauge $\gamma(M)=\lambda_{\max}(M)$, one has $\mathcal B_\gamma=\enscond{A\succeq0}{\tr(A)\leq1}$; if $S_\alpha(g)\succ0$, then

\[
A_\alpha^\star
=
\frac{S_\alpha(g)^{1/2}}{\tr(S_\alpha(g)^{1/2})},
\qquad
\mathcal J_{\gamma}^{\alpha}(g)(x)
=
-\tr(S_\alpha(g)^{1/2})\,S_\alpha(g)^{-1/2}g(x).
\]
\paragraph{JKO interpretation.}
\begin{equation}\label{eq-normalized-spectral-jko}
\alpha^{k+1}
\in
\uargmin{\alpha\in\Pp_2(\RR^d)}
\frac{1}{2\tau}\Wass_\gamma(\alpha^k,\alpha)^2+f(\alpha).
\end{equation}
\(\Wass_\gamma(\alpha^k,(\Id+\tau v)_\sharp\alpha^k)^2 = \tau^2\mathcal N_{\gamma,\alpha^k}(v)^2+o(\tau^2).\)
\begin{prop}[Formal normalized spectral gradient flow]\label{prop-normalized-spectral-gradient-flow}
Assume that $f$ has a smooth first variation and that the tangent expansion of $\Wass_\gamma$ above is valid along smooth perturbations. Then the formal small-step limit of \eqref{eq-normalized-spectral-jko} solves
\begin{equation}\label{eq-normalized-spectral-flow}
\partial_t\alpha_t
+
\diverg\!\left(
\alpha_t\,\mathcal J_{\gamma}^{\alpha_t}(\Wgrad f(\alpha_t))
\right)=0.
\end{equation}
Equivalently, if $A_t^\star$ solves the inverse-trace problem for
\(S_t=\int \Wgrad f(\alpha_t)(x)\Wgrad f(\alpha_t)(x)^\top\d\alpha_t(x),\)
then the velocity is \(v_t(x)=-(A_t^\star)^{-1}\Wgrad f(\alpha_t)(x).\)
Moreover, along smooth solutions, \(\frac{\d}{\d t}f(\alpha_t) = -\mathcal N_{\gamma,\alpha_t}(v_t)^2.\)
\end{prop}
\begin{proof}
Write $\alpha=(\Id+\tau v)_\sharp\alpha_t$. The spectral JKO objective equals, up to terms independent of $v$ and $o(\tau)$,
\[
\tau\left[
\frac12\mathcal N_{\gamma,\alpha_t}(v)^2
+
\int\dotp{\Wgrad f(\alpha_t)(x)}{v(x)}\d\alpha_t(x)
\right].
\]
Minimizing the bracket gives $v=\mathcal J_{\gamma}^{\alpha_t}(\Wgrad f(\alpha_t))$. Substituting this velocity into the transport expansion
\((\Id+\tau v)_\sharp\alpha_t = \alpha_t-\tau\diverg(\alpha_t v)+o(\tau)\)
gives \eqref{eq-normalized-spectral-flow}.
Finally, the variation formula gives $\frac{\d}{\d t}f(\alpha_t)=\int\dotp{\Wgrad f(\alpha_t)}{v_t}\d\alpha_t$. Since $v_t$ minimizes a $2$-homogeneous penalized objective, the one-dimensional rescaling $s\mapsto sv_t$ has derivative zero at $s=1$, hence
\(\int\dotp{\Wgrad f(\alpha_t)}{v_t}\d\alpha_t + \mathcal N_{\gamma,\alpha_t}(v_t)^2 =0,\)
which proves the dissipation identity.
\end{proof}
\paragraph{Empirical and Muon limits.}
For empirical measures $\alpha_X=n^{-1}\sum_i\delta_{x_i}$, stack the velocities $v(x_i)$ into a matrix $V\in\RR^{n\times d}$ and the gradients $g(x_i)$ into a matrix $G\in\RR^{n\times d}$. By $1$-homogeneity of $\gamma$,

\[
\mathcal N_{\gamma,\alpha_X}(v)^2
=
\gamma\!\left(\frac1n V^\top V\right)
=
\frac1n\gamma(V^\top V),
\qquad
\int\dotp{g}{v}\d\alpha_X
=
\frac1n\tr(G^\top V).
\]
Thus \eqref{eq-normalized-spectral-lmo} is, up to the harmless factor $1/n$, the finite-dimensional matrix LMO

\begin{equation}\label{eq-normalized-matrix-lmo}
J_\gamma(G)
\in
\uargmin{V\in\RR^{n\times d}}
\left\{
\tr(G^\top V)+\frac12\gamma(V^\top V)
\right\}.
\end{equation}
\(\dot X(t)=J_\gamma(G_X(t)), \qquad (G_X(t))_i=\Wgrad f(\alpha_{X(t)})(x_i(t)).\)
\begin{equation}\label{eq-normalized-muon-lmo}
J_\gamma(G)
=
-\norm{G}_{S_1}UW^\top
=
-\tr\!\left((G^\top G)^{1/2}\right)G(G^\top G)^{\dagger/2}.
\end{equation}