% !TEX root = ../OT4ML.tex

\chapter{Notation Table}
\label{app-notation-table}

This appendix collects the main notation used throughout the book. The last
column points to the first section, equation, definition, proposition or theorem
where the notation is defined or first used in a mathematically meaningful way.

{\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.16}
\begin{longtable}{>{\raggedright\arraybackslash}p{.24\textwidth}
                  >{\raggedright\arraybackslash}p{.45\textwidth}
                  >{\raggedright\arraybackslash}p{.24\textwidth}}
\hline
\textbf{Notation} & \textbf{Meaning} & \textbf{First reference} \\
\hline
\endfirsthead
\hline
\textbf{Notation} & \textbf{Meaning} & \textbf{First reference} \\
\hline
\endhead
\hline
\endfoot

\multicolumn{3}{l}{\textbf{Ambient spaces, measures and elementary objects}}\\
\hline
$\RR^d$ & Euclidean ambient space. & Section~\ref{sec-measures} \\
$\X,\Y$ & Source and target spaces. & Eq.~\eqref{eq-monge-continuous} \\
$\Mm(\X)$ & Finite signed Radon measures on $\X$. & Section~\ref{sec-measures} \\
$\Mm_+(\X),\Mm_+^1(\X)$ & Positive finite measures and probability measures. & Section~\ref{sec-measures} \\
$\Pp(\X),\Pp_p(\X)$ & Probability measures, with finite $p$-moment for $\Pp_p$. & Section~\ref{sec-kantorovich-continuous} \\
$\simplex_n$ & Probability simplex of histograms of length $n$. & Definition~\ref{def-probability-simplex} \\
$\de_x$ & Dirac mass at $x$. & Definition~\ref{def-discrete-measure} \\
$\al,\be,\ga$ & Source, target and auxiliary probability measures. & Eq.~\eqref{eq-monge-continuous} \\
$\density{\al}$ & Density of $\al$ with respect to a reference measure. & Definition~\ref{def-relative-density} \\
$\d\al,\d x$ & Integration against $\al$ and against Lebesgue measure. & Section~\ref{sec-measures} \\
$\EE$ & Expectation of a random variable. & Section~\ref{sec-measures} \\
$\supp(\pi)$ & Topological support of a measure. & Definition~\ref{def:support} \\
$\Supp(\b)$ & Index support of a histogram. & Eq.~\eqref{eq-discr-diverg} \\
$\Cc(\X)$ & Continuous real-valued functions on $\X$. & Section~\ref{sec-measures} \\
$\norm{\cdot}$ & Euclidean norm or the norm indicated by a subscript. & Chapter~\ref{sec-matching} \\
$\dotp{\cdot}{\cdot}$ & Euclidean/Frobenius pairing or measure-function pairing. & Section~\ref{sec-measures} \\

\hline
\multicolumn{3}{l}{\textbf{Discrete matching and discrete Kantorovich OT}}\\
\hline
$(x_i)_i,(y_j)_j$ & Source and target point clouds. & Eq.~\eqref{eq-optimal-assignment} \\
$\C=(\C_{i,j})$ & Cost matrix between source and target points. & Eq.~\eqref{eq-optimal-assignment} \\
$\sigma\in\Perm(n)$ & Permutation encoding a one-to-one matching. & Eq.~\eqref{eq-optimal-assignment} \\
$P_\sigma,\mathcal P_n^{\mathrm{perm}}$ & Permutation matrix and the set of all such matrices. & Definition~\ref{def-permutation-matrices} \\
$\mathcal B_n$ & Birkhoff polytope of bistochastic matrices. & Definition~\ref{def-birkhoff-polytope} \\
$\a,\b$ & Discrete probability histograms. & Eq.~\eqref{eq-discr-couplings} \\
$\P$ & Discrete transport/coupling matrix. & Eq.~\eqref{eq-discr-couplings} \\
$\CouplingsD(\a,\b)$ & Polytope of discrete couplings with marginals $\a,\b$. & Eq.~\eqref{eq-discr-couplings} \\
$\ones_n,\transp{\P}$ & All-ones vector and transpose of $\P$. & Eq.~\eqref{eq-discr-couplings} \\
$\MKD_\C(\a,\b)$ & Discrete Kantorovich optimal value with cost $\C$. & Eq.~\eqref{eq-kanto-discr} \\
$\distD$ & Ground distance matrix for discrete Wasserstein distances. & Definition~\ref{def-discrete-wasserstein-distance} \\
$\WassD_p(\a,\b)$ & Discrete $p$-Wasserstein distance. & Definition~\ref{def-discrete-wasserstein-distance} \\

\hline
\multicolumn{3}{l}{\textbf{Monge maps, one-dimensional OT and Gaussians}}\\
\hline
$T,\T$ & Transport map. & Eq.~\eqref{eq-monge-continuous} \\
$T_\sharp\al$ & Push-forward of $\al$ by $T$. & Definition~\ref{defn-pushfwd} \\
$T^\sharp g$ & Pullback of a test function, $T^\sharp g=g\circ T$. & Remark~\ref{rem-pullback-pushforward} \\
$\Id$ & Identity map. & Definition~\ref{defn-pushfwd} \\
$\tilde\Wass_p$ & Directed Monge transport distance. & Eq.~\eqref{eq-monge-distance} \\
$\nabla\phi$ & Brenier map for quadratic cost. & Theorem~\ref{thm-brenier} \\
$\cumul{\al}$ & Cumulative distribution function of a 1-D measure. & Eq.~\eqref{eq-cumul-defn} \\
$\cumul{\al}^{-1}$ & Quantile function of a 1-D measure. & Eq.~\eqref{eq-OT-map-1d} \\
$\Gaussian(\mean,\cov)$ & Gaussian law with mean $\mean$ and covariance $\cov$. & Eq.~\eqref{eq-gauss-pf} \\
$\mean_\al,\cov_\al$ & Mean and covariance of a Gaussian measure $\al$. & Eq.~\eqref{eq-dist-gauss} \\
$\Bb(\cov_\al,\cov_\be)$ & Bures covariance distance. & Definition~\ref{def-bures-metric} \\
$\Phi_2(\mu)$ & Raw second-moment matrix of a probability measure. & Proposition~\ref{prop-bures-second-moment-lift} \\
$D_{\Phi_2}$ & Second-moment quotient distance, equal to the Bures metric. & Proposition~\ref{prop-bures-second-moment-lift} \\
$\tr(\cov)$ & Trace of a matrix. & Eq.~\eqref{eq-dist-gauss} \\

\hline
\multicolumn{3}{l}{\textbf{Continuous Kantorovich OT and Wasserstein distances}}\\
\hline
$\pi$ & Coupling or transport plan. & Definition~\ref{def-continuous-couplings} \\
$\Couplings(\al,\be)$ & Set of couplings between $\al$ and $\be$. & Eq.~\eqref{eq-coupling-generic} \\
$\MK_\c(\al,\be)$ & Kantorovich optimal value with ground cost $\c$. & Eq.~\eqref{eq-mk-generic} \\
$\dist$ & Ground distance on the underlying metric space. & Eq.~\eqref{eq-defn-wass-dist} \\
$\Wass_p(\al,\be)$ & $p$-Wasserstein distance. & Definition~\ref{def-wasserstein-distance} \\
$\Wass_\infty(\al,\be)$ & Worst-displacement Wasserstein distance. & Eq.~\eqref{eq-wass-infty} \\
$\mathfrak A,\mathfrak B$ & Probability laws over probability measures. & Eq.~\eqref{eq-wow-parametric-law} \\
$\bar\alpha_{\mathfrak A}$ & Collapsed mixture associated with a law over measures. & Definition~\ref{def-collapsed-barycentric-mixture} \\
$\mathbb W_2$ & Wasserstein distance on the Wasserstein space. & Eq.~\eqref{eq-wow-distance} \\
$\Gamma$ & $c$-cyclically monotone subset of $\X\times\Y$. & Definition~\ref{def:ccm} \\
$\rho$ & Glued or composed coupling. & Lemma~\ref{lem-gluing-general} \\
$\rightharpoonup$ & Weak$^*$ convergence of measures. & Definition~\ref{dfn-weak-conv} \\
$\TV,\norm{\cdot}_{\TV}$ & Total variation divergence/norm. & Section~\ref{sec-measures} \\

\hline
\multicolumn{3}{l}{\textbf{Duality, transforms and weak norms}}\\
\hline
$\fD,\gD$ & Discrete dual potentials. & Eq.~\eqref{eq-dual} \\
$\f,\g$ & Continuous dual potentials. & Eq.~\eqref{eq-dual-generic} \\
$\PotentialsD(\a,\b)$ & Feasible set of discrete dual potentials. & Eq.~\eqref{eq-feasible-potential} \\
$\Potentials(\c)$ & Feasible set of continuous dual potentials. & Eq.~\eqref{eq-dfn-pot-dual} \\
$f^c,g^c$ & $c$-transform of a potential. & Definition~\ref{def-c-transform} \\
$\Laguerre_j(\gD)$ & Laguerre/power cell in semi-discrete OT. & Eq.~\eqref{eq-laguerre-cells} \\
$\Qq_m(\al)$ & Optimal $m$-point quantization error. & Eq.~\eqref{eq-optimal-quantization} \\
$\Lip(f)$ & Lipschitz constant of $f$. & Eq.~\eqref{eq-lip-constant} \\
$\Wass_1$ & Kantorovich--Rubinstein distance/norm. & Eq.~\eqref{eq-w1-metric} \\
$\Wass_{1,G}$ & Graph Wasserstein-1/transshipment distance. & Proposition~\ref{prop-graph-w1-beckmann} \\
$d_G,\nabla_G,\operatorname{div}_G$ & Graph geodesic distance, gradient and divergence. & Proposition~\ref{prop-graph-w1-beckmann} \\
$\norm{\cdot}_B$ & Dual norm induced by a discriminator class $B$. & Eq.~\eqref{eq-dual-norm-cont} \\
$\RKHS,\Krkhs$ & Reproducing kernel Hilbert space and its kernel. & Definition~\ref{def-kernel-mmd-norm} \\
$\MMD_k$ & Maximum mean discrepancy/kernel norm for $k$. & Definition~\ref{def-kernel-mmd-norm} \\
$\Divergm_\phi,\DivergmD_\phi$ & Continuous and discrete $\phi$-divergences. & Eq.~\eqref{eq-phi-div} \\
$\phi'_\infty$ & Recession slope of an entropy function. & Definition~\ref{def_entropy} \\
$\phi^\star$ & Legendre transform of $\phi$. & Eq.~\eqref{eq-legendre} \\
$\KL,\KLD$ & Continuous and discrete Kullback--Leibler divergences. & Definitions~\ref{def-discrete-relative-entropy}, \ref{def-measure-relative-entropy} \\
$\Hellinger$ & Hellinger divergence/distance. & Section~\ref{sec-phi-div} \\
$\JS$ & Jensen--Shannon divergence. & Section~\ref{sec-phi-div} \\

\hline
\multicolumn{3}{l}{\textbf{Entropic regularization and Sinkhorn algorithms}}\\
\hline
$\epsilon$ & Entropic regularization strength. & Eq.~\eqref{eq-regularized-discr} \\
$\HD(\P)$ & Shannon--Boltzmann entropy of a matrix. & Definition~\ref{def-discrete-shannon-boltzmann-entropy} \\
$\MKD_\C^\epsilon(\a,\b)$ & Discrete entropic OT value. & Eq.~\eqref{eq-regularized-discr} \\
$\MK_\c^\epsilon(\al,\be)$ & Continuous entropic OT value. & Eq.~\eqref{eq-entropic-generic} \\
$\K$ & Gibbs kernel $e^{-\C/\epsilon}$. & Eq.~\eqref{eq-scaling-form} \\
$\uD,\vD$ & Left and right Sinkhorn scalings. & Eq.~\eqref{eq-scaling-form} \\
$\diag(\uD)\K\diag(\vD)$ & Scaling form of the entropic coupling. & Eq.~\eqref{eq-sink-matrix} \\
$\odot$ & Entrywise product of vectors. & Eq.~\eqref{eq-dualsinkhorn-constraints2} \\
$\it{\uD},\itt{\uD}$ & Current and next Sinkhorn iterates. & Eq.~\eqref{eq-sinkhorn} \\
$\Hilbert$ & Hilbert projective metric on positive vectors. & Definition~\ref{def-hilbert-metric} \\
$\Proj^\KLD$ & KL/Bregman projection. & Eq.~\eqref{eq-kl-proj} \\
$\bar\MK_\c^\epsilon(\al,\be)$ & Debiased Sinkhorn divergence. & Eq.~\eqref{eq-sinkhorn-divergence} \\

\hline
\multicolumn{3}{l}{\textbf{Extensions of OT}}\\
\hline
$\psi_1,\psi_2$ & Entropy functions penalizing marginal mismatch. & Eq.~\eqref{eq-unbalanced-primal} \\
$\UW_c,\UW_{c,\tau}$ & Relaxed unbalanced OT value with marginal penalties. & Eq.~\eqref{eq-unbalanced-primal} \\
$\operatorname{POT}_m$ & Partial OT value transporting exactly mass $m$. & Section~\ref{sec-unbalanced} \\
$\operatorname{POT}^{\TV}_\lambda$ & TV-penalized partial OT value with mass price $\lambda$. & Proposition~\ref{prop-tv-partial-ot-lagrange} \\
$L_c$ & Reverse-formulation local unbalanced cost. & Eq.~\eqref{eq-unbalanced-reverse-local-cost} \\
$H_c$ & Homogeneous perspective of the local cost $L_c$. & Eq.~\eqref{eq-unbalanced-homogeneous-local-cost} \\
$\HW$ & Homogeneous unbalanced formulation. & Eq.~\eqref{eq-homogeneous} \\
$\mathfrak{C}[\X]$ & Cone over the metric space $\X$. & Section~\ref{sec-unbalanced} \\
$\CW,\CW_\kappa$ & Cone formulation of unbalanced OT, with $\CW_\kappa$ using growth scale $\kappa$. & Theorem~\ref{thm-cone-unbalanced-ot}, Eq.~\eqref{eq-dynamic-unbalanced-ot} \\
$\mathcal A_\kappa$ & Dynamic unbalanced perspective action for transport and growth. & Eq.~\eqref{eq-dynamic-unbalanced-ot} \\
$\WFR_\kappa$ & Wasserstein--Fisher--Rao dynamic distance with growth scale $\kappa$. & Eq.~\eqref{eq-dynamic-unbalanced-ot} \\
$\be_s,\la_s$ & Input measures and weights in barycenter problems. & Eq.~\eqref{eq-barycenter-generic} \\
$\al^\star$ & Optimal measure, often a barycenter. & Section~\ref{sec-barycenters} \\
$\SW_p$ & Sliced Wasserstein distance. & Definition~\ref{def-sliced-wasserstein} \\
$\Sphere^{d-1}$ & Unit sphere of projection directions. & Definition~\ref{def-sliced-wasserstein} \\
$P_\theta$ & Projection on direction $\theta$. & Definition~\ref{def-sliced-wasserstein} \\
$\MaxSW_p$ & Max-sliced Wasserstein distance. & Definition~\ref{def-sliced-variants} \\
$\operatorname{MSWGG}_2$ & Min-sliced lifted-plan upper bound on $\Wass_2$. & Section~\ref{sec-sliced-wasserstein} \\
$\SW_{p,k},\MaxSW_{p,k}$ & Average and max Wasserstein distances over $k$-dimensional projections. & Definition~\ref{def-subspace-sliced-wasserstein} \\
$\Wass_\gamma$ & Spectral Wasserstein distance associated with a matrix gauge $\gamma$. & Eq.~\eqref{eq-spectral-wasserstein} \\
$\mathcal B_\gamma$ & Polar set defining the robust projected form of $\Wass_\gamma$. & Eq.~\eqref{eq-spectral-polar-set} \\
$\Wass_{2,A}$ & Quadratic Wasserstein pseudodistance after projection by $A^{1/2}$. & Eq.~\eqref{eq-quadratic-projected-cost} \\
$\SRW_{2,k}$ & Paty--Cuturi subspace robust Wasserstein distance. & Section~\ref{sec-spectral-subspace-wasserstein} \\
$\LOT_\rho$ & Linear OT distance around reference $\rho$. & Eq.~\eqref{eq-lot-embedding} \\
$\bar T_\pi$ & Barycentric projection of a coupling $\pi$. & Eq.~\eqref{eq-barycentric-projection} \\
$\bar\beta_\pi$ & Pushforward of $\alpha$ by the barycentric projection. & Eq.~\eqref{eq-barycentric-projection} \\
$\WOT_C$ & Weak OT value with conditional-law cost $C$. & Eq.~\eqref{eq-weak-ot} \\
$g^C$ & Weak $C$-transform in weak OT duality. & Proposition~\ref{prop-weak-ot-duality} \\
$\Couplings_{\mathrm{mart}}(\alpha,\beta)$ & Martingale couplings between $\alpha$ and $\beta$. & Definition~\ref{def-martingale-coupling} \\
$\preceq_{\mathrm{st}},\preceq_{\mathrm{cx}}$ & Stochastic order and convex order. & Section~\ref{sec-martingale-ot} \\
$u_t,V_t$ & Positive vector-valued density and spatial flux. & Eqs.~\eqref{eq-vector-valued-bb}, \eqref{eq-vector-valued-continuity} \\
$\mathcal W_{\Phi}$ & Dynamic vector-valued BB-type cost. & Eq.~\eqref{eq-vector-valued-bb} \\
$\distD,\distD'$ & Intra-domain distance matrices in discrete GW. & Eq.~\eqref{eq-gw-def} \\
$\De$ & Discrepancy between intra-domain distances. & Eq.~\eqref{eq-gw-def} \\
$\GWD$ & Discrete Gromov--Wasserstein cost. & Eq.~\eqref{eq-gw-def} \\
$\XX,\YY$ & Metric-measure spaces. & Definition~\ref{def-metric-measure-space} \\
$\GW$ & Continuous Gromov--Wasserstein distance. & Eq.~\eqref{eq-gw-generic} \\
$d_{\mathrm H},d_{\mathrm{GH}}$ & Hausdorff and Gromov--Hausdorff distances. & Section~\ref{sec-gromov-wasserstein} \\
$\operatorname{FGW}_{\lambda,p}$ & Fused Gromov--Wasserstein distance. & Section~\ref{sec-gromov-wasserstein} \\
$\mathbb S^m,\mathbb S_+^m$ & Real symmetric matrices and their positive semidefinite cone. & Definition~\ref{def-positive-matrix-valued-measure} \\
$A_t,P_t$ & Positive matrix-valued density and spatial matrix flux. & Eqs.~\eqref{eq-matrix-valued-bb}, \eqref{eq-matrix-valued-continuity} \\
$\mathcal W_{\mathrm{mat}}$ & Conservative matrix-valued BB-type cost. & Eq.~\eqref{eq-matrix-valued-bb} \\
$\mathbb H_n,\mathbb H_n^+,\mathbb H_n^{+,1}$ & Hermitian matrices, positive semidefinite Hermitian matrices and density matrices. & Definition~\ref{def-hermitian-density-matrices} \\
$\operatorname{Tr}_A,\operatorname{Tr}_B$ & Partial traces of a bipartite matrix. & Eq.~\eqref{eq-qot-partial-traces} \\
$\mathrm{QOT}_C(A,B)$ & Finite-dimensional quantum OT value with cost observable $C$. & Eq.~\eqref{eq-qot-primal} \\
$\mathrm{QOT}_C^\epsilon(A,B)$ & Entropically regularized quantum OT value. & Eq.~\eqref{eq-qot-entropic-primal} \\
$T_e(F,G),T_s(F,G)$ & Exact Gibbs coupling and symmetric Gurvits-scaling surrogate. & Eqs.~\eqref{eq-qot-gibbs-coupling}, \eqref{eq-qot-symmetric-scaling} \\

\hline
\multicolumn{3}{l}{\textbf{Dynamic OT and Wasserstein gradient flows}}\\
\hline
$\alpha_t$ & Time-dependent curve of probability measures. & Eq.~\eqref{eq:eulerian-advection} \\
$v_t$ & Eulerian velocity field transporting $\alpha_t$. & Eq.~\eqref{eq:eulerian-advection} \\
$T_t$ & Lagrangian particle flow map. & Eq.~\eqref{eq:lagrangian-advection} \\
$P_t$ & Interpolant map in flow matching; later also path evaluation. & Eq.~\eqref{eq:interp-coupling} \\
$\Wass_2^2$ via action & Benamou--Brenier dynamic formulation. & Eq.~\eqref{eq:benamou-brenier} \\
$\Wgrad f(\alpha)$ & Wasserstein gradient of a functional. & Proposition~\ref{prop-formal-wass-gradient} \\
$\Pp_{2,\mathbf m}(\Omega;\RR_+^p)$ & Positive multi-species measures with fixed component masses. & Eq.~\eqref{eq-multispecies-space} \\
$\Wass_{2,\oplus}$ & Mass-weighted product Wasserstein distance for independent species transport. & Eq.~\eqref{eq-multispecies-product-metric} \\
$\delta f(\alpha)$ & First variation of $f$ at $\alpha$. & Proposition~\ref{prop-formal-wass-gradient} \\
$\partial_t\alpha+\diverg(\alpha v)=0$ & Continuity equation. & Eq.~\eqref{eq:eulerian-advection} \\
$\alpha_{t+\tau}$ & One JKO/minimizing-movement step. & Eq.~\eqref{eq:jko-discr} \\
$\Ss=C([0,1];\RR^d)$ & Path space in the superposition formulation. & Chapter~\ref{sec-wasserstein-flows} \\
\end{longtable}
}
